---
title: "The Ultimate R Basics Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(knitr)
```

# How to Use This Cheat Sheet

This is a complete R reference. Every section is independent. Find what you need, copy the pattern, change the names.

| Topic | Section |
|---|---|
| Assign variables | 3 |
| Know data types | 4 |
| Build vectors, lists, data frames | 5 |
| Subset data with `[`, `[[`, `$` | 6 |
| Arithmetic, comparison, logic | 7 |
| Pipe `|>` and `%>%` | 8 |
| Write functions | 9 |
| If/else, loops, apply family | 10 |
| Strings | 11 |
| Factors | 12 |
| Dates | 13 |
| Missing values | 14 |
| Read/write files | 15 |
| dplyr: filter, select, mutate, arrange | 16 |
| Grouped summaries | 17 |
| tidyr: pivot, separate, unite | 18 |
| Joins | 19 |
| purrr: map family | 20 |
| Base R plots | 21 |
| ggplot2 plots | 22 |
| Statistical functions | 23 |
| Random numbers | 24 |
| Vectorization vs loops | 25 |
| Debugging | 26 |
| Common errors | 27 |
| Packages | 28 |
| RStudio shortcuts | 29 |

---

# Part 0 — The 30-Second Survival Guide

If you only remember five things:

1. **Assign with `<-`**: `x <- 5`.
2. **Call functions with parentheses**: `mean(x)`.
3. **Index with square brackets**: `x[1]`, `df[1, 2]`, `df$col`.
4. **Pipe chains operations**: `x |> mean() |> round(2)`.
5. **Get help with `?`**: `?mean`.

Everything else is syntax you can look up.

---

# Part 1 — Installing and Loading Packages

## 1.1 Install once

```{r eval=FALSE}
install.packages("tidyverse")
install.packages(c("forecast", "car", "lmtest"))
```

## 1.2 Load every session

```{r eval=FALSE}
library(tidyverse)
library(forecast)
```

**Rules:**
- `install.packages()` needs quotes around the name.
- `library()` does **not** need quotes.
- You only install once; you load every session.

## 1.3 See what's installed and loaded

```{r}
# All installed packages
length(rownames(installed.packages()))

# Currently loaded
search()
```

## 1.4 Update

```{r eval=FALSE}
update.packages(ask = FALSE)
```

---

# Part 2 — Getting Help

| What you want | How |
|---|---|
| Function help | `?mean` |
| Function help (alt) | `help("mean")` |
| Search help | `??regression` |
| See arguments | `args(lm)` |
| See examples | `example(lm)` |
| See source code | `lm` (just type the name, no parens) |
| Vignettes | `vignette("dplyr")` |
| Package help | `help(package = "dplyr")` |

```{r}
# --- Self-contained: help examples ---
?mean
args(lm)
```

**Tip:** In RStudio, highlight a function name and press **F1** to open help.

---

# Part 3 — Variables and Assignment

## 3.1 Assignment operators

```{r}
# --- Self-contained: assignment ---
p3_x <- 5         # standard
p3_y = 10         # works, but not idiomatic
15 -> p3_z        # right-to-left (rare)

p3_x
p3_y
p3_z
```

**Use `<-`**. Reserve `=` for function arguments.

## 3.2 Naming rules

- Start with a letter (or `.` not followed by a digit).
- Can contain letters, digits, `.`, `_`.
- **Case-sensitive**: `x` and `X` are different.
- Avoid `c`, `T`, `F`, `mean` — they're built-in.

**Valid:** `my_var`, `x1`, `myVar`, `data.2024`
**Invalid:** `1x`, `my-var`, `my var`

## 3.3 List and remove objects

```{r}
# --- Self-contained: list and remove objects ---
p3_a <- 1
p3_b <- 2
p3_c <- 3

ls()                    # list objects (in this session, many exist)
ls(pattern = "^p3_")    # list by pattern
rm(p3_a)                # remove one
rm(list = ls(pattern = "^p3_"))   # remove all p3_ objects
```

## 3.4 Data types overview

| Type | Example | Check with |
|---|---|---|
| `numeric` (double) | `3.14`, `-1.5e3` | `is.numeric(x)` |
| `integer` | `5L`, `-2L` | `is.integer(x)` |
| `character` | `"hello"`, `'hi'` | `is.character(x)` |
| `logical` | `TRUE`, `FALSE`, `T`, `F` | `is.logical(x)` |
| `complex` | `3 + 2i` | `is.complex(x)` |
| `factor` | `factor(c("a","b"))` | `is.factor(x)` |
| `Date` | `as.Date("2024-01-01")` | `is.Date(x)` |

---

# Part 4 — Data Types in Detail

## 4.1 Numeric

```{r}
# --- Self-contained: numeric ---
p4_a <- 3.14
p4_b <- 1e3        # 1000
p4_c <- -2.5
p4_d <- 5L         # integer (add L)

class(p4_a)        # "numeric"
class(p4_d)        # "integer"
is.numeric(p4_a)   # TRUE
is.integer(p4_a)   # FALSE
```

**Note:** By default, numbers are `numeric` (double). Use `L` suffix for integers.

## 4.2 Character

```{r}
# --- Self-contained: character ---
p4_name <- "Parsa"
p4_greeting <- 'Hello'
p4_sentence <- "She said \"hi\""

nchar(p4_name)         # number of characters
toupper(p4_name)
tolower(p4_name)
paste(p4_greeting, p4_name)
paste0(p4_greeting, ", ", p4_name, "!")
```

**Rule:** Single and double quotes are equivalent. Use `\"` to escape quotes inside a string.

## 4.3 Logical

```{r}
# --- Self-contained: logical ---
p4_t <- TRUE
p4_f <- FALSE
p4_cmp <- 3 > 2     # TRUE
p4_neg <- !p4_t     # FALSE

p4_t & p4_f         # AND -> FALSE
p4_t | p4_f         # OR  -> TRUE
p4_t && p4_f        # AND (scalar) -> FALSE
p4_t || p4_f        # OR  (scalar) -> TRUE
```

`&` and `|` are vectorized; `&&` and `||` short-circuit and return one value.

## 4.4 Coercion

R converts between types when forced. Order of precedence: **logical < integer < numeric < character**.

```{r}
# --- Self-contained: coercion ---
p4_mixed <- c(1, 2, "three")   # all become character
p4_mixed
class(p4_mixed)

as.numeric("3.14")             # 3.14
as.numeric("abc")              # NA with warning
as.character(42)               # "42"
as.logical(0)                  # FALSE
as.logical(1)                  # TRUE
```

**Force conversion explicitly** with `as.*()` to avoid surprises.

## 4.5 Type check and convert

```{r}
# --- Self-contained: type checks ---
p4_v <- c(1, 2, 3)

is.numeric(p4_v)
is.character(p4_v)
is.vector(p4_v)
as.character(p4_v)
```

---

# Part 5 — Data Structures

## 5.1 Vector (1D, same type)

```{r}
# --- Self-contained: vectors ---
p5_num  <- c(1, 2, 3, 4, 5)
p5_chr  <- c("a", "b", "c")
p5_lgl  <- c(TRUE, FALSE, TRUE)
p5_seq  <- 1:10
p5_rep  <- rep(1:3, times = 2)     # 1 2 3 1 2 3
p5_rep2 <- rep(1:3, each = 2)      # 1 1 2 2 3 3
p5_seq2 <- seq(0, 1, by = 0.25)    # 0.00 0.25 0.50 0.75 1.00

length(p5_num)
length(p5_chr)
```

## 5.2 Matrix (2D, same type)

```{r}
# --- Self-contained: matrices ---
p5_mat <- matrix(1:12, nrow = 3, ncol = 4)
p5_mat
p5_mat2 <- matrix(1:12, nrow = 3, byrow = TRUE)
p5_mat2

dim(p5_mat)          # 3 4
nrow(p5_mat); ncol(p5_mat)
t(p5_mat)            # transpose
p5_mat %*% t(p5_mat) # matrix multiplication
```

## 5.3 Array (nD, same type)

```{r}
# --- Self-contained: arrays ---
p5_arr <- array(1:24, dim = c(2, 3, 4))
dim(p5_arr)
p5_arr[1, 2, 3]      # one element
```

## 5.4 List (1D, mixed types)

```{r}
# --- Self-contained: lists ---
p5_list <- list(
  name  = "Parsa",
  age   = 25,
  scores = c(90, 85, 88),
  passed = TRUE
)
p5_list
p5_list$name
p5_list[["age"]]
p5_list[[3]]           # the scores vector
length(p5_list)
names(p5_list)
```

## 5.5 Data frame (2D, mixed types, columns same length)

```{r}
# --- Self-contained: data frame ---
p5_df <- data.frame(
  id   = 1:4,
  name = c("A", "B", "C", "D"),
  score = c(90, 85, 88, 92)
)
p5_df
str(p5_df)
summary(p5_df)
nrow(p5_df); ncol(p5_df)
names(p5_df)
```

## 5.6 Tibble (modern data frame)

```{r}
# --- Self-contained: tibble ---
p5_tib <- tibble(
  id    = 1:4,
  name  = c("A", "B", "C", "D"),
  score = c(90, 85, 88, 92)
)
p5_tib
class(p5_tib)
```

**Tibble vs data.frame:**
- Tibbles print nicely (no row-name clutter, only first 10 rows shown).
- Tibbles don't auto-convert strings to factors (R ≥ 4.0).
- Tibbles allow list-columns.

## 5.7 Summary: which structure to use

| Need | Use |
|---|---|
| One column of data | vector |
| Table of same-type numbers (math) | matrix |
| Multi-dimensional numeric array | array |
| Mixed-type container | list |
| Table with named columns | tibble (or data.frame) |

---

# Part 6 — Indexing and Subsetting

## 6.1 Vectors

```{r}
# --- Self-contained: vector indexing ---
p6_v <- c(10, 20, 30, 40, 50)

p6_v[1]         # 10 (first)
p6_v[c(1, 3)]   # 10 30
p6_v[-2]        # remove 2nd -> 10 30 40 50
p6_v[2:4]       # 20 30 40
p6_v[p6_v > 25] # 30 40 50
p6_v[c(TRUE, FALSE, TRUE, FALSE, TRUE)]  # 10 30 50
```

**Warning:** R indexing starts at 1, not 0.

## 6.2 Matrices

```{r}
# --- Self-contained: matrix indexing ---
p6_m <- matrix(1:12, nrow = 3)
p6_m
p6_m[1, 1]        # row 1, column 1
p6_m[1, ]         # row 1
p6_m[, 2]         # column 2
p6_m[1:2, 2:3]    # submatrix
p6_m[p6_m > 5]    # vector of elements > 5
```

## 6.3 Data frames / tibbles

```{r}
# --- Self-contained: data frame indexing ---
p6_df <- tibble(
  id    = 1:4,
  name  = c("A", "B", "C", "D"),
  score = c(90, 85, 88, 92)
)
p6_df
p6_df[1, ]            # row 1
p6_df[, "score"]      # column by name (returns tibble)
p6_df$score           # column by $ (returns vector)
p6_df[["score"]]      # column by [[ (returns vector)
p6_df$score[2]        # 2nd element of score
p6_df[p6_df$score > 87, ]   # rows where score > 87
```

## 6.4 Lists

```{r}
# --- Self-contained: list indexing ---
p6_l <- list(a = 1, b = 2:4, c = "hi")

p6_l[1]         # sub-list with one element
p6_l[[1]]       # the element itself (numeric 1)
p6_l$b          # the vector 2:4
p6_l[["b"]]     # same
p6_l$b[2]       # 3
```

**Rule:** `[` returns a container, `[[` returns the contents.

## 6.5 `$` vs `[[` vs `[`

| Access | Returns | Use for |
|---|---|---|
| `df$col` | vector | one column by name |
| `df[["col"]]` | vector | one column, programmatic |
| `df["col"]` | tibble | one column, keep as tibble |
| `df[1]` | tibble | first column |
| `df[1, ]` | tibble | first row |
| `df[[1]]` | vector | first column, contents |

## 6.6 Named vectors

```{r}
# --- Self-contained: named vectors ---
p6_named <- c(a = 1, b = 2, c = 3)
p6_named["b"]     # 2
p6_named[["b"]]   # 2
names(p6_named)   # "a" "b" "c"
```

---

# Part 7 — Operators

## 7.1 Arithmetic

```{r}
# --- Self-contained: arithmetic ---
7 + 3      # 10
7 - 3      # 4
7 * 3      # 21
7 / 3      # 2.333...
7 %/% 3    # 2  (integer division)
7 %% 3     # 1  (modulo)
7 ^ 3      # 343
sqrt(16)   # 4
abs(-5)    # 5
exp(1)     # 2.718...
log(100)   # natural log
log10(100) # 2
log2(8)    # 3
```

## 7.2 Comparison

```{r}
# --- Self-contained: comparison ---
3 == 3     # TRUE
3 != 4     # TRUE
3 < 4      # TRUE
3 <= 4     # TRUE
3 > 4      # FALSE
3 >= 4     # FALSE
"a" < "b"  # TRUE (alphabetical)
```

**Warning:** `==` is comparison, `=` is assignment.

## 7.3 Logical

```{r}
# --- Self-contained: logical ---
TRUE & TRUE     # TRUE
TRUE & FALSE    # FALSE
TRUE | FALSE    # TRUE
!TRUE           # FALSE

# Short-circuit (scalar, stops early)
TRUE && FALSE   # FALSE
TRUE || FALSE   # TRUE

# Vectorized
c(TRUE, FALSE) & c(TRUE, TRUE)   # TRUE FALSE
```

## 7.4 Membership and matching

```{r}
# --- Self-contained: membership ---
3 %in% c(1, 2, 3)       # TRUE
"a" %in% c("a", "b")    # TRUE
c(1, 2, 3) %in% c(2, 3) # FALSE TRUE TRUE

match(c("b", "a"), c("a", "b", "c"))   # 2 1
```

## 7.5 Pipe operators

```{r}
# --- Self-contained: pipes ---
set.seed(701)
p7_v <- rnorm(10)

# Without pipe
round(mean(p7_v), 2)

# Base pipe (R >= 4.1)
p7_v |> mean() |> round(2)

# magrittr pipe (tidyverse)
p7_v %>% mean() %>% round(2)
```

**Rule:** `x |> f()` is the same as `f(x)`. Pipes pass the left side as the **first** argument of the right side.

**Placeholder for non-first argument:**
```{r}
# --- Self-contained: pipe with placeholder ---
p7_x <- c(1, NA, 3)
p7_x |> mean(na.rm = TRUE)   # works
# p7_x |> mean(x = _)         # _ only with base pipe, R >= 4.2
```

## 7.6 Special values

```{r}
# --- Self-contained: special values ---
Inf          # positive infinity
-Inf
NaN          # Not a Number (0/0)
NA           # missing
NULL         # empty / no value

1/0          # Inf
0/0          # NaN
NA + 1       # NA
is.na(NA)    # TRUE
is.nan(NaN)  # TRUE
is.null(NULL)# TRUE
```

---

# Part 8 — The Pipe in Depth

## 8.1 Why pipe

Without pipe:
```{r}
# --- Self-contained: without pipe ---
set.seed(801)
p8_v <- rnorm(20)
round(sqrt(mean(p8_v^2)), 3)
```

With pipe:
```{r}
# --- Self-contained: with pipe ---
set.seed(802)
p8_v <- rnorm(20)
p8_v |> (\(x) x^2)() |> mean() |> sqrt() |> round(3)
```

The pipe helps readability when operations are nested.

## 8.2 Pipe with data frames

```{r}
# --- Self-contained: pipe with dplyr ---
set.seed(803)
p8_df <- tibble(x = rnorm(20), y = rnorm(20))

p8_df |>
  filter(x > 0) |>
  mutate(z = x + y) |>
  summarise(mean_z = mean(z))
```

## 8.3 Common pipe mistake

```{r eval=FALSE}
# --- WRONG ---
# p8_df |> filter(x > 0) |> lm(y ~ x)   # lm expects data = first arg, but gets filtered df

# --- RIGHT ---
p8_df |> filter(x > 0) |> (\(d) lm(y ~ x, data = d))()
```

**Rule:** Pipe passes left side as **first argument**. If you need a different argument, wrap the function in an anonymous function `\(x) ...` (R ≥ 4.1) or `function(x) ...`.

---

# Part 9 — Functions

## 9.1 Calling functions

```{r}
# --- Self-contained: calling functions ---
mean(c(1, 2, 3))                     # 2
mean(x = c(1, 2, 3))                 # same, named arg
round(3.14159, digits = 2)           # 3.14
sample(1:10, size = 3, replace = TRUE)
```

**Named arguments** can be in any order. **Positional arguments** must be in order.

## 9.2 Common math/stat functions

```{r}
# --- Self-contained: common functions ---
p9_x <- c(3, 1, 4, 1, 5, 9, 2, 6)

mean(p9_x)
median(p9_x)
sd(p9_x)          # standard deviation
var(p9_x)         # variance
min(p9_x)
max(p9_x)
range(p9_x)       # min and max
sum(p9_x)
prod(p9_x)
length(p9_x)
quantile(p9_x, c(0.25, 0.5, 0.75))
summary(p9_x)
```

## 9.3 Writing your own function

```{r}
# --- Self-contained: user function ---
p9_add <- function(a, b) {
  a + b
}
p9_add(2, 3)

p9_square <- function(x) x^2
p9_square(4)

# With default argument
p9_power <- function(x, p = 2) x^p
p9_power(3)       # 9
p9_power(3, 3)    # 27

# Returning multiple values (list)
p9_stats <- function(x) {
  list(mean = mean(x), sd = sd(x), n = length(x))
}
p9_stats(p9_x)
```

## 9.4 Anonymous functions

```{r}
# --- Self-contained: anonymous functions ---
# R >= 4.1
p9_result1 <- (\(x) x^2)(5)
p9_result1

# Classic
p9_result2 <- (function(x) x^2)(5)
p9_result2

# With sapply
sapply(1:3, \(x) x^2)
```

## 9.5 `...` (dots)

Pass extra arguments through:

```{r}
# --- Self-contained: dots ---
p9_mean_safe <- function(x, ...) {
  mean(x, ...)
}
p9_mean_safe(c(1, NA, 3), na.rm = TRUE)
```

## 9.6 Scope

```{r}
# --- Self-contained: scope ---
p9_global <- 10

p9_test <- function() {
  p9_local <- 5
  p9_global + p9_local   # can read global
}
p9_test()
# p9_local   # ERROR: not found (it's local to p9_test)
```

**Rule:** Functions can read globals but cannot modify them without `<<-`.

---

# Part 10 — Control Flow

## 10.1 if / else

```{r}
# --- Self-contained: if/else ---
p10_x <- 7

if (p10_x > 5) {
  p10_result <- "big"
} else if (p10_x > 2) {
  p10_result <- "medium"
} else {
  p10_result <- "small"
}
p10_result
```

**Vectorized version:** `ifelse(test, yes, no)`.

```{r}
# --- Self-contained: ifelse ---
p10_v <- c(-1, 0, 1, 2)
ifelse(p10_v > 0, "pos", "non-pos")
```

**Note:** `ifelse` evaluates both branches. For type-stable, use `dplyr::if_else()`.

```{r}
# --- Self-contained: dplyr::if_else ---
p10_v2 <- c(-1, 0, 1, 2)
dplyr::if_else(p10_v2 > 0, "pos", "non-pos")
```

## 10.2 for loop

```{r}
# --- Self-contained: for loop ---
p10_sum <- 0
for (i in 1:5) {
  p10_sum <- p10_sum + i
}
p10_sum

# Loop over a vector
p10_fruits <- c("apple", "banana", "cherry")
for (f in p10_fruits) {
  cat("Fruit:", f, "\n")
}
```

## 10.3 while loop

```{r}
# --- Self-contained: while loop ---
p10_i <- 1
while (p10_i <= 5) {
  cat("i =", p10_i, "\n")
  p10_i <- p10_i + 1
}
```

## 10.4 break and next

```{r}
# --- Self-contained: break and next ---
for (i in 1:10) {
  if (i == 3) next    # skip 3
  if (i == 7) break   # stop at 7
  cat(i, " ")
}
```

## 10.5 Vectorization (preferred over loops)

```{r}
# --- Self-contained: vectorization beats loops ---
set.seed(1001)
p10_v <- rnorm(1e6)

# Bad: loop
p10_slow <- numeric(length(p10_v))
for (i in seq_along(p10_v)) p10_slow[i] <- p10_v[i] * 2

# Good: vectorized
p10_fast <- p10_v * 2

identical(p10_slow, p10_fast)
```

## 10.6 The apply family

```{r}
# --- Self-contained: apply family ---
p10_mat <- matrix(1:12, nrow = 3)

# apply: matrix -> vector
apply(p10_mat, 1, sum)   # row sums
apply(p10_mat, 2, sum)   # column sums

# lapply: list/vector -> list
p10_list <- list(a = 1:3, b = 4:6, c = 7:9)
lapply(p10_list, sum)

# sapply: list/vector -> vector (when possible)
sapply(p10_list, sum)

# vapply: like sapply but type-safe
vapply(p10_list, sum, numeric(1))

# tapply: split-apply-combine on groups
p10_groups <- c("A", "A", "B", "B", "A")
p10_values <- c(1, 2, 3, 4, 5)
tapply(p10_values, p10_groups, mean)

# mapply: multivariate apply
mapply(function(x, y) x + y, 1:3, 4:6)
```

---

# Part 11 — Strings

## 11.1 Basics

```{r}
# --- Self-contained: strings ---
p11_s <- "Hello, World!"

nchar(p11_s)             # 13
toupper(p11_s)
tolower(p11_s)
substr(p11_s, 1, 5)      # "Hello"
substring(p11_s, 8)      # "World!"
strsplit(p11_s, ", ")[[1]]   # c("Hello", "World!")

paste("a", "b", "c")             # "a b c"
paste0("a", "b", "c")            # "abc"
paste(c("a", "b"), 1:2, sep = "-")   # "a-1" "b-2"
```

## 11.2 Pattern matching

```{r}
# --- Self-contained: pattern matching ---
p11_s <- c("apple", "banana", "cherry", "apricot")

grep("ap", p11_s)              # 1 4 (indices)
grep("ap", p11_s, value = TRUE) # "apple" "apricot"
grepl("ap", p11_s)             # TRUE FALSE FALSE TRUE

sub("a", "A", p11_s)           # replace first
gsub("a", "A", p11_s)          # replace all
```

## 11.3 stringr (tidyverse)

```{r}
# --- Self-contained: stringr ---
p11_s <- c("apple", "banana", "cherry", "apricot")

str_detect(p11_s, "an")        # FALSE TRUE FALSE FALSE
str_replace(p11_s, "a", "A")
str_replace_all(p11_s, "a", "A")
str_sub(p11_s, 1, 3)           # first 3 characters
str_length(p11_s)
str_to_upper(p11_s)
str_trim("  hello  ")
str_pad("hi", 5, side = "left")
str_split("a,b,c", ",")[[1]]
```

## 11.4 Printing strings

```{r}
# --- Self-contained: printing ---
p11_name <- "Parsa"
p11_age  <- 25

cat("My name is", p11_name, "and I am", p11_age, "years old.\n")
print(paste("My name is", p11_name))

# sprintf for formatting
sprintf("Name: %s, Age: %d", p11_name, p11_age)
sprintf("Pi to 2 decimals: %.2f", pi)
```

**`cat`** for messages, **`print`** for objects, **`sprintf`** for formatted output.

## 11.5 Glue

```{r}
# --- Self-contained: glue ---
library(glue)
p11_name <- "Parsa"
p11_age  <- 25
glue("My name is {p11_name} and I am {p11_age} years old.")
```

---

# Part 12 — Factors

## 12.1 Creating factors

```{r}
# --- Self-contained: factors ---
p12_f <- factor(c("low", "high", "medium", "low", "high"))
p12_f
levels(p12_f)     # alphabetical by default
nlevels(p12_f)
```

## 12.2 Setting level order

```{r}
# --- Self-contained: factor levels ---
p12_f <- factor(c("low", "high", "medium", "low", "high"),
                levels = c("low", "medium", "high"))
p12_f
levels(p12_f)
```

## 12.3 Changing the reference level

```{r}
# --- Self-contained: relevel ---
p12_f <- factor(c("A", "B", "C", "A", "B"))
p12_f2 <- relevel(p12_f, ref = "B")
levels(p12_f2)
```

## 12.4 Table and counts

```{r}
# --- Self-contained: factor tables ---
p12_f <- factor(c("low", "high", "medium", "low", "high"))

table(p12_f)
prop.table(table(p12_f))
```

## 12.5 Convert to/from factor

```{r}
# --- Self-contained: factor conversion ---
p12_chr <- c("a", "b", "a", "c")
p12_f   <- factor(p12_chr)
as.character(p12_f)
as.numeric(p12_f)        # 1 2 1 3 (level codes)
as.integer(p12_f)
```

**Warning:** `as.numeric(factor)` returns level codes, not the original numbers. To get original: `as.numeric(as.character(f))`.

## 12.6 forcats (tidyverse)

```{r}
# --- Self-contained: forcats ---
library(forcats)
p12_f <- factor(c("low", "high", "medium", "low", "high"))

fct_relevel(p12_f, "high")
fct_reorder(p12_f, c(1, 3, 2, 1, 3))   # reorder by a numeric
fct_infreq(p12_f)                       # most frequent first
fct_lump(p12_f, n = 2)                  # keep top 2, lump rest
```

---

# Part 13 — Dates

## 13.1 Creating dates

```{r}
# --- Self-contained: dates ---
p13_d <- as.Date("2024-01-15")
p13_d
class(p13_d)

p13_d2 <- as.Date("15/01/2024", format = "%d/%m/%Y")
p13_d2

p13_dt <- as.POSIXct("2024-01-15 10:30:00", tz = "UTC")
p13_dt
```

## 13.2 Components

```{r}
# --- Self-contained: date components ---
p13_d <- as.Date("2024-01-15")

format(p13_d, "%Y")   # "2024"
format(p13_d, "%m")   # "01"
format(p13_d, "%d")   # "15"
format(p13_d, "%A")   # "Monday"
format(p13_d, "%B")   # "January"

weekdays(p13_d)
months(p13_d)
quarters(p13_d)
```

## 13.3 Arithmetic

```{r}
# --- Self-contained: date arithmetic ---
p13_d <- as.Date("2024-01-15")

p13_d + 7                # one week later
p13_d - 30               # 30 days earlier
p13_d2 <- as.Date("2024-02-15")
p13_d2 - p13_d           # Time difference of 31 days
as.numeric(p13_d2 - p13_d)
```

## 13.4 Sequences

```{r}
# --- Self-contained: date sequences ---
p13_seq <- seq(as.Date("2024-01-01"), by = "month", length.out = 6)
p13_seq

seq(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "quarter")
```

## 13.5 lubridate

```{r}
# --- Self-contained: lubridate ---
library(lubridate)
p13_d <- ymd("2024-01-15")
p13_d
year(p13_d)
month(p13_d)
day(p13_d)
wday(p13_d, label = TRUE)
p13_d + weeks(2)
p13_d %m+% months(1)     # month arithmetic with rollover
```

---

# Part 14 — Missing Values

## 14.1 Detecting

```{r}
# --- Self-contained: missing values ---
p14_v <- c(1, NA, 3, NA, 5)

is.na(p14_v)          # FALSE TRUE FALSE TRUE FALSE
anyNA(p14_v)          # TRUE
sum(is.na(p14_v))     # 2
mean(is.na(p14_v))    # 0.4 (fraction missing)
complete.cases(p14_v) # TRUE FALSE TRUE FALSE TRUE
```

## 14.2 Removing

```{r}
# --- Self-contained: remove NAs ---
p14_v <- c(1, NA, 3, NA, 5)
p14_v[!is.na(p14_v)]     # 1 3 5
na.omit(p14_v)           # same, with attribute
```

## 14.3 Functions with `na.rm`

```{r}
# --- Self-contained: na.rm ---
p14_v <- c(1, NA, 3, NA, 5)
mean(p14_v)                # NA
mean(p14_v, na.rm = TRUE)  # 3
sum(p14_v, na.rm = TRUE)
sd(p14_v, na.rm = TRUE)
```

## 14.4 Filling / replacing

```{r}
# --- Self-contained: fill NAs ---
p14_v <- c(1, NA, 3, NA, 5)

# Replace with zero
p14_v[is.na(p14_v)] <- 0

# Replace with mean
p14_v2 <- c(1, NA, 3, NA, 5)
p14_v2[is.na(p14_v2)] <- mean(p14_v2, na.rm = TRUE)

# Forward fill (last observation carried forward)
p14_v3 <- c(1, NA, NA, 4, NA)
p14_filled <- zoo::na.locf(p14_v3)

# Interpolation
p14_interp <- approx(seq_along(p14_v3), p14_v3, xout = seq_along(p14_v3))$y
```

## 14.5 NA in data frames

```{r}
# --- Self-contained: NA in tibbles ---
p14_df <- tibble(
  x = c(1, NA, 3),
  y = c("a", "b", NA)
)
p14_df

tidyr::drop_na(p14_df)          # drop rows with any NA
tidyr::replace_na(p14_df, list(x = 0, y = "unknown"))
```

## 14.6 NA vs NULL vs NaN

| Value | Meaning | In vector |
|---|---|---|
| `NA` | missing | stays in vector |
| `NULL` | empty / no value | disappears |
| `NaN` | not a number (0/0) | stays |
| `Inf` | infinity | stays |

```{r}
# --- Self-contained: NA vs NULL vs NaN ---
length(c(1, NA, 3))    # 3
length(c(1, NULL, 3))  # 2
length(c(1, NaN, 3))   # 3
```

---

# Part 15 — Reading and Writing Files

## 15.1 CSV

```{r eval=FALSE}
# --- Read ---
df <- read.csv("data.csv")
df <- read_csv("data.csv")          # tidyverse, faster
df <- read.csv("data.csv", stringsAsFactors = FALSE)

# --- Write ---
write.csv(df, "out.csv", row.names = FALSE)
write_csv(df, "out.csv")
```

## 15.2 Excel

```{r eval=FALSE}
library(readxl)
df <- read_excel("data.xlsx", sheet = 1)
```

## 15.3 RDS (R's native format — preserves types)

```{r eval=FALSE}
saveRDS(df, "data.rds")
df <- readRDS("data.rds")
```

## 15.4 RData (multiple objects)

```{r eval=FALSE}
save(df, fit, file = "workspace.RData")
load("workspace.RData")
```

## 15.5 Text files

```{r eval=FALSE}
writeLines(c("line 1", "line 2"), "out.txt")
readLines("out.txt")
```

## 15.6 From clipboard

```{r eval=FALSE}
df <- read.table("clipboard", header = TRUE)   # Windows
df <- read.table(pipe("pbpaste"), header = TRUE)  # macOS
```

## 15.7 Inspect files

```{r}
# --- Self-contained: inspect a data frame ---
set.seed(1501)
p15_df <- tibble(x = rnorm(5), y = letters[1:5])

head(p15_df)          # first 6 rows
head(p15_df, 3)       # first 3 rows
tail(p15_df)
str(p15_df)           # structure
glimpse(p15_df)       # tidyverse structure
summary(p15_df)
dim(p15_df)
names(p15_df)
```

---

# Part 16 — Data Manipulation with dplyr

## 16.1 The five verbs

```{r}
# --- Self-contained: dplyr verbs ---
library(dplyr)
set.seed(1601)
p16_df <- tibble(
  id    = 1:6,
  name  = c("A", "B", "C", "D", "E", "F"),
  score = c(90, 85, 88, 92, 79, 95),
  group = c("X", "X", "Y", "Y", "X", "Y")
)

# filter: keep rows
p16_df |> filter(score > 85)

# select: keep columns
p16_df |> select(name, score)

# mutate: add/modify columns
p16_df |> mutate(passed = score >= 85)

# arrange: sort rows
p16_df |> arrange(desc(score))

# summarise: reduce to one row
p16_df |> summarise(mean_score = mean(score), n = n())
```

## 16.2 filter

```{r}
# --- Self-contained: filter ---
set.seed(1602)
p16_df <- tibble(id = 1:6, score = c(90, 85, 88, 92, 79, 95),
                 group = c("X", "X", "Y", "Y", "X", "Y"))

p16_df |> filter(score > 85)
p16_df |> filter(score > 85 & group == "X")
p16_df |> filter(score > 85 | group == "X")
p16_df |> filter(group %in% c("X", "Y"))
p16_df |> filter(!is.na(score))
```

## 16.3 select

```{r}
# --- Self-contained: select ---
set.seed(1603)
p16_df <- tibble(id = 1:3, name = c("A","B","C"),
                 score = c(90,85,88), group = c("X","X","Y"))

p16_df |> select(name, score)
p16_df |> select(-id)                    # drop
p16_df |> select(starts_with("s"))
p16_df |> select(where(is.numeric))
p16_df |> select(name:score)             # range
p16_df |> relocate(score, .before = name)
```

## 16.4 mutate

```{r}
# --- Self-contained: mutate ---
set.seed(1604)
p16_df <- tibble(x = c(1, 2, 3), y = c(4, 5, 6))

p16_df |> mutate(z = x + y)
p16_df |> mutate(z = x + y, w = z * 2)      # uses z just created
p16_df |> mutate(x2 = x^2, x_cat = if_else(x > 1, "big", "small"))
p16_df |> transmute(z = x + y)              # keep only new cols
```

## 16.5 arrange

```{r}
# --- Self-contained: arrange ---
set.seed(1605)
p16_df <- tibble(name = c("A","B","C","D"), score = c(90,85,88,92),
                 group = c("X","X","Y","Y"))

p16_df |> arrange(score)
p16_df |> arrange(desc(score))
p16_df |> arrange(group, desc(score))
```

## 16.6 summarise + group_by

```{r}
# --- Self-contained: group_by + summarise ---
set.seed(1606)
p16_df <- tibble(group = c("X","X","Y","Y","X","Y"),
                 score = c(90,85,88,92,79,95))

p16_df |>
  group_by(group) |>
  summarise(mean_score = mean(score), n = n(), .groups = "drop")

# Multiple summaries
p16_df |>
  group_by(group) |>
  summarise(
    mean = mean(score),
    sd   = sd(score),
    min  = min(score),
    max  = max(score),
    n    = n(),
    .groups = "drop"
  )
```

## 16.7 count and tally

```{r}
# --- Self-contained: count ---
set.seed(1607)
p16_df <- tibble(group = c("X","X","Y","Y","X","Y"))

p16_df |> count(group)
p16_df |> count(group, sort = TRUE)
```

## 16.8 slice and distinct

```{r}
# --- Self-contained: slice and distinct ---
set.seed(1608)
p16_df <- tibble(x = c(1, 2, 2, 3, 3, 3))

p16_df |> distinct()
p16_df |> distinct(x)
p16_df |> slice(1:2)
p16_df |> slice_head(n = 2)
p16_df |> slice_max(x, n = 2)
p16_df |> slice_min(x, n = 2)
p16_df |> slice_sample(n = 2)
```

## 16.9 pull and rename

```{r}
# --- Self-contained: pull and rename ---
set.seed(1609)
p16_df <- tibble(id = 1:3, score = c(90, 85, 88))

p16_df |> pull(score)              # extract a column as vector
p16_df |> rename(points = score)   # rename columns
```

## 16.10 across (apply to many columns)

```{r}
# --- Self-contained: across ---
set.seed(1610)
p16_df <- tibble(a = rnorm(10), b = rnorm(10), g = rep(c("X","Y"), 5))

p16_df |>
  group_by(g) |>
  summarise(across(where(is.numeric), mean), .groups = "drop")

p16_df |> mutate(across(where(is.numeric), ~ .x * 2))
```

## 16.11 Summary of dplyr verbs

| Verb | Purpose |
|---|---|
| `filter()` | keep rows matching condition |
| `select()` | keep columns |
| `mutate()` | add / modify columns |
| `arrange()` | sort rows |
| `summarise()` | reduce to summary rows |
| `group_by()` | group for summary |
| `count()` | count rows by group |
| `distinct()` | unique rows |
| `slice()` | pick rows by position |
| `rename()` | rename columns |
| `relocate()` | move columns |
| `pull()` | extract a column as vector |
| `across()` | apply function to many columns |

---

# Part 17 — Reshaping with tidyr

## 17.1 pivot_longer (wide → long)

```{r}
# --- Self-contained: pivot_longer ---
set.seed(1701)
p17_wide <- tibble(
  id  = 1:3,
  q1  = c(10, 11, 12),
  q2  = c(20, 21, 22),
  q3  = c(30, 31, 32)
)
p17_wide

p17_long <- p17_wide |>
  pivot_longer(cols = starts_with("q"),
               names_to = "quarter",
               values_to = "value")
p17_long
```

## 17.2 pivot_wider (long → wide)

```{r}
# --- Self-contained: pivot_wider ---
p17_long <- tibble(
  id      = rep(1:3, each = 3),
  quarter = rep(c("q1", "q2", "q3"), 3),
  value   = c(10, 20, 30, 11, 21, 31, 12, 22, 32)
)
p17_long

p17_wide <- p17_long |>
  pivot_wider(names_from = quarter, values_from = value)
p17_wide
```

## 17.3 separate and unite

```{r}
# --- Self-contained: separate and unite ---
p17_df <- tibble(full = c("A-1", "B-2", "C-3"))

p17_df |> separate(full, into = c("letter", "number"), sep = "-")

p17_df2 <- tibble(letter = c("A","B"), number = c(1, 2))
p17_df2 |> unite("full", letter, number, sep = "-")
```

## 17.4 Complete and fill

```{r}
# --- Self-contained: complete ---
p17_df <- tibble(
  year = c(2020, 2021, 2023),
  val  = c(1, 2, 4)
)
p17_df |> tidyr::complete(year = 2020:2023)

# Fill missing values from previous row
p17_df2 <- tibble(x = c(1, NA, NA, 4, NA))
p17_df2 |> fill(x, .direction = "down")
```

---

# Part 18 — Joins

## 18.1 The four main joins

```{r}
# --- Self-contained: joins ---
p18_a <- tibble(id = c(1, 2, 3), x = c("a","b","c"))
p18_b <- tibble(id = c(2, 3, 4), y = c("B","C","D"))

# Inner: only matching rows
p18_a |> inner_join(p18_b, by = "id")

# Left: all rows in a
p18_a |> left_join(p18_b, by = "id")

# Right: all rows in b
p18_a |> right_join(p18_b, by = "id")

# Full: all rows in both
p18_a |> full_join(p18_b, by = "id")
```

## 18.2 Anti and semi joins

```{r}
# --- Self-contained: anti and semi ---
p18_a <- tibble(id = c(1, 2, 3), x = c("a","b","c"))
p18_b <- tibble(id = c(2, 3, 4), y = c("B","C","D"))

# Anti: rows in a NOT in b
p18_a |> anti_join(p18_b, by = "id")

# Semi: rows in a that ARE in b (only a's columns)
p18_a |> semi_join(p18_b, by = "id")
```

## 18.3 Joining by different names

```{r}
# --- Self-contained: join with different key names ---
p18_a <- tibble(id_a = c(1, 2, 3), x = c("a","b","c"))
p18_b <- tibble(id_b = c(2, 3, 4), y = c("B","C","D"))

p18_a |> inner_join(p18_b, by = c("id_a" = "id_b"))
```

## 18.4 Join summary

| Join | Keeps |
|---|---|
| `inner_join()` | matches only |
| `left_join()` | all rows from left |
| `right_join()` | all rows from right |
| `full_join()` | all rows from both |
| `anti_join()` | rows in left NOT in right |
| `semi_join()` | rows in left that match right |

---

# Part 19 — purrr (Functional Programming)

## 19.1 map

```{r}
# --- Self-contained: map ---
library(purrr)
p19_list <- list(a = 1:3, b = 4:6, c = 7:9)

map(p19_list, sum)          # list
map_dbl(p19_list, sum)      # numeric vector
map_chr(p19_list, ~ paste(.x, collapse = "-"))
map_lgl(p19_list, ~ any(.x > 5))
```

## 19.2 map over data frame columns

```{r}
# --- Self-contained: map over columns ---
set.seed(1902)
p19_df <- tibble(a = rnorm(10), b = rnorm(10), c = rnorm(10))

map_dbl(p19_df, mean)
map_dbl(p19_df, sd)
```

## 19.3 map2 and pmap

```{r}
# --- Self-contained: map2 and pmap ---
map2_dbl(1:3, 4:6, ~ .x + .y)

pmap_dbl(list(1:3, 4:6, 7:9), ~ ..1 + ..2 + ..3)
```

## 19.4 walk (for side effects)

```{r}
# --- Self-contained: walk ---
p19_list <- list(1, 2, 3)
walk(p19_list, ~ cat("Value:", .x, "\n"))
```

---

# Part 20 — Base R Plots

## 20.1 Scatter plot

```{r}
# --- Self-contained: base scatter ---
set.seed(2001)
p20_x <- rnorm(50)
p20_y <- 2 + 3 * p20_x + rnorm(50)

plot(p20_x, p20_y,
     main = "Scatter plot",
     xlab = "X", ylab = "Y",
     col = "steelblue", pch = 19)
abline(lm(p20_y ~ p20_x), col = "red", lwd = 2)
```

## 20.2 Line plot

```{r}
# --- Self-contained: base line plot ---
set.seed(2002)
p20_y <- cumsum(rnorm(50))
plot(p20_y, type = "l", main = "Line plot", col = "blue", lwd = 2)
```

## 20.3 Histogram

```{r}
# --- Self-contained: histogram ---
set.seed(2003)
p20_v <- rnorm(200)
hist(p20_v, breaks = 20, col = "lightblue",
     main = "Histogram", xlab = "Value")
```

## 20.4 Boxplot

```{r}
# --- Self-contained: boxplot ---
set.seed(2004)
p20_df <- data.frame(
  group = rep(c("A","B","C"), each = 20),
  value = c(rnorm(20), rnorm(20, 2), rnorm(20, 4))
)
boxplot(value ~ group, data = p20_df, col = c("red","green","blue"))
```

## 20.5 Bar plot

```{r}
# --- Self-contained: barplot ---
p20_counts <- c(A = 10, B = 20, C = 15)
barplot(p20_counts, col = "steelblue", main = "Counts")
```

## 20.6 Layout: multiple plots

```{r}
# --- Self-contained: multiple plots ---
par(mfrow = c(1, 2))     # 1 row, 2 cols
plot(1:10, main = "Left")
plot(10:1, main = "Right")
par(mfrow = c(1, 1))     # reset
```

## 20.7 Save a plot

```{r eval=FALSE}
png("plot.png", width = 800, height = 600)
plot(1:10)
dev.off()

pdf("plot.pdf", width = 8, height = 6)
plot(1:10)
dev.off()
```

---

# Part 21 — ggplot2

## 21.1 Grammar of graphics

Every ggplot has:

1. **Data:** `ggplot(data = df)`
2. **Aesthetics:** `aes(x = x, y = y, color = group)`
3. **Geometries:** `+ geom_point()`, `+ geom_line()`, etc.
4. **Facets:** `+ facet_wrap(~ group)`
5. **Scales:** `+ scale_color_manual(...)`
6. **Theme:** `+ theme_minimal()`

## 21.2 Scatter plot

```{r}
# --- Self-contained: ggplot scatter ---
set.seed(2101)
p21_df <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50),
                 group = sample(c("A","B"), 50, replace = TRUE))

ggplot(p21_df, aes(x = x, y = y, color = group)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "Scatter", x = "X", y = "Y") +
  theme_minimal()
```

## 21.3 Line plot

```{r}
# --- Self-contained: ggplot line ---
set.seed(2102)
p21_df <- tibble(t = 1:50, y = cumsum(rnorm(50)))

ggplot(p21_df, aes(t, y)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(title = "Line", x = "Time", y = "Value") +
  theme_minimal()
```

## 21.4 Histogram

```{r}
# --- Self-contained: ggplot histogram ---
set.seed(2103)
p21_df <- tibble(x = rnorm(200))

ggplot(p21_df, aes(x)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  theme_minimal()
```

## 21.5 Boxplot

```{r}
# --- Self-contained: ggplot boxplot ---
set.seed(2104)
p21_df <- tibble(
  group = rep(c("A","B","C"), each = 20),
  value = c(rnorm(20), rnorm(20, 2), rnorm(20, 4))
)

ggplot(p21_df, aes(group, value, fill = group)) +
  geom_boxplot() +
  theme_minimal()
```

## 21.6 Bar plot

```{r}
# --- Self-contained: ggplot bar ---
p21_df <- tibble(group = c("A","B","C"), count = c(10, 20, 15))

ggplot(p21_df, aes(group, count)) +
  geom_col(fill = "steelblue") +
  theme_minimal()
```

## 21.7 Facets

```{r}
# --- Self-contained: facets ---
set.seed(2107)
p21_df <- tibble(
  x = rnorm(100),
  y = rnorm(100),
  group = rep(c("A","B"), 50)
)

ggplot(p21_df, aes(x, y)) +
  geom_point() +
  facet_wrap(~ group) +
  theme_minimal()
```

## 21.8 Saving

```{r eval=FALSE}
ggsave("plot.png", width = 6, height = 4, dpi = 300)
ggsave("plot.pdf", width = 6, height = 4)
```

## 21.9 Common geoms

| Geom | Use for |
|---|---|
| `geom_point()` | scatter |
| `geom_line()` | line |
| `geom_bar()` / `geom_col()` | bar |
| `geom_histogram()` | histogram |
| `geom_boxplot()` | boxplot |
| `geom_smooth(method = "lm")` | regression line |
| `geom_text()` | labels |
| `geom_hline(yintercept = ...)` | horizontal line |
| `geom_vline(xintercept = ...)` | vertical line |
| `geom_ribbon()` | confidence bands |

## 21.10 Common theme tweaks

```{r eval=FALSE}
+ theme_minimal()
+ theme_bw()
+ theme_classic()
+ labs(title = "...", x = "...", y = "...")
+ scale_color_manual(values = c("red", "blue"))
+ scale_x_continuous(limits = c(0, 10))
+ coord_flip()
```

---

# Part 22 — Statistical Functions

## 22.1 Descriptive

```{r}
# --- Self-contained: descriptive stats ---
set.seed(2201)
p22_x <- rnorm(100)

mean(p22_x)
median(p22_x)
sd(p22_x)
var(p22_x)
range(p22_x)
quantile(p22_x, c(0.025, 0.5, 0.975))
IQR(p22_x)
summary(p22_x)
```

## 22.2 Distributions

For each distribution, R has 4 functions: `d` (density), `p` (CDF), `q` (quantile), `r` (random).

| Distribution | Functions |
|---|---|
| Normal | `dnorm`, `pnorm`, `qnorm`, `rnorm` |
| Binomial | `dbinom`, `pbinom`, `qbinom`, `rbinom` |
| Poisson | `dpois`, `ppois`, `qpois`, `rpois` |
| Uniform | `dunif`, `punif`, `qunif`, `runif` |
| Exponential | `dexp`, `pexp`, `qexp`, `rexp` |
| t | `dt`, `pt`, `qt`, `rt` |
| Chi-square | `dchisq`, `pchisq`, `qchisq`, `rchisq` |
| F | `df`, `pf`, `qf`, `rf` |

```{r}
# --- Self-contained: distribution functions ---
# Density at x = 0 for N(0, 1)
dnorm(0)

# CDF at x = 1.96
pnorm(1.96)

# 97.5% quantile of N(0, 1)
qnorm(0.975)

# 5 random draws from N(0, 1)
set.seed(2202)
rnorm(5)
```

## 22.3 Hypothesis tests

```{r}
# --- Self-contained: hypothesis tests ---
set.seed(2203)
p22_x <- rnorm(30)

# One-sample t-test
t.test(p22_x, mu = 0)

# Two-sample t-test
p22_a <- rnorm(30, mean = 1)
p22_b <- rnorm(30, mean = 0)
t.test(p22_a, p22_b)

# Paired t-test
t.test(p22_a, p22_b, paired = TRUE)

# Correlation test
cor.test(p22_x, rnorm(30))

# Chi-square test
chisq.test(matrix(c(10, 20, 30, 40), nrow = 2))
```

## 22.4 Regression

```{r}
# --- Self-contained: regression ---
set.seed(2204)
p22_df <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50))

p22_fit <- lm(y ~ x, data = p22_df)
summary(p22_fit)
coef(p22_fit)
confint(p22_fit)
predict(p22_fit, newdata = data.frame(x = 1))
```

## 22.5 Correlation

```{r}
# --- Self-contained: correlation ---
set.seed(2205)
p22_df <- tibble(x = rnorm(50), y = rnorm(50))

cor(p22_df$x, p22_df$y)                # Pearson
cor(p22_df$x, p22_df$y, method = "spearman")
cor(p22_df$x, p22_df$y, method = "kendall")
cor(p22_df)                             # correlation matrix
```

## 22.6 Hypothesis test decision table

| Question | Function |
|---|---|
| Compare one mean to a value | `t.test(x, mu = 0)` |
| Compare two independent means | `t.test(x, y)` |
| Compare paired means | `t.test(x, y, paired = TRUE)` |
| Compare >2 means | `aov(y ~ g)` or `anova(lm(y ~ g))` |
| Test normality | `shapiro.test(x)` |
| Test variance equality | `var.test(x, y)` |
| Test a proportion | `prop.test(x, n, p)` |
| Test association (categorical) | `chisq.test(table)` |
| Correlation | `cor.test(x, y)` |

---

# Part 23 — Random Numbers and Sampling

## 23.1 Setting the seed

```{r}
# --- Self-contained: set.seed ---
set.seed(42)
rnorm(3)

set.seed(42)
rnorm(3)   # same values
```

**Rule:** Always `set.seed()` before simulation or sampling for reproducibility.

## 23.2 Common generators

```{r}
# --- Self-contained: generators ---
set.seed(2302)
rnorm(5)            # N(0, 1)
rnorm(5, mean = 10, sd = 2)
runif(5)            # U(0, 1)
runif(5, min = 0, max = 100)
rbinom(5, size = 10, prob = 0.3)
rpois(5, lambda = 3)
rexp(5, rate = 1)
rt(5, df = 5)
```

## 23.3 Sampling

```{r}
# --- Self-contained: sampling ---
set.seed(2303)
p23_v <- 1:10

sample(p23_v, size = 3)                 # without replacement
sample(p23_v, size = 3, replace = TRUE) # with replacement
sample(p23_v)                            # permutation
sample(c("H", "T"), size = 10, replace = TRUE, prob = c(0.5, 0.5))
```

## 23.4 Split into train/test

```{r}
# --- Self-contained: train/test split ---
set.seed(2304)
p23_df <- tibble(x = rnorm(100), y = rnorm(100))

p23_idx  <- sample(1:nrow(p23_df), size = 0.8 * nrow(p23_df))
p23_train <- p23_df[p23_idx, ]
p23_test  <- p23_df[-p23_idx, ]

nrow(p23_train); nrow(p23_test)
```

---

# Part 24 — Vectorization vs Loops

## 24.1 Why vectorize

R is optimized for vector operations. Loops are slow.

```{r}
# --- Self-contained: timing ---
set.seed(2401)
p24_v <- rnorm(1e6)

# Slow: loop
system.time({
  p24_out <- numeric(length(p24_v))
  for (i in seq_along(p24_v)) p24_out[i] <- p24_v[i]^2
})

# Fast: vectorized
system.time({
  p24_out2 <- p24_v^2
})
```

## 24.2 Vectorized operations

```{r}
# --- Self-contained: vectorized operations ---
p24_a <- c(1, 2, 3)
p24_b <- c(10, 20, 30)

p24_a + p24_b           # element-wise
p24_a * p24_b
p24_a / p24_b
p24_a ^ 2
sqrt(p24_a)
log(p24_a)
p24_a > 1.5
```

## 24.3 Recycling

R recycles shorter vectors:

```{r}
# --- Self-contained: recycling ---
c(1, 2, 3) + c(10, 20)   # warning: longer object length not multiple
c(1, 2, 3, 4) + c(10, 20)  # 11 22 13 24
```

## 24.4 When to use loops

Loops are fine when:
- You need to update values based on previous iterations (e.g., recursive simulation).
- The computation is inherently sequential.
- You use `lapply` / `purrr::map` instead.

---

# Part 25 — Debugging

## 25.1 Errors vs warnings vs messages

- **Error:** code stops. E.g., `log("a")`.
- **Warning:** code runs but something is off. E.g., `as.numeric("abc")`.
- **Message:** informational. E.g., package loading.

```{r eval=FALSE}
log("a")            # Error
as.numeric("abc")   # Warning: NAs introduced by coercion
```

## 25.2 Reading error messages

Example: `Error in x[i] : subscript out of bounds`.

- Function: `x[i]`
- Problem: `i` is larger than `length(x)` or negative invalid.

**Check:**
```{r}
# --- Self-contained: check indices ---
p25_x <- 1:5
length(p25_x)
p25_x[10]   # NA
# p25_x[[10]] # Error
```

## 25.3 Debugging tools

```{r eval=FALSE}
# Print statements
print(x); cat("x =", x, "\n"); str(x); class(x); dim(x)

# Traceback after error
traceback()

# Debug a function step-by-step
debug(lm)
lm(y ~ x, data = df)
undebug(lm)

# Browser inside a function
# Add browser() where you want to pause

# try / tryCatch
try(log("a"))
result <- tryCatch(log("a"), error = function(e) NA)
```

## 25.4 Common checks

```{r}
# --- Self-contained: inspect ---
p25_df <- tibble(x = c(1, 2, NA), y = c("a", "b", "c"))

str(p25_df)
class(p25_df)
dim(p25_df)
names(p25_df)
head(p25_df)
summary(p25_df)
sapply(p25_df, class)
anyNA(p25_df)
```

## 25.5 `stopifnot`

```{r eval=FALSE}
my_fun <- function(x) {
  stopifnot(is.numeric(x), length(x) > 0)
  mean(x)
}
my_fun(c(1, 2, 3))   # works
my_fun("a")          # error
```

---

# Part 26 — Common Errors and Fixes

| Error | Cause | Fix |
|---|---|---|
| `object 'x' not found` | Variable not defined or misspelled | Check spelling, define `x` first |
| `could not find function "f"` | Package not loaded or typo | `library(pkg)` or fix spelling |
| `unexpected symbol` | Syntax error (missing `)`, `,`) | Check parentheses and commas |
| `subscript out of bounds` | Index beyond vector length | Check `length()`, use `[` not `[[` for out-of-range |
| `non-numeric argument to binary operator` | Wrong type | `as.numeric()` or check `class()` |
| `NAs introduced by coercion` | Converting non-numeric string | Check the string, handle `NA` |
| `replacement has X rows, data has Y` | Length mismatch in assignment | Match lengths |
| `object of type 'closure' is not subsettable` | You called a function like a variable | Don't name variables the same as functions |
| `cannot open file` | Wrong path | Use absolute path or `setwd()` |
| `there is no package called 'X'` | Not installed | `install.packages("X")` |
| `contrasts can be applied only to factors with 2 or more levels` | Factor with 1 level | Check `unique(df$var)`, use `factor()` |
| `model frame and formula mismatch` | Removed rows in one variable | Use `na.omit()` or `drop_na()` first |
| `singular fit` / `system is computationally singular` | Perfect collinearity | Remove one predictor |
| `longer object length is not a multiple of shorter object length` | Recycling | Match lengths or use `rep()` |
| `$ operator is invalid for atomic vectors` | Using `$` on a vector | Use `[` or `[[` |

## 26.1 Quick fixes

```{r}
# --- Self-contained: fix patterns ---
# Check types
p26_x <- c(1, 2, 3)
class(p26_x)

# Rename variable to avoid function clash
p26_mean <- c(1, 2, 3)   # good
# mean <- c(1, 2, 3)     # bad: overwrites function

# Use [[ not [ for lists when extracting elements
p26_l <- list(a = 1:3)
p26_l[["a"]]     # correct: returns vector
# p26_l[["b"]]   # Error: subscript out of bounds
```

---

# Part 27 — RStudio Shortcuts

| Shortcut | Action |
|---|---|
| `Ctrl+Enter` | Run current line / selection |
| `Ctrl+Shift+Enter` | Run entire script |
| `Ctrl+Alt+R` | Run entire script (alt) |
| `Ctrl+Shift+K` | Knit (RMarkdown) |
| `Ctrl+1` | Focus source editor |
| `Ctrl+2` | Focus console |
| `Alt+-` | Insert `<-` |
| `Ctrl+Shift+M` | Insert `%>%` |
| `Ctrl+Shift+A` | Reformat code |
| `Ctrl+D` | Delete current line |
| `Ctrl+Shift+D` | Duplicate line |
| `Ctrl+L` | Clear console |
| `Ctrl+Shift+C` | Comment/uncomment |
| `Ctrl+F` | Find |
| `Ctrl+Shift+F` | Find in files |
| `F1` | Help for function at cursor |
| `F2` | Go to definition |
| `Tab` | Autocomplete |
| `Ctrl+Up` | Command history |
| `Alt+Shift+K` | Keyboard shortcut help |

---

# Part 28 — Quick Reference Table

| Task | Code |
|---|---|
| Assign | `x <- 5` |
| Sequence | `1:10`, `seq(0, 1, by = 0.1)` |
| Repeat | `rep(1:3, times = 2)` |
| Vector | `c(1, 2, 3)` |
| List | `list(a = 1, b = "x")` |
| Matrix | `matrix(1:9, nrow = 3)` |
| Data frame | `data.frame(x = 1:3, y = 4:6)` |
| Tibble | `tibble(x = 1:3, y = 4:6)` |
| Access column | `df$x` or `df[["x"]]` |
| Access element | `df[1, 2]` or `v[3]` |
| Filter | `df |> filter(x > 5)` |
| Select cols | `df |> select(a, b)` |
| Add column | `df |> mutate(z = x + y)` |
| Sort | `df |> arrange(x)` |
| Group + summarize | `df |> group_by(g) |> summarise(m = mean(x))` |
| Join | `a |> inner_join(b, by = "id")` |
| Reshape long | `pivot_longer(cols, names_to, values_to)` |
| Reshape wide | `pivot_wider(names_from, values_from)` |
| Mean | `mean(x, na.rm = TRUE)` |
| Median | `median(x)` |
| SD | `sd(x)` |
| Quantiles | `quantile(x, c(.25, .75))` |
| T-test | `t.test(x, y)` |
| Correlation | `cor(x, y)` / `cor.test(x, y)` |
| Regression | `lm(y ~ x, data = df)` |
| Predict | `predict(fit, newdata)` |
| Set seed | `set.seed(42)` |
| Sample | `sample(x, size = n, replace = FALSE)` |
| Map | `map(x, f)` / `map_dbl(x, f)` |
| Scatter plot | `ggplot(df, aes(x, y)) + geom_point()` |
| Save plot | `ggsave("plot.png")` |
| Read CSV | `read_csv("file.csv")` |
| Write CSV | `write_csv(df, "file.csv")` |

---

# Part 29 — Theory Q&A (What to Say)

**Q: What's the difference between `<-` and `=`?**
A: Both assign in most contexts, but `<-` is idiomatic and preferred. Use `=` only for function arguments (`mean(x, na.rm = TRUE)`).

**Q: What's the difference between `[`, `[[`, and `$`?**
A: `[` returns a **subset** (vector, list, data frame). `[[` returns the **contents** of a single element (unwrapped). `$` is shorthand for `[[` on a named element.

**Q: What's the difference between a vector and a list?**
A: A vector holds elements of the **same type**. A list can hold elements of **different types** (including other lists).

**Q: What's the difference between `NA` and `NULL`?**
A: `NA` is a missing value placeholder that occupies a position. `NULL` means "no value at all" — assigning to a vector element with `NULL` removes it.

**Q: What's the difference between a factor and a character?**
A: Character is just text. A factor stores categorical data with predefined **levels**, which matters for regression (`lm` treats factors as dummies) and plotting.

**Q: What's the difference between `&` and `&&`?**
A: `&` is vectorized (returns a vector). `&&` is scalar and short-circuits (returns one value, stops at first FALSE).

**Q: Why use the pipe `|>`?**
A: It makes nested function calls readable: instead of `f(g(h(x)))`, write `x |> h() |> g() |> f()`.

**Q: What is vectorization?**
A: Performing an operation on an entire vector at once, instead of element-by-element. It's faster and cleaner than loops in R.

**Q: What is the difference between `apply`, `lapply`, `sapply`, `tapply`, `mapply`?**
A: `apply` works on matrix margins. `lapply` returns a list. `sapply` simplifies to a vector. `tapply` applies by group. `mapply` handles multiple inputs.

**Q: How do you handle missing values?**
A: Detect with `is.na()`, remove with `na.omit()` or `drop_na()`, exclude in functions with `na.rm = TRUE`, or fill with mean/median/`zoo::na.locf()`.

**Q: What's the difference between `data.frame` and `tibble`?**
A: Tibbles print nicely, don't auto-convert strings to factors, don't use row names, and allow list-columns.

**Q: What's the difference between `read.csv` and `read_csv`?**
A: `read.csv` is base R, slower, uses `stringsAsFactors` (defaults changed in R 4.0). `read_csv` is tidyverse, faster, returns a tibble, doesn't convert strings.

---

# Part 30 — The Single Most Important Idea

R is:

1. **Vectorized** — operations apply to whole vectors.
2. **Functional** — functions take inputs, return outputs.
3. **Pipe-friendly** — chains read top-to-bottom.
4. **Type-aware** — every object has a class (`numeric`, `character`, `factor`, `tbl`, ...).
5. **Subsettable** — `[`, `[[`, `$` extract pieces.

If you know these five things and can look up functions, you can do anything in R.

The pipeline for any R task:

1. **Load** the data (`read_csv`, `readRDS`).
2. **Inspect** (`str`, `glimpse`, `summary`, `head`).
3. **Clean** (`filter`, `mutate`, `drop_na`).
4. **Transform** (`group_by`, `summarise`, `pivot_*`).
5. **Analyze** (`lm`, `t.test`, `glm`).
6. **Visualize** (`ggplot` or base `plot`).
7. **Report** (`kable`, `glue`, `sprintf`).

If you can do these seven steps, you can complete any R task.

---

*End of cheat sheet. Good luck in the competition.*
```

---

## What's inside and why

| Part | Covers |
|---|---|
| 0 | Five survival rules |
| 1–2 | Packages, help |
| 3–4 | Variables, assignment, data types (numeric, integer, character, logical, complex) |
| 5 | Vectors, matrices, arrays, lists, data frames, tibbles |
| 6 | Indexing with `[`, `[[`, `$`, logical/negative/name indexing |
| 7 | Arithmetic, comparison, logical, membership, pipes, `Inf`/`NaN`/`NA`/`NULL` |
| 8 | Pipes in depth + placeholder + anonymous functions |
| 9 | Functions: define, default args, `...`, anonymous, scope |
| 10 | if/else, for, while, `break`/`next`, apply family |
| 11 | Strings: base, `grep`/`gsub`, stringr, `sprintf`, glue |
| 12 | Factors: levels, relevel, tables, forcats |
| 13 | Dates: `as.Date`, components, arithmetic, lubridate |
| 14 | Missing values: detect, remove, fill, `NA` vs `NULL` vs `NaN` |
| 15 | Read/write CSV, Excel, RDS, RData |
| 16 | Full dplyr toolkit: filter, select, mutate, arrange, group_by, summarise, across, slice, pull, relocate |
| 17 | tidyr: pivot_longer, pivot_wider, separate, unite, complete, fill |
| 18 | Joins: inner, left, right, full, anti, semi |
| 19 | purrr: map, map_dbl, map2, pmap, walk |
| 20 | Base plots: plot, hist, boxplot, barplot, layout, save |
| 21 | ggplot2: scatter, line, hist, boxplot, bar, facets, saving |
| 22 | Statistical functions: descriptive, distributions, tests, regression, correlation |
| 23 | Random numbers, sampling, seed, train/test split |
| 24 | Vectorization vs loops + recycling |
| 25 | Debugging tools: traceback, debug, tryCatch, stopifnot |
| 26 | 15 common errors with fixes |
| 27 | RStudio shortcuts |
| 28 | Quick reference table |
| 29 | Theory Q&A |
| 30 | Single most important idea |

