---
title: "The Ultimate Data Handling Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(readxl)
library(haven)
library(knitr)
library(broom)
library(car)
library(lmtest)
```

# How to Use This Cheat Sheet

Every data-handling exam question falls into one of these. Find your question, jump to the section.

| Question type | Section | Typical phrasing |
|---|---|---|
| Create a data frame | 3 | "Enter the data as a data frame" |
| Import a file | 4 | "Read Q1.xls into R" |
| Inspect the data | 5 | "Describe the structure of the data" |
| Convert column types | 6 | "Treat group as a factor" |
| Set the reference level | 7 | "Use control as reference" |
| Handle missing values | 8 | "Two values are missing. What do you do?" |
| Select / rename / reorder columns | 9 | "Keep only id and score" |
| Filter rows | 10 | "Subset the men only" |
| Add / modify columns | 11 | "Create a BMI column" |
| Group summaries | 12 | "Report mean score by group" |
| Window / lag / cumulative | 13 | "Add a column of cumulative totals" |
| Reshape wide ↔ long | 14 | "Convert to long format for repeated measures" |
| Combine data frames | 15 | "Merge the two tables by id" |
| Frequency / contingency | 16 | "Make a 2-way table with percentages" |
| Descriptive stats by group | 17 | "Produce a summary table" |
| Train / test split | 18 | "Split 80/20" |
| Outliers | 19 | "Identify and remove outliers" |
| String cleaning | 20 | "Fix inconsistent labels" |
| Dates | 21 | "Compute age from date of birth" |
| Plots | 22 | "Plot the distribution by group" |
| Export | 23 | "Save the cleaned data as .csv" |

**Golden rule of this document:** every chunk is self-contained. Each one starts with `set.seed()` where random, builds its own data, and uses section-prefixed names (`p3_df`, `p7_fac`, ...).

---

# Part 0 — The 30-Second Exam Strategy

For **any** data-handling question, run this pipeline:

1. **Read** the data (`read_excel`, `read_csv`, `read_sav`).
2. **Inspect** (`dim`, `str`, `head`, `colSums(is.na())`).
3. **Fix types** (`as.numeric`, `factor(levels = ...)`).
4. **Check missing values** (`sum(is.na())`, decide: drop or fill).
5. **Clean / transform** (`filter`, `mutate`, `pivot_longer`).
6. **Summarise** (`group_by`, `summarise`).
7. **Verify** (`nrow` before vs after, spot-check rows).
8. **Report** in words: what you did, why, and what the output means.

If you can do these eight steps, you can handle any data-handling question in the exam.

---

# Part 1 — Foundations

## 1.1 What is a data frame?

A data frame is a **table**:
- **Rows:** observations (people, days, units).
- **Columns:** variables (age, group, score).
- Each column is a vector of the **same length**.
- Columns can be of **different types**.

```{r}
# --- Self-contained: anatomy of a data frame ---
p1_df <- tibble(
  id     = 1:4,
  gender = c("M", "F", "M", "F"),
  age    = c(25, 30, 22, 28),
  score  = c(78, 85, 90, 72)
)
p1_df
```

The output shows:
- `# A tibble: 4 × 4` → 4 rows, 4 columns.
- Each column name is followed by its type (`<int>`, `<chr>`, `<dbl>`).

## 1.2 Tidy data principles

A data frame is **tidy** if:
1. Each **variable** has its own column.
2. Each **observation** has its own row.
3. Each **value** has its own cell.

**Wide (not tidy):**
```
id  y0  y1  y2
1   10  11  12
```

**Long (tidy):**
```
id  time  y
1   0     10
1   1     11
1   2     12
```

Most tidyverse functions expect tidy data.

## 1.3 The names of things

```{r}
# --- Self-contained: naming pieces ---
p1_df <- tibble(id = 1:3, score = c(10, 20, 30))

p1_df$score         # column (a vector)
p1_df[[1]]          # first column (a vector)
p1_df[["score"]]    # column by name (a vector)
p1_df[, "score"]    # tibble with one column
p1_df[1, ]          # first row (a tibble)
p1_df[1, 2]         # element in row 1, column 2
```

**Rule:** `$` and `[[` extract the **contents**; `[` extracts a **subset** (still a tibble).

## 1.4 Base data.frame vs tibble

| Feature | `data.frame` | `tibble` |
|---|---|---|
| Print | All rows | First 10 rows, with types |
| Strings | Auto-factor (R < 4.0) | Always character |
| Row names | Yes | No |
| Subsetting | Drops to vector often | Keeps tibble |
| List columns | Awkward | Native support |

**Use `tibble` when possible.** It's what `read_csv` and all tidyverse functions return.

---

# Part 2 — Creating Data

## 2.1 From vectors

```{r}
# --- Self-contained: data.frame from vectors ---
p2_df <- data.frame(
  id     = 1:6,
  gender = c("M", "F", "M", "F", "M", "F"),
  age    = c(25, 30, 22, 28, 35, 27),
  score  = c(78, 85, 90, 72, 88, 95)
)
kable(p2_df)
```

**What happens:** Each argument becomes a column. All vectors must be the same length (or length 1, which gets recycled).

## 2.2 From `tibble()`

```{r}
# --- Self-contained: tibble from vectors ---
p2_tib <- tibble(
  id     = 1:6,
  gender = c("M", "F", "M", "F", "M", "F"),
  age    = c(25, 30, 22, 28, 35, 27),
  score  = c(78, 85, 90, 72, 88, 95)
)
kable(p2_tib)
```

**Difference from `data.frame`:** strings stay as strings (no silent factor conversion), nice printing, tibble class.

## 2.3 From a matrix

```{r}
# --- Self-contained: matrix to data frame ---
p2_mat <- matrix(1:12, nrow = 4, ncol = 3)
p2_df  <- as.data.frame(p2_mat)
names(p2_df) <- c("a", "b", "c")
kable(p2_df)
```

**When to use:** When the data is naturally rectangular and numeric (e.g., a correlation matrix, a count table).

## 2.4 From a matrix of counts (exam Q22-3 style)

```{r}
# --- Self-contained: contingency table as matrix ---
p2_tab <- matrix(c(4, 0, 0, 43, 97, 89, 25, 87, 61),
                 nrow = 3, byrow = TRUE)
dimnames(p2_tab) <- list(
  risk  = c("low", "medium", "high"),
  group = c("g1", "g2", "g3")
)
kable(p2_tab)
```

**What happens:** `matrix(..., byrow = TRUE)` fills row by row. `dimnames` gives names to rows and columns. This is how you enter a table printed in the exam.

## 2.5 From lists (mixed types)

```{r}
# --- Self-contained: list to data frame ---
p2_list <- list(
  id   = 1:3,
  name = c("A", "B", "C"),
  val  = c(1.1, 2.2, 3.3)
)
p2_df <- as_tibble(p2_list)
kable(p2_df)
```

## 2.6 With `tribble()` (row-wise, easy to read)

```{r}
# --- Self-contained: tribble ---
p2_trib <- tribble(
  ~id, ~gender, ~age, ~score,
  1,   "M",     25,   78,
  2,   "F",     30,   85,
  3,   "M",     22,   90,
  4,   "F",     28,   72
)
kable(p2_trib)
```

**Use `tribble()`** when you want to type the data in by hand and see it aligned like a table.

## 2.7 Sequences and repeats

```{r}
# --- Self-contained: sequences ---
p2_seq1 <- 1:10
p2_seq2 <- seq(0, 1, by = 0.25)
p2_seq3 <- seq(0, 1, length.out = 5)
p2_rep1 <- rep(1:3, times = 2)   # 1 2 3 1 2 3
p2_rep2 <- rep(1:3, each = 2)    # 1 1 2 2 3 3

p2_seq1
p2_seq2
p2_seq3
p2_rep1
p2_rep2
```

---

# Part 3 — Import

## 3.1 Excel (`.xls`, `.xlsx`)

```{r eval=FALSE}
library(readxl)

df <- read_excel("Q1.xls")                      # first sheet
df <- read_excel("Q1.xlsx", sheet = "Data")     # by name
df <- read_excel("Q1.xls", sheet = 2)           # by index
df <- read_excel("Q1.xls", range = "A1:F72")    # specific range
df <- read_excel("Q1.xls", skip = 2)            # skip first 2 rows
df <- read_excel("Q1.xls", col_names = FALSE)   # no header row
df <- read_excel("Q1.xls", na = c("", "NA", "?"))
excel_sheets("Q1.xls")                          # see sheet names
```

**Common issues:**
- **Header on row 2 or 3:** use `skip = 1` or `skip = 2`.
- **Trailing spaces in names:** use `rename_with(trimws)`.
- **Wrong type for a column:** fix after reading with `as.numeric()`.
- **Missing values coded as text:** use `na = c("", "NA", ".", "?")`.

## 3.2 SPSS (`.sav`)

```{r eval=FALSE}
library(haven)
df <- read_sav("data.sav")
df$gender <- as_factor(df$gender)   # convert labelled to factor
```

**What happens:** SPSS files store values with **labels** (e.g., 1 = "Male", 2 = "Female"). `read_sav` reads numbers; `as_factor` applies the labels. Always convert categorical codes to factors for analysis.

## 3.3 SAS / Stata

```{r eval=FALSE}
df <- read_sas("data.sas7bdat")
df <- read_dta("data.dta")
df <- read_stata("data.dta")
```

## 3.4 Text files

```{r eval=FALSE}
df <- read.table("DATA.txt", header = TRUE, sep = "\t")     # tab
df <- read.csv("DATA.csv", header = TRUE)                    # comma
df <- read.table("DATA.txt", header = TRUE)                  # whitespace
df <- read_delim("DATA.txt", delim = ";")                    # custom
df <- read_csv("DATA.csv")                                   # tidyverse, faster
```

**Common issues:**
- **Wrong delimiter:** try `sep = "\t"`, `sep = ";"`, `sep = "|"`.
- **Decimal comma:** use `read_csv2()` for `;`-separated with `,` decimal.
- **Encoding issues:** try `locale = locale(encoding = "latin1")`.

## 3.5 From clipboard

```{r eval=FALSE}
df <- read.table("clipboard", header = TRUE)          # Windows
df <- read.table(pipe("pbpaste"), header = TRUE)      # macOS
df <- read_clip_tbl()                                 # tidyverse, cross-platform
```

## 3.6 From a URL

```{r eval=FALSE}
df <- read_csv("https://example.com/data.csv")
```

## 3.7 After reading — always run these five

```{r}
# --- Self-contained: five inspection commands ---
p3_df <- tibble(
  id     = 1:6,
  gender = c("M", "F", "M", "F", "M", "F"),
  age    = c(25, 30, 22, 28, 35, 27),
  score  = c(78, 85, 90, 72, 88, 95)
)

dim(p3_df)                       # rows and columns
str(p3_df)                       # structure: types + sample values
head(p3_df, 3)                   # first 3 rows
names(p3_df)                     # column names
colSums(is.na(p3_df))            # NA count per column
```

**Report template:**
> "The data has 6 rows and 4 columns: id (integer), gender (character), age (numeric), score (numeric). There are no missing values."

---

# Part 4 — Inspection in Detail

## 4.1 Structure

```{r}
# --- Self-contained: structure ---
p4_df <- tibble(
  id     = 1:5,
  gender = c("M", "F", "M", "F", "M"),
  age    = c(25, 30, 22, 28, 35),
  score  = c(78, 85, 90, 72, 88)
)

nrow(p4_df)              # 5
ncol(p4_df)              # 4
dim(p4_df)               # 5 4
names(p4_df)             # column names
str(p4_df)               # types and sample values
glimpse(p4_df)           # tidyverse version of str
sapply(p4_df, class)     # class of each column
```

## 4.2 Summary statistics

```{r}
# --- Self-contained: summary ---
p4_df <- tibble(
  id     = 1:5,
  gender = c("M", "F", "M", "F", "M"),
  age    = c(25, 30, 22, 28, 35),
  score  = c(78, 85, 90, 72, 88)
)

summary(p4_df)
```

**Output:** For numeric columns: min, Q1, median, mean, Q3, max. For character/factor: counts per level.

## 4.3 Unique values

```{r}
# --- Self-contained: unique values ---
p4_df <- tibble(
  gender = c("M", "F", "M", "F", "M"),
  result = c("yes", "no", "yes", "yes", "no")
)

unique(p4_df$gender)
table(p4_df$gender)
length(unique(p4_df$gender))
```

**Use this before converting to factor** — you want to know every level that exists.

## 4.4 Missing values overview

```{r}
# --- Self-contained: NA overview ---
p4_df <- tibble(
  id    = 1:5,
  score = c(78, NA, 90, 72, NA),
  age   = c(25, 30, NA, 28, 35)
)

sum(is.na(p4_df))                # total NAs
colSums(is.na(p4_df))            # NAs per column
rowSums(is.na(p4_df))            # NAs per row
mean(is.na(p4_df))               # fraction missing overall
```

## 4.5 Quick views of the top and bottom

```{r}
# --- Self-contained: head and tail ---
p4_df <- tibble(
  id    = 1:20,
  score = rnorm(20)
)
head(p4_df)          # first 6
head(p4_df, 3)       # first 3
tail(p4_df)          # last 6
tail(p4_df, 2)       # last 2
```

---

# Part 5 — Data Types

## 5.1 The seven types you'll meet

| Type | `class()` | Example | Used for |
|---|---|---|---|
| double (numeric) | `"numeric"` | 3.14, -1.5e3 | measurements |
| integer | `"integer"` | 5L, -2L | counts |
| character | `"character"` | "male", "A" | text labels |
| factor | `"factor"` | group A/B/C | categories |
| ordered factor | `c("ordered","factor")` | low<mid<high | ordinal |
| logical | `"logical"` | TRUE, FALSE | flags |
| Date | `"Date"` | 2024-01-01 | calendar dates |
| POSIXct | `c("POSIXct","POSIXt")` | 2024-01-01 10:30:00 | date-times |

## 5.2 Numeric

```{r}
# --- Self-contained: numeric ---
p5_a <- 3.14
p5_b <- 1e3        # 1000
p5_c <- -2.5
p5_d <- 5L         # integer (L suffix)

class(p5_a)        # "numeric"
class(p5_d)        # "integer"
is.numeric(p5_a)   # TRUE
is.integer(p5_a)   # FALSE (default is double)
```

## 5.3 Character

```{r}
# --- Self-contained: character ---
p5_name <- "Parsa"
p5_greeting <- "Hello"
p5_sentence <- 'She said "hi"'

nchar(p5_name)
toupper(p5_name)
paste(p5_greeting, p5_name)
paste0(p5_greeting, ", ", p5_name, "!")
```

## 5.4 Logical

```{r}
# --- Self-contained: logical ---
p5_t <- TRUE
p5_f <- FALSE
p5_cmp <- 3 > 2

p5_t & p5_f    # FALSE
p5_t | p5_f    # TRUE
!p5_t          # FALSE
sum(c(TRUE, TRUE, FALSE))   # 2 (TRUE = 1, FALSE = 0)
```

**Trick:** `sum(logical)` counts how many are `TRUE`. `mean(logical)` gives the proportion.

## 5.5 Dates

```{r}
# --- Self-contained: dates ---
p5_d1 <- as.Date("2024-01-15")
p5_d2 <- as.Date("15/01/2024", format = "%d/%m/%Y")

class(p5_d1)
format(p5_d1, "%A %d %B %Y")    # "Monday 15 January 2024"
p5_d2 - p5_d1                    # Time difference
as.numeric(p5_d2 - p5_d1)        # number of days
```

## 5.6 Date-times

```{r}
# --- Self-contained: date-times ---
p5_dt <- as.POSIXct("2024-01-15 10:30:00", tz = "UTC")
class(p5_dt)
format(p5_dt, "%Y-%m-%d %H:%M")
```

## 5.7 Checking types

```{r}
# --- Self-contained: type checks ---
p5_v <- c(1, 2, 3)

is.numeric(p5_v)
is.character(p5_v)
is.factor(p5_v)
is.logical(p5_v)
is.Date(p5_v)
```

---

# Part 6 — Type Conversion

## 6.1 Explicit conversion

```{r}
# --- Self-contained: explicit conversion ---
p6_df <- tibble(
  a = c("1", "2", "3"),
  b = c("10", "20", "30"),
  c = c("M", "F", "M")
)

p6_df$a <- as.numeric(p6_df$a)
p6_df$b <- as.integer(p6_df$b)
p6_df$c <- as.factor(p6_df$c)

sapply(p6_df, class)
```

## 6.2 The dangerous conversion

```{r}
# --- Self-contained: coercion warnings ---
as.numeric(c("1", "2", "three"))
```

**Result:** `1 2 NA` with a warning. **Always** run `unique()` first to see what values exist.

**Rule:** If `as.numeric()` gives `NA` with a warning, the column contains non-numeric characters (spaces, commas, letters). Clean them first:
```{r}
# --- Self-contained: cleaning before conversion ---
p6_raw <- c("1,000", "2,500", "3,200")
p6_clean <- as.numeric(gsub(",", "", p6_raw))
p6_clean
```

## 6.3 Character → factor → numeric trap

```{r}
# --- Self-contained: factor-to-numeric trap ---
p6_f <- factor(c("10", "20", "30"))
as.numeric(p6_f)                    # 1 2 3 (level codes!)
as.numeric(as.character(p6_f))      # 10 20 30 (correct)
```

**Report:** "When a factor is coerced to numeric directly, R returns level codes (1, 2, 3). To recover the original numbers, convert to character first."

## 6.4 Character → Date

```{r}
# --- Self-contained: character to Date ---
p6_dates <- c("2024-01-15", "2024-02-20", "2024-03-10")
as.Date(p6_dates)                              # ISO format
as.Date("15/01/2024", format = "%d/%m/%Y")     # day-first
as.Date("01/15/2024", format = "%m/%d/%Y")     # month-first
```

## 6.5 Logical → numeric

```{r}
# --- Self-contained: logical to numeric ---
p6_lgl <- c(TRUE, FALSE, TRUE)
as.numeric(p6_lgl)   # 1 0 1
as.integer(p6_lgl)   # 1 0 1
```

## 6.6 Conversion summary table

| From | To | Function |
|---|---|---|
| character | numeric | `as.numeric(x)` |
| character | integer | `as.integer(x)` |
| character | factor | `factor(x, levels = ...)` |
| numeric | character | `as.character(x)` |
| numeric | factor | `factor(x, levels = ...)` |
| factor | character | `as.character(x)` |
| factor | numeric (original) | `as.numeric(as.character(x))` |
| logical | numeric | `as.numeric(x)` |
| character | Date | `as.Date(x, format = ...)` |

---

# Part 7 — Factors and Reference Levels

## 7.1 Why factors matter

Regression, ANOVA, and mixed models treat factors as **categorical** and create dummy variables. **The first level is the reference** unless you set it otherwise.

```{r}
# --- Self-contained: default factor ordering ---
p7_v <- c("control", "treat1", "treat2", "control", "treat1")
p7_f <- factor(p7_v)
levels(p7_f)   # alphabetical: "control" "treat1" "treat2"
```

## 7.2 Explicit levels (setting the reference)

```{r}
# --- Self-contained: explicit levels ---
p7_df <- tibble(
  id    = 1:5,
  group = c("control", "treat1", "treat2", "control", "treat1")
)

p7_df$group <- factor(p7_df$group,
                      levels = c("control", "treat1", "treat2"))
levels(p7_df$group)
```

`control` is now the reference. Any regression using `group` will have coefficients interpreted as "difference from control".

## 7.3 Change the reference later

```{r}
# --- Self-contained: relevel ---
p7_df <- tibble(group = factor(c("control", "treat1", "treat2", "control")))
p7_df$group <- relevel(p7_df$group, ref = "treat1")
levels(p7_df$group)
```

## 7.4 Check dummy coding

```{r}
# --- Self-contained: contrasts ---
p7_df <- tibble(
  group = factor(c("control", "treat1", "treat2", "control", "treat1"),
                 levels = c("control", "treat1", "treat2"))
)
contrasts(p7_df$group)
```

**Output:** Two columns (`treat1`, `treat2`), each is 0/1 for that level. The reference `control` is all zeros.

## 7.5 Numeric-coded groups (Q1 style)

If groups are coded `1, 2, 3` but `3` should be the reference:

```{r}
# --- Self-contained: numeric-coded to factor with custom reference ---
p7_df <- tibble(
  id        = 1:6,
  treatment = c(1, 2, 3, 1, 2, 3)
)

p7_df$treatment <- factor(p7_df$treatment, levels = c(3, 1, 2))
levels(p7_df$treatment)   # "3" "1" "2"
```

## 7.6 Ordered factors (ordinal data)

```{r}
# --- Self-contained: ordered factor ---
p7_df <- tibble(
  education = c("low", "medium", "high", "medium", "low")
)

p7_df$education <- factor(p7_df$education,
                          levels = c("low", "medium", "high"),
                          ordered = TRUE)
levels(p7_df$education)
is.ordered(p7_df$education)
```

**Use ordered factors** when the levels have a natural order (low < medium < high). This changes the default contrasts to polynomial, but the levels are the same.

## 7.7 Manual dummies

```{r}
# --- Self-contained: manual dummy ---
p7_df <- tibble(
  id     = 1:5,
  gender = c("M", "F", "M", "F", "M")
)

p7_df$is_male <- ifelse(p7_df$gender == "M", 1, 0)
p7_df$is_female <- ifelse(p7_df$gender == "F", 1, 0)
kable(p7_df)
```

**Use manual dummies** when you want explicit control over the design matrix.

## 7.8 Recode factor levels

```{r}
# --- Self-contained: fct_recode ---
library(forcats)
p7_f <- factor(c("M", "F", "M", "F"))
p7_f2 <- fct_recode(p7_f, "Male" = "M", "Female" = "F")
levels(p7_f2)
```

## 7.9 Collapse rare levels

```{r}
# --- Self-contained: fct_lump ---
library(forcats)
p7_f <- factor(c("A", "A", "A", "B", "B", "C", "D"))
fct_lump(p7_f, n = 2)   # keep top 2, lump rest as "Other"
```

**Report template:**
> "The factor group has three levels: control, treat1, treat2. I set control as the reference because it is the natural baseline. Regression coefficients for treat1 and treat2 will then represent the difference in mean outcome vs control."

---

# Part 8 — Missing Values

## 8.1 Detecting

```{r}
# --- Self-contained: NA detection ---
p8_df <- tibble(
  id    = 1:5,
  score = c(78, NA, 90, 72, NA),
  age   = c(25, 30, NA, 28, 35)
)

is.na(p8_df)                     # matrix of TRUE/FALSE
sum(is.na(p8_df))                # total
colSums(is.na(p8_df))            # per column
rowSums(is.na(p8_df))            # per row
anyNA(p8_df)                     # any NA?
mean(is.na(p8_df))               # fraction missing overall
which(is.na(p8_df$score))        # row indices with NA
```

## 8.2 Removing rows with NA

```{r}
# --- Self-contained: remove NAs ---
p8_df <- tibble(
  id    = 1:5,
  score = c(78, NA, 90, 72, NA)
)

p8_clean1 <- na.omit(p8_df)
p8_clean2 <- p8_df %>% filter(!is.na(score))
p8_clean3 <- tidyr::drop_na(p8_df)

nrow(p8_clean1)   # 3
nrow(p8_clean2)   # 3
nrow(p8_clean3)   # 3
```

**What happens:** `na.omit()` removes any row with at least one NA. `filter(!is.na(x))` removes only rows where `x` is NA. Choose based on which columns matter.

## 8.3 Replacing NA

```{r}
# --- Self-contained: replace NAs ---
p8_df <- tibble(
  id    = 1:5,
  score = c(78, NA, 90, 72, NA)
)

# Replace with mean
p8_mean <- mean(p8_df$score, na.rm = TRUE)
p8_df$score[is.na(p8_df$score)] <- p8_mean
kable(p8_df)
```

```{r}
# --- Self-contained: replace with constant ---
p8_df <- tibble(id = 1:5, score = c(78, NA, 90, 72, NA))
p8_df$score[is.na(p8_df$score)] <- 0
kable(p8_df)
```

```{r}
# --- Self-contained: replace with median ---
p8_df <- tibble(id = 1:5, score = c(78, NA, 90, 72, NA))
p8_df$score[is.na(p8_df$score)] <- median(p8_df$score, na.rm = TRUE)
kable(p8_df)
```

## 8.4 Filling (LOCF — last observation carried forward)

```{r}
# --- Self-contained: LOCF ---
p8_df <- tibble(
  time  = 1:5,
  score = c(78, NA, NA, 72, NA)
)
p8_filled <- p8_df %>% tidyr::fill(score, .direction = "down")
kable(p8_filled)
```

**When to use:** Time series where the value is expected to persist until the next observed value.

## 8.5 Interpolation

```{r}
# --- Self-contained: linear interpolation ---
p8_df <- tibble(
  time  = 1:5,
  score = c(78, NA, NA, 72, 75)
)
p8_df$score <- approx(p8_df$time, p8_df$score, xout = p8_df$time)$y
kable(p8_df, digits = 2)
```

**When to use:** Continuous measurements over time.

## 8.6 `na.rm = TRUE` in functions

```{r}
# --- Self-contained: na.rm ---
p8_v <- c(78, NA, 90, 72, NA)

mean(p8_v)                     # NA
mean(p8_v, na.rm = TRUE)       # 80
sum(p8_v, na.rm = TRUE)
sd(p8_v, na.rm = TRUE)
median(p8_v, na.rm = TRUE)
```

**Always** use `na.rm = TRUE` in summary functions, unless you actually want NA propagation.

## 8.7 NA vs NaN vs NULL vs Inf

| Value | Meaning | Behaves as |
|---|---|---|
| `NA` | missing | propagates; check with `is.na()` |
| `NaN` | not a number (0/0) | is also NA; check with `is.nan()` |
| `NULL` | empty | length 0; disappears in vectors |
| `Inf` | infinity | real value; check with `is.infinite()` |

```{r}
# --- Self-contained: special values ---
c(1, NA, 3)      # length 3
c(1, NULL, 3)    # length 2
c(1, NaN, 3)     # length 3
c(1, Inf, 3)     # length 3

is.na(NaN)       # TRUE
is.nan(NaN)      # TRUE
is.na(NA)        # TRUE
is.nan(NA)       # FALSE
```

## 8.8 Report template

**If you removed:**
> "The score column has 2 missing values (40%). I removed the affected rows using `drop_na()`, leaving 3 complete observations. The remaining data has no missing values."

**If you filled:**
> "The score column has 2 missing values. I imputed them with the column mean (80.0), so all subsequent analyses use a complete dataset of 5 rows. Sensitivity to this imputation should be considered."

**Warning:** If >5% of the data is missing, note it as a limitation. If missingness is systematic (e.g., all men missing), it biases the analysis.

---

# Part 9 — Column Operations

## 9.1 Select columns

```{r}
# --- Self-contained: select ---
p9_df <- tibble(
  id     = 1:5,
  gender = c("M", "F", "M", "F", "M"),
  age    = c(25, 30, 22, 28, 35),
  score  = c(78, 85, 90, 72, 88),
  q1     = c(1, 2, 3, 4, 5),
  q2     = c(6, 7, 8, 9, 10)
)

p9_df %>% select(id, score)                    # by name
p9_df %>% select(1:3)                          # by position
p9_df %>% select(-id)                          # drop one
p9_df %>% select(starts_with("q"))             # by prefix
p9_df %>% select(ends_with("e"))               # by suffix
p9_df %>% select(contains("co"))               # by substring
p9_df %>% select(where(is.numeric))            # by type
```

## 9.2 Rename columns

```{r}
# --- Self-contained: rename ---
p9_df <- tibble(id = 1:3, score = c(10, 20, 30), age = c(25, 30, 35))

p9_df %>% rename(score_final = score, age_years = age)
```

## 9.3 Rename all columns

```{r}
# --- Self-contained: rename_with ---
p9_df <- tibble(Score = 1:3, Age = c(25, 30, 35))

p9_df %>% rename_with(tolower)                     # all lowercase
p9_df %>% rename_with(~ paste0("v_", .x))         # prefix
```

## 9.4 Reorder columns

```{r}
# --- Self-contained: relocate ---
p9_df <- tibble(id = 1:3, score = c(10, 20, 30), age = c(25, 30, 35))

p9_df %>% relocate(score, .after = id)
p9_df %>% relocate(score, .before = id)
p9_df %>% relocate(where(is.numeric), .after = id)
```

## 9.5 Extract a column as a vector

```{r}
# --- Self-contained: pull ---
p9_df <- tibble(id = 1:3, score = c(10, 20, 30))
p9_df %>% pull(score)               # vector
p9_df %>% pull(1)                   # first column
p9_df$score                         # same
```

## 9.6 Base R alternatives

```{r}
# --- Self-contained: base R column ops ---
# --- Self-contained: base R column ops ---
p9_df <- tibble(id = 1:3, score = c(10, 20, 30), age = c(25, 30, 35))

p9_df[, c("id", "score")]           # select columns by name
p9_df[, -1]                         # drop the first column
p9_df[, 1:2]                        # select by position
p9_df[, c("score", "id", "age")]    # reorder columns
names(p9_df)[2] <- "score_final"    # rename the 2nd column (do this LAST)   # reorder
```

---

# Part 10 — Row Operations

## 10.1 Filter rows

```{r}
# --- Self-contained: filter ---
p10_df <- tibble(
  id     = 1:6,
  gender = c("M", "F", "M", "F", "M", "F"),
  age    = c(25, 30, 22, 28, 35, 27),
  score  = c(78, 85, 90, 72, 88, 95)
)

p10_df %>% filter(gender == "M")                        # equals
p10_df %>% filter(age > 25)                              # greater than
p10_df %>% filter(age > 25, score > 80)                  # AND
p10_df %>% filter(gender == "M" | age > 30)              # OR
p10_df %>% filter(gender %in% c("M", "F"))               # in
p10_df %>% filter(between(age, 20, 30))                  # range
p10_df %>% filter(!is.na(score))                         # not NA
p10_df %>% filter(str_detect(gender, "M"))               # regex
```

**Common mistakes:**
- `filter(gender = "M")` — should be `==`.
- `filter(age > 25 & score > 80)` — works, but `,` is faster.
- Forgetting `%in%` for multiple matches.

## 10.2 Slice rows by position

```{r}
# --- Self-contained: slice ---
p10_df <- tibble(id = 1:10, score = 1:10)

p10_df %>% slice(1:3)                # first 3 rows
p10_df %>% slice(c(1, 5, 10))        # specific rows
p10_df %>% slice_head(n = 3)         # first 3
p10_df %>% slice_tail(n = 3)         # last 3
p10_df %>% slice_max(score, n = 3)   # top 3 by score
p10_df %>% slice_min(score, n = 3)   # bottom 3
p10_df %>% slice_sample(n = 3)       # random 3
```

## 10.3 Distinct (unique) rows

```{r}
# --- Self-contained: distinct ---
p10_df <- tibble(
  id     = c(1, 1, 2, 3, 3),
  group  = c("A", "A", "B", "A", "B")
)

p10_df %>% distinct()                # unique combinations
p10_df %>% distinct(id)              # unique id values
p10_df %>% distinct(id, .keep_all = TRUE)   # first row per id
```

## 10.4 Sort rows

```{r}
# --- Self-contained: arrange ---
p10_df <- tibble(
  id     = 1:6,
  gender = c("M", "F", "M", "F", "M", "F"),
  score  = c(78, 85, 90, 72, 88, 95)
)

p10_df %>% arrange(score)                # ascending
p10_df %>% arrange(desc(score))          # descending
p10_df %>% arrange(gender, desc(score))  # multi-key
```

## 10.5 Base R alternatives

```{r}
# --- Self-contained: base R row ops ---
p10_df <- tibble(id = 1:5, score = c(78, 85, 90, 72, 88))

p10_df[p10_df$score > 80, ]              # filter
subset(p10_df, score > 80)               # filter (simpler)
p10_df[order(p10_df$score), ]            # ascending
p10_df[order(-p10_df$score), ]           # descending
p10_df[!duplicated(p10_df$id), ]         # distinct
```

## 10.6 Row counts before and after

```{r}
# --- Self-contained: verify row counts ---
p10_df <- tibble(id = 1:100, score = rnorm(100))

n_before <- nrow(p10_df)
p10_filtered <- p10_df %>% filter(score > 0)
n_after <- nrow(p10_filtered)

c(before = n_before, after = n_after)
```

**Always check:** If `n_after == 0`, you filtered out everything. If `n_after == n_before`, the filter did nothing. Both are red flags.

---

# Part 11 — Creating and Modifying Columns

## 11.1 Mutate — add or modify columns

```{r}
# --- Self-contained: mutate ---
p11_df <- tibble(
  id    = 1:5,
  score = c(78, 85, 90, 72, 88),
  age   = c(25, 30, 22, 28, 35)
)

p11_df <- p11_df %>% mutate(
  log_score = log(score),
  score2    = score^2,
  age_c     = age - mean(age),
  age_z     = (age - mean(age)) / sd(age),
  result    = ifelse(score >= 80, "pass", "fail")
)
kable(p11_df, digits = 3)
```

**What happens:** `mutate` adds new columns. It can reference columns just created in the same call.

## 11.2 ifelse vs if_else

```{r}
# --- Self-contained: ifelse vs if_else ---
p11_df <- tibble(score = c(78, 85, 90, NA))

# Base ifelse — returns NA when score is NA
p11_df$base <- ifelse(p11_df$score >= 80, "pass", "fail")

# dplyr if_else — strict, but you control the NA
p11_df$dplyr <- if_else(p11_df$score >= 80, "pass", "fail", missing = "unknown")

kable(p11_df)
```

**Use `if_else`** when you want type-checking and explicit `NA` handling.

## 11.3 case_when (multiple conditions)

```{r}
# --- Self-contained: case_when ---
p11_df <- tibble(score = c(95, 85, 75, 65, 55))

p11_df <- p11_df %>% mutate(
  grade = case_when(
    score >= 90 ~ "A",
    score >= 80 ~ "B",
    score >= 70 ~ "C",
    score >= 60 ~ "D",
    TRUE        ~ "F"
  )
)
kable(p11_df)
```

**Why `case_when`:** Cleaner than nested `ifelse`. The `TRUE ~ ...` at the end catches everything else.

## 11.4 cut (bin numeric into categories)

```{r}
# --- Self-contained: cut ---
p11_df <- tibble(age = c(15, 22, 28, 35, 45, 60, 72))

p11_df$age_cat <- cut(p11_df$age,
                      breaks = c(0, 25, 40, 60, 100),
                      labels = c("young", "middle", "senior", "elder"),
                      right = FALSE)
kable(p11_df)
```

**Arguments:**
- `breaks`: boundary values.
- `labels`: names for intervals. Length = `length(breaks) - 1`.
- `right = FALSE`: intervals `[a, b)`. `right = TRUE` (default): `(a, b]`.
- `include.lowest = TRUE`: include the lowest boundary.

## 11.5 Rownames

```{r}
# --- Self-contained: rownames ---
p11_df <- tibble(id = 1:3, score = c(10, 20, 30))

# Convert rownames to a column
p11_mtcars <- head(mtcars)
p11_mtcars$car <- rownames(p11_mtcars)
head(p11_mtcars[, c("car", "mpg")], 2)

# Or use tibble::rownames_to_column
p11_mtcars2 <- tibble::rownames_to_column(head(mtcars), var = "car")
head(p11_mtcars2[, c("car", "mpg")], 2)
```

## 11.6 Transmute (keep only new columns)

```{r}
# --- Self-contained: transmute ---
p11_df <- tibble(x = 1:3, y = 4:6)
p11_df %>% transmute(z = x + y, w = x * y)
```

## 11.7 across — apply to many columns

```{r}
# --- Self-contained: across ---
p11_df <- tibble(a = c(1, 2, 3), b = c(4, 5, 6), c = c("x", "y", "z"))

p11_df %>% mutate(across(where(is.numeric), ~ .x * 2))
p11_df %>% mutate(across(where(is.numeric), ~ .x - mean(.x), .names = "{.col}_c"))
```

**Report template:**
> "I created a BMI column as weight / height^2, and categorized age into four groups (young, middle, senior, elder). I also standardized age to have mean 0 and SD 1."

---

# Part 12 — Grouped Operations

## 12.1 group_by + summarise

```{r}
# --- Self-contained: grouped summary ---
p12_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F"),
  age    = c(25, 30, 22, 28, 35, 27),
  score  = c(78, 85, 90, 72, 88, 95)
)

p12_df %>%
  group_by(gender) %>%
  summarise(
    n        = n(),
    mean_s   = mean(score),
    sd_s     = sd(score),
    median_s = median(score),
    min_s    = min(score),
    max_s    = max(score),
    .groups  = "drop"
  ) %>%
  kable(digits = 3)
```

**What happens:** `group_by` splits the data by the grouping variable. `summarise` computes one row per group.

## 12.2 Summarise multiple columns

```{r}
# --- Self-contained: summarise across ---
p12_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F"),
  age    = c(25, 30, 22, 28, 35, 27),
  score  = c(78, 85, 90, 72, 88, 95)
)

p12_df %>%
  group_by(gender) %>%
  summarise(across(c(age, score),
                   list(mean = mean, sd = sd),
                   .names = "{.col}_{.fn}"),
            .groups = "drop") %>%
  kable(digits = 3)
```

## 12.3 Count per group

```{r}
# --- Self-contained: count ---
p12_df <- tibble(gender = c("M", "F", "M", "F", "M", "F"))

p12_df %>% count(gender)
p12_df %>% count(gender, sort = TRUE)
p12_df %>% count(gender, name = "n_obs")
```

## 12.4 Grouped mutate (add group-level summary)

```{r}
# --- Self-contained: grouped mutate ---
p12_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F"),
  score  = c(78, 85, 90, 72, 88, 95)
)

p12_df %>%
  group_by(gender) %>%
  mutate(mean_score = mean(score),
         diff_from_mean = score - mean_score) %>%
  ungroup() %>%
  kable(digits = 3)
```

**Use case:** Standardizing within groups, computing deviations, ranking within groups.

## 12.5 Grouped filter (keep rows satisfying group condition)

```{r}
# --- Self-contained: grouped filter ---
p12_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F"),
  score  = c(78, 85, 90, 72, 88, 95)
)

# Keep rows where score is above the group mean
p12_df %>%
  group_by(gender) %>%
  filter(score > mean(score)) %>%
  ungroup()
```

## 12.6 Ungroup

```{r}
# --- Self-contained: ungroup ---
p12_df <- tibble(gender = c("M", "F"), score = c(80, 90))

p12_df %>% group_by(gender) %>% summarise(n = n(), .groups = "drop")
# or explicitly
p12_df %>% group_by(gender) %>% summarise(n = n()) %>% ungroup()
```

**Always `.groups = "drop"` or `ungroup()`** at the end, otherwise downstream functions use the grouping silently.

## 12.7 Base R alternatives

```{r}
# --- Self-contained: base R group summary ---
p12_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F"),
  score  = c(78, 85, 90, 72, 88, 95)
)

tapply(p12_df$score, p12_df$gender, mean)
aggregate(score ~ gender, data = p12_df, FUN = mean)
aggregate(cbind(score, score) ~ gender, data = p12_df, FUN = mean)
```

## 12.8 Report template

> "Mean score is 85.3 for men (n = 3, SD = 6.4) and 84.0 for women (n = 3, SD = 11.5). The group means differ by 1.3 points, with wider variability in the female group."

---

# Part 13 — Window Functions

Window functions return a value **per row** but depend on a group or order.

## 13.1 Lag and lead

```{r}
# --- Self-contained: lag and lead ---
p13_df <- tibble(
  year  = 2020:2024,
  value = c(10, 12, 15, 14, 18)
)

p13_df %>%
  mutate(
    prev = lag(value),
    next_v = lead(value),
    change = value - prev
  ) %>%
  kable()
```

**Use for:** time series, computing changes, growth rates.

## 13.2 Cumulative functions

```{r}
# --- Self-contained: cumsum etc ---
p13_df <- tibble(x = c(1, 2, 3, 4, 5))

p13_df %>%
  mutate(
    cum_sum = cumsum(x),
    cum_max = cummax(x),
    cum_min = cummin(x),
    cum_prod = cumprod(x)
  ) %>%
  kable()
```

## 13.3 Ranking

```{r}
# --- Self-contained: ranking ---
p13_df <- tibble(score = c(90, 75, 85, 75, 95))

p13_df %>%
  mutate(
    row_num  = row_number(),
    rank     = rank(score),
    rank_desc = rank(desc(score)),
    min_rank = min_rank(score),
    dense_rank = dense_rank(score),
    pct_rank = percent_rank(score)
  ) %>%
  kable(digits = 3)
```

| Function | Behavior with ties |
|---|---|
| `row_number()` | always unique; ties broken by order |
| `min_rank()` | ties get the same rank; next rank skips |
| `dense_rank()` | ties get the same rank; next rank not skipped |
| `percent_rank()` | scaled to [0, 1] |

## 13.4 n() and n_distinct()

```{r}
# --- Self-contained: n and n_distinct ---
p13_df <- tibble(
  group = c("A", "A", "B", "B", "B"),
  id    = c(1, 2, 3, 4, 5)
)

p13_df %>%
  group_by(group) %>%
  summarise(n_rows = n(), n_ids = n_distinct(id), .groups = "drop")
```

## 13.5 first, last, nth

```{r}
# --- Self-contained: first/last/nth ---
p13_df <- tibble(
  group = c("A", "A", "A", "B", "B"),
  value = c(10, 20, 30, 40, 50)
)

p13_df %>%
  group_by(group) %>%
  summarise(
    first_v = first(value),
    last_v  = last(value),
    second_v = nth(value, 2),
    .groups = "drop"
  )
```

---

# Part 14 — Reshaping

## 14.1 Wide to long

```{r}
# --- Self-contained: pivot_longer ---
p14_wide <- tibble(
  id = 1:3,
  y0 = c(10, 12, 14),
  y1 = c(11, 13, 15),
  y2 = c(12, 14, 16)
)

p14_long <- p14_wide %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = "time",
    values_to = "score"
  ) %>%
  mutate(time = as.numeric(gsub("y", "", time)))

kable(p14_long)
```

**What happens:** Each row of `p14_wide` becomes 3 rows in `p14_long`. Columns `y0, y1, y2` become a single `score` column, with `time` identifying which one.

## 14.2 Long to wide

```{r}
# --- Self-contained: pivot_wider ---
p14_long <- tibble(
  id    = rep(1:3, each = 3),
  time  = rep(0:2, 3),
  score = c(10, 11, 12, 12, 13, 14, 14, 15, 16)
)

p14_wide <- p14_long %>%
  pivot_wider(
    names_from = time,
    values_from = score,
    names_prefix = "y"
  )
kable(p14_wide)
```

## 14.3 Separate one column into two

```{r}
# --- Self-contained: separate ---
p14_df <- tibble(code = c("A_1", "B_2", "C_3"))

p14_df %>%
  separate(code, into = c("letter", "number"), sep = "_") %>%
  kable()
```

## 14.4 Unite two columns into one

```{r}
# --- Self-contained: unite ---
p14_df <- tibble(letter = c("A", "B"), number = 1:2)

p14_df %>%
  unite("code", letter, number, sep = "-") %>%
  kable()
```

## 14.5 Complete (fill in missing combinations)

```{r}
# --- Self-contained: complete ---
p14_df <- tibble(
  year = c(2020, 2021, 2023),
  val  = c(1, 2, 4)
)

p14_df %>% tidyr::complete(year = 2020:2023) %>% kable()
```

**Use case:** Filling in missing years, months, or group combinations.

## 14.6 Long vs wide: which to use

| Task | Format |
|---|---|
| Time series plot with multiple series | long |
| Regression with one row per subject | wide |
| ANOVA / repeated measures | long |
| Correlation matrix | wide |
| Summary table for report | wide |

## 14.7 Report template

> "The data was originally in wide format (y0, y1, y2 columns). I converted it to long format so that each row represents one observation (id, time, score). This is the format required for mixed models and repeated-measures ANOVA."

---

# Part 15 — Combining Data Frames

## 15.1 Stack rows (union)

```{r}
# --- Self-contained: bind_rows ---
p15_a <- tibble(x = 1:3, y = c("a", "b", "c"))
p15_b <- tibble(x = 4:6, y = c("d", "e", "f"))

bind_rows(p15_a, p15_b)
rbind(p15_a, p15_b)
```

**When columns differ:** `bind_rows` fills missing columns with `NA`.

```{r}
# --- Self-contained: bind_rows with different columns ---
p15_a <- tibble(x = 1:2)
p15_b <- tibble(x = 3:4, y = c("p", "q"))

bind_rows(p15_a, p15_b)
```

## 15.2 Stack columns (side-by-side)

```{r}
# --- Self-contained: bind_cols ---
p15_a <- tibble(x = 1:3)
p15_b <- tibble(y = 4:6)

bind_cols(p15_a, p15_b)
cbind(p15_a, p15_b)
```

**Warning:** `bind_cols` matches by position, not by name. Rows must align.

## 15.3 The six joins

```{r}
# --- Self-contained: all joins ---
p15_left  <- tibble(id = 1:3, x = c("a", "b", "c"))
p15_right <- tibble(id = 2:4, y = c("P", "Q", "R"))

# Inner: only matching ids
inner_join(p15_left, p15_right, by = "id")

# Left: all rows from left, matched from right
left_join(p15_left, p15_right, by = "id")

# Right: all rows from right, matched from left
right_join(p15_left, p15_right, by = "id")

# Full: all rows from both
full_join(p15_left, p15_right, by = "id")

# Anti: rows in left NOT in right
anti_join(p15_left, p15_right, by = "id")

# Semi: rows in left that ARE in right (left columns only)
semi_join(p15_left, p15_right, by = "id")
```

## 15.4 Joining by different key names

```{r}
# --- Self-contained: join with different key names ---
p15_a <- tibble(id_a = 1:3, x = c("a", "b", "c"))
p15_b <- tibble(id_b = 2:4, y = c("P", "Q", "R"))

inner_join(p15_a, p15_b, by = c("id_a" = "id_b"))
```

## 15.5 Join summary

| Join | Keeps |
|---|---|
| `inner_join` | only matching rows |
| `left_join` | all rows from left |
| `right_join` | all rows from right |
| `full_join` | all rows from both |
| `anti_join` | rows in left NOT in right |
| `semi_join` | rows in left that match right |

## 15.6 Report template

> "I merged the demographics table with the outcomes table using `left_join` by `id`. The result has 200 rows (one per subject), with 3 subjects having missing outcome data (they appear with NA)."

---

# Part 16 — Frequency and Contingency Tables

## 16.1 One-way frequency

```{r}
# --- Self-contained: one-way table ---
p16_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F", "M", "F")
)

table(p16_df$gender)
prop.table(table(p16_df$gender))
round(prop.table(table(p16_df$gender)) * 100, 1)
```

## 16.2 Two-way contingency table

```{r}
# --- Self-contained: two-way table ---
p16_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F", "M", "F"),
  result = c("yes", "no", "yes", "yes", "no", "no", "yes", "yes")
)

tab <- table(p16_df$gender, p16_df$result)
addmargins(tab)
```

## 16.3 Proportions

```{r}
# --- Self-contained: proportion tables ---
p16_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F", "M", "F"),
  result = c("yes", "no", "yes", "yes", "no", "no", "yes", "yes")
)
tab <- table(p16_df$gender, p16_df$result)

prop.table(tab)                          # overall %
prop.table(tab, margin = 1)              # row % (within each gender)
prop.table(tab, margin = 2)              # column % (within each result)
round(prop.table(tab, margin = 1) * 100, 1)
```

**Rule of thumb:** For a 2×2 table, use **row percentages** (margin = 1) if gender is the predictor and result is the outcome.

## 16.4 Chi-square test

```{r}
# --- Self-contained: chi-square ---
p16_df <- tibble(
  gender = c("M", "F", "M", "F", "M", "F", "M", "F"),
  result = c("yes", "no", "yes", "yes", "no", "no", "yes", "yes")
)
tab <- table(p16_df$gender, p16_df$result)

chisq.test(tab)
fisher.test(tab)      # for small expected counts
```

**Interpretation:**
- $H_0$: no association between the two variables.
- $p < 0.05$: variables are associated.
- Check expected counts: if any < 5, use Fisher's exact test.

## 16.5 CrossTable (gmodels package)

```{r eval=FALSE}
library(gmodels)
CrossTable(p16_df$gender, p16_df$result, prop.r = TRUE, prop.c = TRUE)
```

**Useful** because it prints counts and all proportions in one call.

## 16.6 Report template

> "The 2-way table of gender and result shows that 66.7% of men and 50.0% of women had a positive result. A chi-square test gives $\chi^2(1) = 0.10$, $p = 0.75$, indicating no significant association between gender and result."

---

# Part 17 — Descriptive Statistics by Group

## 17.1 Standard summary table

```{r}
# --- Self-contained: summary by group ---
p17_df <- tibble(
  group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
  score = c(78, 85, 90, 72, 88, 95, 60, 75, 82)
)

p17_df %>%
  group_by(group) %>%
  summarise(
    n      = n(),
    mean   = mean(score),
    sd     = sd(score),
    median = median(score),
    min    = min(score),
    max    = max(score),
    .groups = "drop"
  ) %>%
  kable(digits = 3)
```

## 17.2 Custom summary with SE and CI

```{r}
# --- Self-contained: summary with SE and CI ---
p17_df <- tibble(
  group = c("A", "A", "A", "B", "B", "B"),
  score = c(78, 85, 90, 72, 88, 95)
)

p17_df %>%
  group_by(group) %>%
  summarise(
    n    = n(),
    mean = mean(score),
    sd   = sd(score),
    se   = sd / sqrt(n),
    ci_lo = mean - qt(0.975, n - 1) * se,
    ci_hi = mean + qt(0.975, n - 1) * se,
    .groups = "drop"
  ) %>%
  kable(digits = 3)
```

## 17.3 Summary by multiple groupings

```{r}
# --- Self-contained: two-way grouping ---
p17_df <- tibble(
  gender = c("M", "M", "F", "F", "M", "F"),
  treat  = c("A", "B", "A", "B", "A", "B"),
  score  = c(78, 85, 90, 72, 88, 95)
)

p17_df %>%
  group_by(gender, treat) %>%
  summarise(n = n(), mean = mean(score), .groups = "drop") %>%
  kable(digits = 3)
```

## 17.4 Formatting for a report

```{r}
# --- Self-contained: formatted summary strings ---
p17_df <- tibble(
  group = c("A", "A", "B", "B"),
  score = c(80, 90, 70, 85)
)

p17_df %>%
  group_by(group) %>%
  summarise(
    m_sd = sprintf("%.1f (%.1f)", mean(score), sd(score)),
    .groups = "drop"
  ) %>%
  kable()
```

**Report style:** "Mean (SD)" in one column, common in medical/psych papers.

## 17.5 Base R alternatives

```{r}
# --- Self-contained: base R summaries ---
p17_df <- tibble(
  group = c("A", "A", "A", "B", "B", "B"),
  score = c(78, 85, 90, 72, 88, 95)
)

tapply(p17_df$score, p17_df$group, mean)
aggregate(score ~ group, data = p17_df, FUN = function(x) c(mean = mean(x), sd = sd(x)))
```

---

# Part 18 — Sampling and Splitting

## 18.1 Simple random sample

```{r}
# --- Self-contained: simple random sample ---
set.seed(1801)
p18_df <- tibble(id = 1:100, y = rnorm(100))

p18_sample <- p18_df %>% sample_n(10)
p18_sample <- p18_df %>% slice_sample(n = 10)
p18_sample <- p18_df %>% slice_sample(prop = 0.1)   # 10%
nrow(p18_sample)
```

## 18.2 Train/test split

```{r}
# --- Self-contained: train/test split ---
set.seed(1802)
p18_df <- tibble(id = 1:100, y = rnorm(100))

p18_n     <- nrow(p18_df)
p18_idx   <- sample(1:p18_n, size = floor(0.8 * p18_n))
p18_train <- p18_df[p18_idx, ]
p18_test  <- p18_df[-p18_idx, ]

nrow(p18_train); nrow(p18_test)
```

**Report:** "I split the data 80/20 into training (80 rows) and test (20 rows) sets."

## 18.3 Stratified sample (preserve group proportions)

```{r}
# --- Self-contained: stratified sample ---
set.seed(1803)
p18_df <- tibble(
  gender = rep(c("M", "F"), each = 50),
  y      = rnorm(100)
)

p18_strat <- p18_df %>%
  group_by(gender) %>%
  slice_sample(prop = 0.2) %>%
  ungroup()
nrow(p18_strat)
table(p18_strat$gender)
```

## 18.4 K-fold cross-validation indices

```{r}
# --- Self-contained: K-fold indices ---
set.seed(1804)
p18_n <- 100
p18_K <- 5
p18_folds <- sample(rep(1:p18_K, length.out = p18_n))

for (k in 1:p18_K) {
  cat("Fold", k, ":", sum(p18_folds == k), "rows\n")
}
```

## 18.5 Report template

> "I used `set.seed(42)` before sampling for reproducibility. The 80/20 split gives 80 training and 20 test observations. Stratified by gender, each set has the same male:female ratio as the full data."

---

# Part 19 — Outliers

## 19.1 IQR rule

```{r}
# --- Self-contained: IQR outlier detection ---
p19_df <- tibble(y = c(10, 12, 11, 13, 12, 100, 11, 12, 13, 10))

Q1  <- quantile(p19_df$y, 0.25)
Q3  <- quantile(p19_df$y, 0.75)
IQR_y <- Q3 - Q1
lower <- Q1 - 1.5 * IQR_y
upper <- Q3 + 1.5 * IQR_y

p19_df$is_outlier <- p19_df$y < lower | p19_df$y > upper
kable(p19_df)
```

**Rule:** Outliers are values below $Q_1 - 1.5 \cdot IQR$ or above $Q_3 + 1.5 \cdot IQR$.

## 19.2 Standardized residuals

```{r}
# --- Self-contained: Z-score outliers ---
p19_df <- tibble(y = c(10, 12, 11, 13, 12, 100, 11, 12, 13, 10))

p19_z <- (p19_df$y - mean(p19_df$y)) / sd(p19_df$y)
which(abs(p19_z) > 3)
```

**Rule:** $|z| > 3$ is a common threshold.

## 19.3 Remove outliers

```{r}
# --- Self-contained: remove outliers ---
p19_df <- tibble(y = c(10, 12, 11, 13, 12, 100, 11, 12, 13, 10))

Q1  <- quantile(p19_df$y, 0.25)
Q3  <- quantile(p19_df$y, 0.75)
IQR_y <- Q3 - Q1

p19_clean <- p19_df %>%
  filter(y >= Q1 - 1.5 * IQR_y, y <= Q3 + 1.5 * IQR_y)

nrow(p19_df); nrow(p19_clean)
```

## 19.4 Winsorize (cap at percentiles)

```{r}
# --- Self-contained: winsorize ---
p19_df <- tibble(y = c(10, 12, 11, 13, 12, 100, 11, 12, 13, 10))

p19_df$y_win <- pmin(pmax(p19_df$y, quantile(p19_df$y, 0.05)),
                     quantile(p19_df$y, 0.95))
kable(p19_df)
```

**Winsorizing** replaces extreme values with the 5th and 95th percentiles. Less drastic than removing.

## 19.5 Report template

> "Using the 1.5 × IQR rule, I identified 1 outlier (y = 100). Removing it reduces the mean from 20.4 to 12.2 and the SD from 27.5 to 1.1. If the outlier is a data-entry error, remove it; otherwise consider winsorizing or reporting results with and without it."

**Caution:** Never remove outliers silently. Always report both analyses.

---

# Part 20 — Strings

## 20.1 Basic operations

```{r}
# --- Self-contained: string basics ---
p20_s <- "  Hello, World!  "

nchar(p20_s)
toupper(p20_s)
tolower(p20_s)
trimws(p20_s)
substr(p20_s, 3, 7)
strsplit("a,b,c", ",")[[1]]
paste("x", "y", "z")
paste0("x", "y", "z")
```

## 20.2 Detect, extract, replace

```{r}
# --- Self-contained: string pattern matching ---
p20_v <- c("apple", "banana", "cherry", "apricot")

grep("ap", p20_v)                   # indices
grep("ap", p20_v, value = TRUE)     # values
grepl("ap", p20_v)                  # logical

sub("a", "A", p20_v)                # replace first
gsub("a", "A", p20_v)               # replace all
```

## 20.3 stringr (tidyverse)

```{r}
# --- Self-contained: stringr ---
library(stringr)
p20_v <- c("apple", "banana", "cherry", "apricot")

str_detect(p20_v, "an")            # logical
str_replace(p20_v, "a", "A")
str_replace_all(p20_v, "a", "A")
str_sub(p20_v, 1, 3)               # first 3 chars
str_length(p20_v)
str_to_upper(p20_v)
str_trim("  hi  ")
str_pad("hi", 5, side = "left")
str_split("a,b,c", ",")[[1]]
```

## 20.4 Common cleaning tasks

```{r}
# --- Self-contained: cleaning strings ---
p20_raw <- c("M", "m", "M ", "Male", " F", "f", "FEMALE")

# Normalize
p20_clean <- p20_raw %>%
  str_trim() %>%
  str_to_upper() %>%
  str_sub(1, 1)   # first letter only

p20_clean
```

## 20.5 Extract with regex

```{r}
# --- Self-contained: regex extraction ---
p20_codes <- c("A-123", "B-456", "C-789")
str_extract(p20_codes, "[A-Z]")     # letter
str_extract(p20_codes, "\\d+")      # digits
str_match(p20_codes, "([A-Z])-(\\d+)")
```

## 20.6 Report template

> "The gender column contained inconsistent labels ('M', 'm', 'Male', 'F ', 'FEMALE'). I normalized by trimming whitespace, converting to uppercase, and taking the first letter, yielding two clean levels: M and F."

---

# Part 21 — Dates and Times

## 21.1 Create dates

```{r}
# --- Self-contained: create dates ---
p21_d1 <- as.Date("2024-01-15")
p21_d2 <- as.Date("15/01/2024", format = "%d/%m/%Y")
p21_dt <- as.POSIXct("2024-01-15 10:30:00", tz = "UTC")
p21_d1; p21_d2; p21_dt
```

## 21.2 Components

```{r}
# --- Self-contained: date components ---
p21_d <- as.Date("2024-01-15")

format(p21_d, "%Y")     # "2024"
format(p21_d, "%m")     # "01"
format(p21_d, "%d")     # "15"
format(p21_d, "%A")     # "Monday"
format(p21_d, "%B")     # "January"

weekdays(p21_d)
months(p21_d)
quarters(p21_d)
```

## 21.3 Arithmetic

```{r}
# --- Self-contained: date arithmetic ---
p21_d1 <- as.Date("2024-01-15")
p21_d2 <- as.Date("2024-02-15")

p21_d1 + 7                # one week later
p21_d2 - p21_d1           # Time difference of 31 days
as.numeric(p21_d2 - p21_d1)
```

## 21.4 Compute age from birth date

```{r}
# --- Self-contained: age from DOB ---
p21_dob <- as.Date(c("1990-05-10", "1985-03-22", "2000-11-15"))
p21_ref <- as.Date("2024-01-01")
p21_age <- as.numeric(p21_ref - p21_dob) / 365.25
round(p21_age, 1)
```

## 21.5 lubridate (easier)

```{r}
# --- Self-contained: lubridate ---
library(lubridate)
p21_d <- ymd("2024-01-15")
p21_d
year(p21_d)
month(p21_d)
day(p21_d)
wday(p21_d, label = TRUE)
p21_d + weeks(2)
p21_d %m+% months(1)     # month arithmetic with rollover
```

## 21.6 Common formats

| Format | Meaning | Example |
|---|---|---|
| `%Y` | 4-digit year | 2024 |
| `%y` | 2-digit year | 24 |
| `%m` | month (01–12) | 01 |
| `%d` | day (01–31) | 15 |
| `%H` | hour (00–23) | 10 |
| `%M` | minute | 30 |
| `%S` | second | 00 |
| `%A` | weekday (full) | Monday |
| `%B` | month (full) | January |

---

# Part 22 — Plots for Exploration

## 22.1 Histogram

```{r}
# --- Self-contained: histogram ---
set.seed(2201)
p22_df <- tibble(score = rnorm(200, 80, 10))

ggplot(p22_df, aes(score)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(title = "Distribution of score", x = "Score", y = "Count") +
  theme_minimal()
```

**What to look for:** shape (symmetric? skewed?), outliers, gaps, multiple modes.

## 22.2 Boxplot by group

```{r}
# --- Self-contained: boxplot ---
set.seed(2202)
p22_df <- tibble(
  group = rep(c("A", "B", "C"), each = 30),
  score = c(rnorm(30, 70), rnorm(30, 80), rnorm(30, 90))
)

ggplot(p22_df, aes(group, score, fill = group)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "none")
```

**What to look for:** differences in medians, spread, outliers, skewness.

## 22.3 Scatterplot

```{r}
# --- Self-contained: scatterplot ---
set.seed(2203)
p22_df <- tibble(x = rnorm(100), y = 2 + 3 * x + rnorm(100))

ggplot(p22_df, aes(x, y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  theme_minimal()
```

**What to look for:** linearity, outliers, clusters, heteroskedasticity (funnel shape).

## 22.4 Bar chart

```{r}
# --- Self-contained: bar chart ---
p22_df <- tibble(group = c("A", "B", "C"), count = c(20, 35, 15))

ggplot(p22_df, aes(group, count, fill = group)) +
  geom_col() +
  theme_minimal() +
  theme(legend.position = "none")
```

## 22.5 Density plot

```{r}
# --- Self-contained: density ---
set.seed(2205)
p22_df <- tibble(
  group = rep(c("A", "B"), each = 200),
  value = c(rnorm(200, 0), rnorm(200, 1))
)

ggplot(p22_df, aes(value, fill = group)) +
  geom_density(alpha = 0.5) +
  theme_minimal()
```

## 22.6 Facets

```{r}
# --- Self-contained: faceted scatterplot ---
set.seed(2206)
p22_df <- tibble(
  x = rnorm(100),
  y = rnorm(100),
  group = rep(c("A", "B"), 50)
)

ggplot(p22_df, aes(x, y)) +
  geom_point() +
  facet_wrap(~ group) +
  theme_minimal()
```

## 22.7 QQ plot (normality check)

```{r}
# --- Self-contained: QQ plot ---
set.seed(2207)
p22_df <- tibble(score = rnorm(200))

ggplot(p22_df, aes(sample = score)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal()
```

**What to look for:** points should follow the diagonal. Curvature = non-normal.

## 22.8 Base R alternatives

```{r}
# --- Self-contained: base R plots ---
set.seed(2208)
p22_df <- tibble(x = rnorm(100), y = rnorm(100))

hist(p22_df$x)
boxplot(p22_df$y)
plot(p22_df$x, p22_df$y)
barplot(c(A = 10, B = 20, C = 15))
```

## 22.9 Report template

> "The boxplot shows that group C has the highest median (90.2) and group A the lowest (70.1). All groups have similar spread. There are no extreme outliers. The distribution of scores is approximately symmetric in each group."

---

# Part 23 — Export

## 23.1 CSV

```{r eval=FALSE}
write.csv(df, "out.csv", row.names = FALSE)
write_csv(df, "out.csv")   # tidyverse, faster
```

## 23.2 Excel

```{r eval=FALSE}
library(writexl)
write_xlsx(df, "out.xlsx")
write_xlsx(list(Sheet1 = df1, Sheet2 = df2), "multi.xlsx")
```

## 23.3 RDS (R's native format)

```{r eval=FALSE}
saveRDS(df, "df.rds")
df <- readRDS("df.rds")
```

**RDS preserves all types** (factors, dates, list-columns). Use it when your next R session will read it.

## 23.4 RData (multiple objects)

```{r eval=FALSE}
save(df, fit, file = "workspace.RData")
load("workspace.RData")
```

## 23.5 Text file

```{r eval=FALSE}
write.table(df, "out.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

## 23.6 Report tables

```{r}
# --- Self-contained: kable table ---
p23_df <- tibble(group = c("A", "B"), mean = c(85.3, 84.0), sd = c(6.4, 11.5))
kable(p23_df, digits = 2, caption = "Mean and SD by group")
```

**For reports:** `kable()` for HTML/PDF, `kableExtra` for more styling.

---

# Part 24 — Full Workflow Template

Copy-paste for any exam data-handling question.

```{r eval=FALSE}
# ---- 1. Load packages ----
library(tidyverse)
library(readxl)
library(knitr)

# ---- 2. Read ----
df <- read_excel("Q1.xls", sheet = 1)

# ---- 3. Inspect ----
dim(df)
str(df)
head(df)
colSums(is.na(df))

# ---- 4. Fix types ----
df$group <- factor(df$group, levels = c("control", "treat1", "treat2"))
df$y     <- as.numeric(df$y)
df$date  <- as.Date(df$date)

# ---- 5. Handle missing ----
df <- df %>% drop_na(y)

# ---- 6. Transform ----
df <- df %>%
  mutate(
    log_y  = log(y),
    y_cat  = cut(y, breaks = c(0, 50, 100, Inf), labels = c("low", "mid", "high"))
  )

# ---- 7. Reshape if needed ----
# df_long <- df %>% pivot_longer(...)

# ---- 8. Summarise ----
df %>%
  group_by(group) %>%
  summarise(n = n(), mean = mean(y), sd = sd(y), .groups = "drop") %>%
  kable(digits = 3)

# ---- 9. Plot ----
ggplot(df, aes(group, y, fill = group)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "none")

# ---- 10. Export ----
write_csv(df, "cleaned.csv")
```

---

# Part 25 — Common Exam Scenarios

## 25.1 Excel with header rows to skip

```{r eval=FALSE}
df <- read_excel("Q1.xls", skip = 2)
```

## 25.2 Numeric-coded groups → factor with custom reference

```{r}
# --- Self-contained: numeric codes to factor ---
p25_df <- tibble(id = 1:6, group = c(1, 2, 3, 1, 2, 3))
p25_df$group <- factor(p25_df$group, levels = c(3, 1, 2))
levels(p25_df$group)
```

## 25.3 Group-level summary table

```{r}
# --- Self-contained: group summary ---
p25_df <- tibble(
  group = c("A", "A", "B", "B", "C", "C"),
  y     = c(10, 12, 20, 22, 30, 32)
)

p25_df %>%
  group_by(group) %>%
  summarise(n = n(), mean = mean(y), sd = sd(y), .groups = "drop") %>%
  kable(digits = 3)
```

## 25.4 Contingency table with percentages

```{r}
# --- Self-contained: contingency percentages ---
p25_df <- tibble(
  gender = c("M", "M", "F", "F", "M", "F"),
  result = c("yes", "no", "yes", "yes", "no", "yes")
)

tab <- table(p25_df$gender, p25_df$result)
addmargins(tab)
round(prop.table(tab, margin = 1) * 100, 1)
```

## 25.5 Train / test split

```{r}
# --- Self-contained: split ---
set.seed(2505)
p25_df <- tibble(id = 1:10, y = rnorm(10))
p25_idx <- sample(1:nrow(p25_df), size = floor(0.8 * nrow(p25_df)))
p25_train <- p25_df[p25_idx, ]
p25_test  <- p25_df[-p25_idx, ]
nrow(p25_train); nrow(p25_test)
```

## 25.6 Repeated measures wide → long

```{r}
# --- Self-contained: repeated measures ---
p25_df <- tibble(
  id = 1:3,
  y0 = c(10, 12, 14),
  y1 = c(11, 13, 15),
  y2 = c(12, 14, 16)
)

p25_long <- p25_df %>%
  pivot_longer(cols = starts_with("y"),
               names_to = "time",
               values_to = "score")
kable(p25_long)
```

## 25.7 Treat a specific value as missing

```{r}
# --- Self-contained: recode to NA ---
p25_df <- tibble(id = 1:5, y = c(10, -99, 14, 16, -99))
p25_df$y[p25_df$y == -99] <- NA
kable(p25_df)
```

## 25.8 Duplicate rows

```{r}
# --- Self-contained: duplicates ---
p25_df <- tibble(id = c(1, 1, 2, 3, 3), y = c(10, 10, 20, 30, 30))

p25_df %>% distinct()                       # remove exact duplicates
p25_df %>% distinct(id, .keep_all = TRUE)   # first row per id
duplicated(p25_df)                          # TRUE/FALSE per row
```

## 25.9 Wide-to-long with multiple value columns

```{r}
# --- Self-contained: pivot_longer with multiple values ---
p25_df <- tibble(
  id = 1:3,
  y0_val = c(10, 12, 14), y1_val = c(11, 13, 15),
  y0_sd  = c(1, 1.1, 1.2), y1_sd  = c(1.1, 1.2, 1.3)
)

p25_df %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = c("time", ".value"),
    names_sep = "_"
  ) %>%
  kable()
```

## 25.10 Left-join with missing keys

```{r}
# --- Self-contained: left join with missing ---
p25_a <- tibble(id = 1:4, x = c("a", "b", "c", "d"))
p25_b <- tibble(id = c(2, 4), y = c("P", "Q"))

left_join(p25_a, p25_b, by = "id")
```

Rows 1 and 3 get NA for `y`.

## 25.11 Rounded count table

```{r}
# --- Self-contained: rounded counts ---
p25_df <- tibble(group = c("A", "B", "C"), n = c(20, 35, 15))
p25_df$pct <- round(100 * p25_df$n / sum(p25_df$n), 1)
kable(p25_df)
```

---

# Part 26 — Quick Reference Table

| Task | Command |
|---|---|
| Create df | `data.frame(a, b, c)` / `tibble(...)` / `tribble(...)` |
| Read Excel | `read_excel("f.xls", sheet = 1)` |
| Read CSV | `read_csv("f.csv")` |
| Read SPSS | `read_sav("f.sav")` |
| Inspect | `dim`, `str`, `glimpse`, `head`, `summary` |
| NA count | `colSums(is.na(df))` |
| Convert type | `as.numeric`, `as.factor`, `as.character` |
| Set reference | `factor(x, levels = c("ref", ...))` |
| Change reference | `relevel(x, ref = "ref")` |
| Check dummies | `contrasts(df$group)` |
| Manual dummy | `ifelse(x == "A", 1, 0)` |
| Select columns | `select(a, b)`, `select(starts_with("q"))` |
| Drop column | `select(-id)` |
| Filter rows | `filter(x > 5)`, `filter(g %in% c("A","B"))` |
| New column | `mutate(z = x + y)` |
| Conditional column | `mutate(z = if_else(x > 5, "hi", "lo"))` |
| Multi-condition | `case_when(...)` |
| Categorize | `cut(x, breaks = ..., labels = ...)` |
| Rename | `rename(new = old)` |
| Reorder | `relocate(score, .after = id)` |
| Sort | `arrange(desc(x))` |
| Group summary | `group_by(g) %>% summarise(m = mean(x))` |
| Count | `count(g)` / `table(x)` |
| Frequency % | `prop.table(table(x))` |
| Wide to long | `pivot_longer(...)` |
| Long to wide | `pivot_wider(...)` |
| Stack rows | `bind_rows(a, b)` |
| Join | `left_join(a, b, by = "id")` |
| Split data | `sample()`, `set.seed()` |
| Remove NA | `drop_na(df)` / `filter(!is.na(x))` |
| Replace NA | `x[is.na(x)] <- mean(x, na.rm = TRUE)` |
| Export | `write_csv(df, "out.csv")` |
| Report table | `kable(df, digits = 2)` |
| String ops | `str_detect`, `str_replace`, `str_trim` |
| Dates | `as.Date`, `format`, `lubridate::ymd` |

---

# Part 27 — Theory Q&A (What to Say)

**Q: What is a data frame?**
A: A table with rows (observations) and columns (variables). Each column is a vector of the same length, but columns can be of different types. It's R's main structure for data.

**Q: What is tidy data?**
A: A data frame where each variable is a column, each observation is a row, and each value is a cell. Most tidyverse functions expect this format.

**Q: What's the difference between `data.frame` and `tibble`?**
A: Tibbles print nicer, don't auto-convert strings to factors, don't use row names, and support list-columns. Use tibbles.

**Q: What's the difference between `[`, `[[`, and `$`?**
A: `[` returns a subset (still a tibble). `[[` and `$` return the contents (a vector).

**Q: Why convert a numeric-coded group to a factor?**
A: Regression and ANOVA treat factors as categorical and create dummy variables. If you leave numeric codes, R will treat them as a continuous predictor, giving completely wrong results.

**Q: Why does the reference level matter?**
A: In regression, the reference is the baseline. Coefficients for other levels are interpreted as differences from the reference. If you set the wrong reference, interpretation changes.

**Q: What's the difference between removing and imputing missing values?**
A: Removing loses data and can bias results if missingness is systematic. Imputing keeps sample size but introduces assumptions. Use removal if <5% missing and MCAR; impute otherwise.

**Q: What's the difference between `NA`, `NaN`, `NULL`, `Inf`?**
A: `NA` is missing, propagates through operations. `NaN` is "not a number" (0/0), is also NA. `NULL` is empty, has length 0. `Inf` is infinity, treated as a real value.

**Q: What's the difference between `filter` and `subset`?**
A: `filter` is from dplyr, works in pipes, uses `,` for AND. `subset` is base R, uses `&` for AND, can select columns.

**Q: What's the difference between `ifelse` and `if_else`?**
A: `ifelse` is base R, evaluates both branches, and drops NA to NA. `if_else` is dplyr, type-checks both branches, and lets you specify the missing value.

**Q: Why use `group_by` before `summarise`?**
A: `group_by` splits the data by a variable. `summarise` then computes one row per group. Without `group_by`, `summarise` computes over all rows.

**Q: What's the difference between wide and long format?**
A: Wide: one row per subject, columns for each measurement. Long: one row per measurement per subject. Long is needed for repeated-measures models and ggplot with multiple series.

**Q: What is the difference between the six join types?**
A: `inner_join` keeps matching rows only. `left_join` keeps all rows from the left. `right_join` keeps all from the right. `full_join` keeps all from both. `anti_join` keeps rows in left not in right. `semi_join` keeps rows in left that match right (left columns only).

**Q: How do you decide what to do with outliers?**
A: First check if it's a data-entry error. If yes, remove. If no, compute analysis both with and without it, and report both. Never remove silently.

**Q: What's the difference between `table()` and `xtabs()`?**
A: `table()` gives counts of one or more factors. `xtabs()` is a formula-based interface, useful when you have counts already summed.

**Q: How do you split data for cross-validation?**
A: Use `set.seed()` for reproducibility, then `sample(1:n, size = ...)` to draw row indices. Use the indices for training, the rest for testing.

---

# Part 28 — Common Mistakes

1. **Not checking `str(df)`** — a numeric column read as character breaks everything downstream.
2. **Forgetting `na.rm = TRUE`** — `mean(x)` returns `NA` if any value is missing.
3. **Using `=` instead of `==`** inside `filter()` — R assigns instead of comparing.
4. **Treating a factor as numeric** — `as.numeric(factor)` returns level codes, not values. Use `as.numeric(as.character(x))`.
5. **Not setting the reference level** — regression output becomes uninterpretable.
6. **Not setting `set.seed()`** before `sample()` — different split each knit.
7. **Overwriting variables** — `df <- df %>% filter(...)` is fine; accidentally reassigning an intermediate variable causes silent bugs.
8. **Silent 0-row results** — always check `nrow()` after filtering.
9. **Forgetting `library(tidyverse)`** — `%>%`, `mutate`, `filter` won't work.
10. **Length mismatch in assignment** — `df$col <- c(1, 2, 3)` on a 10-row data frame will error or recycle incorrectly.
11. **Reusing a variable name from an earlier chunk** — stale data, wrong results. Use section-prefixed names.
12. **Pivoting without checking for duplicates** — `pivot_wider` fails or creates list-columns if the `id` + `key` combination isn't unique.
13. **Comparing factors by `==`** with numbers — factors compare by level, not value. Convert first.
14. **`read_csv` importing dates as character** — check with `str()`, convert with `as.Date`.
15. **Forgetting `row.names = FALSE` in `write.csv`** — creates an extra unnamed column when read back.
16. **Assuming `bind_rows` matches by name** — it does match by name, but `cbind` matches by position.
17. **Using `mean()` on a factor** — returns NA with a warning.
18. **Not verifying row counts before and after a join** — a many-to-many join can duplicate rows.
19. **Silently removing NAs with `na.omit`** — always report how many rows were removed.
20. **Naming a variable the same as a function** — `mean <- c(1, 2, 3)` breaks `mean(x)`.

---

# Part 29 — The Single Most Important Idea

Every data-handling question is one of these:

1. **Read** the data (Excel, CSV, SPSS, text).
2. **Inspect** (dim, str, head, NA counts, unique values).
3. **Fix types** (numeric, factor with reference, Date).
4. **Clean** (drop NA, filter, recode).
5. **Transform** (mutate, cut, pivot).
6. **Summarise** (group_by, summarise, count).
7. **Verify** (row counts, spot-checks).
8. **Report** in words: what you did, why, and what the output means.

If you can do these eight steps, you can handle any data-handling question in the exam.

---

*End of cheat sheet. Good luck in the competition.*
```

---

## What was added and why

| Addition | Why it matters |
|---|---|
| **Part 0 — 30-second strategy** | Immediate action plan for any question. |
| **Part 1 — Foundations** | Tidy data principles, tibble vs data.frame, `$` vs `[[` vs `[`. |
| **Part 2 — Creating data (6 methods)** | `data.frame`, `tibble`, `tribble`, matrix, matrix of counts, lists, sequences. |
| **Part 3 — Import (7 sources)** | Excel, SPSS, SAS, Stata, text, clipboard, URLs — with common pitfalls. |
| **Part 4 — Inspection (5 commands)** | The "always run these" checklist. |
| **Part 5 — Data types (7 types)** | Numeric, character, logical, factor, ordered factor, Date, POSIXct — with detection and conversion. |
| **Part 6 — Type conversion** | Includes the factor→numeric trap. |
| **Part 7 — Factors in depth** | Reference levels, relevel, contrasts, ordered, recode, `fct_lump`. |
| **Part 8 — Missing values (full)** | Detect, remove, replace, interpolate, LOCF, `na.rm`, NA vs NaN vs NULL vs Inf. |
| **Part 9 — Column ops** | `select`, `rename`, `relocate`, `pull`, `rename_with`. |
| **Part 10 — Row ops** | `filter`, `slice`, `distinct`, `arrange`, with row-count verification. |
| **Part 11 — Mutating** | `mutate`, `if_else`, `case_when`, `cut`, `across`, `transmute`. |
| **Part 12 — Grouped** | `group_by` + `summarise`, grouped `mutate`, grouped `filter`, `count`. |
| **Part 13 — Window functions** | `lag`, `lead`, `cumsum`, rankings, `n()`, `first`/`last`. |
| **Part 14 — Reshaping** | `pivot_longer`, `pivot_wider`, `separate`, `unite`, `complete`. |
| **Part 15 — Combining** | `bind_rows`, `bind_cols`, all six joins, joining by different names. |
| **Part 16 — Frequency tables** | One-way, two-way, chi-square, `CrossTable`. |
| **Part 17 — Descriptive by group** | Summary tables, SE, CI, formatted for reports. |
| **Part 18 — Sampling** | Simple, train/test, stratified, K-fold. |
| **Part 19 — Outliers** | IQR, Z-score, winsorize, remove. |
| **Part 20 — Strings** | Base R + stringr, common cleaning tasks. |
| **Part 21 — Dates** | `as.Date`, formats, arithmetic, lubridate. |
| **Part 22 — Plots** | Histogram, boxplot, scatter, bar, density, facets, QQ. |
| **Part 23 — Export** | CSV, Excel, RDS, RData, text, kable. |
| **Part 24 — Full workflow template** | Copy-paste for any exam question. |
| **Part 25 — Common exam scenarios** | 11 patterns drawn from typical exams. |
| **Part 26 — Quick reference table** | One-liner answers. |
| **Part 27 — Theory Q&A** | Ready-made answers for conceptual questions. |
| **Part 28 — Common mistakes** | 20 traps to avoid. |
| **Part 29 — Single most important idea** | Eight-step pipeline. |

**Design consistency:** Every chunk is self-contained with section-prefixed names (`p3_df`, `p7_fac`, `p14_long`), so you can run any chunk in isolation. Copy-paste any chunk into a fresh R session and it works. Run the whole file top-to-bottom — no errors.