---
title: "The Ultimate Multivariate Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(broom)
library(car)
library(knitr)
```

# How to Use This Cheat Sheet

Every multivariate question falls into one of these. Find your question, jump to the section.

| Question type | Section | Typical phrasing |
|---|---|---|
| Compare two mean vectors | 3 | "Do the two groups differ in mean vector?" |
| Compare 3+ mean vectors | 4 | "MANOVA for group differences" |
| Test multivariate normality | 5 | "Is the data MVN?" |
| Estimate covariance structure | 6 | "Compound symmetry / unstructured" |
| Reduce dimension | 7 | "Run PCA and interpret" |
| Number of components | 8 | "How many components to keep?" |
| Latent factors | 9 | "Factor analysis / Bartlett test" |
| Classify observations | 10 | "Discriminant analysis" |
| Cluster observations | 11 | "Hierarchical / k-means clustering" |
| Correlation between two sets | 12 | "Canonical correlation" |
| Detect multivariate outliers | 13 | "Mahalanobis distance" |
| Check sphericity | 14 | "Test for sphericity" |
| Theory questions | 15 | "What is Hotelling T²?" |
| Exam scenarios | 16 | Mixed |
| Quick reference | 17 | — |
| Common mistakes | 18 | — |

**Golden rule of this document:** every chunk is self-contained.

---

# Part 0 — The 30-Second Exam Strategy

For **any** multivariate question, run this pipeline:

1. **Identify the goal:** compare means, reduce dimensions, find structure, classify, cluster?
2. **Extract the numeric matrix** `X <- as.matrix(df %>% select(numeric_vars))`.
3. **Standardize** if variables have different units (`scale(X)`).
4. **Fit the right method:**
   - Two groups, mean vectors → Hotelling T²
   - 3+ groups, mean vectors → MANOVA
   - Distribution → Mardia's test
   - Dimension reduction → `prcomp` (PCA) or `factanal` (FA)
   - Classification → `MASS::lda` / `MASS::qda`
   - Clustering → `hclust` / `kmeans`
   - Outliers → Mahalanobis distance
5. **Report** eigenvalues, loadings, test statistic, p-value.
6. **Interpret** components/factors by naming them from the loadings.

If you can do these six steps, you can answer any multivariate question.

---

# Part 1 — Foundations

## 1.1 What is multivariate data?

Each observation is a **vector** of $p$ responses: $\mathbf{y}_i = (y_{i1}, \dots, y_{ip})^\top$.

- **Univariate:** one response per unit.
- **Multivariate:** multiple responses per unit, potentially correlated.

**Examples:** measurements on people (height, weight, BP), scores on tests (math, verbal, science), gene expression.

## 1.2 Why not just run $p$ univariate tests?

1. **Inflated Type I error** — $p$ tests inflate family-wise error to $1 - (1-\alpha)^p$.
2. **Ignores correlation** — univariate tests ignore the covariance between responses, losing power.
3. **Doesn't answer the joint question** — "do groups differ on the *vector* of responses?" is different from "do groups differ on *any* response?"

## 1.3 Notation

| Symbol | Meaning |
|---|---|
| $\mathbf{y}_i$ | response vector for unit $i$ ($p \times 1$) |
| $\mathbf{X}$ | $n \times p$ data matrix |
| $\bar{\mathbf{y}}$ | sample mean vector ($p \times 1$) |
| $\mathbf{S}$ | sample covariance matrix ($p \times p$) |
| $\mathbf{R}$ | sample correlation matrix ($p \times p$) |
| $\boldsymbol{\mu}$ | population mean vector |
| $\boldsymbol{\Sigma}$ | population covariance matrix |
| $\mathbf{S}_p$ | pooled covariance matrix |

## 1.4 The sample covariance matrix

$$\mathbf{S} = \frac{1}{n-1} \sum_{i=1}^n (\mathbf{y}_i - \bar{\mathbf{y}})(\mathbf{y}_i - \bar{\mathbf{y}})^\top$$

- Diagonal entries: variances.
- Off-diagonal entries: covariances.
- $\mathbf{R}$ = correlation matrix = $\mathbf{D}^{-1/2} \mathbf{S} \mathbf{D}^{-1/2}$ where $\mathbf{D}$ = diag of $\mathbf{S}$.

## 1.5 Mean vector test (intuition)

To test $\mathbf{H}_0: \boldsymbol{\mu} = \boldsymbol{\mu}_0$ (one-sample):

$$T^2 = n (\bar{\mathbf{y}} - \boldsymbol{\mu}_0)^\top \mathbf{S}^{-1} (\bar{\mathbf{y}} - \boldsymbol{\mu}_0)$$

Converted to F: $F = \frac{n - p}{p (n-1)} T^2 \sim F_{p, n-p}$.

**Interpretation:** Large $T^2$ → mean vector differs from $\boldsymbol{\mu}_0$.

---

# Part 2 — Preparing Multivariate Data

## 2.1 A basic multivariate data frame

```{r}
# --- Self-contained: multivariate data frame ---
set.seed(201)
p2_n  <- 120
p2_df <- tibble(
  id     = 1:p2_n,
  gender = factor(rep(c("M", "F"), each = 60)),
  q1 = rnorm(p2_n, 50, 10),
  q2 = rnorm(p2_n, 55, 12),
  q3 = rnorm(p2_n, 48,  9),
  q4 = rnorm(p2_n, 60, 11),
  q5 = rnorm(p2_n, 52, 10),
  q6 = rnorm(p2_n, 58, 13)
)
head(p2_df)
```

## 2.2 Extract the numeric matrix

```{r}
# --- Self-contained: numeric matrix ---
set.seed(202)
p2_df <- tibble(
  id = 1:100, g = factor(rep(c("A", "B"), each = 50)),
  q1 = rnorm(100), q2 = rnorm(100), q3 = rnorm(100)
)

p2_X <- p2_df %>% select(starts_with("q")) %>% as.matrix()
dim(p2_X)
class(p2_X)
```

**Why:** Almost every multivariate method needs a numeric matrix, not a data frame with factors.

## 2.3 Mean vector and covariance matrix

```{r}
# --- Self-contained: mean and covariance ---
set.seed(203)
p2_X <- matrix(rnorm(60 * 3), ncol = 3)

p2_mean <- colMeans(p2_X)
p2_S    <- cov(p2_X)
p2_R    <- cor(p2_X)

round(p2_mean, 3)
round(p2_S, 3)
round(p2_R, 3)
```

## 2.4 Standardize (when units differ)

```{r}
# --- Self-contained: standardization ---
set.seed(204)
p2_X  <- cbind(height = rnorm(100, 170, 10),
               weight = rnorm(100, 70, 15),
               bmi    = rnorm(100, 24, 3))

p2_Xs <- scale(p2_X)         # mean 0, sd 1
round(colMeans(p2_Xs), 5)
round(apply(p2_Xs, 2, sd), 5)
```

**When to standardize:**
- Variables have different units (cm vs kg).
- You plan to use PCA on the correlation matrix.
- You plan LDA or clustering with mixed scales.

**When not to:** all variables on the same scale and variance differences are meaningful.

## 2.5 Group summary (for comparison designs)

```{r}
# --- Self-contained: group means ---
set.seed(205)
p2_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 30)),
  y1 = c(rnorm(30, 10, 2), rnorm(30, 12, 2), rnorm(30, 15, 2)),
  y2 = c(rnorm(30, 20, 3), rnorm(30, 22, 3), rnorm(30, 25, 3))
)

p2_df %>%
  group_by(group) %>%
  summarise(across(c(y1, y2), mean), .groups = "drop") %>%
  kable(digits = 3)
```

**Report template:** "Group A has mean vector (10.1, 20.2), group B (11.9, 22.1), group C (15.2, 25.0). Group C appears highest on both responses."

---

# Part 3 — Two-Sample Hotelling T²

## 3.1 Model

Test $H_0: \boldsymbol{\mu}_1 = \boldsymbol{\mu}_2$ (two group mean vectors equal).

$$T^2 = \frac{n_1 n_2}{n_1 + n_2} (\bar{\mathbf{y}}_1 - \bar{\mathbf{y}}_2)^\top \mathbf{S}_p^{-1} (\bar{\mathbf{y}}_1 - \bar{\mathbf{y}}_2)$$

where $\mathbf{S}_p = \frac{(n_1-1)\mathbf{S}_1 + (n_2-1)\mathbf{S}_2}{n_1 + n_2 - 2}$ is the pooled covariance.

**F conversion:**
$$F = \frac{n_1 + n_2 - p - 1}{(n_1 + n_2 - 2) p} T^2 \sim F_{p, n_1 + n_2 - p - 1}$$

## 3.2 Manual implementation

```{r}
# --- Self-contained: Hotelling T2 ---
set.seed(301)
p3_df <- tibble(
  group = factor(rep(c("M", "F"), each = 60)),
  q1 = c(rnorm(60, 50, 10), rnorm(60, 53, 10)),
  q2 = c(rnorm(60, 55, 12), rnorm(60, 58, 12)),
  q3 = c(rnorm(60, 48,  9), rnorm(60, 47,  9))
)

p3_X1 <- p3_df %>% filter(group == "M") %>% select(-group) %>% as.matrix()
p3_X2 <- p3_df %>% filter(group == "F") %>% select(-group) %>% as.matrix()

p3_n1 <- nrow(p3_X1); p3_n2 <- nrow(p3_X2); p3_p <- ncol(p3_X1)

p3_xbar1 <- colMeans(p3_X1)
p3_xbar2 <- colMeans(p3_X2)
p3_S1    <- cov(p3_X1)
p3_S2    <- cov(p3_X2)
p3_Sp    <- ((p3_n1 - 1) * p3_S1 + (p3_n2 - 1) * p3_S2) / (p3_n1 + p3_n2 - 2)

p3_diff <- p3_xbar1 - p3_xbar2
p3_T2   <- as.numeric((p3_n1 * p3_n2) / (p3_n1 + p3_n2) *
                      t(p3_diff) %*% solve(p3_Sp) %*% p3_diff)

p3_F  <- (p3_n1 + p3_n2 - p3_p - 1) * p3_T2 /
         ((p3_n1 + p3_n2 - 2) * p3_p)
p3_pv <- 1 - pf(p3_F, p3_p, p3_n1 + p3_n2 - p3_p - 1)

tibble(
  Statistic = c("Hotelling T2", "F", "df1", "df2", "p-value"),
  Value     = round(c(p3_T2, p3_F, p3_p, p3_n1 + p3_n2 - p3_p - 1, p3_pv), 4)
) %>% kable()
```

**Interpretation:** "Hotelling's $T^2 = ..., F(p, df) = ..., p = ...$. Reject/fail to reject $H_0$ at the 5% level."

## 3.3 Which variables drive the difference

After rejecting $H_0$, run univariate t-tests with Bonferroni:

```{r}
# --- Self-contained: univariate follow-up ---
set.seed(302)
p3_df <- tibble(
  group = factor(rep(c("M", "F"), each = 60)),
  q1 = c(rnorm(60, 50, 10), rnorm(60, 53, 10)),
  q2 = c(rnorm(60, 55, 12), rnorm(60, 58, 12)),
  q3 = c(rnorm(60, 48,  9), rnorm(60, 47,  9))
)

p3_df %>%
  pivot_longer(cols = starts_with("q"), names_to = "var", values_to = "value") %>%
  group_by(var) %>%
  summarise(
    t = t.test(value ~ group)$statistic,
    p = t.test(value ~ group)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_bonf = p.adjust(p, method = "bonferroni")) %>%
  kable(digits = 4)
```

**Rule:** Only interpret follow-up t-tests if $T^2$ was significant.

## 3.4 Confidence ellipse for the mean difference

```{r}
# --- Self-contained: confidence ellipse ---
set.seed(303)
p3_df <- tibble(
  group = factor(rep(c("A", "B"), each = 40)),
  q1 = c(rnorm(40, 10, 2), rnorm(40, 12, 2)),
  q2 = c(rnorm(40, 20, 3), rnorm(40, 22, 3))
)

p3_diff <- colMeans(subset(p3_df, group == "A")[, -1]) -
           colMeans(subset(p3_df, group == "B")[, -1])

# 95% CI ellipse parameters (approximate)
p3_S <- cov(subset(p3_df, group == "A")[, -1])
p3_eig <- eigen(p3_S)
round(p3_diff, 3)
round(p3_eig$values, 3)
```

## 3.5 Assumptions

1. **Multivariate normality** in each group.
2. **Equal covariance matrices** across groups (homoscedasticity).
3. **Independence** of observations.

**Test equal covariance:**

```{r}
# --- Self-contained: Box's M test ---
set.seed(304)
p3_df <- tibble(
  group = factor(rep(c("A", "B"), each = 40)),
  q1 = c(rnorm(40, 10, 2), rnorm(40, 12, 2)),
  q2 = c(rnorm(40, 20, 3), rnorm(40, 22, 3))
)

# Box's M via heplots is available, but a simpler check:
p3_S1 <- cov(subset(p3_df, group == "A")[, -1])
p3_S2 <- cov(subset(p3_df, group == "B")[, -1])
round(p3_S1, 3)
round(p3_S2, 3)
# Rule of thumb: max/min variance ratio < 4 -> OK
```

**If equal-covariance fails:** use separate-covariance version (Welch-like):

```{r}
# --- Self-contained: separate-covariance T2 ---
set.seed(305)
p3_df <- tibble(
  group = factor(rep(c("A", "B"), each = 40)),
  q1 = c(rnorm(40, 10, 1), rnorm(40, 12, 4)),
  q2 = c(rnorm(40, 20, 1), rnorm(40, 22, 4))
)

p3_X1 <- subset(p3_df, group == "A")[, -1] %>% as.matrix()
p3_X2 <- subset(p3_df, group == "B")[, -1] %>% as.matrix()

p3_T2_sep <- as.numeric(
  t(colMeans(p3_X1) - colMeans(p3_X2)) %*%
  solve(cov(p3_X1)/nrow(p3_X1) + cov(p3_X2)/nrow(p3_X2)) %*%
  (colMeans(p3_X1) - colMeans(p3_X2))
)
round(p3_T2_sep, 4)
```

**Report template:** "Hotelling's $T^2 = 15.4$, $F(2, 117) = 7.62$, $p < 0.001$. Reject $H_0$; the two groups differ in their mean vector. Follow-up t-tests (Bonferroni) show q1 and q2 differ between groups ($p < 0.001$), while q3 does not."

---

# Part 4 — MANOVA (3+ Groups)

## 4.1 Model

One-way MANOVA tests $H_0: \boldsymbol{\mu}_1 = \boldsymbol{\mu}_2 = \dots = \boldsymbol{\mu}_g$.

**Decomposition:** $\mathbf{T} = \mathbf{B} + \mathbf{W}$ (total = between + within).

**Four test statistics:**

| Statistic | Formula | Distribution |
|---|---|---|
| Wilks' $\Lambda$ | $\det(\mathbf{W}) / \det(\mathbf{T})$ | Exact F for some cases |
| Pillai's trace | $\text{tr}(\mathbf{B}(\mathbf{B}+\mathbf{W})^{-1})$ | Approximate F |
| Hotelling-Lawley | $\text{tr}(\mathbf{W}^{-1}\mathbf{B})$ | Approximate F |
| Roy's largest root | largest eigenvalue of $\mathbf{W}^{-1}\mathbf{B}$ | Upper bound |

**Recommendation:** Use **Wilks' Lambda** for balanced designs; **Pillai's trace** when assumptions are uncertain (most robust).

## 4.2 Fit MANOVA

```{r}
# --- Self-contained: one-way MANOVA ---
set.seed(401)
p4_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 30)),
  y1 = c(rnorm(30, 10, 2), rnorm(30, 12, 2), rnorm(30, 15, 2)),
  y2 = c(rnorm(30, 20, 3), rnorm(30, 22, 3), rnorm(30, 25, 3)),
  y3 = c(rnorm(30, 30, 4), rnorm(30, 28, 4), rnorm(30, 35, 4))
)

p4_fit <- manova(cbind(y1, y2, y3) ~ group, data = p4_df)
summary(p4_fit, test = "Wilks")
summary(p4_fit, test = "Pillai")
summary(p4_fit, test = "Hotelling-Lawley")
summary(p4_fit, test = "Roy")
```

**Output columns:** Df, test statistic, approximate F, num Df, den Df, Pr(>F).

**Report template:** "MANOVA shows a significant effect of group on the joint response vector, Wilks' $\Lambda = 0.61$, $F(6, 170) = 7.8$, $p < 0.001$."

## 4.3 Follow-up univariate ANOVAs

```{r}
# --- Self-contained: follow-up univariate ANOVAs ---
set.seed(402)
p4_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 30)),
  y1 = c(rnorm(30, 10, 2), rnorm(30, 12, 2), rnorm(30, 15, 2)),
  y2 = c(rnorm(30, 20, 3), rnorm(30, 22, 3), rnorm(30, 25, 3)),
  y3 = c(rnorm(30, 30, 4), rnorm(30, 28, 4), rnorm(30, 35, 4))
)
p4_fit <- manova(cbind(y1, y2, y3) ~ group, data = p4_df)
summary.aov(p4_fit)
```

**Report:** "Follow-up univariate ANOVAs show y1 and y2 differ by group (p < 0.01), while y3 does not (p = 0.12)."

## 4.4 Post-hoc on group means (per response)

```{r}
# --- Self-contained: post-hoc on y1 ---
set.seed(403)
p4_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 30)),
  y1 = c(rnorm(30, 10, 2), rnorm(30, 12, 2), rnorm(30, 15, 2))
)

TukeyHSD(aov(y1 ~ group, data = p4_df))
```

## 4.5 Assumptions

1. **MVN** within each group.
2. **Equal covariance matrices** across groups (Box's M test).
3. **Independence.**

**Box's M test:**

```{r}
# --- Self-contained: Box's M ---
# heplots package provides boxM(); if not available, use manual check.
set.seed(404)
p4_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 30)),
  y1 = c(rnorm(30, 10, 2), rnorm(30, 12, 2), rnorm(30, 15, 2)),
  y2 = c(rnorm(30, 20, 3), rnorm(30, 22, 3), rnorm(30, 25, 3))
)

by(p4_df[, c("y1", "y2")], p4_df$group, cov)
```

**If equal covariance fails:** use separate-covariance MANOVA or apply a transformation. In practice, MANOVA is fairly robust if group sizes are equal.

---

# Part 5 — Multivariate Normality

## 5.1 Why it matters

Most multivariate methods (Hotelling T², MANOVA, LDA) assume **multivariate normality (MVN)**. Violation can:
- Inflate Type I error in MANOVA.
- Bias the F approximations.
- Distort Mahalanobis-based outlier detection.

## 5.2 What MVN looks like

$\mathbf{y}_i \sim N_p(\boldsymbol{\mu}, \boldsymbol{\Sigma})$.

Density:
$$f(\mathbf{y}) = \frac{1}{(2\pi)^{p/2} |\boldsymbol{\Sigma}|^{1/2}} \exp\!\left(-\tfrac{1}{2}(\mathbf{y}-\boldsymbol{\mu})^\top \boldsymbol{\Sigma}^{-1} (\mathbf{y}-\boldsymbol{\mu})\right)$$

**Properties:**
- Every linear combination is univariate normal.
- Every subset of variables is MVN.
- Squared Mahalanobis distances follow $\chi^2_p$.

## 5.3 Mardia's test (manual)

**Skewness** and **kurtosis** are tested separately.

$$\hat{b}_{1,p} = \frac{1}{n^2} \sum_{i,j} \left[(\mathbf{y}_i - \bar{\mathbf{y}})^\top \mathbf{S}^{-1} (\mathbf{y}_j - \bar{\mathbf{y}})\right]^3$$
$$\hat{b}_{2,p} = \frac{1}{n} \sum_i \left[(\mathbf{y}_i - \bar{\mathbf{y}})^\top \mathbf{S}^{-1} (\mathbf{y}_i - \bar{\mathbf{y}})\right]^2$$

Under MVN:
- $n \hat{b}_{1,p}/6 \sim \chi^2_{p(p+1)(p+2)/6}$
- $\hat{b}_{2,p} \approx N(p(p+2), \sqrt{8p(p+2)/n})$

```{r}
# --- Self-contained: Mardia's test ---
set.seed(501)
p5_df <- tibble(
  y1 = rnorm(150),
  y2 = rnorm(150),
  y3 = rnorm(150)
)

p5_mardia <- function(X) {
  X <- as.matrix(X)
  n <- nrow(X); p <- ncol(X)
  S <- cov(X); Sinv <- solve(S)
  Xc <- scale(X, center = TRUE, scale = FALSE)
  D <- Xc %*% Sinv %*% t(Xc)

  b1p <- mean(D^3)
  chi_skew <- n * b1p / 6
  df_skew  <- p * (p + 1) * (p + 2) / 6
  p_skew   <- 1 - pchisq(chi_skew, df_skew)

  b2p <- mean(diag(D)^2)
  expected <- p * (p + 2)
  var_kurt <- 8 * p * (p + 2) / n
  z_kurt   <- (b2p - expected) / sqrt(var_kurt)
  p_kurt   <- 2 * (1 - pnorm(abs(z_kurt)))

  tibble(
    Test      = c("Mardia skewness", "Mardia kurtosis"),
    Statistic = round(c(chi_skew, z_kurt), 4),
    df        = c(df_skew, NA),
    p_value   = round(c(p_skew, p_kurt), 4)
  )
}

p5_mardia(p5_df) %>% kable()
```

**Interpretation:**
- **Both $p > 0.05$:** fail to reject MVN; data is consistent with MVN.
- **Skewness $p < 0.05$:** asymmetric — try transformations (log, sqrt).
- **Kurtosis $p < 0.05$:** heavy-tailed or light-tailed.
- **Both fail:** strongly non-normal — use non-parametric alternatives.

## 5.4 Chi-square Q-Q plot for MVN

```{r}
# --- Self-contained: chi-square Q-Q plot ---
set.seed(502)
p5_X <- matrix(rnorm(150 * 3), ncol = 3)

p5_md <- mahalanobis(p5_X, colMeans(p5_X), cov(p5_X))
p5_qq <- qqplot(qchisq(ppoints(length(p5_md)), df = 3),
                sort(p5_md),
                plot.it = FALSE)

ggplot(tibble(theoretical = p5_qq$x, sample = p5_qq$y),
       aes(theoretical, sample)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(title = "Chi-square Q-Q plot for MVN",
       x = "Theoretical chi-square quantiles",
       y = "Squared Mahalanobis distances")
```

**Interpretation:** Points on the red line → MVN. Curvature at the tails → heavy-tailed or light-tailed.

## 5.5 Practical decision

| Situation | Action |
|---|---|
| $n$ large (> 30 per group), mild non-normality | Proceed — MANOVA is robust |
| $n$ small, non-normal | Use permutation test or transformation |
| Skewed data | log/Box-Cox transform then retest |
| Outliers dominate | Remove or use robust methods |

## 5.6 Report template

> "Mardia's test for multivariate normality gives skewness $\chi^2(20) = 22.1$, $p = 0.33$ and kurtosis $z = 0.48$, $p = 0.63$. Both $p$-values exceed 0.05, so the data is consistent with multivariate normality. The chi-square Q-Q plot shows points on the diagonal."

---

# Part 6 — Covariance Structures

## 6.1 Why model the covariance?

Repeated measures and longitudinal data have structured covariance. Common structures:

| Structure | Parameters | Assumption |
|---|---|---|
| **Unstructured (UN)** | $p(p+1)/2$ | No constraint |
| **Compound symmetry (CS)** | 2 | Equal variances, equal covariances |
| **AR(1)** | 2 | Correlations decay with distance |
| **Toeplitz** | $p$ | Constant along diagonals |
| **Diagonal** | $p$ | Zero covariances |

## 6.2 Compound symmetry (CS)

$$\boldsymbol{\Sigma} = \sigma^2 \begin{pmatrix} 1 & \rho & \rho \\ \rho & 1 & \rho \\ \rho & \rho & 1 \end{pmatrix} = \sigma^2 [(1-\rho)\mathbf{I}_p + \rho \mathbf{J}_p]$$

**Two parameters:** $\sigma^2$ (common variance), $\rho$ (common correlation).

**Estimation from sample:**

```{r}
# --- Self-contained: CS estimation ---
set.seed(601)
p6_X <- matrix(rnorm(120 * 4), ncol = 4)

p6_S <- cov(p6_X)
p6_p <- ncol(p6_S)

p6_sigma2_hat <- mean(diag(p6_S))
p6_rho_hat    <- mean(p6_S[lower.tri(p6_S)]) / p6_sigma2_hat

p6_Sigma_hat <- p6_sigma2_hat * ((1 - p6_rho_hat) * diag(p6_p) +
                                 p6_rho_hat * matrix(1, p6_p, p6_p))

round(p6_sigma2_hat, 3)
round(p6_rho_hat, 3)
round(p6_Sigma_hat, 3)
```

**Interpretation:** "Under CS, all variances are estimated as $\hat\sigma^2 = ...$ and all covariances as $\hat\sigma^2 \hat\rho = ...$. The estimated correlation between any two time points is $\hat\rho = ...$."

## 6.3 Testing sphericity

**Sphericity:** the variances of all pairwise differences are equal. Equivalent to CS for the covariance structure of differences.

**Mauchly's test:**
- $H_0$: sphericity holds.
- $p > 0.05$: sphericity is reasonable.
- $p < 0.05$: violated — use Greenhouse-Geisser correction.

**Direct sphericity check for repeated measures:**

```{r}
# --- Self-contained: sphericity check ---
set.seed(602)
p6_df <- expand.grid(
  subject = factor(1:20),
  time    = factor(c("t1", "t2", "t3"))
) %>%
  mutate(y = 10 + as.numeric(time) * 2 +
             rep(rnorm(20, 0, 1), each = 3) + rnorm(60, 0, 1))

# Check sphericity via afex (if available)
# afex::aov_ez(id = "subject", dv = "y", data = p6_df, within = "time")

# Manual: variances of pairwise differences
p6_wide <- p6_df %>%
  pivot_wider(names_from = time, values_from = y)

p6_diffs <- cbind(p6_wide$t1 - p6_wide$t2,
                  p6_wide$t1 - p6_wide$t3,
                  p6_wide$t2 - p6_wide$t3)
apply(p6_diffs, 2, var)
```

**Interpretation:** If the three variances are approximately equal, sphericity holds.

## 6.4 Compound symmetry vs unstructured

Test: likelihood ratio or AIC/BIC comparison.

```{r}
# --- Self-contained: compare CS to UN ---
# Use nlme::gls (base R companion to lme4)
# library(nlme)
# gls_cs <- gls(y ~ time, data = p6_df, correlation = corCompSymm(form = ~ 1 | subject))
# gls_un <- gls(y ~ time, data = p6_df, correlation = corSymm(form = ~ 1 | subject))
# anova(gls_cs, gls_un)
```

**Report template:** "Under compound symmetry, $\hat\sigma^2 = 2.14$ and $\hat\rho = 0.52$. The AIC of the CS model (215.3) is higher than the UN model (212.1), suggesting CS is adequate given the extra parameters."

## 6.5 When to use each structure

| Data type | Recommended |
|---|---|
| Balanced repeated measures, few time points | CS |
| Longitudinal, decreasing correlations | AR(1) |
| Exploratory, few time points | UN |
| Many time points | AR(1) or Toeplitz |
| Independent responses | Diagonal (I) |

---

# Part 7 — Principal Component Analysis

## 7.1 What PCA does

**Goal:** Find a small number of **linear combinations** of the original variables that capture most of the variance.

**PC1** = direction of maximum variance.
**PC2** = direction of maximum variance orthogonal to PC1.
...

$$Z_k = a_{k1} X_1 + a_{k2} X_2 + \dots + a_{kp} X_p$$

where $\mathbf{a}_k$ is the $k$-th eigenvector of the covariance (or correlation) matrix.

## 7.2 Running PCA

```{r}
# --- Self-contained: PCA on standardized data ---
set.seed(701)
p7_n  <- 150
p7_df <- tibble(
  q1 = rnorm(p7_n, 50, 10),
  q2 = rnorm(p7_n, 55, 12),
  q3 = rnorm(p7_n, 48,  9),
  q4 = rnorm(p7_n, 60, 11),
  q5 = rnorm(p7_n, 52, 10),
  q6 = rnorm(p7_n, 58, 13)
)

p7_pca <- prcomp(p7_df, scale. = TRUE)
summary(p7_pca)
```

**Why `scale. = TRUE`:** standardizes each variable to unit variance, so PCA is on the correlation matrix. Use this when variables have different units.

## 7.3 Eigenvalues and variance explained

```{r}
# --- Self-contained: eigenvalues ---
set.seed(702)
p7_df <- tibble(
  q1 = rnorm(150, 50, 10), q2 = rnorm(150, 55, 12),
  q3 = rnorm(150, 48,  9), q4 = rnorm(150, 60, 11),
  q5 = rnorm(150, 52, 10), q6 = rnorm(150, 58, 13)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)

p7_eig <- p7_pca$sdev^2
p7_prop <- p7_eig / sum(p7_eig)
p7_cum  <- cumsum(p7_prop)

tibble(
  PC          = 1:length(p7_eig),
  Eigenvalue  = round(p7_eig, 4),
  PropVar     = round(p7_prop, 4),
  CumProp     = round(p7_cum, 4)
) %>% kable()
```

**Reading the output:**
- Eigenvalue = variance of the component.
- PropVar = fraction of total variance explained.
- CumProp = running total.

## 7.4 How many components to keep?

Three common criteria:

**Kaiser criterion:** keep eigenvalues > 1 (when using correlation matrix).

```{r}
# --- Self-contained: Kaiser ---
set.seed(703)
p7_df <- tibble(
  q1 = rnorm(150), q2 = rnorm(150), q3 = rnorm(150),
  q4 = rnorm(150), q5 = rnorm(150), q6 = rnorm(150)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)
p7_eig <- p7_pca$sdev^2
sum(p7_eig > 1)
```

**Scree plot:** look for the "elbow."

```{r}
# --- Self-contained: scree plot ---
set.seed(704)
p7_df <- tibble(
  q1 = rnorm(150), q2 = rnorm(150), q3 = rnorm(150),
  q4 = rnorm(150), q5 = rnorm(150), q6 = rnorm(150)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)

# Option A: plot eigenvalues manually (most reliable)
p7_eig <- p7_pca$sdev^2
plot(1:length(p7_eig), p7_eig, type = "b",
     xlab = "Component", ylab = "Eigenvalue",
     main = "Scree plot")
abline(h = 1, col = "red", lty = 2)   # Kaiser reference line

# Option B (alternative): base screeplot
# screeplot(p7_pca, type = "lines", main = "Scree plot")
```

**Parallel analysis:** compare to eigenvalues from random data (more rigorous, but more complex).

**Rule of thumb:**
- Keep enough for ≥ 70–80% cumulative variance.
- Kaiser criterion tends to overestimate in large datasets.
- Scree plot is subjective.

## 7.5 Loadings (component equations)

```{r}
# --- Self-contained: loadings ---
set.seed(705)
p7_df <- tibble(
  q1 = rnorm(150), q2 = rnorm(150), q3 = rnorm(150),
  q4 = rnorm(150), q5 = rnorm(150), q6 = rnorm(150)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)
round(p7_pca$rotation, 3)
```

**Reading loadings:**
- Each column = one PC.
- Each entry = weight (coefficient) of that variable in that PC.
- Loadings are the eigenvectors, normalized to unit length.

**Rule of thumb for interpretation:** consider loadings with absolute value > 0.4 as "significant."

**Example interpretation:**
> "PC1 has strong positive loadings on q1, q3, q5 (all > 0.6), which suggests PC1 captures an *overall score* dimension. PC2 has strong positive loadings on q2 and q4 and negative on q6, suggesting a contrast between early and late items."

## 7.6 Naming components

After examining loadings, name each component based on the pattern:

| Loading pattern | Name |
|---|---|
| All positive, similar magnitude | Overall size / general factor |
| Half positive, half negative | Contrast |
| Concentrated on a few variables | Specific factor |
| Alternating signs | Oscillation |

## 7.7 Orthogonality check

PCs are orthogonal by construction:

```{r}
# --- Self-contained: orthogonality ---
set.seed(706)
p7_df <- tibble(
  q1 = rnorm(150), q2 = rnorm(150), q3 = rnorm(150)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)
round(t(p7_pca$rotation) %*% p7_pca$rotation, 5)
```

**Should be the identity matrix.** This confirms the components are uncorrelated.

## 7.8 Component scores

The scores are the coordinates of each observation in the new PC space:

```{r}
# --- Self-contained: scores ---
set.seed(707)
p7_df <- tibble(
  q1 = rnorm(150), q2 = rnorm(150), q3 = rnorm(150)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)
p7_scores <- as_tibble(p7_pca$x)
head(p7_scores)
```

**Properties:**
- Mean = 0 for each PC.
- Variance = eigenvalue.
- Uncorrelated across PCs.

## 7.9 Group comparison on scores

If you have groups, test whether components separate groups:

```{r}
# --- Self-contained: group comparison on PC scores ---
set.seed(708)
p7_df <- tibble(
  group = factor(rep(c("A", "B"), each = 75)),
  q1 = c(rnorm(75, 50, 10), rnorm(75, 55, 10)),
  q2 = c(rnorm(75, 55, 12), rnorm(75, 60, 12)),
  q3 = c(rnorm(75, 48,  9), rnorm(75, 50,  9))
)
p7_pca <- prcomp(p7_df %>% select(-group), scale. = TRUE)
p7_scores <- as_tibble(p7_pca$x) %>% mutate(group = p7_df$group)

t.test(PC1 ~ group, data = p7_scores)
t.test(PC2 ~ group, data = p7_scores)
```

**Report:** "PC1 differs significantly between groups (t = -2.8, p = 0.006), suggesting the groups differ in the direction captured by PC1. PC2 does not differ (p = 0.42)."

## 7.10 Biplot

```{r}
# --- Self-contained: biplot ---
set.seed(709)
p7_df <- tibble(
  q1 = rnorm(100), q2 = rnorm(100), q3 = rnorm(100)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)

biplot(p7_pca)
```

**Reading a biplot:**
- Points = observations.
- Arrows = variables (direction = loading, length = importance).
- Arrows pointing the same direction = positively correlated variables.
- Arrows at right angles = uncorrelated.
- Arrows pointing opposite = negatively correlated.

**ggplot version:**

```{r}
# --- Self-contained: PCA plot with ggplot ---
set.seed(710)
p7_df <- tibble(
  q1 = rnorm(100), q2 = rnorm(100), q3 = rnorm(100)
)
p7_pca <- prcomp(p7_df, scale. = TRUE)
p7_scores <- as_tibble(p7_pca$x)

ggplot(p7_scores, aes(PC1, PC2)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "PCA scores")
```

## 7.11 Covariance vs correlation PCA

| Use | `prcomp` argument |
|---|---|
| Variables have same units and variance matters | `scale. = FALSE` (covariance) |
| Variables have different units | `scale. = TRUE` (correlation) |
| Variables on different scales and no clear priority | `scale. = TRUE` |

**Diagnostic:** if standard deviations differ by a factor > 5, use correlation.

```{r}
# --- Self-contained: SD diagnostic ---
set.seed(711)
p7_df <- tibble(
  a = rnorm(100, 10, 1),
  b = rnorm(100, 100, 30)
)
apply(p7_df, 2, sd)
# Ratio 30/1 = 30 -> use scale. = TRUE
```

## 7.12 Reporting template

> "PCA on the standardized variables gives the following eigenvalues: 2.31, 1.54, 1.08, 0.62, 0.30, 0.15. Using the Kaiser criterion (eigenvalue > 1), we retain 3 components explaining 82.2% of the total variance. PC1 (38.5%) loads strongly and positively on all six items (loadings > 0.55), suggesting a general *achievement* factor. PC2 (25.7%) contrasts the first three items with the last three, suggesting a *verbal vs quantitative* dimension."

---

# Part 8 — Choosing Number of Components (Formal Methods)

## 8.1 Kaiser criterion

```{r}
# --- Self-contained: Kaiser ---
set.seed(801)
p8_X <- matrix(rnorm(200 * 5), ncol = 5)
p8_eig <- prcomp(p8_X, scale. = TRUE)$sdev^2
sum(p8_eig > 1)
```

## 8.2 Broken-stick model

Compare each eigenvalue to the expected value under a null (broken-stick) distribution:

```{r}
# --- Self-contained: broken-stick ---
set.seed(802)
p8_X <- matrix(rnorm(200 * 5), ncol = 5)
p8_eig <- prcomp(p8_X, scale. = TRUE)$sdev^2
p8_p   <- length(p8_eig)

p8_broken <- sapply(1:p8_p, function(k) sum(1 / (k:p8_p)) / p8_p * p8_p)
tibble(PC = 1:p8_p, Eigenvalue = p8_eig,
       BrokenStick = round(p8_broken, 3),
       Keep = p8_eig > p8_broken) %>% kable()
```

**Interpretation:** Keep components whose eigenvalue exceeds the broken-stick value.

## 8.3 Cumulative variance threshold

```{r}
# --- Self-contained: cumulative threshold ---
set.seed(803)
p8_X <- matrix(rnorm(200 * 5), ncol = 5)
p8_pca <- prcomp(p8_X, scale. = TRUE)
p8_cum <- cumsum(p8_pca$sdev^2 / sum(p8_pca$sdev^2))
min(which(p8_cum >= 0.80))
```

**Rule of thumb:** 70–80% cumulative variance.

## 8.4 Parallel analysis (manual approximation)

Compare to eigenvalues from random data of the same size:

```{r}
# --- Self-contained: parallel analysis (approximation) ---
set.seed(804)
p8_X <- matrix(rnorm(200 * 5), ncol = 5)
p8_eig <- prcomp(p8_X, scale. = TRUE)$sdev^2

# Simulate random data eigenvalues
p8_rand <- replicate(100, {
  Xr <- matrix(rnorm(200 * 5), ncol = 5)
  prcomp(Xr, scale. = TRUE)$sdev^2
})
p8_rand_mean <- rowMeans(p8_rand)

tibble(
  PC = 1:5,
  Real = round(p8_eig, 3),
  Random = round(p8_rand_mean, 3),
  Keep = p8_eig > p8_rand_mean
) %>% kable()
```

**Rule:** Keep components where real eigenvalue > mean random eigenvalue.

---

# Part 9 — Factor Analysis

## 9.1 What FA does

**Goal:** Explain correlations among observed variables using a **smaller number of latent factors**.

Model: $\mathbf{x} = \boldsymbol{\Lambda} \mathbf{f} + \boldsymbol{\varepsilon}$
- $\mathbf{x}$: observed vector ($p \times 1$).
- $\mathbf{f}$: latent factors ($k \times 1$, $k \ll p$).
- $\boldsymbol{\Lambda}$: loadings matrix ($p \times k$).
- $\boldsymbol{\varepsilon}$: unique errors.

**PCA vs FA:**

| | PCA | FA |
|---|---|---|
| Goal | Maximize variance | Explain covariance |
| Input | Covariance/correlation matrix | Covariance/correlation matrix |
| Output | Components (observed) | Factors (latent) |
| Number | Determined by data | Chosen by theory |

## 9.2 Fit factor analysis

```{r}
# --- Self-contained: factor analysis ---
set.seed(901)
p9_n <- 200
p9_f1 <- rnorm(p9_n); p9_f2 <- rnorm(p9_n)

p9_df <- tibble(
  q1 = p9_f1 + rnorm(p9_n, 0, 0.5),
  q2 = p9_f1 + rnorm(p9_n, 0, 0.5),
  q3 = p9_f1 + rnorm(p9_n, 0, 0.5),
  q4 = p9_f2 + rnorm(p9_n, 0, 0.5),
  q5 = p9_f2 + rnorm(p9_n, 0, 0.5),
  q6 = p9_f2 + rnorm(p9_n, 0, 0.5)
)

p9_fit <- factanal(p9_df, factors = 2, rotation = "varimax")
print(p9_fit)
```

## 9.3 Reading the output

| Field | Meaning |
|---|---|
| `Uniquenesses` | Variance of each variable NOT explained by factors |
| `Loadings` | Correlation of each variable with each factor |
| `SS loadings` | Sum of squared loadings per factor |
| `Proportion Var` | Fraction of total variance explained by each factor |
| `Cumulative Var` | Running total |
| `Test of the hypothesis` | Tests whether k factors are sufficient |

**Rule for interpretation:** consider loadings > 0.4 (or < -0.4) as "significant."

## 9.4 Choosing the number of factors

```{r}
# --- Self-contained: try 1 to 3 factors ---
set.seed(902)
p9_n <- 200
p9_f1 <- rnorm(p9_n); p9_f2 <- rnorm(p9_n)
p9_df <- tibble(
  q1 = p9_f1 + rnorm(p9_n, 0, 0.5),
  q2 = p9_f1 + rnorm(p9_n, 0, 0.5),
  q3 = p9_f1 + rnorm(p9_n, 0, 0.5),
  q4 = p9_f2 + rnorm(p9_n, 0, 0.5),
  q5 = p9_f2 + rnorm(p9_n, 0, 0.5),
  q6 = p9_f2 + rnorm(p9_n, 0, 0.5)
)

# Only try k values where df >= 0 (df = ((p-k)^2 - (p+k)) / 2)
p9_p <- ncol(p9_df)
p9_max_k <- floor((2 * p9_p + 1 - sqrt(8 * p9_p + 1)) / 2)   # largest feasible k

for (k in 1:p9_max_k) {
  fit <- tryCatch(
    factanal(p9_df, factors = k, rotation = "varimax"),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    cat("k =", k, "  model failed\n")
  } else if (is.null(fit$PVAL)) {
    cat("k =", k, "  df =", fit$dof,
        "  just-identified (no p-value)\n")
  } else {
    cat("k =", k, "  df =", fit$dof,
        "  p-value =", round(fit$PVAL, 4), "\n")
  }
}
```

**Rule:** Choose the **smallest k** with $p > 0.05$ (factor model fits).

## 9.5 Rotation

| Rotation | When |
|---|---|
| `"varimax"` | Factors should be uncorrelated (orthogonal) |
| `"promax"` | Factors may be correlated |
| `"none"` | Exploratory only |

**Varimax** is the default for most applications.

## 9.6 Bartlett's test of sphericity

**Question:** Are the variables correlated enough to justify factor analysis?

$$H_0: \mathbf{R} = \mathbf{I} \text{ (identity correlation matrix)}$$

$$\chi^2 = -\left(n - 1 - \frac{2p+5}{6}\right) \log \det(\mathbf{R}) \sim \chi^2_{p(p-1)/2}$$

```{r}
# --- Self-contained: Bartlett's sphericity ---
set.seed(903)
p9_df <- tibble(
  q1 = rnorm(100), q2 = rnorm(100), q3 = rnorm(100),
  q4 = rnorm(100), q5 = rnorm(100)
)

p9_R <- cor(p9_df)
p9_n <- nrow(p9_df)
p9_p <- ncol(p9_df)
p9_chi2 <- -(p9_n - 1 - (2 * p9_p + 5) / 6) * log(det(p9_R))
p9_df_chi <- p9_p * (p9_p - 1) / 2
p9_pv <- 1 - pchisq(p9_chi2, p9_df_chi)

tibble(
  Test    = "Bartlett sphericity",
  Chi_sq  = round(p9_chi2, 4),
  df      = p9_df_chi,
  p_value = round(p9_pv, 4)
) %>% kable()
```

**Interpretation:**
- $p < 0.05$: correlations are large enough → factor analysis justified.
- $p > 0.05$: correlations too small → don't run FA.

## 9.7 KMO (Kaiser-Meyer-Olkin) — sampling adequacy

Manual calculation:

```{r}
# --- Self-contained: KMO ---
set.seed(904)
p9_df <- tibble(
  q1 = rnorm(100), q2 = rnorm(100), q3 = rnorm(100),
  q4 = rnorm(100), q5 = rnorm(100)
)

p9_kmo <- function(X) {
  R <- cor(X)
  Rinv <- solve(R)
  # Partial correlations
  D <- diag(1 / sqrt(diag(Rinv)))
  P <- -D %*% Rinv %*% D
  diag(P) <- 0
  # Off-diagonal sums
  R2 <- R^2; diag(R2) <- 0
  P2 <- P^2
  sum(R2) / (sum(R2) + sum(P2))
}
round(p9_kmo(p9_df), 3)
```

**Benchmark:**

| KMO | Adequacy |
|---|---|
| ≥ 0.90 | Marvelous |
| 0.80–0.89 | Meritorious |
| 0.70–0.79 | Middling |
| 0.60–0.69 | Mediocre |
| 0.50–0.59 | Miserable |
| < 0.50 | Unacceptable |

## 9.8 Reporting template

> "Bartlett's test of sphericity gives $\chi^2(15) = 92.4$, $p < 0.001$, indicating the correlation matrix is not identity. The KMO measure of sampling adequacy is 0.71 (middling). Factor analysis with $k = 2$ factors gives $p = 0.34$ for the model test, so 2 factors are sufficient. After varimax rotation, factor 1 loads strongly on q1, q2, q3 (loadings > 0.6) and factor 2 on q4, q5, q6 (loadings > 0.6). We label factor 1 as *verbal ability* and factor 2 as *quantitative ability*."

---

# Part 10 — Discriminant Analysis (LDA, QDA)

## 10.1 What it does

**Goal:** Find linear combinations of predictors that best separate known groups, and use them to classify new observations.

**Two methods:**
- **Linear Discriminant Analysis (LDA):** assumes equal covariance across groups.
- **Quadratic Discriminant Analysis (QDA):** allows unequal covariance.

## 10.2 Fit LDA

```{r}
# --- Self-contained: LDA ---
set.seed(1001)
p10_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 50)),
  x1 = c(rnorm(50, 0, 1), rnorm(50, 3, 1), rnorm(50, 6, 1)),
  x2 = c(rnorm(50, 0, 1), rnorm(50, 1, 1), rnorm(50, 2, 1))
)

p10_lda <- MASS::lda(group ~ x1 + x2, data = p10_df)
p10_lda
```

**Output:**
- `Prior probabilities of groups`: assumed prior (equal or sample-based).
- `Group means`: mean of each predictor per group.
- `Coefficients of linear discriminants`: LD1, LD2 = linear combinations of predictors.
- `Proportion of trace`: variance explained by each LD.

## 10.3 Predictions and confusion matrix

```{r}
# --- Self-contained: LDA predictions ---
set.seed(1002)
p10_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 50)),
  x1 = c(rnorm(50, 0, 1), rnorm(50, 3, 1), rnorm(50, 6, 1)),
  x2 = c(rnorm(50, 0, 1), rnorm(50, 1, 1), rnorm(50, 2, 1))
)
p10_lda <- MASS::lda(group ~ x1 + x2, data = p10_df)

p10_pred <- predict(p10_lda)
table(Predicted = p10_pred$class, Actual = p10_df$group)

mean(p10_pred$class == p10_df$group)   # accuracy
```

**Report:** "LDA correctly classifies X of 150 observations (accuracy 0.XX). The confusion matrix shows the most confusion is between groups B and C."

## 10.4 QDA

```{r}
# --- Self-contained: QDA ---
set.seed(1003)
p10_df <- tibble(
  group = factor(rep(c("A", "B"), each = 50)),
  x1 = c(rnorm(50, 0, 1), rnorm(50, 3, 2)),
  x2 = c(rnorm(50, 0, 1), rnorm(50, 1, 2))
)

p10_qda <- MASS::qda(group ~ x1 + x2, data = p10_df)
p10_pred_qda <- predict(p10_qda)
table(Predicted = p10_pred_qda$class, Actual = p10_df$group)
```

**When to use QDA:** group covariances clearly differ (Box's M significant). But QDA has more parameters — needs more data.

## 10.5 Testing LDA assumptions

**Equal group covariances (Box's M):**

```{r}
# --- Self-contained: group covariance check ---
set.seed(1004)
p10_df <- tibble(
  group = factor(rep(c("A", "B"), each = 50)),
  x1 = c(rnorm(50, 0, 1), rnorm(50, 3, 1)),
  x2 = c(rnorm(50, 0, 1), rnorm(50, 1, 1))
)

by(p10_df[, c("x1", "x2")], p10_df$group, cov)
```

**If covariances differ substantially:** use QDA.

## 10.6 Reporting template

> "LDA on x1 and x2 produces two discriminant functions (LD1, LD2). LD1 explains 92% of the between-group variance and primarily contrasts group C against A and B (coefficient on x1 = 0.92). Leave-one-out cross-validation gives 94% classification accuracy."

---

# Part 11 — Cluster Analysis

## 11.1 Hierarchical clustering

**Goal:** Group observations into clusters by successive merging/splitting.

**Steps:**
1. Compute pairwise distances.
2. Use a linkage method to merge clusters.
3. Cut the dendrogram at a chosen height.

**Linkage methods:**
- **Complete:** max distance between clusters.
- **Average:** mean distance.
- **Single:** min distance (chaining).
- **Ward:** minimizes within-cluster variance (recommended).

```{r}
# --- Self-contained: hierarchical clustering ---
set.seed(1101)
p11_X <- rbind(
  matrix(rnorm(50 * 2, mean = 0), ncol = 2),
  matrix(rnorm(50 * 2, mean = 5), ncol = 2)
)

p11_d  <- dist(p11_X)
p11_hc <- hclust(p11_d, method = "ward.D2")

plot(p11_hc, main = "Dendrogram",
     xlab = "", ylab = "Distance", labels = FALSE)
```

**Cut the tree:**

```{r}
# --- Self-contained: cut dendrogram ---
set.seed(1102)
p11_X <- rbind(
  matrix(rnorm(50 * 2, mean = 0), ncol = 2),
  matrix(rnorm(50 * 2, mean = 5), ncol = 2)
)
p11_hc <- hclust(dist(p11_X), method = "ward.D2")
p11_clusters <- cutree(p11_hc, k = 2)
table(p11_clusters)
```

## 11.2 K-means clustering

**Requires the number of clusters $k$ in advance.**

```{r}
# --- Self-contained: k-means ---
set.seed(1103)
p11_X <- rbind(
  matrix(rnorm(50 * 2, mean = 0), ncol = 2),
  matrix(rnorm(50 * 2, mean = 5), ncol = 2)
)

p11_km <- kmeans(p11_X, centers = 2, nstart = 25)
p11_km$centers
table(p11_km$cluster)
```

**Choose $k$ via the elbow method:**

```{r}
# --- Self-contained: elbow plot ---
set.seed(1104)
p11_X <- rbind(
  matrix(rnorm(50 * 2, mean = 0), ncol = 2),
  matrix(rnorm(50 * 2, mean = 5), ncol = 2)
)

p11_wss <- sapply(1:8, function(k) kmeans(p11_X, k, nstart = 25)$tot.withinss)
plot(1:8, p11_wss, type = "b",
     xlab = "Number of clusters", ylab = "Total within-SS",
     main = "Elbow plot")
```

**Look for the "elbow" — where the curve flattens.**

## 11.3 Silhouette coefficient

**Average silhouette width** measures cluster quality. Higher is better.

```{r}
# --- Self-contained: silhouette (manual for k=2) ---
set.seed(1105)
p11_X <- rbind(
  matrix(rnorm(50 * 2, mean = 0), ncol = 2),
  matrix(rnorm(50 * 2, mean = 5), ncol = 2)
)
p11_d  <- dist(p11_X)
p11_km <- kmeans(p11_X, centers = 2, nstart = 25)
p11_cl <- p11_km$cluster

# For each point: (b - a) / max(a, b), where a = mean dist to own cluster, b = min mean dist to other
p11_dmat <- as.matrix(p11_d)
p11_sil <- sapply(1:nrow(p11_X), function(i) {
  own <- which(p11_cl == p11_cl[i])
  a <- mean(p11_dmat[i, own[own != i]])
  other <- which(p11_cl != p11_cl[i])
  b <- min(sapply(unique(p11_cl[other]), function(c) mean(p11_dmat[i, other[p11_cl[other] == c]])))
  (b - a) / max(a, b)
})
mean(p11_sil)
```

**Benchmark:** > 0.5 strong structure, 0.25–0.5 moderate, < 0.25 weak.

## 11.4 Comparing methods

| Method | Strength | Weakness |
|---|---|---|
| Hierarchical (Ward) | Visual dendrogram, no k in advance | Greedy merges |
| K-means | Fast, scalable | Need k in advance, spherical clusters |
| DBSCAN | Finds arbitrary shapes, handles outliers | Sensitive to eps |
| Model-based | Probabilistic | Assumes distribution |

## 11.5 Reporting template

> "Ward's hierarchical clustering on standardized variables suggests 3 clusters based on the dendrogram (cut at height = 3.5). K-means with k = 3 gives similar cluster assignments. The average silhouette width is 0.52, indicating strong cluster structure. Cluster 1 has high values on x1 and x2, cluster 2 has low values, and cluster 3 has moderate values."

---

# Part 12 — Canonical Correlation Analysis (CCA)

## 12.1 What it does

**Goal:** Find the maximum correlation between two **sets** of variables.

Given set $X$ (variables $x_1, \dots, x_p$) and set $Y$ (variables $y_1, \dots, y_q$):
- Find linear combination $\mathbf{a}^\top \mathbf{x}$ and $\mathbf{b}^\top \mathbf{y}$ with maximum correlation.
- Repeat for subsequent pairs (orthogonal to previous).

## 12.2 Fit CCA

```{r}
# --- Self-contained: CCA ---
set.seed(1201)
p12_n <- 100
p12_df <- tibble(
  x1 = rnorm(p12_n), x2 = rnorm(p12_n),
  y1 = rnorm(p12_n), y2 = rnorm(p12_n), y3 = rnorm(p12_n)
)
# Induce some correlation
p12_df$y1 <- p12_df$y1 + 0.7 * p12_df$x1

p12_X <- p12_df %>% select(x1, x2) %>% as.matrix()
p12_Y <- p12_df %>% select(y1, y2, y3) %>% as.matrix()

p12_cca <- cancor(p12_X, p12_Y)
p12_cca$cor          # canonical correlations
p12_cca$xcoef        # X coefficients
p12_cca$ycoef        # Y coefficients
```

## 12.3 Test significance

**Wilks' Lambda** tests whether the canonical correlations beyond the first are zero:

```{r}
# --- Self-contained: CCA significance ---
set.seed(1202)
p12_n <- 100
p12_X <- matrix(rnorm(p12_n * 2), ncol = 2)
p12_Y <- matrix(rnorm(p12_n * 3), ncol = 3)
p12_Y[, 1] <- p12_Y[, 1] + 0.7 * p12_X[, 1]

p12_cca <- cancor(p12_X, p12_Y)

# Wilks Lambda for the full set
p12_p <- ncol(p12_X); p12_q <- ncol(p12_Y)
p12_r <- length(p12_cca$cor)
p12_Lambda <- prod(1 - p12_cca$cor^2)
p12_chi2   <- -(p12_n - 1 - (p12_p + p12_q + 1) / 2) * log(p12_Lambda)
p12_df     <- p12_p * p12_q
p12_pv     <- 1 - pchisq(p12_chi2, p12_df)

tibble(
  Canonical_Correlation = round(p12_cca$cor, 3),
  Wilks_Lambda          = round(p12_Lambda, 4),
  Chi_sq                = round(p12_chi2, 4),
  df                    = p12_df,
  p_value               = round(p12_pv, 4)
) %>% kable()
```

**Interpretation:** If $p < 0.05$, at least one canonical correlation is non-zero.

## 12.4 Reporting template

> "CCA between the (x1, x2) set and the (y1, y2, y3) set yields canonical correlations 0.72, 0.31, 0.08. Wilks' $\Lambda = 0.43$, $\chi^2(6) = 79.2$, $p < 0.001$. The first canonical variate pair is highly correlated (r = 0.72). The X-variate loads primarily on x1 and the Y-variate on y1, indicating that the strongest relationship is between x1 and y1."

---

# Part 13 — Multivariate Outliers

## 13.1 Mahalanobis distance

For observation $\mathbf{y}_i$:
$$D_i^2 = (\mathbf{y}_i - \bar{\mathbf{y}})^\top \mathbf{S}^{-1} (\mathbf{y}_i - \bar{\mathbf{y}})$$

Under MVN, $D_i^2 \sim \chi^2_p$. Values exceeding $\chi^2_{p, 1-\alpha}$ are outliers.

```{r}
# --- Self-contained: Mahalanobis outliers ---
set.seed(1301)
p13_X <- matrix(rnorm(100 * 3), ncol = 3)
p13_X[5, ] <- c(10, 10, 10)   # inject an outlier
p13_X[80, ] <- c(-8, -8, -8)  # inject another

p13_md <- mahalanobis(p13_X, colMeans(p13_X), cov(p13_X))
p13_threshold <- qchisq(0.975, df = ncol(p13_X))
which(p13_md > p13_threshold)
```

**Report template:** "Two observations (5 and 80) exceed the $\chi^2_3$ 97.5% threshold of 9.35, indicating multivariate outliers. Removing them reduces the covariance condition number from X to Y."

## 13.2 Robust covariance

If outliers are present, use a robust covariance matrix (e.g., MCD). Not in base R, but a simple approach is to compute MD iteratively:

```{r}
# --- Self-contained: iterative MD removal ---
set.seed(1302)
p13_X <- matrix(rnorm(100 * 3), ncol = 3)
p13_X[5, ] <- c(10, 10, 10)

p13_Xc <- p13_X
repeat {
  md <- mahalanobis(p13_Xc, colMeans(p13_Xc), cov(p13_Xc))
  thr <- qchisq(0.975, df = ncol(p13_Xc))
  if (all(md <= thr)) break
  p13_Xc <- p13_Xc[md <= thr, , drop = FALSE]
}
nrow(p13_X) - nrow(p13_Xc)
```

## 13.3 Diagnostic plot

```{r}
# --- Self-contained: MD plot ---
set.seed(1303)
p13_X <- matrix(rnorm(100 * 3), ncol = 3)
p13_X[5, ] <- c(10, 10, 10)

p13_md <- mahalanobis(p13_X, colMeans(p13_X), cov(p13_X))
p13_thr <- qchisq(0.975, df = ncol(p13_X))

ggplot(tibble(idx = 1:length(p13_md), md = p13_md),
       aes(idx, md)) +
  geom_point() +
  geom_hline(yintercept = p13_thr, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Mahalanobis distances", y = "Squared MD")
```

---

# Part 14 — Testing Sphericity and Compound Symmetry

## 14.1 Sphericity test (Mauchly)

Manual approximation via Box's epsilon:

```{r}
# --- Self-contained: sphericity check ---
set.seed(1401)
p14_df <- expand.grid(
  subject = factor(1:20),
  time    = factor(c("t1", "t2", "t3", "t4"))
) %>%
  mutate(y = rnorm(80) + as.numeric(time) * 0.5)

p14_wide <- p14_df %>%
  pivot_wider(names_from = time, values_from = y) %>%
  select(-subject) %>%
  as.matrix()

p14_S <- cov(p14_wide)
p14_p <- ncol(p14_S)

# Box's epsilon (lower bound for Greenhouse-Geisser)
p14_Strace <- sum(diag(p14_S)) / p14_p
p14_Sbar   <- p14_S - diag(p14_Strace, p14_p)   # contrast covariance
p14_eps    <- (sum(diag(p14_Sbar))^2) / ((p14_p - 1) * sum(p14_Sbar^2))
round(p14_eps, 3)
```

**Interpretation:** $\epsilon = 1$ → perfect sphericity. $\epsilon \geq 0.75$ → okay. $\epsilon < 0.75$ → use Greenhouse-Geisser correction.

## 14.2 Compound symmetry test

Test CS by comparing the sample covariance matrix to the CS structure:

```{r}
# --- Self-contained: CS test ---
set.seed(1402)
p14_wide <- matrix(rnorm(100 * 4), ncol = 4)
p14_S <- cov(p14_wide)
p14_p <- ncol(p14_S)

p14_sigma2 <- mean(diag(p14_S))
p14_rho    <- mean(p14_S[lower.tri(p14_S)]) / p14_sigma2
p14_CS     <- p14_sigma2 * ((1 - p14_rho) * diag(p14_p) +
                             p14_rho * matrix(1, p14_p, p14_p))

round(p14_S, 3)
round(p14_CS, 3)
```

**Visually compare** — if the sample matrix looks like CS, it's adequate.

---

# Part 15 — Theory Q&A

**Q: What is multivariate data?**
A: Data where each unit has multiple correlated responses. Analysis treats the responses jointly to account for correlation.

**Q: Why not use multiple univariate tests?**
A: Inflated Type I error (family-wise), ignores covariance structure, doesn't answer the joint hypothesis.

**Q: What is Hotelling's T²?**
A: The multivariate generalization of the t-statistic. Tests whether two mean vectors differ. Converted to an F-statistic for exact p-values.

**Q: What is Wilks' Lambda?**
A: A test statistic for MANOVA. $\Lambda = \det(\mathbf{W}) / \det(\mathbf{T})$ (within/total). Smaller $\Lambda$ → more group separation.

**Q: When do you use MANOVA vs multiple ANOVAs?**
A: MANOVA tests whether the **mean vector** differs across groups, accounting for correlations. Use MANOVA when responses are correlated and you want to control Type I error across the joint response. Follow up with univariate ANOVAs to identify which response drives the effect.

**Q: What is multivariate normality?**
A: $\mathbf{y}_i \sim N_p(\boldsymbol{\mu}, \boldsymbol{\Sigma})$. Implies all linear combinations are univariate normal and squared Mahalanobis distances follow $\chi^2_p$.

**Q: What is Mardia's test?**
A: Tests MVN via multivariate skewness and kurtosis. Skewness tests asymmetry; kurtosis tests tail weight. Both p-values should exceed 0.05 for MVN.

**Q: What is compound symmetry?**
A: A covariance structure where all variances are equal and all covariances are equal. $\Sigma = \sigma^2 [(1-\rho)I + \rho J]$. Common in repeated-measures ANOVA.

**Q: What is sphericity?**
A: The variances of all pairwise differences are equal. Required for univariate repeated-measures ANOVA. Tested by Mauchly's test. If violated, use Greenhouse-Geisser.

**Q: What is PCA?**
A: A dimension-reduction technique that finds orthogonal linear combinations of the original variables that maximize variance. The components are uncorrelated.

**Q: PCA vs factor analysis?**
A: PCA creates observed components that maximize variance. FA creates latent factors that explain correlations among observed variables. Use PCA for data reduction, FA for theory testing.

**Q: How do you choose the number of components?**
A: Kaiser criterion (eigenvalue > 1), scree plot elbow, cumulative variance (≥70–80%), parallel analysis, broken-stick.

**Q: What do loadings mean?**
A: Correlations between variables and components. Large absolute values (> 0.4) indicate which variables are important for that component.

**Q: What is Bartlett's test of sphericity?**
A: Tests whether the correlation matrix is the identity. If yes, no correlations → don't run factor analysis. p < 0.05 means sphericity is rejected → correlations exist → FA justified.

**Q: What is KMO?**
A: Kaiser-Meyer-Olkin measure of sampling adequacy. Compares simple correlations to partial correlations. Values ≥ 0.6 usually acceptable for factor analysis.

**Q: What is LDA?**
A: Linear discriminant analysis. Finds linear combinations of predictors that maximize between-group separation. Assumes equal covariance across groups.

**Q: LDA vs QDA?**
A: LDA assumes equal covariance; QDA allows unequal. QDA has more parameters — needs more data.

**Q: What is clustering?**
A: Unsupervised grouping of similar observations. Hierarchical clustering builds a dendrogram; k-means partitions into $k$ clusters.

**Q: What is canonical correlation?**
A: Correlation between linear combinations of two sets of variables. Extends correlation to the multivariate case.

**Q: What is Mahalanobis distance?**
A: A measure of how far an observation is from the mean, accounting for covariance. Used for multivariate outlier detection.

**Q: Why does standardization matter?**
A: PCA and clustering are sensitive to scale. Variables with large variance dominate unless standardized (correlation matrix).

**Q: What is the relationship between Hotelling T² and F?**
A: $F = \frac{n-p}{p(n-1)} T^2 \sim F_{p, n-p}$ for the one-sample case. For two-sample: $F = \frac{n_1+n_2-p-1}{(n_1+n_2-2)p} T^2$.

---

# Part 16 — Exam Scenarios

## 16.1 "Are the two groups different on all responses jointly?"

1. Extract numeric matrix, split by group.
2. Compute Hotelling $T^2$ (Part 3.2).
3. Convert to F, compute p.
4. If significant, do univariate follow-ups with Bonferroni.

## 16.2 "MANOVA with 3+ groups"

1. `manova(cbind(y1, y2, ...) ~ group, data = df)`.
2. Report Wilks' Lambda, approximate F, p.
3. Follow up with `summary.aov(fit)`.

## 16.3 "Test multivariate normality"

1. Run Mardia's manual test (Part 5.3).
2. Report skewness and kurtosis p-values.
3. Draw chi-square Q-Q plot.

## 16.4 "Run PCA and interpret"

1. Extract numeric matrix.
2. `prcomp(X, scale. = TRUE)`.
3. Show eigenvalues, proportion of variance, cumulative.
4. Show loadings, name each component.
5. Show scores plot (if groups are present).

## 16.5 "How many components?"

1. Kaiser: `sum(eigenvalues > 1)`.
2. Cumulative: `min(which(cumsum(prop) >= 0.80))`.
3. Scree plot elbow.
4. Choose the smallest consistent answer.

## 16.6 "Factor analysis with k factors"

1. Bartlett's sphericity to justify FA.
2. KMO to check adequacy.
3. `factanal(df, factors = k, rotation = "varimax")`.
4. Report loadings, name factors.
5. Choose smallest $k$ with model p > 0.05.

## 16.7 "Discriminant analysis"

1. `MASS::lda(group ~ x1 + x2, data = df)`.
2. Report group means, discriminant coefficients, proportion of trace.
3. Confusion matrix via `predict()`.

## 16.8 "Cluster the observations"

1. Standardize variables.
2. Hierarchical clustering with Ward's method.
3. Inspect dendrogram, choose $k$.
4. Confirm with k-means.
5. Report cluster sizes and centroids.

## 16.9 "Detect outliers"

1. Compute squared Mahalanobis distances.
2. Threshold at $\chi^2_{p, 0.975}$.
3. Report which observations exceed the threshold.
4. Consider removing or using robust methods.

## 16.10 "Test compound symmetry / sphericity"

1. Extract wide-format data.
2. Compute sample covariance.
3. Compare diagonal entries (variances) and off-diagonals (covariances).
4. If approximately equal → CS holds.

---

# Part 17 — Quick Reference

| Task | Command |
|---|---|
| Numeric matrix | `as.matrix(df %>% select(numeric_vars))` |
| Mean vector | `colMeans(X)` |
| Covariance | `cov(X)` |
| Correlation | `cor(X)` |
| Standardize | `scale(X)` |
| Hotelling T² | manual (Part 3.2) |
| MANOVA | `manova(cbind(y1, y2) ~ group, data = df)` |
| Follow-up ANOVAs | `summary.aov(manova_fit)` |
| Mardia's test | manual function (Part 5.3) |
| Chi-square Q-Q | `qqplot(qchisq(ppoints(n), df = p), sort(md))` |
| CS covariance | mean of diag, mean of off-diag (Part 6.2) |
| PCA | `prcomp(df, scale. = TRUE)` |
| Eigenvalues | `pca$sdev^2` |
| Prop. variance | `pca$sdev^2 / sum(pca$sdev^2)` |
| Loadings | `pca$rotation` |
| Scores | `pca$x` |
| Scree plot | `plot(pca, type = "l")` |
| Biplot | `biplot(pca)` |
| Kaiser | `sum(eigenvalues > 1)` |
| Broken-stick | manual (Part 8.2) |
| Parallel analysis | manual (Part 8.4) |
| Factor analysis | `factanal(df, factors = k, rotation = "varimax")` |
| Bartlett sphericity | manual (Part 9.6) |
| KMO | manual (Part 9.7) |
| LDA | `MASS::lda(group ~ ., data = df)` |
| QDA | `MASS::qda(group ~ ., data = df)` |
| Hierarchical cluster | `hclust(dist(X), method = "ward.D2")` |
| K-means | `kmeans(X, centers = k, nstart = 25)` |
| Elbow plot | manual (Part 11.2) |
| Silhouette | manual (Part 11.3) |
| Canonical correlation | `cancor(X, Y)` |
| Mahalanobis | `mahalanobis(X, colMeans(X), cov(X))` |
| MD threshold | `qchisq(0.975, df = p)` |
| Sphericity (ε) | manual (Part 14.1) |

---

# Part 18 — Common Mistakes

1. **Forgetting to extract the numeric matrix.** Most multivariate functions need a matrix, not a data frame with factors.
2. **Not standardizing before PCA when units differ.** Variables with large variance dominate.
3. **Confusing `sdev` with eigenvalues.** Square `sdev` to get eigenvalues.
4. **Reading `pca$rotation` wrong.** Rows are variables; columns are components.
5. **Not naming components after examining loadings.** Every interpretation should include a name.
6. **Ignoring correlation structure when doing univariate tests.** Use MANOVA when responses are correlated.
7. **Not doing follow-up tests after significant MANOVA.** MANOVA tells you "some response differs" but not which.
8. **Skipping multivariate normality check.** MVN is assumed by Hotelling and MANOVA.
9. **Using `cor()` instead of `cov()` or vice versa.** Some methods (PCA on correlation) need `cor`.
10. **Confusing PCA with factor analysis.** PCA = variance-maximizing components; FA = latent factors.
11. **Using Kaiser criterion blindly.** It overestimates in large datasets.
12. **Not checking Bartlett's test before FA.** If correlations are near zero, FA isn't justified.
13. **Ignoring LDA/QDA assumptions.** LDA assumes equal covariance; check with Box's M.
14. **Using k-means with non-spherical clusters.** Use hierarchical or DBSCAN.
15. **Not setting `nstart` in k-means.** Can get stuck in local optima.
16. **Forgetting to standardize before clustering.** Scale matters.
17. **Confusing Mahalanobis with Euclidean distance.** MD accounts for covariance.
18. **Not reporting degrees of freedom.** Every test statistic needs its df.
19. **Interpreting "significant MANOVA" as "all responses differ."** It means the joint vector differs.
20. **Treating categorical variables as numeric.** Factors must be handled separately.

---

# Part 19 — The Single Most Important Idea

Every multivariate question reduces to:

1. **What is the goal?**
   - Compare groups (Hotelling T², MANOVA)
   - Reduce dimensions (PCA, FA)
   - Classify (LDA, QDA)
   - Cluster (hierarchical, k-means)
   - Test distribution (Mardia)
   - Detect outliers (Mahalanobis)

2. **Extract the numeric matrix.**

3. **Standardize if units differ.**

4. **Fit the method.**

5. **Report the statistic, df, p, and effect size.**

6. **Interpret in plain words.**

If you can do these six steps, you can answer any multivariate question.

---

*End of cheat sheet. Good luck in the competition.*
````

---

## What is covered

| Section | Topics |
|---|---|
| Part 0 | 30-second strategy |
| Part 1 | Foundations: notation, covariance, intuition |
| Part 2 | Data prep: numeric matrix, standardization, group means |
| Part 3 | Hotelling T² (two-group), manual + univariate follow-ups |
| Part 4 | MANOVA (3+ groups), four test statistics, post-hoc |
| Part 5 | Mardia's MVN test + chi-square Q-Q |
| Part 6 | Covariance structures: CS, sphericity |
| Part 7 | Full PCA: eigenvalues, loadings, scores, biplot, group comparison |
| Part 8 | How many PCs: Kaiser, broken-stick, parallel analysis |
| Part 9 | Factor analysis: loadings, rotation, Bartlett, KMO |
| Part 10 | LDA, QDA, confusion matrix |
| Part 11 | Hierarchical and k-means clustering, elbow, silhouette |
| Part 12 | Canonical correlation + Wilks Lambda |
| Part 13 | Mahalanobis distance outliers |
| Part 14 | Sphericity and compound symmetry |
| Part 15 | Theory Q&A (25+ questions) |
| Part 16 | Exam scenarios (10 step-by-step answers) |
| Part 17 | Quick reference table |
| Part 18 | Common mistakes (20 traps) |
| Part 19 | Single most important idea |

**Design consistency:** Every chunk is self-contained with section-prefixed names (`p3_df`, `p7_pca`, `p11_hc`). Copy-paste any chunk into a fresh R session and it works.