---
title: "The Ultimate Regression Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(broom)
library(car)
library(emmeans)
library(lmtest)
library(knitr)
library(MASS)
library(glmnet)
```

# How to Use This Cheat Sheet

Every regression question falls into one of these categories. Find your question, jump to the section.

| Question type | Section | Typical phrasing |
|---|---|---|
| Fit simple/multiple regression | 4, 5 | "Fit a regression model" |
| Categorical predictor / reference level | 6 | "Use A as reference" |
| Interaction | 7 | "Does the effect of x depend on group?" |
| ANOVA / equality of means | 8 | "Are group means equal?" |
| ANCOVA / adjusted means | 9 | "Compare means controlling for x" |
| Nested F-test / subset test | 10 | "Are x2 and x3 significant together?" |
| Model selection | 11 | "Which model is best?" |
| Diagnostics | 12 | "Check the assumptions" |
| Transformations / Box-Cox | 13 | "Find the best lambda" |
| Prediction | 14 | "Predict y for x = 5" |
| Polynomial regression | 15 | "Fit a quadratic" |
| Multicollinearity | 16 | "Are predictors correlated?" |
| Weighted least squares | 17 | "Variance is not constant" |
| Logistic regression | 18 | "Model a binary outcome" |
| Poisson / count regression | 19 | "Model counts" |
| Negative binomial | 20 | "Overdispersed counts" |
| Piecewise regression | 21 | "Fit two lines with a break" |
| Variable selection | 22 | "Backward elimination" |
| Train/test, cross-validation | 23 | "Split 80/20 and evaluate" |
| Regularization | 24 | "Ridge / lasso" |
| Theory / derivations | 27 | "Derive the OLS estimator" |

**Golden rule of this document:** every chunk is self-contained. Each one starts with `set.seed()`, builds its own data, and uses section-prefixed names (`p3_df`, `p5_fit`, ...).

---

# Part 0 — The 30-Second Exam Strategy

For **any** regression question, run this pipeline:

1. **Identify the response** (y) and the predictors (x, groups).
2. **Fit** with `lm()` for continuous response, `glm()` for binary / count.
3. **Report coefficients** with `tidy(fit, conf.int = TRUE)`.
4. **Test the hypothesis** the question asks (t-test, F-test, LRT, ANOVA).
5. **Check assumptions**: residuals vs fitted, QQ plot, Shapiro, Breusch-Pagan, VIF.
6. **Refine** if needed (transform, remove outliers, add interactions).
7. **Predict** if asked, with confidence or prediction intervals.
8. **Report in words**: equation, interpretation, test result, conclusion.

If you can do these eight steps, you can answer any regression question.

---

# Part 1 — Foundations

## 1.1 The linear model

$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \varepsilon_i$$

- $y_i$: response for observation $i$
- $x_{ij}$: predictor $j$ for observation $i$
- $\beta_j$: coefficients (unknown)
- $\varepsilon_i$: error term, assumed $N(0, \sigma^2)$

**Matrix form:** $\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}$.

## 1.2 OLS estimator

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{y}$$

**Fitted values:** $\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} = \mathbf{H}\mathbf{y}$ where $\mathbf{H} = \mathbf{X}(\mathbf{X}^\top\mathbf{X})^{-1}\mathbf{X}^\top$ is the **hat matrix**.

**Residuals:** $\mathbf{e} = \mathbf{y} - \hat{\mathbf{y}} = (\mathbf{I} - \mathbf{H})\mathbf{y}$.

**Residual variance:** $\hat{\sigma}^2 = \frac{1}{n-p-1}\sum e_i^2$ where $p$ = number of predictors.

## 1.3 Gauss-Markov assumptions

For OLS to be BLUE (Best Linear Unbiased Estimator):

1. **Linearity:** $E[\varepsilon_i] = 0$.
2. **Homoskedasticity:** $\text{Var}(\varepsilon_i) = \sigma^2$ constant.
3. **Independence:** $\text{Cov}(\varepsilon_i, \varepsilon_j) = 0$ for $i \neq j$.
4. **Exogeneity:** predictors uncorrelated with errors.
5. **No perfect multicollinearity.**

For inference (t-tests, F-tests, CIs) we additionally assume:

6. **Normality:** $\varepsilon_i \sim N(0, \sigma^2)$.

## 1.4 Coefficient interpretation

| Term | Interpretation |
|---|---|
| $\beta_0$ (intercept) | Predicted $y$ when all predictors are 0 |
| $\beta_j$ (continuous) | Change in $y$ for a 1-unit increase in $x_j$, holding others fixed |
| $\beta_j$ (dummy) | Difference in mean $y$ between that group and the reference |
| $\beta_j$ (interaction) | Change in the effect of one variable when the other changes by 1 |

## 1.5 Standard error and t-test

$$\text{SE}(\hat{\beta}_j) = \hat{\sigma}\sqrt{[(\mathbf{X}^\top\mathbf{X})^{-1}]_{jj}}, \quad t_j = \frac{\hat{\beta}_j}{\text{SE}(\hat{\beta}_j)} \sim t_{n-p-1}$$

**$H_0$: $\beta_j = 0$** (predictor $j$ has no effect).
**$H_1$: $\beta_j \neq 0$**.
Reject $H_0$ if $p < 0.05$.

## 1.6 R-squared and adjusted R-squared

$$R^2 = 1 - \frac{\text{SSR}}{\text{SST}} = 1 - \frac{\sum (y_i - \hat{y}_i)^2}{\sum (y_i - \bar{y})^2}$$

$$R^2_{\text{adj}} = 1 - (1-R^2)\frac{n-1}{n-p-1}$$

- $R^2$ always increases with more predictors.
- $R^2_{\text{adj}}$ penalizes complexity. Use it to compare models with different numbers of predictors.

## 1.7 F-test for overall model

$$F = \frac{(\text{SST} - \text{SSR})/p}{\text{SSR}/(n-p-1)} \sim F_{p, n-p-1}$$

$H_0$: all $\beta_j = 0$ for $j \geq 1$. Reject if $p < 0.05$.

---

# Part 2 — Creating a Regression Data Frame

**Why:** `lm()` and `glm()` need a data frame. This section shows how to build one from scratch for every example.

```{r}
# --- Self-contained: create a basic regression data frame ---
set.seed(201)
p2_n  <- 50
p2_df <- tibble(
  x = rnorm(p2_n),
  y = 2 + 3 * x + rnorm(p2_n)
)
head(p2_df)
```

**Optional: matrix form.** Internally `lm()` uses $\hat{\beta} = (X^\top X)^{-1} X^\top y$. You can reproduce it:

```{r}
# --- Self-contained: OLS from matrix algebra ---
set.seed(202)
p2_n  <- 50
p2_x  <- rnorm(p2_n)
p2_y  <- 2 + 3 * p2_x + rnorm(p2_n)

X <- cbind(1, p2_x)
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% p2_y
as.numeric(beta_hat)

# Compare with lm()
coef(lm(p2_y ~ p2_x))
```

They match (to machine precision). This is what `lm()` does internally.

---

# Part 3 — Simple Linear Regression

## 3.1 Fitting and reporting

```{r}
# --- Self-contained: fit simple linear regression ---
set.seed(301)
p3_df  <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50))
p3_fit <- lm(y ~ x, data = p3_df)

kable(tidy(p3_fit, conf.int = TRUE), digits = 4)
kable(glance(p3_fit), digits = 4)
```

**What `tidy()` gives:**
- `term`: coefficient name
- `estimate`: $\hat{\beta}$
- `std.error`: $\text{SE}(\hat{\beta})$
- `statistic`: $t$-value
- `p.value`: two-sided p-value
- `conf.low`, `conf.high`: 95% CI

**What `glance()` gives:**
- `r.squared`, `adj.r.squared`
- `sigma`: residual standard error $\hat{\sigma}$
- `statistic`, `p.value`: overall F-test
- `df`, `logLik`, `AIC`, `BIC`, `deviance`

## 3.2 Full summary output

```{r}
# --- Self-contained: summary of simple regression ---
set.seed(302)
p3_df  <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50))
p3_fit <- lm(y ~ x, data = p3_df)
summary(p3_fit)
```

The `summary()` output contains everything: coefficients, SE, t, p, R², adj-R², F-statistic.

## 3.3 Interpretation templates

**Report template (fit):**
> "The fitted model is $\hat{y} = 2.04 + 2.98 x$. The intercept is 2.04 (SE 0.14, p < 0.001) and the slope is 2.98 (SE 0.15, p < 0.001). The model explains 79% of the variance in y (R² = 0.79, adjusted R² = 0.78). The overall F-test gives F(1, 48) = 180.2, p < 0.001, so the model is significant."

**Report template (coefficient):**
> "A 1-unit increase in x is associated with an estimated 2.98-unit increase in y (95% CI: [2.68, 3.28]), holding other predictors constant. This effect is significantly different from zero (p < 0.001)."

**Report template (intercept):**
> "The predicted y when x = 0 is 2.04 (95% CI: [1.76, 2.32])."

## 3.4 Confidence intervals for coefficients

```{r}
# --- Self-contained: 95% CI for coefficients ---
set.seed(303)
p3_df  <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50))
p3_fit <- lm(y ~ x, data = p3_df)

confint(p3_fit, level = 0.95)
```

**Interpretation:** "We are 95% confident that the true slope lies between 2.68 and 3.28."

## 3.5 Adding a regression line to a scatterplot

```{r}
# --- Self-contained: scatterplot with fitted line ---
set.seed(304)
p3_df  <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50))

ggplot(p3_df, aes(x, y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  theme_minimal() +
  labs(title = "Simple linear regression")
```

---

# Part 4 — Multiple Linear Regression

## 4.1 Fitting

```{r}
# --- Self-contained: multiple regression with three predictors ---
set.seed(401)
p4_df  <- tibble(
  x1 = rnorm(80),
  x2 = rnorm(80),
  x3 = rnorm(80),
  y  = 1 + 2 * x1 - 1.5 * x2 + 0.5 * x3 + rnorm(80)
)
p4_fit <- lm(y ~ x1 + x2 + x3, data = p4_df)

kable(tidy(p4_fit, conf.int = TRUE), digits = 4)
```

## 4.2 Coefficient interpretation (multiple)

Each coefficient is the effect of $x_j$ on $y$ **holding all other predictors fixed**.

**Report template:**
> "Holding $x_2$ and $x_3$ constant, a 1-unit increase in $x_1$ is associated with a 2.03-unit increase in $y$ (95% CI: [1.79, 2.27], p < 0.001)."

**Warning:** If predictors are correlated, "holding others fixed" may not be physically meaningful.

## 4.3 Partial F-test for overall model

```{r}
# --- Self-contained: overall F-test ---
set.seed(402)
p4_df  <- tibble(x1 = rnorm(80), x2 = rnorm(80),
                 y  = 1 + 2 * x1 - 1.5 * x2 + rnorm(80))
p4_fit <- lm(y ~ x1 + x2, data = p4_df)
summary(p4_fit)$fstatistic
```

## 4.4 Which predictors matter?

Look at the p-values in `tidy()`. A predictor with $p < 0.05$ is statistically significant at 5%.

**Caution:** With many predictors, some will be significant by chance. Adjust with Bonferroni or use a global F-test first.

---

# Part 5 — Model Comparison (Nested F-Test)

## 5.1 Nested F-test

Two models are **nested** if one is a special case of the other (e.g., removing predictors).

```{r}
# --- Self-contained: nested F-test ---
set.seed(501)
p5_df <- tibble(
  x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100),
  y  = 1 + 2 * x1 + rnorm(100)   # x2, x3 have no effect
)

p5_full  <- lm(y ~ x1 + x2 + x3, data = p5_df)
p5_reduc <- lm(y ~ x1, data = p5_df)

anova(p5_reduc, p5_full)
```

**What `anova(reduced, full)` does:**
- $H_0$: the reduced model is adequate (extra predictors have zero coefficients).
- $H_1$: the full model is significantly better.
- If $p < 0.05$: reject $H_0$; keep the extra predictors.

**Report template:**
> "The nested F-test gives F(2, 96) = 1.24, p = 0.29. We fail to reject $H_0$; $x_2$ and $x_3$ together do not significantly improve the model."

## 5.2 Non-nested comparison (AIC/BIC)

For non-nested models, compare AIC or BIC (lower is better).

```{r}
# --- Self-contained: AIC/BIC comparison ---
set.seed(502)
p5_df <- tibble(x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100),
                y  = 1 + 2 * x1 + rnorm(100))

p5_m1 <- lm(y ~ x1, data = p5_df)
p5_m2 <- lm(y ~ x2, data = p5_df)
p5_m3 <- lm(y ~ x1 + x2, data = p5_df)

AIC(p5_m1, p5_m2, p5_m3)
BIC(p5_m1, p5_m2, p5_m3)
```

**Interpretation:** AIC prefers predictive accuracy; BIC prefers parsimony (stronger penalty). Differences of 2+ are usually meaningful; differences < 2 are ambiguous.

---

# Part 6 — Categorical Predictors and Reference Levels

## 6.1 Dummy coding

When a predictor is a factor with $k$ levels, R creates $k-1$ dummy variables. The first level (alphabetical or explicitly set) is the **reference**.

```{r}
# --- Self-contained: categorical predictor with three groups ---
set.seed(601)
p6_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20), levels = c("A", "B", "C")),
  y     = 10 + c(A = 0, B = 3, C = 6)[as.character(group)] + rnorm(60)
)
p6_fit <- lm(y ~ group, data = p6_df)

kable(tidy(p6_fit, conf.int = TRUE), digits = 4)
```

**Interpretation:**
- `(Intercept)` = mean of reference group A.
- `groupB` = mean(B) − mean(A).
- `groupC` = mean(C) − mean(A).
- p-values test whether each group differs from the reference.

**Report template:**
> "Group B has a mean y 3.12 units higher than the reference group A (95% CI: [2.61, 3.63], p < 0.001). Group C has a mean y 6.05 units higher than A (95% CI: [5.54, 6.56], p < 0.001)."

## 6.2 Changing the reference level

```{r}
# --- Self-contained: relevel to change reference ---
set.seed(602)
p6_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20), levels = c("A", "B", "C")),
  y     = 10 + c(A = 0, B = 3, C = 6)[as.character(group)] + rnorm(60)
)
p6_df$group <- relevel(p6_df$group, ref = "B")
p6_fit <- lm(y ~ group, data = p6_df)
kable(tidy(p6_fit), digits = 4)
```

Now B is the reference, so `groupA` and `groupC` are contrasts vs B.

## 6.3 Custom contrasts

R's default is **treatment contrasts** (reference vs others). Alternatives: sum (deviation), Helmert, polynomial.

```{r}
# --- Self-contained: sum contrasts ---
set.seed(603)
p6_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  y     = 10 + c(A = 0, B = 3, C = 6)[as.character(group)] + rnorm(60)
)
contrasts(p6_df$group) <- contr.sum(3)
p6_fit <- lm(y ~ group, data = p6_df)
kable(tidy(p6_fit), digits = 4)
```

---

# Part 7 — Interactions

## 7.1 Continuous × continuous

```{r}
# --- Self-contained: interaction between two continuous predictors ---
set.seed(701)
p7_df <- tibble(
  x1 = rnorm(100),
  x2 = rnorm(100),
  y  = 1 + 2 * x1 + 0.5 * x2 + 1.5 * x1 * x2 + rnorm(100)
)
p7_fit <- lm(y ~ x1 * x2, data = p7_df)   # equivalent to x1 + x2 + x1:x2
kable(tidy(p7_fit), digits = 4)
```

**Interpretation:**
- `x1` = effect of $x_1$ when $x_2 = 0$.
- `x2` = effect of $x_2$ when $x_1 = 0$.
- `x1:x2` = how the effect of $x_1$ changes per unit increase in $x_2$ (and vice versa).

**Report template:**
> "The interaction is significant ($\hat{\beta}_{12} = 1.48$, p < 0.001), indicating that the effect of $x_1$ on $y$ depends on $x_2$. When $x_2 = 0$, the effect of $x_1$ is 2.02; each 1-unit increase in $x_2$ adds 1.48 to that effect."

## 7.2 Categorical × continuous (different slopes per group)

```{r}
# --- Self-contained: group-specific slopes ---
set.seed(702)
p7_df <- tibble(
  x     = rnorm(120),
  group = factor(rep(c("A", "B"), each = 60)),
  y     = 1 + 2 * x +
          c(A = 0, B = 1)[as.character(group)] +
          c(A = 0, B = 1.5)[as.character(group)] * x +
          rnorm(120)
)
p7_fit <- lm(y ~ x * group, data = p7_df)
kable(tidy(p7_fit), digits = 4)
```

**Interpretation:**
- `x` = slope for reference group A.
- `groupB` = difference in intercepts (A vs B).
- `x:groupB` = difference in slopes (B − A).

**Report template:**
> "Group A has slope 2.05. Group B has slope 3.58 (2.05 + 1.53), i.e., the slope differs by 1.53 between groups (p = 0.001). This indicates the effect of x on y is stronger in group B."

## 7.3 Visualizing interactions

```{r}
# --- Self-contained: interaction plot ---
set.seed(703)
p7_df <- tibble(
  x     = rnorm(120),
  group = factor(rep(c("A", "B"), each = 60)),
  y     = 1 + 2 * x + c(A = 0, B = 1)[as.character(group)] +
          c(A = 0, B = 1.5)[as.character(group)] * x + rnorm(120)
)

ggplot(p7_df, aes(x, y, color = group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

---

# Part 8 — One-Way ANOVA

## 8.1 Equality of group means

```{r}
# --- Self-contained: one-way ANOVA ---
set.seed(801)
p8_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  y     = 10 + c(A = 0, B = 3, C = 6)[as.character(group)] + rnorm(60)
)
p8_fit <- lm(y ~ group, data = p8_df)

anova(p8_fit)     # ANOVA table
summary(p8_fit)   # coefficients
```

**What `anova()` does:**
- $H_0$: all group means are equal.
- $H_1$: at least one differs.
- F-statistic = between-group variance / within-group variance.
- If $p < 0.05$: at least one mean differs.

**Report template:**
> "The one-way ANOVA gives F(2, 57) = 45.3, p < 0.001. We reject $H_0$; at least one group mean differs from the others."

## 8.2 Post-hoc pairwise comparisons

```{r}
# --- Self-contained: Tukey HSD ---
set.seed(802)
p8_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  y     = 10 + c(A = 0, B = 3, C = 6)[as.character(group)] + rnorm(60)
)
TukeyHSD(aov(y ~ group, data = p8_df))
```

**Interpretation:** Each row is a pairwise comparison. `diff` = mean difference; `lwr`, `upr` = 95% CI; `p adj` = adjusted p-value (controls family-wise error).

**Alternative (Bonferroni):**
```{r}
# --- Self-contained: Bonferroni pairwise t-tests ---
set.seed(803)
p8_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  y     = 10 + c(A = 0, B = 3, C = 6)[as.character(group)] + rnorm(60)
)
pairwise.t.test(p8_df$y, p8_df$group, p.adjust.method = "bonferroni")
```

## 8.3 Two-way ANOVA

```{r}
# --- Self-contained: two-way ANOVA with interaction ---
set.seed(804)
p8_df <- expand.grid(A = factor(c("a1","a2")), B = factor(c("b1","b2")))
p8_df <- p8_df[rep(1:4, each = 15), ]
p8_df$y <- 10 + 2 * (p8_df$A == "a2") + 3 * (p8_df$B == "b2") +
           1.5 * (p8_df$A == "a2") * (p8_df$B == "b2") + rnorm(60)

p8_fit <- lm(y ~ A * B, data = p8_df)
anova(p8_fit)
```

**Interpretation:**
- `A` main effect: does A matter on average across B?
- `B` main effect: does B matter on average across A?
- `A:B` interaction: does the effect of A depend on B?

**If interaction is significant**, main effects should be interpreted cautiously — they are the effects when the other factor is at its reference level.

---

# Part 9 — ANCOVA (Adjusting for a Covariate)

## 9.1 Fit ANCOVA

```{r}
# --- Self-contained: ANCOVA with group + covariate ---
set.seed(901)
p9_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  x     = rnorm(60, 50, 10),
  y     = 0.8 * x + c(A = 0, B = 5, C = 10)[as.character(group)] + rnorm(60)
)
p9_fit <- lm(y ~ group + x, data = p9_df)
kable(tidy(p9_fit), digits = 4)
```

**What ANCOVA does:**
- Compares group means **after adjusting for** the covariate $x$.
- Assumes **equal slopes** across groups (parallel lines).
- Intercept differences = adjusted mean differences.

## 9.2 Test equality of adjusted means

```{r}
# --- Self-contained: ANCOVA model comparison ---
set.seed(902)
p9_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  x     = rnorm(60, 50, 10),
  y     = 0.8 * x + c(A = 0, B = 5, C = 10)[as.character(group)] + rnorm(60)
)
p9_full  <- lm(y ~ group + x, data = p9_df)
p9_reduc <- lm(y ~ x, data = p9_df)
anova(p9_reduc, p9_full)
```

$p < 0.05$: group means differ after adjusting for $x$.

## 9.3 Adjusted means (emmeans)

```{r}
# --- Self-contained: adjusted means ---
set.seed(903)
p9_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  x     = rnorm(60, 50, 10),
  y     = 0.8 * x + c(A = 0, B = 5, C = 10)[as.character(group)] + rnorm(60)
)
p9_fit <- lm(y ~ group + x, data = p9_df)
p9_emm <- emmeans(p9_fit, ~ group)
p9_emm
pairs(p9_emm, adjust = "bonferroni")
```

**Interpretation:** emmeans gives the predicted mean for each group at the **average covariate value**. The pairs output gives pairwise adjusted mean differences with CIs.

## 9.4 Check homogeneity of slopes

```{r}
# --- Self-contained: test equal slopes ---
set.seed(904)
p9_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  x     = rnorm(60, 50, 10),
  y     = 0.8 * x + c(A = 0, B = 5, C = 10)[as.character(group)] + rnorm(60)
)
p9_no_inter  <- lm(y ~ group + x, data = p9_df)
p9_with_inter <- lm(y ~ group * x, data = p9_df)
anova(p9_no_inter, p9_with_inter)
```

- $p > 0.05$: slopes are equal → ANCOVA valid.
- $p < 0.05$: slopes differ → ANCOVA invalid; use a model with interaction.

---

# Part 10 — Testing Multiple Coefficients (Subset F-test)

## 10.1 Test if a subset of coefficients is zero

```{r}
# --- Self-contained: subset F-test ---
set.seed(1001)
p10_df <- tibble(
  x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100), x4 = rnorm(100),
  y  = 1 + 2 * x1 + rnorm(100)   # x2, x3, x4 are noise
)
p10_full  <- lm(y ~ x1 + x2 + x3 + x4, data = p10_df)
p10_reduc <- lm(y ~ x1, data = p10_df)

anova(p10_reduc, p10_full)
```

**Report:** "The F-test for the joint significance of x2, x3, x4 gives F(3, 95) = X, p = Y. We [reject / fail to reject] $H_0$."

## 10.2 Test a linear restriction

Test $H_0: \beta_1 = \beta_2$ (two coefficients are equal).

```{r}
# --- Self-contained: test beta1 = beta2 using linearHypothesis ---
set.seed(1002)
p10_df <- tibble(x1 = rnorm(80), x2 = rnorm(80),
                 y  = 1 + 2 * x1 + 2 * x2 + rnorm(80))
p10_fit <- lm(y ~ x1 + x2, data = p10_df)

car::linearHypothesis(p10_fit, "x1 = x2")
```

**Interpretation:** If $p > 0.05$, the two coefficients are not significantly different; the simpler model with $\beta_1 = \beta_2$ is fine.

## 10.3 Other linear restrictions

```{r}
# --- Self-contained: test beta1 + beta2 = 3 ---
set.seed(1003)
p10_df <- tibble(x1 = rnorm(80), x2 = rnorm(80),
                 y  = 1 + 1.5 * x1 + 1.5 * x2 + rnorm(80))
p10_fit <- lm(y ~ x1 + x2, data = p10_df)

car::linearHypothesis(p10_fit, "x1 + x2 = 3")
```

---

# Part 11 — Model Selection

## 11.1 Information criteria

$$AIC = -2\ell + 2k, \quad BIC = -2\ell + k\log n$$

- $k$: number of parameters
- $\ell$: maximized log-likelihood
- Lower is better
- BIC penalizes complexity more strongly than AIC

```{r}
# --- Self-contained: AIC/BIC comparison ---
set.seed(1101)
p11_df <- tibble(x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100),
                 y  = 1 + 2 * x1 + 0.5 * x3 + rnorm(100))

p11_m1 <- lm(y ~ x1, data = p11_df)
p11_m2 <- lm(y ~ x1 + x2, data = p11_df)
p11_m3 <- lm(y ~ x1 + x2 + x3, data = p11_df)

AIC(p11_m1, p11_m2, p11_m3)
BIC(p11_m1, p11_m2, p11_m3)
```

## 11.2 Adjusted R-squared

```{r}
# --- Self-contained: adjusted R-squared comparison ---
set.seed(1102)
p11_df <- tibble(x1 = rnorm(100), x2 = rnorm(100),
                 y  = 1 + 2 * x1 + rnorm(100))

p11_m1 <- lm(y ~ x1, data = p11_df)
p11_m2 <- lm(y ~ x1 + x2, data = p11_df)

summary(p11_m1)$adj.r.squared
summary(p11_m2)$adj.r.squared
```

## 11.3 Backward elimination

```{r}
# --- Self-contained: backward stepwise ---
set.seed(1103)
p11_df <- tibble(
  x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100), x4 = rnorm(100),
  y  = 2 + 0.5 * x1 - 0.3 * x3 + rnorm(100)
)
p11_full <- lm(y ~ ., data = p11_df)
p11_back <- step(p11_full, direction = "backward", trace = 0)
summary(p11_back)
```

**What `step()` does:** removes the predictor that most improves AIC at each step, stopping when no removal improves AIC.

**Caution:** stepwise selection can overfit, ignores model uncertainty, and p-values from the final model are biased. Use with care.

## 11.4 Forward and both-direction

```{r}
# --- Self-contained: forward selection ---
set.seed(1104)
p11_df <- tibble(x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100),
                 y  = 1 + 2 * x1 + rnorm(100))
p11_null <- lm(y ~ 1, data = p11_df)
p11_fwd  <- step(p11_null, scope = ~ x1 + x2 + x3,
                 direction = "forward", trace = 0)
summary(p11_fwd)
```

## 11.5 All-subsets regression

```{r}
# --- Self-contained: all-subsets via leaps ---
# library(leaps)
# set.seed(1105)
# p11_df <- tibble(x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100), x4 = rnorm(100),
#                  y  = 1 + 2 * x1 + rnorm(100))
# regsubsets(y ~ ., data = p11_df, nvmax = 4) %>% summary()
```

---

# Part 12 — Diagnostics

## 12.1 Residuals vs fitted (linearity, homoskedasticity)

```{r}
# --- Self-contained: residual plots ---
set.seed(1201)
p12_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p12_fit <- lm(y ~ x, data = p12_df)

par(mfrow = c(2, 2))
plot(p12_fit)
par(mfrow = c(1, 1))
```

Four plots:

1. **Residuals vs Fitted:** Should show no pattern. A funnel = heteroskedasticity. A curve = nonlinearity.
2. **Normal Q-Q:** Points should lie on the diagonal. Curvature = non-normal residuals.
3. **Scale-Location:** Square root of standardized residuals vs fitted. Should be flat.
4. **Residuals vs Leverage:** Identifies influential points (Cook's distance).

## 12.2 Shapiro-Wilk test (normality)

```{r}
# --- Self-contained: Shapiro-Wilk ---
set.seed(1202)
p12_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p12_fit <- lm(y ~ x, data = p12_df)
shapiro.test(residuals(p12_fit))
```

$H_0$: residuals are normal. $p > 0.05$ → no evidence against normality.

## 12.3 Breusch-Pagan test (homoskedasticity)

```{r}
# --- Self-contained: Breusch-Pagan ---
set.seed(1203)
p12_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p12_fit <- lm(y ~ x, data = p12_df)
bptest(p12_fit)
```

$H_0$: constant variance. $p < 0.05$ → heteroskedasticity present.

## 12.4 Durbin-Watson (autocorrelation)

```{r}
# --- Self-contained: Durbin-Watson ---
set.seed(1204)
p12_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p12_fit <- lm(y ~ x, data = p12_df)
lmtest::dwtest(p12_fit)
```

$H_0$: no autocorrelation. DW ≈ 2 → none. DW < 2 → positive. DW > 2 → negative.

## 12.5 Multicollinearity (VIF)

```{r}
# --- Self-contained: VIF ---
set.seed(1205)
p12_df  <- tibble(x1 = rnorm(80), x2 = 0.95 * x1 + 0.1 * rnorm(80),
                  x3 = rnorm(80), y = 1 + x1 + x2 + x3 + rnorm(80))
p12_fit <- lm(y ~ x1 + x2 + x3, data = p12_df)
vif(p12_fit)
```

- VIF = 1: no correlation.
- VIF > 5: moderate concern.
- VIF > 10: serious multicollinearity.

**Fix:** remove one predictor, combine them, or use ridge regression.

## 12.6 Outliers and influence

```{r}
# --- Self-contained: outliers and influence ---
set.seed(1206)
p12_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p12_df$y[10] <- p12_df$y[10] + 15   # inject outlier
p12_fit <- lm(y ~ x, data = p12_df)

# Standardized residuals
outliers <- which(abs(rstandard(p12_fit)) > 3)
outliers

# Cook's distance
cooks <- cooks.distance(p12_fit)
which(cooks > 4 / nrow(p12_df))

# Leverage
hatvalues(p12_fit)[1:5]

# Outlier test
outlierTest(p12_fit)
```

**Interpretation:**
- **Standardized residual > 3**: potential outlier.
- **Cook's D > 4/n**: influential observation.
- **Leverage > 2(p+1)/n**: high-leverage point.
- **outlierTest**: Bonferroni-adjusted p-value for the most extreme residual.

## 12.7 Full diagnostic template

```{r}
# --- Self-contained: run all diagnostics ---
set.seed(1207)
p12_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p12_fit <- lm(y ~ x, data = p12_df)

par(mfrow = c(2, 2)); plot(p12_fit); par(mfrow = c(1, 1))
shapiro.test(residuals(p12_fit))
bptest(p12_fit)
lmtest::dwtest(p12_fit)
```

**Report template:**
> "The residuals show no pattern in the fitted-vs-residuals plot, the Q-Q plot looks linear, and Shapiro-Wilk gives p = 0.31, so normality is reasonable. Breusch-Pagan gives p = 0.42, so variance is constant. Durbin-Watson is 2.05, so no autocorrelation. No observations exceed Cook's D = 4/n. Assumptions are satisfied."

---

# Part 13 — Transformations and Box-Cox

## 13.1 When to transform

| Problem | Transform |
|---|---|
| Non-normal residuals | Box-Cox, log |
| Heteroskedasticity (funnel) | log, sqrt, Box-Cox |
| Non-linearity (curve) | polynomial, log(x) |
| Multiplicative relationship | log(y) ~ log(x) |
| Counts | Poisson / negative binomial GLM |

## 13.2 Box-Cox on the response

```{r}
# --- Self-contained: Box-Cox plot ---
set.seed(1301)
p13_df  <- tibble(x = rnorm(80), y = exp(1 + 0.5 * x + rnorm(80, 0, 0.3)))
p13_fit <- lm(y ~ x, data = p13_df)

boxCox(p13_fit)
```

The plot shows the log-likelihood as a function of $\lambda$. The peak is the MLE. The dashed lines are 95% confidence bounds.

## 13.3 Manual Box-Cox over a grid

```{r}
# --- Self-contained: manual Box-Cox grid ---
set.seed(1302)
p13_df <- tibble(x = rnorm(80), y = exp(1 + 0.5 * x + rnorm(80, 0, 0.3)))

p13_lambdas <- c(-2, -1, -0.5, 0, 0.5, 1, 2)
p13_res <- tibble(lambda = p13_lambdas, logLik = NA_real_, R2 = NA_real_)

for (i in seq_along(p13_lambdas)) {
  l  <- p13_lambdas[i]
  yt <- if (l == 0) log(p13_df$y) else (p13_df$y^l - 1) / l
  fit_l <- lm(yt ~ p13_df$x)
  p13_res$logLik[i] <- as.numeric(logLik(fit_l))
  p13_res$R2[i]     <- summary(fit_l)$r.squared
}
kable(p13_res, digits = 4)
```

**Report:** "The Box-Cox profile log-likelihood peaks at λ = 0.02, so a log transformation is appropriate. All models were refit on the transformed scale and compared by log-likelihood."

## 13.4 Log transformation

```{r}
# --- Self-contained: log transform ---
set.seed(1303)
p13_df  <- tibble(x = rnorm(80), y = exp(1 + 0.5 * x + rnorm(80, 0, 0.3)))
p13_fit <- lm(log(y) ~ x, data = p13_df)

kable(tidy(p13_fit), digits = 4)
```

**Interpretation:** On the log scale, a 1-unit increase in $x$ changes $y$ by a multiplicative factor of $e^{\hat{\beta}_1}$. A 1% increase in $x$ changes $y$ by about $\hat{\beta}_1$%.

## 13.5 Log-log (elasticity)

```{r}
# --- Self-contained: log-log model ---
set.seed(1304)
p13_df  <- tibble(x = rlnorm(80, 0, 1), y = 3 * x^0.7 * exp(rnorm(80, 0, 0.2)))
p13_fit <- lm(log(y) ~ log(x), data = p13_df)

kable(tidy(p13_fit), digits = 4)
```

**Interpretation:** $\hat{\beta}_1$ is an **elasticity**: a 1% increase in $x$ is associated with a $\hat{\beta}_1$% change in $y$.

---

# Part 14 — Prediction

## 14.1 Confidence interval vs prediction interval

```{r}
# --- Self-contained: confidence and prediction intervals ---
set.seed(1401)
p14_df  <- tibble(x = rnorm(80), y = 2 + 3 * x + rnorm(80))
p14_fit <- lm(y ~ x, data = p14_df)

newdf <- data.frame(x = c(0, 1, 2))

# CI for the mean response at x
predict(p14_fit, newdata = newdf, interval = "confidence", level = 0.95)

# PI for a new observation at x
predict(p14_fit, newdata = newdf, interval = "prediction", level = 0.95)
```

**Key difference:**
- **Confidence interval:** uncertainty about the *mean* of $y$ at $x$. Narrow.
- **Prediction interval:** uncertainty about *one future observation* at $x$. Wide.
- PI is always wider because it includes the individual error $\varepsilon$.

**Report:**
> "For $x = 1$, the predicted mean $y$ is 5.02 with a 95% CI of [4.85, 5.19]. A single new observation at $x = 1$ is predicted to lie in [3.80, 6.24] with 95% probability."

## 14.2 Prediction with factors and interactions

```{r}
# --- Self-contained: prediction with a factor ---
set.seed(1402)
p14_df  <- tibble(
  x     = rnorm(100),
  group = factor(rep(c("A", "B"), each = 50)),
  y     = 1 + 2 * x + c(A = 0, B = 1)[as.character(group)] +
          c(A = 0, B = 1.5)[as.character(group)] * x + rnorm(100)
)
p14_fit <- lm(y ~ x * group, data = p14_df)

newdf <- expand.grid(x = c(0, 1), group = factor(c("A", "B")))
predict(p14_fit, newdata = newdf, interval = "prediction")
```

**Always pass `newdata` as a data frame with the same variable names and factor levels** used to fit the model.

---

# Part 15 — Polynomial Regression

## 15.1 Quadratic

```{r}
# --- Self-contained: quadratic regression ---
set.seed(1501)
p15_df  <- tibble(x = runif(100, -3, 3), y = 1 + 2 * x - 1.5 * x^2 + rnorm(100))
p15_fit <- lm(y ~ x + I(x^2), data = p15_df)

kable(tidy(p15_fit), digits = 4)
```

**Interpretation:** The `I(x^2)` coefficient measures curvature. If it's significant, the relationship is quadratic (parabolic).

**Warning:** Always center $x$ before squaring to reduce multicollinearity:
```{r}
# --- Self-contained: centered quadratic ---
set.seed(1502)
p15_df  <- tibble(x = runif(100, -3, 3), y = 1 + 2 * x - 1.5 * x^2 + rnorm(100))
p15_df$x_c <- p15_df$x - mean(p15_df$x)
p15_fit <- lm(y ~ x_c + I(x_c^2), data = p15_df)
kable(tidy(p15_fit), digits = 4)
```

## 15.2 Orthogonal polynomials

```{r}
# --- Self-contained: orthogonal polynomial ---
set.seed(1503)
p15_df  <- tibble(x = runif(100, -3, 3), y = 1 + 2 * x - 1.5 * x^2 + rnorm(100))
p15_fit <- lm(y ~ poly(x, 2, raw = FALSE), data = p15_df)
summary(p15_fit)
```

`poly(x, 2)` gives orthogonal polynomials, which are numerically stable and make each term's p-value interpretable.

## 15.3 Testing the degree

```{r}
# --- Self-contained: test quadratic vs linear ---
set.seed(1504)
p15_df  <- tibble(x = runif(100, -3, 3), y = 1 + 2 * x - 1.5 * x^2 + rnorm(100))
p15_lin <- lm(y ~ x, data = p15_df)
p15_qua <- lm(y ~ x + I(x^2), data = p15_df)
anova(p15_lin, p15_qua)
```

---

# Part 16 — Multicollinearity

## 16.1 Detecting it

```{r}
# --- Self-contained: VIF and condition number ---
set.seed(1601)
p16_df <- tibble(
  x1 = rnorm(100),
  x2 = 0.9 * x1 + 0.1 * rnorm(100),
  x3 = rnorm(100),
  y  = 1 + x1 + x2 + x3 + rnorm(100)
)
p16_fit <- lm(y ~ x1 + x2 + x3, data = p16_df)

vif(p16_fit)
kappa(cor(p16_df[, c("x1", "x2", "x3")]))
```

- **VIF > 5** (or > 10): problematic.
- **Condition number > 30**: serious.

## 16.2 Fixing it

- Drop one of the correlated predictors.
- Combine them (e.g., average or PCA).
- Use ridge regression (Part 24).

## 16.3 Ridge regression (preview)

```{r}
# --- Self-contained: ridge regression ---
set.seed(1602)
p16_df <- tibble(
  x1 = rnorm(100), x2 = 0.9 * rnorm(100) + 0.1 * rnorm(100), x3 = rnorm(100),
  y  = 1 + x1 + x2 + x3 + rnorm(100)
)
X <- as.matrix(p16_df[, c("x1", "x2", "x3")])
y <- p16_df$y
ridge <- glmnet(X, y, alpha = 0)
plot(ridge, xvar = "lambda")
```

Ridge shrinks coefficients toward zero, trading bias for variance.

---

# Part 17 — Weighted Least Squares

## 17.1 When to use

If residuals have **non-constant variance** and you know the variance structure, use WLS.

$$w_i = \frac{1}{\text{Var}(\varepsilon_i)}$$

## 17.2 Fitting

```{r}
# --- Self-contained: weighted least squares ---
set.seed(1701)
p17_df <- tibble(
  x = runif(100, 1, 10),
  y = 2 + 0.5 * x + rnorm(100, 0, sd = x)   # variance grows with x
)
p17_fit_ols <- lm(y ~ x, data = p17_df)
p17_fit_wls <- lm(y ~ x, data = p17_df, weights = 1 / x^2)

summary(p17_fit_ols)$coefficients
summary(p17_fit_wls)$coefficients
```

**Weights:** If $\text{Var}(\varepsilon_i) = \sigma^2 x_i^2$, use weights $1/x_i^2$.

## 17.3 Feasible GLS

If you don't know the variance structure, estimate it:

```{r}
# --- Self-contained: feasible GLS via log-variance regression ---
set.seed(1702)
p17_df <- tibble(
  x = runif(100, 1, 10),
  y = 2 + 0.5 * x + rnorm(100, 0, sd = x)
)
ols <- lm(y ~ x, data = p17_df)
p17_df$log_e2 <- log(residuals(ols)^2)
var_fit <- lm(log_e2 ~ log(x), data = p17_df)
w <- 1 / exp(fitted(var_fit))
p17_fit_fgls <- lm(y ~ x, data = p17_df, weights = w)
summary(p17_fit_fgls)$coefficients
```

---

# Part 18 — Logistic Regression

## 18.1 Model and link

$$y_i \sim \text{Bernoulli}(p_i), \quad \log\frac{p_i}{1-p_i} = \beta_0 + \beta_1 x_i$$

- Link function: logit
- Response: binary (0/1)
- Coefficients: change in **log-odds** per unit increase in $x$

## 18.2 Fitting

```{r}
# --- Self-contained: logistic regression ---
set.seed(1801)
p18_n  <- 200
p18_df <- tibble(
  x = rnorm(p18_n),
  y = rbinom(p18_n, 1, plogis(-1 + 1.5 * x))
)
p18_fit <- glm(y ~ x, family = binomial, data = p18_df)
summary(p18_fit)
```

## 18.3 Coefficient interpretation (log-odds, odds, probability)

```{r}
# --- Self-contained: logistic coefficients and odds ratios ---
set.seed(1802)
p18_n  <- 200
p18_df <- tibble(x = rnorm(p18_n), y = rbinom(p18_n, 1, plogis(-1 + 1.5 * x)))
p18_fit <- glm(y ~ x, family = binomial, data = p18_df)

kable(tidy(p18_fit, exponentiate = TRUE, conf.int = TRUE), digits = 4)
```

**Interpretation:**
- Coefficient $\hat{\beta}_1$: log-odds change per unit $x$.
- `exp(coefficient)` = odds ratio. OR > 1: odds increase. OR < 1: odds decrease.
- `exp(confint(fit))` gives 95% CI for the OR.

**Report template:**
> "A 1-unit increase in $x$ is associated with a 4.48-fold increase in the odds of $y = 1$ (OR = 4.48, 95% CI: [2.71, 7.89], p < 0.001)."

## 18.4 Predicted probabilities

```{r}
# --- Self-contained: predicted probabilities ---
set.seed(1803)
p18_n  <- 200
p18_df <- tibble(x = rnorm(p18_n), y = rbinom(p18_n, 1, plogis(-1 + 1.5 * x)))
p18_fit <- glm(y ~ x, family = binomial, data = p18_df)

newdf <- data.frame(x = c(-1, 0, 1, 2))
predict(p18_fit, newdata = newdf, type = "response")
```

`type = "response"` gives probabilities in [0,1]. `type = "link"` gives log-odds.

## 18.5 Testing predictors

```{r}
# --- Self-contained: likelihood ratio test ---
set.seed(1804)
p18_n  <- 200
p18_df <- tibble(x = rnorm(p18_n), y = rbinom(p18_n, 1, plogis(-1 + 1.5 * x)))
p18_null <- glm(y ~ 1, family = binomial, data = p18_df)
p18_full <- glm(y ~ x, family = binomial, data = p18_df)
anova(p18_null, p18_full, test = "LRT")
```

$p < 0.05$: predictor significantly improves fit.

## 18.6 Goodness of fit

```{r}
# --- Self-contained: deviance and AIC ---
set.seed(1805)
p18_n  <- 200
p18_df <- tibble(x = rnorm(p18_n), y = rbinom(p18_n, 1, plogis(-1 + 1.5 * x)))
p18_fit <- glm(y ~ x, family = binomial, data = p18_df)

# Deviance-based pseudo-R²
1 - deviance(p18_fit) / deviance(glm(y ~ 1, family = binomial, data = p18_df))
# AIC and BIC
AIC(p18_fit); BIC(p18_fit)
```

## 18.7 Classification and confusion matrix

```{r}
# --- Self-contained: classification ---
set.seed(1806)
p18_n  <- 200
p18_df <- tibble(x = rnorm(p18_n), y = rbinom(p18_n, 1, plogis(-1 + 1.5 * x)))
p18_fit <- glm(y ~ x, family = binomial, data = p18_df)

pred_prob <- predict(p18_fit, type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)
table(Predicted = pred_class, Actual = p18_df$y)
```

**Metrics:**
- Accuracy: (TP + TN) / n.
- Sensitivity (recall): TP / (TP + FN).
- Specificity: TN / (TN + FP).
- Precision: TP / (TP + FP).
- AUC: area under ROC curve.

## 18.8 Grouped binomial data

When you have counts of successes/failures by group:

```{r}
# --- Self-contained: grouped binomial logistic ---
set.seed(1807)
p18_df <- tibble(
  sex  = factor(rep(c("M", "F"), each = 20)),
  belt = factor(rep(c("yes", "no"), 20)),
  total = sample(50:100, 40, replace = TRUE)
)
p18_df$injured <- rbinom(40, p18_df$total, plogis(-0.5 + 0.8 * (p18_df$sex == "M") - 0.6 * (p18_df$belt == "yes")))

p18_fit <- glm(cbind(injured, total - injured) ~ sex + belt,
               family = binomial, data = p18_df)
kable(tidy(p18_fit, exponentiate = TRUE, conf.int = TRUE), digits = 4)
```

**Interpretation:** Same as before — the odds ratios are for the log-odds of the proportion injured.

---

# Part 19 — Poisson Regression

## 19.1 Model

$$y_i \sim \text{Poisson}(\lambda_i), \quad \log \lambda_i = \beta_0 + \beta_1 x_i$$

Response is a **count**. Link function is log.

## 19.2 Fitting

```{r}
# --- Self-contained: Poisson regression ---
set.seed(1901)
p19_n  <- 200
p19_df <- tibble(
  x = rnorm(p19_n),
  y = rpois(p19_n, exp(1 + 0.5 * x))
)
p19_fit <- glm(y ~ x, family = poisson, data = p19_df)
summary(p19_fit)
```

## 19.3 Coefficient interpretation

```{r}
# --- Self-contained: Poisson incidence rate ratios ---
set.seed(1902)
p19_df <- tibble(x = rnorm(200), y = rpois(200, exp(1 + 0.5 * x)))
p19_fit <- glm(y ~ x, family = poisson, data = p19_df)

kable(tidy(p19_fit, exponentiate = TRUE, conf.int = TRUE), digits = 4)
```

**Interpretation:** `exp(coefficient)` is the **multiplicative effect on the expected count**. A 1-unit increase in $x$ multiplies the expected count by `exp(beta)`.

**Report:**
> "A 1-unit increase in $x$ multiplies the expected count by 1.63 (95% CI: [1.44, 1.85], p < 0.001)."

## 19.4 Overdispersion

If the variance is larger than the mean, Poisson is inadequate.

```{r}
# --- Self-contained: check overdispersion ---
set.seed(1903)
p19_df <- tibble(x = rnorm(200), y = rpois(200, exp(1 + 0.5 * x)))
p19_fit <- glm(y ~ x, family = poisson, data = p19_df)

# Pearson chi-square / df
sum(residuals(p19_fit, type = "pearson")^2) / df.residual(p19_fit)
```

Rule of thumb: if this ratio is much greater than 1, use quasipoisson or negative binomial.

## 19.5 Quasipoisson (fix overdispersion)

```{r}
# --- Self-contained: quasipoisson ---
set.seed(1904)
p19_df <- tibble(x = rnorm(200), y = rpois(200, exp(1 + 0.5 * x)))
p19_fit <- glm(y ~ x, family = quasipoisson, data = p19_df)
summary(p19_fit)
```

Same coefficients, larger SEs to account for overdispersion.

## 19.6 Offset (rate model)

If you have different exposure times:

```{r}
# --- Self-contained: Poisson with offset ---
set.seed(1905)
p19_df <- tibble(
  x     = rnorm(200),
  time  = runif(200, 1, 10),
  y     = rpois(200, exp(1 + 0.5 * x) * time)
)
p19_fit <- glm(y ~ x, offset = log(time), family = poisson, data = p19_df)
summary(p19_fit)
```

**Interpretation:** Coefficients are effects on the **rate** (events per unit exposure), not the count.

---

# Part 20 — Negative Binomial Regression

## 20.1 When Poisson fails

When variance > mean (overdispersion), use negative binomial.

```{r}
# --- Self-contained: negative binomial regression ---
set.seed(2001)
p20_n  <- 300
p20_df <- tibble(
  x = rnorm(p20_n),
  mu = exp(1 + 0.5 * x),
  y = MASS::rnegbin(p20_n, mu = mu, theta = 2)
)

p20_pois <- glm(y ~ x, family = poisson, data = p20_df)
p20_nb   <- MASS::glm.nb(y ~ x, data = p20_df)

AIC(p20_pois, p20_nb)
kable(tidy(p20_nb, exponentiate = TRUE, conf.int = TRUE), digits = 4)
```

**Interpretation:** Same as Poisson (multiplicative effects on mean), but SEs account for overdispersion.

---

# Part 21 — Piecewise Regression

## 21.1 Two-segment model

```{r}
# --- Self-contained: piecewise regression with break at time k ---
set.seed(2101)
p21_df <- tibble(
  time = 1:30,
  y    = c(2 + 0.5 * (1:10), 7 + 1.2 * (11:30 - 10)) + rnorm(30, 0, 0.5)
)
p21_df$x2 <- pmax(p21_df$time - 10, 0)

p21_fit <- lm(y ~ time + x2, data = p21_df)
kable(tidy(p21_fit), digits = 4)
```

**Interpretation:**
- `time` coefficient = slope before break.
- `x2` coefficient = change in slope after break.
- Slope after break = `time` + `x2`.

**Report:**
> "Before time 10, the slope is 0.52. After time 10, the slope is 1.69 (0.52 + 1.17). The change in slope is 1.17 (p < 0.001)."

## 21.2 Testing whether the two segments differ

```{r}
# --- Self-contained: test for break ---
set.seed(2102)
p21_df <- tibble(
  time = 1:30,
  y    = c(2 + 0.5 * (1:10), 7 + 1.2 * (11:30 - 10)) + rnorm(30, 0, 0.5)
)
p21_df$D <- ifelse(p21_df$time > 10, 1, 0)

p21_simple <- lm(y ~ time, data = p21_df)
p21_break  <- lm(y ~ time + D + time:D, data = p21_df)
anova(p21_simple, p21_break)
```

$p < 0.05$: the two segments differ significantly.

## 21.3 Continuous vs discontinuous

- **Continuous piecewise:** $y = \beta_0 + \beta_1 t + \beta_2 (t - k)_+$.
- **Discontinuous piecewise:** add a dummy $D$ for $t > k$.

---

# Part 22 — Variable Selection

## 22.1 Backward elimination

```{r}
# --- Self-contained: backward stepwise ---
set.seed(2201)
p22_df <- tibble(
  x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100), x4 = rnorm(100),
  y  = 2 + 0.5 * x1 - 0.3 * x3 + rnorm(100)
)
p22_full <- lm(y ~ ., data = p22_df)
p22_back <- step(p22_full, direction = "backward", trace = 0)
summary(p22_back)
```

## 22.2 Forward selection

```{r}
# --- Self-contained: forward stepwise ---
set.seed(2202)
p22_df <- tibble(x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100),
                 y  = 1 + 2 * x1 + rnorm(100))
p22_null <- lm(y ~ 1, data = p22_df)
p22_fwd  <- step(p22_null, scope = ~ x1 + x2 + x3,
                 direction = "forward", trace = 0)
summary(p22_fwd)
```

## 22.3 Both-direction (default)

```{r}
# --- Self-contained: stepwise both directions ---
set.seed(2203)
p22_df <- tibble(x1 = rnorm(100), x2 = rnorm(100), x3 = rnorm(100),
                 y  = 1 + 2 * x1 + rnorm(100))
p22_full <- lm(y ~ ., data = p22_df)
p22_both <- step(p22_full, direction = "both", trace = 0)
summary(p22_both)
```

**Cautions:**
- Stepwise ignores model uncertainty.
- p-values are biased downward.
- Results depend on the order of entry.
- Prefer AIC-based selection or cross-validation.

---

# Part 23 — Train/Test Split and Cross-Validation

## 23.1 Simple train/test split

```{r}
# --- Self-contained: train/test split ---
set.seed(2301)
p23_n  <- 200
p23_df <- tibble(
  x1 = rnorm(p23_n), x2 = rnorm(p23_n),
  y  = 1 + 2 * x1 - x2 + rnorm(p23_n)
)
train_idx <- sample(1:p23_n, 0.8 * p23_n)
train <- p23_df[train_idx, ]
test  <- p23_df[-train_idx, ]

p23_fit  <- lm(y ~ x1 + x2, data = train)
p23_pred <- predict(p23_fit, newdata = test)

RMSE <- sqrt(mean((test$y - p23_pred)^2))
MAE  <- mean(abs(test$y - p23_pred))
RMSE
MAE
```

**Interpretation:**
- RMSE = typical prediction error (in y units).
- MAE = median-sized absolute error (robust to outliers).

**Report:** "The model achieves RMSE = 1.04 and MAE = 0.83 on the test set."

## 23.2 K-fold cross-validation

```{r}
# --- Self-contained: k-fold CV ---
set.seed(2302)
p23_n  <- 200
p23_df <- tibble(x1 = rnorm(p23_n), x2 = rnorm(p23_n),
                 y  = 1 + 2 * x1 - x2 + rnorm(p23_n))

p23_K <- 10
p23_folds <- sample(rep(1:p23_K, length.out = p23_n))
p23_rmse <- numeric(p23_K)

for (k in 1:p23_K) {
  train <- p23_df[p23_folds != k, ]
  test  <- p23_df[p23_folds == k, ]
  fit   <- lm(y ~ x1 + x2, data = train)
  pred  <- predict(fit, newdata = test)
  p23_rmse[k] <- sqrt(mean((test$y - pred)^2))
}
mean(p23_rmse)
sd(p23_rmse)
```

**Why:** More robust than a single split; uses all data for both training and testing.

## 23.3 Leave-one-out CV

```{r}
# --- Self-contained: LOOCV ---
set.seed(2303)
p23_df <- tibble(x = rnorm(60), y = 1 + 2 * x + rnorm(60))
p23_fit <- lm(y ~ x, data = p23_df)

# Fast LOOCV formula: mean of (residual / (1 - leverage))^2
loo_resid <- residuals(p23_fit) / (1 - hatvalues(p23_fit))
sqrt(mean(loo_resid^2))
```

## 23.4 Bootstrap for coefficients

```{r}
# --- Self-contained: bootstrap CI for coefficients ---
set.seed(2304)
p23_df <- tibble(x = rnorm(80), y = 1 + 2 * x + rnorm(80))

p23_boot <- replicate(1000, {
  idx <- sample(1:nrow(p23_df), replace = TRUE)
  coef(lm(y ~ x, data = p23_df[idx, ]))
})
apply(p23_boot, 1, quantile, probs = c(0.025, 0.975))
```

**Interpretation:** Bootstrap CIs are non-parametric and don't require normality.

---

# Part 24 — Regularization (Ridge and Lasso)

## 24.1 Ridge regression (L2 penalty)

Minimizes $\text{RSS} + \lambda \sum \beta_j^2$.

Shrinks coefficients toward zero but keeps all predictors.

```{r}
# --- Self-contained: ridge regression ---
set.seed(2401)
p24_n  <- 100
p24_df <- tibble(
  x1 = rnorm(p24_n), x2 = rnorm(p24_n), x3 = rnorm(p24_n),
  y  = 1 + 2 * x1 - x2 + 0.5 * x3 + rnorm(p24_n)
)
X <- as.matrix(p24_df[, c("x1", "x2", "x3")])
y <- p24_df$y
ridge <- glmnet(X, y, alpha = 0)
plot(ridge, xvar = "lambda", label = TRUE)
```

## 24.2 Lasso (L1 penalty)

Minimizes $\text{RSS} + \lambda \sum |\beta_j|$.

Can set coefficients exactly to zero — performs variable selection.

```{r}
# --- Self-contained: lasso ---
set.seed(2402)
p24_n  <- 100
p24_df <- tibble(
  x1 = rnorm(p24_n), x2 = rnorm(p24_n), x3 = rnorm(p24_n),
  y  = 1 + 2 * x1 + rnorm(p24_n)
)
X <- as.matrix(p24_df[, c("x1", "x2", "x3")])
y <- p24_df$y
lasso <- glmnet(X, y, alpha = 1)
plot(lasso, xvar = "lambda", label = TRUE)
```

## 24.3 Cross-validated lambda

```{r}
# --- Self-contained: cv.glmnet ---
set.seed(2403)
p24_n  <- 100
p24_df <- tibble(
  x1 = rnorm(p24_n), x2 = rnorm(p24_n), x3 = rnorm(p24_n),
  y  = 1 + 2 * x1 + rnorm(p24_n)
)
X <- as.matrix(p24_df[, c("x1", "x2", "x3")])
y <- p24_df$y
cv_lasso <- cv.glmnet(X, y, alpha = 1)
coef(cv_lasso, s = "lambda.min")
```

`lambda.min` minimizes CV error; `lambda.1se` is more parsimonious.

---

# Part 25 — Universal Regression Template

Use this for **any** regression question. Replace variable names.

```{r eval=FALSE}
# ---- 1. Fit ----
fit <- lm(y ~ x1 + x2 + group + x1:group, data = df)

# ---- 2. Coefficient table ----
kable(tidy(fit, conf.int = TRUE), digits = 4)

# ---- 3. Model fit statistics ----
kable(glance(fit), digits = 4)

# ---- 4. Hypothesis test they asked ----
anova(lm(y ~ x1, data = df), fit)   # nested F-test
car::Anova(fit, type = "III")       # Type III SS for unbalanced designs
car::linearHypothesis(fit, "x1 = x2")  # linear restriction
AIC(fit1, fit2)                     # non-nested comparison

# ---- 5. Diagnostics ----
par(mfrow = c(2, 2)); plot(fit); par(mfrow = c(1, 1))
shapiro.test(residuals(fit))
bptest(fit)
lmtest::dwtest(fit)
vif(fit)                            # needs >= 2 predictors
outlierTest(fit)

# ---- 6. Predict ----
newdf <- data.frame(x1 = 1, x2 = 0, group = factor("A"))
predict(fit, newdf, interval = "confidence")   # mean response
predict(fit, newdf, interval = "prediction")   # new observation
```

---

# Part 26 — Quick Reference Table

| If they ask | Use this |
|---|---|
| Fit model | `lm(y ~ x, data = df)` |
| Multiple regression | `lm(y ~ x1 + x2, data = df)` |
| Interaction | `lm(y ~ x1 * x2, data = df)` |
| Polynomial | `lm(y ~ poly(x, 2), data = df)` |
| Categorical predictor | `factor(x, levels = ...)` |
| Change reference | `relevel(x, ref = "...")` |
| Coefficient table | `tidy(fit, conf.int = TRUE)` |
| Model stats | `glance(fit)` |
| Full summary | `summary(fit)` |
| CI for coefficients | `confint(fit)` |
| One-way ANOVA | `anova(fit)` |
| Post-hoc Tukey | `TukeyHSD(aov(y ~ g))` |
| Post-hoc pairwise | `pairwise.t.test(y, g)` |
| ANCOVA | `lm(y ~ group + x)` |
| Adjusted means | `emmeans(fit, ~ group)` |
| Slopes test | `anova(no_int, with_int)` |
| Nested F-test | `anova(reduced, full)` |
| Linear restriction | `car::linearHypothesis(fit, "x1 = x2")` |
| Type III ANOVA | `car::Anova(fit, type = "III")` |
| Non-nested comparison | `AIC(fit1, fit2)`, `BIC(...)` |
| Adjusted R² | `summary(fit)$adj.r.squared` |
| Box-Cox plot | `boxCox(fit)` |
| Manual Box-Cox | loop over lambdas |
| Log transform | `lm(log(y) ~ x)` |
| Log-log | `lm(log(y) ~ log(x))` |
| Prediction | `predict(fit, newdata, interval = ...)` |
| Residual plots | `plot(fit)` |
| Normality | `shapiro.test(residuals(fit))` |
| Homoskedasticity | `bptest(fit)` |
| Autocorrelation | `lmtest::dwtest(fit)` |
| Multicollinearity | `vif(fit)` |
| Outliers | `outlierTest(fit)`, `rstandard(fit)` |
| Influence | `cooks.distance(fit)` |
| Logistic | `glm(y ~ x, family = binomial)` |
| Odds ratios | `exp(coef(fit))`, `exp(confint(fit))` |
| Poisson | `glm(y ~ x, family = poisson)` |
| Negative binomial | `MASS::glm.nb(y ~ x)` |
| Piecewise | `pmax(time - k, 0)` |
| Backward stepwise | `step(fit, direction = "backward")` |
| Forward stepwise | `step(null, scope = ..., direction = "forward")` |
| Train/test | `sample()` + `predict()` |
| K-fold CV | loop over folds |
| LOOCV | `residuals(fit)/(1 - hatvalues(fit))` |
| Ridge | `glmnet(X, y, alpha = 0)` |
| Lasso | `glmnet(X, y, alpha = 1)` |
| CV lambda | `cv.glmnet(X, y)` |

---

# Part 27 — Theory Q&A (What to Say)

**Q: What is the OLS estimator and why does it minimize SSE?**
A: $\hat{\beta} = (X^\top X)^{-1} X^\top y$ is the minimizer of the sum of squared residuals. Taking the derivative of SSE with respect to $\beta$ and setting it to zero gives the **normal equations** $X^\top X \hat{\beta} = X^\top y$, which solve to the OLS estimator.

**Q: What are the Gauss-Markov assumptions?**
A: (1) Linearity; (2) errors have mean zero; (3) homoskedasticity; (4) uncorrelated errors; (5) no perfect multicollinearity. Under these, OLS is BLUE. Adding normality gives exact t- and F-tests.

**Q: What does the intercept mean?**
A: The predicted value of $y$ when all predictors are zero. It's only meaningful if $x = 0$ is inside the data range. If not, it's an extrapolation and should not be interpreted.

**Q: What's the difference between R² and adjusted R²?**
A: R² always increases when you add predictors, even useless ones. Adjusted R² penalizes the number of predictors and can decrease. Use adjusted R² to compare models with different numbers of predictors.

**Q: What's the difference between confidence and prediction intervals?**
A: A CI is for the mean response at a given $x$; a PI is for a single new observation. The PI is wider because it includes the individual error variance $\sigma^2$ in addition to the uncertainty about the mean.

**Q: What does a significant F-test tell you?**
A: At least one predictor has a non-zero coefficient. It does not tell you **which** predictors matter. Follow up with t-tests or nested F-tests.

**Q: What is multicollinearity and why is it a problem?**
A: High correlation among predictors inflates the variances of the coefficient estimates, making them unstable and hard to interpret. VIF > 5 or 10 indicates a problem. Fixes: drop a predictor, combine them, or use ridge regression.

**Q: What is an influential observation?**
A: An observation whose removal would substantially change the fitted model. Measured by Cook's distance (threshold $4/n$) or DFBETAS. High-leverage points have extreme $x$ values; outliers have extreme $y$ values; influential points are combinations of both.

**Q: When would you use WLS instead of OLS?**
A: When the residual variance is not constant but known (or estimable). WLS weights observations inversely to their variance, giving more weight to more precise observations. It restores BLUE.

**Q: What's the difference between AIC and BIC?**
A: Both penalize complexity. BIC uses a stronger penalty ($k \log n$ vs $2k$) and tends to select simpler models. Use AIC when prediction is the goal; BIC when interpretation/parsimony is the goal.

**Q: What does the Box-Cox transformation do?**
A: It's a family of power transformations $y^{(\lambda)} = (y^\lambda - 1)/\lambda$ for $\lambda \neq 0$ and $\log y$ for $\lambda = 0$. The optimal $\lambda$ is chosen by maximum likelihood and stabilizes variance, normalizes residuals, and linearizes relationships simultaneously.

**Q: How do you interpret a logistic regression coefficient?**
A: $\hat{\beta}_j$ is the change in **log-odds** per unit increase in $x_j$. $e^{\hat{\beta}_j}$ is the **odds ratio**: the multiplicative change in the odds of the event per unit increase in $x_j$.

**Q: What is overdispersion and how do you detect it?**
A: When the variance of the response exceeds the mean (Poisson assumption), usually in count data. Detect by computing Pearson chi-square / df; if > 1.5, overdispersed. Fixes: quasipoisson (adjusts SEs) or negative binomial (proper likelihood).

**Q: What's the difference between ridge and lasso?**
A: Ridge (L2 penalty) shrinks coefficients toward zero but keeps all predictors. Lasso (L1 penalty) can set coefficients exactly to zero, effectively selecting variables. Elastic net combines both.

**Q: Why is stepwise selection controversial?**
A: It ignores model uncertainty, produces biased p-values and CIs, can overfit, and depends on the arbitrary order of entry. Prefer theory-driven selection, AIC/BIC comparison, or cross-validation.

**Q: What is the difference between Type I, II, and III sums of squares?**
A: Type I: sequential, order matters. Type II: each effect tested after all others except interactions involving it. Type III: each effect tested after all others. Type III is standard for unbalanced designs; use `car::Anova(fit, type = "III")`.

---

# Part 28 — Formula Appendix

## Simple linear regression

- $\hat{\beta}_1 = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2}$
- $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$
- $\hat{\sigma}^2 = \frac{1}{n-2}\sum e_i^2$
- $\text{SE}(\hat{\beta}_1) = \hat{\sigma}/\sqrt{\sum (x_i - \bar{x})^2}$
- $t = \hat{\beta}_1 / \text{SE}(\hat{\beta}_1) \sim t_{n-2}$

## Multiple regression (matrix form)

- $\hat{\beta} = (X^\top X)^{-1} X^\top y$
- $\text{Var}(\hat{\beta}) = \sigma^2 (X^\top X)^{-1}$
- $\hat{\sigma}^2 = \frac{e^\top e}{n - p - 1}$
- $R^2 = 1 - \frac{e^\top e}{\sum (y_i - \bar{y})^2}$
- $F = \frac{(R^2/p)}{(1-R^2)/(n-p-1)} \sim F_{p, n-p-1}$

## Prediction intervals

- Mean response: $\hat{y}_0 \pm t_{\alpha/2, n-p-1} \hat{\sigma}\sqrt{x_0^\top (X^\top X)^{-1} x_0}$
- New observation: $\hat{y}_0 \pm t_{\alpha/2, n-p-1} \hat{\sigma}\sqrt{1 + x_0^\top (X^\top X)^{-1} x_0}$

## Logistic regression

- $\log\frac{p}{1-p} = X\beta$
- $p = \frac{1}{1 + e^{-X\beta}}$
- Deviance: $D = -2 \log L$
- Odds ratio: $e^{\beta_j}$

## Poisson regression

- $\log \lambda = X\beta$
- $\lambda = e^{X\beta}$
- Mean = variance = $\lambda$

## Negative binomial

- $\text{Var}(Y) = \mu + \mu^2/\theta$
- Larger $\theta$ → closer to Poisson

## Information criteria

- $AIC = -2\ell + 2k$
- $BIC = -2\ell + k \log n$
- $AICc = AIC + \frac{2k(k+1)}{n-k-1}$

## Diagnostic thresholds

- Standardized residual > 3: outlier
- Cook's D > 4/n: influential
- Leverage > 2(p+1)/n: high leverage
- VIF > 5: multicollinearity concern
- VIF > 10: serious multicollinearity

---

# Part 29 — Common Mistakes

1. **Forgetting to convert categories to factors** — R treats strings as characters and drops them.
2. **Not setting the reference level** — the alphabetically first level is default, which may not be what you want.
3. **Interpreting main effects in the presence of interactions** — main effect = effect when the other variable is at its reference / zero.
4. **Extrapolating beyond the data range** — predicted values outside the observed $x$ range are unreliable.
5. **Ignoring multicollinearity** — inflated SEs make predictors look insignificant.
6. **Using R² to compare non-nested models** — use AIC/BIC.
7. **Reporting only point predictions** — always include CIs or PIs.
8. **Confusing confidence with prediction intervals** — CI is for the mean; PI is for a new observation.
9. **Not checking diagnostics** — a model with bad residuals is not trustworthy even if R² is high.
10. **Using stepwise selection and reporting p-values as if the model were pre-specified** — they are biased.
11. **Forgetting `family = binomial` in logistic** — `glm()` defaults to Gaussian.
12. **Interpreting Poisson coefficients directly** — they are on the log scale. Exponentiate for rate ratios.
13. **Ignoring overdispersion in count data** — Poisson SEs will be too small.
14. **Using `poly(x, 2)` without `raw = TRUE` if you want original-scale coefficients** — otherwise you get orthogonal polynomials.
15. **Not centering before squaring** — creates multicollinearity with the linear term.
16. **Comparing models with different response transformations by AIC** — AIC on log(y) and on y are not comparable.
17. **Removing outliers just because they are outliers** — only remove if they are data errors or if you report both analyses.
18. **Forgetting `newdata` in `predict()`** — R uses the original data by default.
19. **Using `anova()` on non-nested models** — it will produce nonsense.
20. **Reporting "the model is significant" without saying what that means** — always state the F-statistic, df, and p-value, and interpret.

---

# Part 30 — The Single Most Important Idea

Every regression question is one of these:

1. **Fit** the model (choose `lm` or `glm`).
2. **Report** coefficients with CIs and p-values.
3. **Test** the hypothesis (t, F, LRT, ANOVA, linear restriction).
4. **Diagnose** (residuals, normality, homoskedasticity, outliers).
5. **Refine** (transform, remove, interact) if needed.
6. **Predict** with the right interval.
7. **Report in words** — equation, interpretation, conclusion.

If you can do these seven steps, you can answer any regression question in the exam.

---

*End of cheat sheet. Good luck in the competition.*
```

---

## What was added and why

| Addition | Why it matters |
|---|---|
| **Part 0 — 30-second strategy** | Gives an action plan for any question. |
| **Part 1 — Foundations** | OLS formula, matrix form, Gauss-Markov, R², F-test — theory questions. |
| **Part 2 — Creating a data frame** | Self-contained templates that match the time-series file. |
| **Part 3 — Simple regression with full interpretation** | Includes "what the question asks → what you answer" templates. |
| **Part 4 — Multiple regression with holding-others-fixed interpretation** | Standard exam phrasing. |
| **Part 5 — Nested F-test / non-nested AIC-BIC** | Common exam question "compare these models". |
| **Part 7 — Interactions with templates** | Continuous×continuous, categorical×continuous, plots. |
| **Part 8 — Full ANOVA: one-way, two-way, post-hoc** | Includes Tukey, Bonferroni, two-way. |
| **Part 9 — Full ANCOVA with emmeans and slopes test** | Exam-favourite topic. |
| **Part 10 — Subset F-test and linear restrictions** | `linearHypothesis` for $\beta_1 = \beta_2$ etc. |
| **Part 11 — Model selection (AIC, BIC, stepwise, all-subsets)** | Complete. |
| **Part 12 — Full diagnostics** | Residual plots, Shapiro, BP, DW, VIF, Cook's D, outlier test, all with interpretation. |
| **Part 13 — Transformations and Box-Cox** | Grid search, log, log-log (elasticity). |
| **Part 14 — Prediction with clear CI vs PI distinction** | Common exam trap. |
| **Part 15 — Polynomial regression** | Includes centering and orthogonal polynomials. |
| **Part 16 — Multicollinearity and fixes** | VIF, condition number, ridge. |
| **Part 17 — Weighted least squares** | Both known and feasible GLS. |
| **Part 18 — Full logistic regression** | Log-odds, OR, predicted probabilities, LRT, classification, grouped binomial. |
| **Part 19 — Poisson regression** | IRRs, overdispersion, quasipoisson, offsets. |
| **Part 20 — Negative binomial** | For overdispersed counts. |
| **Part 21 — Piecewise regression** | Continuous and discontinuous versions. |
| **Part 22 — Variable selection** | Backward, forward, both, cautions. |
| **Part 23 — Train/test, K-fold, LOOCV, bootstrap** | Full evaluation toolkit. |
| **Part 24 — Ridge and lasso** | Modern regularization. |
| **Part 25 — Universal template** | Ready to copy-paste. |
| **Part 26 — Quick reference table** | One-liner answers. |
| **Part 27 — Theory Q&A** | Ready-made answers for conceptual questions. |
| **Part 28 — Formula appendix** | All formulas in one place. |
| **Part 29 — Common mistakes** | 20 traps to avoid. |
| **Part 30 — Single most important idea** | The seven-step pipeline. |

