---
title: "The Ultimate Sampling Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(knitr)
```

# How to Use This Cheat Sheet

| Topic | Section |
|---|---|
| 30-second strategy | 0 |
| Foundations of sampling theory | 1 |
| Building a finite population | 2 |
| Simple Random Sampling (SRS) | 3 |
| Stratified sampling | 4 |
| Systematic sampling | 5 |
| Cluster sampling | 6 |
| Two-stage sampling | 7 |
| Ratio and regression estimators | 8 |
| Bias, MSE, and relative efficiency | 9 |
| Coverage probability simulations | 10 |
| Sample size determination | 11 |
| Design effects | 12 |
| Horvitz-Thompson estimator | 13 |
| Probability proportional to size (PPS) | 14 |
| Non-response | 15 |
| Bootstrap variance | 16 |
| Theory Q&A | 17 |
| Exam scenarios | 18 |
| Quick reference | 19 |
| Common mistakes | 20 |

**Golden rule:** every chunk is self-contained.

---

# Part 0 — The 30-Second Exam Strategy

For **any** sampling question, run this pipeline:

1. **Identify the design:** SRS, stratified, systematic, cluster, two-stage, PPS?
2. **Identify the parameter:** population mean, total, proportion?
3. **Compute the estimator** (usually a weighted mean or proportion).
4. **Compute the variance** using the design-specific formula, including FPC.
5. **Build a CI** with the normal approximation or t.
6. **Simulate** to check bias, variance, MSE, coverage.
7. **Compare designs** using relative efficiency or design effect.

If you can do these seven steps, you can answer any sampling question.

---

# Part 1 — Foundations

## 1.1 Why sampling?

A **population** has $N$ units. Measuring all of them (census) is expensive. A **sample** of $n$ units estimates the population parameters.

**Goal:** Estimate a population **parameter** ($\mu$, $\tau = N\mu$, $P$) from a **sample statistic** ($\bar{y}$, $\hat{\tau}$, $\hat{p}$).

## 1.2 Key vocabulary

| Term | Meaning |
|---|---|
| **Population** | Complete set of $N$ units |
| **Sampling frame** | List of all units in the population |
| **Sample** | Subset of $n$ units selected from the frame |
| **Parameter** | Population value (fixed, unknown): $\mu$, $\tau$, $P$ |
| **Statistic** | Sample value (random): $\bar{y}$, $\hat{\tau}$, $\hat{p}$ |
| **Sampling design** | Rule for selecting the sample |
| **Inclusion probability** $\pi_i$ | Probability unit $i$ is in the sample |
| **Design weight** $w_i = 1/\pi_i$ | How many population units $i$ represents |

## 1.3 Common parameters and their estimators

| Parameter | Symbol | Estimator |
|---|---|---|
| Population mean | $\mu = \frac{1}{N}\sum_{i=1}^N y_i$ | $\bar{y} = \frac{1}{n}\sum_{i \in S} y_i$ |
| Population total | $\tau = \sum_{i=1}^N y_i = N\mu$ | $\hat{\tau} = N \bar{y}$ |
| Population proportion | $P = \frac{1}{N}\sum_{i=1}^N y_i$ (binary) | $\hat{p} = \bar{y}$ (binary) |
| Population variance | $S^2 = \frac{1}{N-1}\sum_{i=1}^N (y_i - \mu)^2$ | $s^2 = \frac{1}{n-1}\sum_{i \in S}(y_i - \bar{y})^2$ |

## 1.4 Bias, variance, MSE

- **Bias:** $\text{Bias}(\hat{\theta}) = E[\hat{\theta}] - \theta$.
- **Variance:** $\text{Var}(\hat{\theta}) = E[(\hat{\theta} - E[\hat{\theta}])^2]$.
- **Mean Squared Error:** $\text{MSE}(\hat{\theta}) = E[(\hat{\theta} - \theta)^2] = \text{Bias}^2 + \text{Variance}$.

**Unbiased** = $\text{Bias} = 0$.
**Consistent** = bias and variance → 0 as $n \to \infty$.
**Relative efficiency of $\hat{\theta}_1$ vs $\hat{\theta}_2$:** $\text{RE} = \text{MSE}(\hat{\theta}_2)/\text{MSE}(\hat{\theta}_1)$.

## 1.5 Finite Population Correction (FPC)

$$f = \frac{n}{N} \quad \text{(sampling fraction)}$$
$$\text{FPC} = 1 - \frac{n}{N}$$

**Effect:** Reduces the variance of the estimator because when you've sampled most of the population, there's less uncertainty about the mean.

- $n \ll N$: FPC ≈ 1 → ignore it.
- $n = N$: FPC = 0 → zero variance (census, no uncertainty).

## 1.6 Variance of the sample mean (SRS)

$$\text{Var}(\bar{y}) = \left(1 - \frac{n}{N}\right) \frac{S^2}{n}$$

where $S^2$ is the **population** variance. In practice we don't know $S^2$, so we estimate:

$$\widehat{\text{Var}}(\bar{y}) = \left(1 - \frac{n}{N}\right) \frac{s^2}{n}$$

## 1.7 Confidence interval for $\mu$

$$\bar{y} \pm z_{\alpha/2} \sqrt{\left(1 - \frac{n}{N}\right) \frac{s^2}{n}}$$

For small $n$, use $t_{n-1}$ instead of $z$.

## 1.8 Independence of the sample

**Independence** means one unit's selection doesn't affect another's. In SRS without replacement, units are **not independent** (but **exchangeable**). Design-based inference handles this using **inclusion probabilities** instead of independence.

## 1.9 Sampling designs at a glance

| Design | When to use | Key feature |
|---|---|---|
| **SRS** | Simple, no auxiliary info | Equal $\pi_i = n/N$ |
| **Stratified** | Population has distinct subgroups | Sample within strata |
| **Systematic** | Ordered frame, quick | Every $k$-th unit |
| **Cluster** | Units geographically grouped | Sample clusters, not units |
| **Two-stage** | Clusters + units | Sample PSUs then SSUs |
| **PPS** | Sizes vary widely | Larger units more likely |

---

# Part 2 — Building a Finite Population

## 2.1 A simple finite population

```{r}
# --- Self-contained: build finite population ---
set.seed(201)
p2_N    <- 1000
p2_pop  <- rnorm(p2_N, mean = 50, sd = 10)

p2_mu    <- mean(p2_pop)
p2_S2    <- var(p2_pop)
p2_sigma <- sqrt(p2_S2)

round(c(mu = p2_mu, var = p2_S2, sd = p2_sigma), 4)
```

## 2.2 A stratified population

```{r}
# --- Self-contained: stratified population ---
set.seed(202)
p2_N1 <- 100; p2_N2 <- 200; p2_N3 <- 300
p2_N  <- p2_N1 + p2_N2 + p2_N3

p2_s1 <- rnorm(p2_N1, 10, 2)
p2_s2 <- rnorm(p2_N2, 12, 3)
p2_s3 <- rnorm(p2_N3, 15, 4)

p2_pop    <- c(p2_s1, p2_s2, p2_s3)
p2_strata <- c(rep(1, p2_N1), rep(2, p2_N2), rep(3, p2_N3))

# Population parameters
p2_mu       <- mean(p2_pop)
p2_S2       <- var(p2_pop)
p2_strata_N <- c(p2_N1, p2_N2, p2_N3)
p2_strata_W <- p2_strata_N / p2_N

# Stratum-specific means and variances
p2_strata_stats <- tibble(
  stratum = 1:3,
  N_h     = p2_strata_N,
  W_h     = p2_strata_W,
  mu_h    = c(mean(p2_s1), mean(p2_s2), mean(p2_s3)),
  S2_h    = c(var(p2_s1),  var(p2_s2),  var(p2_s3))
)
kable(p2_strata_stats, digits = 4)
```

## 2.3 Population as a discrete distribution (exam style)

```{r}
# --- Self-contained: discrete population with strata ---
set.seed(203)
p2_N1 <- 100; p2_N2 <- 200; p2_N3 <- 300
p2_N  <- p2_N1 + p2_N2 + p2_N3

p2_s1 <- sample(1:5, p2_N1, replace = TRUE, prob = rep(0.2, 5))
p2_s2 <- sample(1:5, p2_N2, replace = TRUE, prob = c(0.1, 0.1, 0.1, 0.1, 0.6))
p2_s3 <- sample(1:5, p2_N3, replace = TRUE, prob = c(0.1, 0.3, 0.3, 0.1, 0.2))

p2_pop    <- c(p2_s1, p2_s2, p2_s3)
p2_strata <- c(rep(1, p2_N1), rep(2, p2_N2), rep(3, p2_N3))
p2_mu     <- mean(p2_pop)

round(p2_mu, 4)
table(p2_strata, p2_pop) %>% kable()
```

## 2.4 Summary table by stratum

```{r}
# --- Self-contained: summary by stratum ---
set.seed(204)
p2_N1 <- 100; p2_N2 <- 200; p2_N3 <- 300
p2_s1 <- rnorm(p2_N1, 10, 2)
p2_s2 <- rnorm(p2_N2, 12, 3)
p2_s3 <- rnorm(p2_N3, 15, 4)
p2_pop    <- c(p2_s1, p2_s2, p2_s3)
p2_strata <- c(rep(1, p2_N1), rep(2, p2_N2), rep(3, p2_N3))

tibble(y = p2_pop, h = factor(p2_strata)) %>%
  group_by(h) %>%
  summarise(n = n(), mean = mean(y), sd = sd(y), .groups = "drop") %>%
  kable(digits = 3)
```

---

# Part 3 — Simple Random Sampling (SRS)

## 3.1 Draw an SRS sample

```{r}
# --- Self-contained: SRS sample ---
set.seed(301)
p3_N   <- 500
p3_pop <- rnorm(p3_N, 50, 10)

p3_n   <- 50
p3_srs <- sample(p3_pop, p3_n, replace = FALSE)

p3_ybar <- mean(p3_srs)
p3_s2   <- var(p3_srs)

round(c(ybar = p3_ybar, s2 = p3_s2), 4)
```

## 3.2 Variance with FPC

```{r}
# --- Self-contained: variance of ybar ---
set.seed(302)
p3_N   <- 500
p3_pop <- rnorm(p3_N, 50, 10)
p3_n   <- 50
p3_srs <- sample(p3_pop, p3_n)
p3_ybar <- mean(p3_srs)
p3_s2   <- var(p3_srs)

p3_Var_ybar <- (1 - p3_n / p3_N) * p3_s2 / p3_n
p3_SE_ybar  <- sqrt(p3_Var_ybar)

round(c(ybar = p3_ybar, var_ybar = p3_Var_ybar, se = p3_SE_ybar), 4)
```

## 3.3 Confidence interval for $\mu$

```{r}
# --- Self-contained: 95% CI for mu ---
set.seed(303)
p3_N   <- 500
p3_pop <- rnorm(p3_N, 50, 10)
p3_n   <- 50
p3_srs <- sample(p3_pop, p3_n)
p3_ybar <- mean(p3_srs)
p3_s2   <- var(p3_srs)
p3_se   <- sqrt((1 - p3_n / p3_N) * p3_s2 / p3_n)
p3_z    <- qnorm(0.975)

p3_ci <- c(p3_ybar - p3_z * p3_se, p3_ybar + p3_z * p3_se)
round(p3_ci, 4)
```

**Report template:** "The 95% CI for the population mean is [$\text{lo}, \text{hi}$]."

## 3.4 Estimate the population total

```{r}
# --- Self-contained: total estimate ---
set.seed(304)
p3_N   <- 500
p3_pop <- rnorm(p3_N, 50, 10)
p3_n   <- 50
p3_srs <- sample(p3_pop, p3_n)

p3_tau_hat <- p3_N * mean(p3_srs)
p3_Var_tau <- p3_N^2 * (1 - p3_n / p3_N) * var(p3_srs) / p3_n

round(c(total = p3_tau_hat, se_total = sqrt(p3_Var_tau)), 4)
```

## 3.5 SRS for a proportion

```{r}
# --- Self-contained: proportion ---
set.seed(305)
p3_N   <- 1000
p3_pop <- rbinom(p3_N, 1, 0.4)   # 40% successes
p3_n   <- 100
p3_srs <- sample(p3_pop, p3_n)

p3_phat    <- mean(p3_srs)
p3_Var_ph  <- (1 - p3_n / p3_N) * p3_phat * (1 - p3_phat) / (p3_n - 1)
p3_SE_ph   <- sqrt(p3_Var_ph)

round(c(phat = p3_phat, se = p3_SE_ph), 4)
```

## 3.6 Sample size for a target margin of error

$$n_0 = \frac{z^2 S^2}{e^2}, \quad n = \frac{n_0}{1 + n_0/N}$$

```{r}
# --- Self-contained: sample size for SRS ---
p3_N  <- 500
p3_S2 <- 100    # assume known or guessed from pilot
p3_e  <- 2      # target margin of error
p3_z  <- qnorm(0.975)

p3_n0 <- p3_z^2 * p3_S2 / p3_e^2
p3_n  <- ceiling(p3_n0 / (1 + p3_n0 / p3_N))
p3_n
```

**Interpretation:** To achieve a margin of error of 2 units, sample at least $n$ units.

## 3.7 SRS summary

| Property | Value |
|---|---|
| Inclusion probability $\pi_i$ | $n/N$ for all $i$ |
| Estimator of mean | $\bar{y}$ |
| Estimator of total | $N\bar{y}$ |
| Var of mean | $(1 - n/N)S^2/n$ |
| Var of total | $N^2(1 - n/N)S^2/n$ |
| Best for | No prior information, symmetric units |

---

# Part 4 — Stratified Sampling

## 4.1 What it is

- Divide the population into $H$ non-overlapping **strata**.
- Sample independently from each stratum.
- Combine stratum estimates using population weights.

**Estimator:**
$$\bar{y}_{st} = \sum_{h=1}^H W_h \bar{y}_h, \quad W_h = N_h/N$$

**Variance:**
$$\text{Var}(\bar{y}_{st}) = \sum_{h=1}^H W_h^2 \left(1 - \frac{n_h}{N_h}\right)\frac{S_h^2}{n_h}$$

## 4.2 Proportional allocation

$$n_h = n \cdot \frac{N_h}{N}$$

Sample size per stratum is proportional to stratum size.

```{r}
# --- Self-contained: proportional allocation ---
set.seed(401)
p4_N1 <- 100; p4_N2 <- 200; p4_N3 <- 300
p4_N  <- p4_N1 + p4_N2 + p4_N3
p4_s1 <- rnorm(p4_N1, 10, 2)
p4_s2 <- rnorm(p4_N2, 12, 3)
p4_s3 <- rnorm(p4_N3, 15, 4)

p4_n    <- 60
p4_W    <- c(p4_N1, p4_N2, p4_N3) / p4_N
p4_n_h  <- round(p4_n * p4_W)
p4_n_h
sum(p4_n_h)
```

## 4.3 Draw the stratified sample and estimate

```{r}
# --- Self-contained: stratified estimate ---
set.seed(402)
p4_N1 <- 100; p4_N2 <- 200; p4_N3 <- 300
p4_N  <- p4_N1 + p4_N2 + p4_N3
p4_s1 <- rnorm(p4_N1, 10, 2)
p4_s2 <- rnorm(p4_N2, 12, 3)
p4_s3 <- rnorm(p4_N3, 15, 4)

p4_n   <- 60
p4_W   <- c(p4_N1, p4_N2, p4_N3) / p4_N
p4_n_h <- round(p4_n * p4_W)

p4_samp1 <- sample(p4_s1, p4_n_h[1])
p4_samp2 <- sample(p4_s2, p4_n_h[2])
p4_samp3 <- sample(p4_s3, p4_n_h[3])

p4_ybar_h <- c(mean(p4_samp1), mean(p4_samp2), mean(p4_samp3))
p4_s2_h   <- c(var(p4_samp1),  var(p4_samp2),  var(p4_samp3))

p4_ybar_st <- sum(p4_W * p4_ybar_h)
p4_Var_st  <- sum(p4_W^2 * (1 - p4_n_h / c(p4_N1, p4_N2, p4_N3)) *
                  p4_s2_h / p4_n_h)
p4_SE_st   <- sqrt(p4_Var_st)

round(c(ybar_st = p4_ybar_st, var_st = p4_Var_st, se_st = p4_SE_st), 4)
```

## 4.4 Confidence interval

```{r}
# --- Self-contained: stratified CI ---
set.seed(403)
p4_N1 <- 100; p4_N2 <- 200; p4_N3 <- 300
p4_N  <- p4_N1 + p4_N2 + p4_N3
p4_s1 <- rnorm(p4_N1, 10, 2); p4_s2 <- rnorm(p4_N2, 12, 3); p4_s3 <- rnorm(p4_N3, 15, 4)
p4_n  <- 60
p4_W  <- c(p4_N1, p4_N2, p4_N3) / p4_N
p4_n_h <- round(p4_n * p4_W)

p4_samp1 <- sample(p4_s1, p4_n_h[1])
p4_samp2 <- sample(p4_s2, p4_n_h[2])
p4_samp3 <- sample(p4_s3, p4_n_h[3])
p4_ybar_h <- c(mean(p4_samp1), mean(p4_samp2), mean(p4_samp3))
p4_s2_h   <- c(var(p4_samp1),  var(p4_samp2),  var(p4_samp3))
p4_ybar_st <- sum(p4_W * p4_ybar_h)
p4_SE_st   <- sqrt(sum(p4_W^2 * (1 - p4_n_h / c(p4_N1, p4_N2, p4_N3)) *
                       p4_s2_h / p4_n_h))
p4_z <- qnorm(0.975)
round(c(p4_ybar_st - p4_z * p4_SE_st,
        p4_ybar_st + p4_z * p4_SE_st), 4)
```

## 4.5 Optimal (Neyman) allocation

**Minimizes variance for fixed $n$:**
$$n_h = n \cdot \frac{N_h S_h}{\sum_{j} N_j S_j}$$

**With different costs per stratum:**
$$n_h = n \cdot \frac{N_h S_h / \sqrt{c_h}}{\sum_{j} N_j S_j / \sqrt{c_j}}$$

```{r}
# --- Self-contained: Neyman allocation ---
set.seed(404)
p4_N1 <- 100; p4_N2 <- 200; p4_N3 <- 300
p4_S1 <- 2;   p4_S2 <- 3;   p4_S3 <- 4   # population SDs
p4_N_h <- c(p4_N1, p4_N2, p4_N3)
p4_S_h <- c(p4_S1, p4_S2, p4_S3)
p4_n   <- 60

p4_alpha_h <- p4_N_h * p4_S_h / sum(p4_N_h * p4_S_h)
p4_n_h_opt <- round(p4_n * p4_alpha_h)
p4_n_h_opt
sum(p4_n_h_opt)
```

## 4.6 Optimal with costs

```{r}
# --- Self-contained: Neyman with costs ---
p4_N_h <- c(100, 200, 300)
p4_S_h <- c(2, 3, 4)
p4_c_h <- c(1, 2, 4)     # cost per unit in each stratum
p4_n   <- 60

p4_alpha <- (p4_N_h * p4_S_h / sqrt(p4_c_h)) /
            sum(p4_N_h * p4_S_h / sqrt(p4_c_h))
p4_n_h_opt_cost <- round(p4_n * p4_alpha)
p4_n_h_opt_cost
sum(p4_n_h_opt_cost)
```

## 4.7 Optimal allocation vs proportional: total sample size for target variance

For a target variance $V$:
$$n = \frac{\left(\sum_h W_h S_h\right)^2}{V + \frac{1}{N}\sum_h W_h S_h^2} \quad \text{(proportional)}$$

$$n = \frac{\left(\sum_h W_h S_h\right)^2}{V + \frac{1}{N}\sum_h W_h S_h^2} \quad \text{(Neyman)}$$

The Neyman formula assumes equal costs and stratifies optimally.

## 4.8 Stratified sampling summary

| Allocation | $n_h$ | Best for |
|---|---|---|
| Proportional | $n \cdot N_h/N$ | Self-weighting, simple |
| Neyman | $n \cdot N_h S_h / \sum N_j S_j$ | Minimize variance, equal cost |
| Optimal cost | $n \cdot \frac{N_h S_h/\sqrt{c_h}}{\sum N_j S_j/\sqrt{c_j}}$ | Different costs |

**Key insight:** Stratification is never worse than SRS (when strata are correctly defined). It's better when between-stratum variance is large relative to within-stratum variance.

---

# Part 5 — Systematic Sampling

## 5.1 What it is

- Order the population in some natural sequence.
- Choose a random start and then select every $k$-th unit.
- $k = N/n$ is the **sampling interval**.

## 5.2 Draw a systematic sample

```{r}
# --- Self-contained: systematic sample ---
set.seed(501)
p5_N   <- 500
p5_pop <- rnorm(p5_N, 50, 10)
p5_n   <- 50
p5_k   <- p5_N / p5_n

p5_start <- sample(1:p5_k, 1)
p5_idx   <- seq(p5_start, by = p5_k, length.out = p5_n)
p5_sys   <- p5_pop[p5_idx]

p5_ybar <- mean(p5_sys)
p5_s2   <- var(p5_sys)

round(c(ybar = p5_ybar, s2 = p5_s2, start = p5_start, k = p5_k), 4)
```

## 5.3 Variance under systematic sampling

If the population is approximately random, use the SRS formula as an approximation:
$$\widehat{\text{Var}}(\bar{y}_{sys}) \approx \left(1 - \frac{n}{N}\right)\frac{s^2}{n}$$

The true variance depends on the correlation between units within a sampling interval.

**Two common approximations:**
1. **Treat as SRS:** conservative if the population has no periodic pattern.
2. **Successive-difference (variance among intervals):** accounts for ordering.

```{r}
# --- Self-contained: systematic variance (successive-difference) ---
set.seed(502)
p5_N   <- 500
p5_pop <- rnorm(p5_N, 50, 10)
p5_n   <- 50
p5_k   <- p5_N / p5_n
p5_start <- sample(1:p5_k, 1)
p5_idx <- seq(p5_start, by = p5_k, length.out = p5_n)
p5_sys <- p5_pop[p5_idx]

# Successive-difference estimator
p5_diffs <- diff(p5_sys)
p5_var_sd <- (1 - p5_n / p5_N) * sum(p5_diffs^2) / (2 * p5_n * (p5_n - 1))
p5_var_srs <- (1 - p5_n / p5_N) * var(p5_sys) / p5_n

round(c(var_sd = p5_var_sd, var_srs = p5_var_srs), 4)
```

## 5.4 When systematic beats SRS

- Population is **ordered** by a variable correlated with $y$.
- Systematic sampling behaves like implicit stratification.
- **Danger:** periodic patterns (e.g., every 7th unit) can bias the estimate.

## 5.5 Systematic sampling summary

| Property | Value |
|---|---|
| Selection | Every $k$-th unit after random start |
| $\pi_i$ | $1/k$ for most units |
| Estimator of mean | $\bar{y}_{sys}$ |
| Variance approximation | FPC $\cdot s^2/n$ |
| When to use | Ordered frame, easy to implement |
| Risk | Periodicity in the frame |

---

# Part 6 — Cluster Sampling

## 6.1 What it is

- Population is divided into **clusters** (groups of units).
- **Sample clusters** (not individual units).
- Measure all units in sampled clusters.

**Example:** Sampling schools (clusters) and testing all students in selected schools.

## 6.2 Cluster sampling vs SRS

**Advantages:**
- Cheaper (travel cost, setup cost).
- Requires only a list of clusters, not units.

**Disadvantage:**
- Higher variance if units within clusters are similar (intracluster correlation).

## 6.3 Build a clustered population

```{r}
# --- Self-contained: clustered population ---
set.seed(601)
p6_N_clusters  <- 50           # 50 clusters
p6_cluster_size <- 20          # 20 units each
p6_N_units      <- p6_N_clusters * p6_cluster_size

p6_cluster_effect <- rnorm(p6_N_clusters, 0, 3)
p6_pop <- rep(p6_cluster_effect, each = p6_cluster_size) +
          rnorm(p6_N_units, 50, 5)
p6_cluster_id <- rep(1:p6_N_clusters, each = p6_cluster_size)

p6_mu <- mean(p6_pop)
round(p6_mu, 4)
```

## 6.4 One-stage cluster sampling

Sample $m$ clusters out of $M$. Measure all units in each.

**Estimator of mean:**
$$\bar{y}_{cl} = \frac{\sum_{i \in S_{clusters}} \sum_{j=1}^{N_i} y_{ij}}{\sum_{i \in S_{clusters}} N_i}$$

**Estimator of total:**
$$\hat{\tau}_{cl} = \frac{M}{m} \sum_{i \in S} t_i, \quad t_i = \sum_{j} y_{ij}$$

```{r}
# --- Self-contained: cluster sample ---
set.seed(602)
p6_N_clusters <- 50
p6_cluster_size <- 20
p6_cluster_effect <- rnorm(p6_N_clusters, 0, 3)
p6_pop <- rep(p6_cluster_effect, each = p6_cluster_size) +
          rnorm(p6_N_clusters * p6_cluster_size, 50, 5)
p6_cluster_id <- rep(1:p6_N_clusters, each = p6_cluster_size)
p6_mu <- mean(p6_pop)

p6_m <- 10                          # sample 10 clusters
p6_sampled_clusters <- sample(1:p6_N_clusters, p6_m)

p6_totals <- sapply(p6_sampled_clusters, function(k) {
  sum(p6_pop[p6_cluster_id == k])
})
p6_sizes  <- sapply(p6_sampled_clusters, function(k) {
  sum(p6_cluster_id == k)
})

# Estimator of mean (ratio estimator)
p6_ybar_cl <- sum(p6_totals) / sum(p6_sizes)

# Total estimator
p6_tau_hat <- p6_N_clusters * mean(p6_totals)

round(c(ybar_cl = p6_ybar_cl, tau_hat = p6_tau_hat), 4)
```

## 6.5 Variance under cluster sampling

$$\widehat{\text{Var}}(\hat{\tau}) = M^2 \left(1 - \frac{m}{M}\right) \frac{s_t^2}{m}$$

where $s_t^2$ is the variance of the cluster totals.

```{r}
# --- Self-contained: cluster variance ---
set.seed(603)
p6_N_clusters <- 50
p6_cluster_size <- 20
p6_cluster_effect <- rnorm(p6_N_clusters, 0, 3)
p6_pop <- rep(p6_cluster_effect, each = p6_cluster_size) +
          rnorm(p6_N_clusters * p6_cluster_size, 50, 5)
p6_cluster_id <- rep(1:p6_N_clusters, each = p6_cluster_size)

p6_m <- 10
p6_sampled_clusters <- sample(1:p6_N_clusters, p6_m)
p6_totals <- sapply(p6_sampled_clusters, function(k) sum(p6_pop[p6_cluster_id == k]))

p6_tau_hat <- p6_N_clusters * mean(p6_totals)
p6_Var_tau <- p6_N_clusters^2 * (1 - p6_m / p6_N_clusters) *
              var(p6_totals) / p6_m

round(c(tau_hat = p6_tau_hat, se_tau = sqrt(p6_Var_tau)), 4)
```

## 6.6 Design effect for cluster sampling

$$\text{deff} = \frac{\text{Var}(\hat{\theta}_{cluster})}{\text{Var}(\hat{\theta}_{SRS})} = 1 + (N_i - 1)\rho$$

where $\rho$ = **intraclass correlation** (similarity of units within clusters).

- $\rho = 0$: deff = 1 (clusters don't hurt).
- $\rho > 0$: deff > 1 (clusters hurt).
- $\rho = 1$: deff = $N_i$ (clusters as bad as $n/N_i$ units).

```{r}
# --- Self-contained: deff for clusters ---
set.seed(604)
p6_N_clusters <- 50
p6_cluster_size <- 20
p6_cluster_effect <- rnorm(p6_N_clusters, 0, 3)
p6_pop <- rep(p6_cluster_effect, each = p6_cluster_size) +
          rnorm(p6_N_clusters * p6_cluster_size, 50, 5)
p6_cluster_id <- rep(1:p6_N_clusters, each = p6_cluster_size)

# Intraclass correlation via ANOVA
p6_df <- tibble(y = p6_pop, cluster = factor(p6_cluster_id))
p6_aov <- aov(y ~ cluster, data = p6_df)
p6_msb <- summary(p6_aov)[[1]]["cluster", "Mean Sq"]
p6_msw <- summary(p6_aov)[[1]]["Residuals", "Mean Sq"]
p6_rho <- (p6_msb - p6_msw) / (p6_msb + (p6_cluster_size - 1) * p6_msw)
p6_deff <- 1 + (p6_cluster_size - 1) * p6_rho

round(c(rho = p6_rho, deff = p6_deff), 4)
```

## 6.7 Cluster sampling summary

| Property | Value |
|---|---|
| Sample unit | Cluster |
| Estimator of total | $M \bar{t}$ |
| Variance | Depends on cluster total variance |
| Design effect | $1 + (N_i-1)\rho$ |
| Best for | Geographically dispersed populations |
| Risk | High intracluster correlation inflates variance |

---

# Part 7 — Two-Stage Sampling

## 7.1 What it is

- **Stage 1:** Sample $m$ primary sampling units (PSUs = clusters).
- **Stage 2:** Sample $n_i$ units (SSUs) from each selected PSU.

**Example:** Sample schools, then sample students within schools.

## 7.2 Two-stage sample

```{r}
# --- Self-contained: two-stage sample ---
set.seed(701)
p7_N_PSU  <- 40
p7_N_SSU  <- 25
p7_N      <- p7_N_PSU * p7_N_SSU

p7_PSU_effect <- rnorm(p7_N_PSU, 0, 3)
p7_pop <- rep(p7_PSU_effect, each = p7_N_SSU) +
          rnorm(p7_N, 50, 5)
p7_PSU_id <- rep(1:p7_N_PSU, each = p7_N_SSU)
p7_mu <- mean(p7_pop)

# Stage 1: sample m PSUs
p7_m <- 8
p7_sampled_PSU <- sample(1:p7_N_PSU, p7_m)

# Stage 2: sample n SSUs from each PSU
p7_n <- 10
p7_sample <- unlist(lapply(p7_sampled_PSU, function(k) {
  idx <- which(p7_PSU_id == k)
  sample(idx, p7_n, replace = FALSE)
}))

p7_ybar_2stage <- mean(p7_pop[p7_sample])
round(c(mu = p7_mu, ybar = p7_ybar_2stage), 4)
```

## 7.3 Variance

$$\widehat{\text{Var}}(\bar{y}_{2st}) = \left(1 - \frac{m}{M}\right)\frac{s_b^2}{m} + \frac{1}{m}\left(1 - \frac{n}{N_i}\right)\frac{s_w^2}{n}$$

where:
- $s_b^2$ = between-PSU variance of PSU means.
- $s_w^2$ = within-PSU variance.

```{r}
# --- Self-contained: two-stage variance ---
set.seed(702)
p7_N_PSU <- 40; p7_N_SSU <- 25
p7_PSU_effect <- rnorm(p7_N_PSU, 0, 3)
p7_pop <- rep(p7_PSU_effect, each = p7_N_SSU) + rnorm(p7_N_PSU * p7_N_SSU, 50, 5)
p7_PSU_id <- rep(1:p7_N_PSU, each = p7_N_SSU)

p7_m <- 8
p7_sampled_PSU <- sample(1:p7_N_PSU, p7_m)
p7_n <- 10
p7_sample <- unlist(lapply(p7_sampled_PSU, function(k) {
  sample(which(p7_PSU_id == k), p7_n)
}))

p7_PSU_means <- sapply(p7_sampled_PSU, function(k) {
  mean(p7_pop[p7_PSU_id == k])
})
p7_PSU_vars <- sapply(p7_sampled_PSU, function(k) {
  var(p7_pop[p7_PSU_id == k])
})

p7_sb2 <- var(p7_PSU_means)
p7_sw2 <- mean(p7_PSU_vars)

p7_Var_2st <- (1 - p7_m / p7_N_PSU) * p7_sb2 / p7_m +
              (1 / p7_m) * (1 - p7_n / p7_N_SSU) * p7_sw2 / p7_n
p7_SE_2st <- sqrt(p7_Var_2st)

round(c(ybar_2st = mean(p7_pop[p7_sample]),
        var_2st = p7_Var_2st,
        se_2st = p7_SE_2st), 4)
```

## 7.4 Two-stage summary

| Property | Value |
|---|---|
| Stage 1 | Sample $m$ PSUs |
| Stage 2 | Sample $n_i$ SSUs per PSU |
| Estimator | $\bar{y}$ over all sampled units |
| Variance | Combines between- and within-PSU variance |
| Use when | Multi-level populations |
| Special case | $n_i = N_i$ → cluster sampling |

---

# Part 8 — Ratio and Regression Estimators

## 8.1 Ratio estimator

When an auxiliary variable $x$ is correlated with $y$ and $\mu_x$ is known:

$$\hat{\mu}_{R} = \frac{\bar{y}}{\bar{x}} \mu_x$$

**Bias:** Approximate bias = $\frac{1}{n}(1 - n/N)(\bar{R} C_x^2 - \rho C_x C_y)$ where $\bar{R} = \mu_y/\mu_x$.

**Variance approximation:**
$$\widehat{\text{Var}}(\hat{\mu}_{R}) \approx \left(1 - \frac{n}{N}\right)\frac{s_d^2}{n}$$
where $d_i = y_i - \bar{R} x_i$ and $\bar{R} = \bar{y}/\bar{x}$.

```{r}
# --- Self-contained: ratio estimator ---
set.seed(801)
p8_N <- 500
p8_x <- rnorm(p8_N, 20, 4)
p8_y <- 3 * p8_x + rnorm(p8_N, 0, 3)
p8_mu_x <- mean(p8_x)
p8_mu_y <- mean(p8_y)

p8_n <- 50
p8_srs <- sample(1:p8_N, p8_n)
p8_x_s <- p8_x[p8_srs]
p8_y_s <- p8_y[p8_srs]

p8_R_hat   <- mean(p8_y_s) / mean(p8_x_s)
p8_mu_R    <- p8_R_hat * p8_mu_x
p8_d       <- p8_y_s - p8_R_hat * p8_x_s
p8_Var_R   <- (1 - p8_n / p8_N) * var(p8_d) / p8_n

round(c(mu_y = p8_mu_y, mu_R = p8_mu_R, se_R = sqrt(p8_Var_R)), 4)
```

**When to use:** $x$ is strongly correlated with $y$, and $\mu_x$ is known.

## 8.2 Regression estimator

$$\hat{\mu}_{lr} = \bar{y} + b(\mu_x - \bar{x}), \quad b = \frac{s_{xy}}{s_x^2}$$

**Variance:**
$$\widehat{\text{Var}}(\hat{\mu}_{lr}) \approx \left(1 - \frac{n}{N}\right)\frac{s_r^2}{n}$$

where $s_r^2$ = residual variance from regressing $y$ on $x$.

```{r}
# --- Self-contained: regression estimator ---
set.seed(802)
p8_N <- 500
p8_x <- rnorm(p8_N, 20, 4)
p8_y <- 3 * p8_x + rnorm(p8_N, 0, 3)
p8_mu_x <- mean(p8_x)
p8_mu_y <- mean(p8_y)

p8_n <- 50
p8_srs <- sample(1:p8_N, p8_n)
p8_x_s <- p8_x[p8_srs]; p8_y_s <- p8_y[p8_srs]

p8_b   <- cov(p8_x_s, p8_y_s) / var(p8_x_s)
p8_mu_lr <- mean(p8_y_s) + p8_b * (p8_mu_x - mean(p8_x_s))

p8_resid <- p8_y_s - (mean(p8_y_s) + p8_b * (p8_x_s - mean(p8_x_s)))
p8_Var_lr <- (1 - p8_n / p8_N) * var(p8_resid) * (p8_n - 1) /
             (p8_n - 2) / p8_n

round(c(mu_y = p8_mu_y, mu_lr = p8_mu_lr, se_lr = sqrt(p8_Var_lr)), 4)
```

**Ratio vs regression:** Ratio assumes the line passes through the origin. Regression doesn't.

## 8.3 Reporting

> "With auxiliary variable x known (mean $\mu_x = 20.1$), the ratio estimator gives $\hat{\mu}_R = 60.2$ with SE 0.85, compared to the SRS estimator $\bar{y} = 60.5$ with SE 1.12. The ratio estimator is more efficient because it uses the correlation between x and y."

---

# Part 9 — Bias, MSE, and Relative Efficiency

## 9.1 Simulate bias, variance, MSE

```{r}
# --- Self-contained: bias/variance/MSE simulation ---
set.seed(901)
p9_N   <- 1000
p9_pop <- rnorm(p9_N, 50, 10)
p9_mu  <- mean(p9_pop)

p9_n  <- 30
p9_B  <- 3000

p9_means   <- numeric(p9_B)
p9_medians <- numeric(p9_B)

for (b in 1:p9_B) {
  s <- sample(p9_pop, p9_n)
  p9_means[b]   <- mean(s)
  p9_medians[b] <- median(s)
}

p9_tab <- tibble(
  Estimator = c("Mean", "Median"),
  Bias      = c(mean(p9_means)   - p9_mu,
                mean(p9_medians) - p9_mu),
  Variance  = c(var(p9_means), var(p9_medians)),
  MSE       = c(mean((p9_means   - p9_mu)^2),
                mean((p9_medians - p9_mu)^2))
)
kable(p9_tab, digits = 4)
```

**Interpretation:**
- Mean is unbiased (bias ≈ 0).
- Median may be biased but often has smaller variance if the population is skewed.
- Compare MSE to choose.

## 9.2 Relative efficiency

```{r}
# --- Self-contained: relative efficiency ---
set.seed(902)
p9_N <- 1000
p9_pop <- rnorm(p9_N, 50, 10)
p9_mu <- mean(p9_pop)
p9_n <- 30; p9_B <- 3000

p9_means   <- replicate(p9_B, mean(sample(p9_pop, p9_n)))
p9_medians <- replicate(p9_B, median(sample(p9_pop, p9_n)))

p9_RE_median_vs_mean <- var(p9_medians) / var(p9_means)
p9_RE_mean_vs_median <- var(p9_means) / var(p9_medians)

round(c(RE_median_vs_mean = p9_RE_median_vs_mean,
        RE_mean_vs_median = p9_RE_mean_vs_median), 4)
```

**Interpretation:** RE > 1 → numerator is more efficient (lower variance).

## 9.3 Bias of the ratio estimator (empirical)

```{r}
# --- Self-contained: ratio estimator bias ---
set.seed(903)
p9_N <- 500
p9_x <- rnorm(p9_N, 20, 4)
p9_y <- 3 * p9_x + rnorm(p9_N, 0, 3)
p9_mu_x <- mean(p9_x)
p9_mu_y <- mean(p9_y)

p9_n <- 30; p9_B <- 3000
p9_R_est <- replicate(p9_B, {
  idx <- sample(1:p9_N, p9_n)
  mean(p9_y[idx]) / mean(p9_x[idx]) * p9_mu_x
})

round(c(true = p9_mu_y,
        mean_R = mean(p9_R_est),
        bias = mean(p9_R_est) - p9_mu_y,
        var_R = var(p9_R_est),
        mse_R = mean((p9_R_est - p9_mu_y)^2)), 4)
```

## 9.4 Comparing estimators summary

| Criterion | Formula | Decision |
|---|---|---|
| Bias | $E[\hat{\theta}] - \theta$ | Closer to 0 is better |
| Variance | $E[(\hat{\theta} - E\hat{\theta})^2]$ | Smaller is better |
| MSE | Bias² + Variance | Smaller is better |
| RE of $\hat{\theta}_1$ vs $\hat{\theta}_2$ | MSE(2)/MSE(1) | > 1 → $\hat{\theta}_1$ better |

---

# Part 10 — Coverage Probability Simulations

## 10.1 Coverage of a 95% CI for $\mu$ (SRS)

```{r}
# --- Self-contained: CI coverage ---
set.seed(1001)
p10_N   <- 500
p10_pop <- rnorm(p10_N, 50, 10)
p10_mu  <- mean(p10_pop)

p10_n   <- 30
p10_B   <- 2000
p10_z   <- qnorm(0.975)

p10_cov <- logical(p10_B)
p10_len <- numeric(p10_B)

for (b in 1:p10_B) {
  s <- sample(p10_pop, p10_n)
  ybar <- mean(s); s2 <- var(s)
  se <- sqrt((1 - p10_n / p10_N) * s2 / p10_n)
  ci <- c(ybar - p10_z * se, ybar + p10_z * se)
  p10_cov[b] <- (p10_mu >= ci[1]) & (p10_mu <= ci[2])
  p10_len[b] <- ci[2] - ci[1]
}

round(c(coverage = mean(p10_cov), mean_length = mean(p10_len)), 4)
```

**Interpretation:** With FPC, coverage should be near 0.95. This validates the CI formula.

## 10.2 Coverage with FPC vs without

```{r}
# --- Self-contained: coverage with and without FPC ---
set.seed(1002)
p10_N   <- 500
p10_pop <- rnorm(p10_N, 50, 10)
p10_mu  <- mean(p10_pop)
p10_n   <- 50     # 10% sampling fraction
p10_B   <- 2000
p10_z   <- qnorm(0.975)

p10_cov_fpc <- logical(p10_B)
p10_cov_nofpc <- logical(p10_B)
p10_len_fpc <- numeric(p10_B)
p10_len_nofpc <- numeric(p10_B)

for (b in 1:p10_B) {
  s <- sample(p10_pop, p10_n)
  ybar <- mean(s); s2 <- var(s)
  se_fpc <- sqrt((1 - p10_n / p10_N) * s2 / p10_n)
  se_nf  <- sqrt(s2 / p10_n)

  ci_f <- c(ybar - p10_z * se_fpc, ybar + p10_z * se_fpc)
  ci_n <- c(ybar - p10_z * se_nf,  ybar + p10_z * se_nf)

  p10_cov_fpc[b]   <- (p10_mu >= ci_f[1]) & (p10_mu <= ci_f[2])
  p10_cov_nofpc[b] <- (p10_mu >= ci_n[1]) & (p10_mu <= ci_n[2])
  p10_len_fpc[b]   <- ci_f[2] - ci_f[1]
  p10_len_nofpc[b] <- ci_n[2] - ci_n[1]
}

tibble(
  CI          = c("With FPC", "Without FPC"),
  Coverage    = round(c(mean(p10_cov_fpc), mean(p10_cov_nofpc)), 4),
  MeanLength  = round(c(mean(p10_len_fpc), mean(p10_len_nofpc)), 4)
) %>% kable()
```

**Interpretation:** The FPC CI is shorter but should have similar coverage. If the sampling fraction is small, both are close.

## 10.3 Coverage under misspecified distribution

```{r}
# --- Self-contained: coverage when population is skewed ---
set.seed(1003)
p10_N   <- 500
p10_pop <- rexp(p10_N, 1)   # skewed
p10_mu  <- mean(p10_pop)

p10_n   <- 30
p10_B   <- 2000
p10_z   <- qnorm(0.975)

p10_cov <- numeric(p10_B)
for (b in 1:p10_B) {
  s <- sample(p10_pop, p10_n)
  se <- sqrt((1 - p10_n / p10_N) * var(s) / p10_n)
  p10_cov[b] <- (p10_mu >= mean(s) - p10_z * se) &
                (p10_mu <= mean(s) + p10_z * se)
}
mean(p10_cov)
```

**Interpretation:** For skewed populations, coverage can be lower than 0.95 for small $n$ — the CLT hasn't kicked in yet.

---

# Part 11 — Sample Size Determination

## 11.1 For a target variance of the mean

$$n = \frac{S^2}{V + S^2/N}$$
where $V$ = target variance of $\bar{y}$.

```{r}
# --- Self-contained: sample size for target variance ---
p11_N  <- 1000
p11_S2 <- 100
p11_V  <- 1     # target variance of ybar

p11_n <- p11_S2 / (p11_V + p11_S2 / p11_N)
ceiling(p11_n)
```

## 11.2 For a target margin of error (with FPC)

$$n_0 = \frac{z^2 S^2}{e^2}, \quad n = \frac{n_0}{1 + n_0/N}$$

```{r}
# --- Self-contained: sample size for margin of error ---
p11_N  <- 1000
p11_S2 <- 100
p11_e  <- 2
p11_z  <- qnorm(0.975)

p11_n0 <- p11_z^2 * p11_S2 / p11_e^2
p11_n  <- ceiling(p11_n0 / (1 + p11_n0 / p11_N))
p11_n
```

## 11.3 For a proportion

$$n_0 = \frac{z^2 p(1-p)}{e^2}, \quad n = \frac{n_0}{1 + n_0/N}$$

Use $p = 0.5$ if unknown (worst case).

```{r}
# --- Self-contained: proportion sample size ---
p11_N  <- 2000
p11_p  <- 0.5    # conservative
p11_e  <- 0.05
p11_z  <- qnorm(0.975)

p11_n0 <- p11_z^2 * p11_p * (1 - p11_p) / p11_e^2
p11_n  <- ceiling(p11_n0 / (1 + p11_n0 / p11_N))
p11_n
```

## 11.4 Stratified sample size for target variance (proportional)

$$n = \frac{\sum_h W_h^2 S_h^2 / w_h}{V + \frac{1}{N}\sum_h W_h S_h^2}$$
where $w_h = n_h/n$ is the fraction allocated to stratum $h$.

For proportional allocation ($w_h = W_h$):
$$n = \frac{\sum_h W_h S_h^2}{V + \frac{1}{N}\sum_h W_h S_h^2}$$

```{r}
# --- Self-contained: stratified sample size ---
p11_N_h <- c(200, 400, 400)
p11_S_h <- c(5, 8, 12)
p11_N   <- sum(p11_N_h)
p11_W_h <- p11_N_h / p11_N
p11_V   <- 1

p11_numerator   <- sum(p11_W_h * p11_S_h^2)
p11_denominator <- p11_V + (1 / p11_N) * sum(p11_W_h * p11_S_h^2)
p11_n <- ceiling(p11_numerator / p11_denominator)
p11_n
```

## 11.5 Sample size for Neyman allocation

$$n = \frac{\left(\sum_h W_h S_h\right)^2}{V + \frac{1}{N}\sum_h W_h S_h^2}$$

```{r}
# --- Self-contained: Neyman sample size ---
p11_N_h <- c(200, 400, 400)
p11_S_h <- c(5, 8, 12)
p11_N   <- sum(p11_N_h)
p11_W_h <- p11_N_h / p11_N
p11_V   <- 1

p11_num <- (sum(p11_W_h * p11_S_h))^2
p11_den <- p11_V + (1 / p11_N) * sum(p11_W_h * p11_S_h^2)
p11_n <- ceiling(p11_num / p11_den)
p11_n
```

## 11.6 Sample size summary

| Goal | Formula |
|---|---|
| Target variance $V$ | $n = S^2/(V + S^2/N)$ |
| Target margin $e$ | $n = z^2 S^2/e^2$, adjust with FPC |
| Proportion | $n = z^2 p(1-p)/e^2$, adjust |
| Stratified proportional | $n = \frac{\sum W_h S_h^2}{V + \frac{1}{N}\sum W_h S_h^2}$ |
| Stratified Neyman | $n = \frac{(\sum W_h S_h)^2}{V + \frac{1}{N}\sum W_h S_h^2}$ |

---

# Part 12 — Design Effects

## 12.1 Definition

$$\text{deff} = \frac{\text{Var}(\hat{\theta}_{\text{design}})}{\text{Var}(\hat{\theta}_{SRS})}$$

- deff = 1: same efficiency as SRS.
- deff > 1: less efficient than SRS.
- deff < 1: more efficient than SRS (stratification does this).

## 12.2 Effective sample size

$$n_{\text{eff}} = \frac{n}{\text{deff}}$$

If you have a cluster sample with deff = 3, your 300 observations give the same precision as 100 SRS observations.

## 12.3 deff for stratified sampling

$$\text{deff} = \frac{\sum_h W_h^2 S_h^2 / n_h}{S^2/n}$$

With proportional allocation, deff ≤ 1 always.

```{r}
# --- Self-contained: deff for stratification ---
set.seed(1201)
p12_N_h <- c(200, 400, 400)
p12_S_h <- c(5, 8, 12)
p12_N   <- sum(p12_N_h)
p12_W_h <- p12_N_h / p12_N
p12_S2  <- sum(p12_W_h * p12_S_h^2) +
           sum(p12_W_h * (c(10, 15, 20) - sum(p12_W_h * c(10, 15, 20)))^2)
p12_n   <- 60
p12_n_h <- round(p12_n * p12_W_h)

p12_var_st <- sum(p12_W_h^2 * p12_S_h^2 / p12_n_h)
p12_var_srs <- p12_S2 / p12_n

p12_deff <- p12_var_st / p12_var_srs
round(p12_deff, 4)
```

## 12.4 deff for cluster sampling

$$\text{deff} = 1 + (N_i - 1) \rho$$

Already covered in Part 6.6.

## 12.5 Reporting

> "The stratified design has a design effect of 0.62, so it is 1.6 times more efficient than SRS. The cluster design has deff = 3.1 due to high intraclass correlation (ρ = 0.22), meaning 300 clustered units give the same precision as 100 SRS units."

---

# Part 13 — Horvitz-Thompson Estimator

## 13.1 Definition

For any sampling design with inclusion probabilities $\pi_i$:

$$\hat{\tau}_{HT} = \sum_{i \in S} \frac{y_i}{\pi_i}$$

This is **unbiased** for the population total $\tau$.

**For the mean:**
$$\hat{\mu}_{HT} = \frac{\hat{\tau}_{HT}}{N}$$

## 13.2 Variance of HT estimator

$$\text{Var}(\hat{\tau}_{HT}) = \sum_{i=1}^N \sum_{j=1}^N (\pi_{ij} - \pi_i \pi_j) \frac{y_i}{\pi_i} \frac{y_j}{\pi_j}$$

where $\pi_{ij}$ = joint inclusion probability of units $i$ and $j$.

**Unbiased variance estimator (Yates-Grundy):**
$$\widehat{\text{Var}}(\hat{\tau}_{HT}) = \sum_{i < j \in S} \frac{\pi_i \pi_j - \pi_{ij}}{\pi_{ij}} \left(\frac{y_i}{\pi_i} - \frac{y_j}{\pi_j}\right)^2$$

## 13.3 For SRS

$\pi_i = n/N$, $\pi_{ij} = n(n-1)/(N(N-1))$.

$$\hat{\tau}_{HT} = \frac{N}{n}\sum_{i \in S} y_i = N \bar{y}$$

Same as the usual estimator.

```{r}
# --- Self-contained: HT for SRS ---
set.seed(1301)
p13_N <- 500
p13_pop <- rnorm(p13_N, 50, 10)
p13_n <- 50

p13_srs <- sample(1:p13_N, p13_n)
p13_pi  <- p13_n / p13_N
p13_tau_hat <- sum(p13_pop[p13_srs] / p13_pi)

round(c(tau_hat = p13_tau_hat, N_ybar = p13_N * mean(p13_pop[p13_srs])), 4)
```

## 13.4 Unequal probability sampling

When units have different $\pi_i$ (e.g., PPS), the HT estimator gives an unbiased estimate even though units are not equally weighted.

```{r}
# --- Self-contained: HT with unequal probabilities ---
set.seed(1302)
p13_N <- 10
p13_y <- c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100)
p13_size <- p13_y   # size proportional to y
p13_n <- 3

# Sample with probabilities proportional to size
p13_srs <- sample(1:p13_N, p13_n,
                  prob = p13_size / sum(p13_size),
                  replace = FALSE)

# Inclusion probabilities (approximate for PPS without replacement)
p13_pi <- p13_n * p13_size / sum(p13_size)
p13_tau_hat <- sum(p13_y[p13_srs] / p13_pi[p13_srs])

round(c(true_tau = sum(p13_y), HT = p13_tau_hat), 4)
```

## 13.5 HT summary

| Property | Value |
|---|---|
| Estimator | $\hat{\tau} = \sum_{i \in S} y_i/\pi_i$ |
| Bias | Unbiased for any design |
| Variance | Depends on joint inclusion probabilities |
| When to use | Design-based inference for complex designs |
| Special case | SRS, stratified: reduces to standard estimators |

---

# Part 14 — Probability Proportional to Size (PPS)

## 14.1 What it is

Sample units with probability proportional to a **size measure** $x_i$. Larger units are more likely to be selected.

**When to use:** Businesses (size = employees), schools (size = enrollment), hospitals (size = beds).

## 14.2 PPS with replacement

Sample $n$ units with probability $\propto x_i$. Since sizes differ, HT estimator simplifies.

**Hansen-Hurwitz estimator of total:**
$$\hat{\tau}_{HH} = \frac{1}{n} \sum_{i=1}^n \frac{y_i}{p_i}, \quad p_i = \frac{x_i}{\sum_{j} x_j}$$

```{r}
# --- Self-contained: PPS with replacement ---
set.seed(1401)
p14_N <- 20
p14_x <- runif(p14_N, 10, 100)   # size
p14_y <- 5 * p14_x + rnorm(p14_N, 0, 20)
p14_tau <- sum(p14_y)

p14_n <- 5
p14_idx <- sample(1:p14_N, p14_n,
                  prob = p14_x / sum(p14_x),
                  replace = TRUE)
p14_p <- p14_x / sum(p14_x)
p14_tau_hat <- mean(p14_y[p14_idx] / p14_p[p14_idx])

round(c(true_tau = p14_tau, HH = p14_tau_hat), 4)
```

## 14.3 Variance of HH estimator

$$\widehat{\text{Var}}(\hat{\tau}_{HH}) = \frac{1}{n(n-1)} \sum_{i=1}^n \left(\frac{y_i}{p_i} - \hat{\tau}_{HH}\right)^2$$

```{r}
# --- Self-contained: PPS variance ---
set.seed(1402)
p14_N <- 20
p14_x <- runif(p14_N, 10, 100)
p14_y <- 5 * p14_x + rnorm(p14_N, 0, 20)
p14_n <- 5
p14_idx <- sample(1:p14_N, p14_n, prob = p14_x / sum(p14_x), replace = TRUE)
p14_p <- p14_x / sum(p14_x)

p14_z <- p14_y[p14_idx] / p14_p[p14_idx]
p14_tau_hat <- mean(p14_z)
p14_Var_HH  <- var(p14_z) / p14_n
p14_SE_HH   <- sqrt(p14_Var_HH)

round(c(HH = p14_tau_hat, se = p14_SE_HH), 4)
```

## 14.4 PPS summary

| Property | Value |
|---|---|
| Selection | Probability $\propto x_i$ |
| Estimator of total | $\frac{1}{n}\sum y_i/p_i$ |
| Variance | $\text{Var}(\bar{z})/n$ where $z_i = y_i/p_i$ |
| Best for | Populations with wide size variation |
| Requires | Known size measure for all units |

---

# Part 15 — Non-Response

## 15.1 Types of non-response

- **Unit non-response:** The whole unit doesn't respond.
- **Item non-response:** The unit responds but skips some questions.

## 15.2 Effects on estimates

If non-responders differ systematically from responders, the estimate is **biased**:

$$\text{Bias} = \frac{\text{NR}}{N}(\bar{y}_{NR} - \bar{y}_{R})$$

where NR = non-responders, R = responders.

```{r}
# --- Self-contained: non-response bias ---
set.seed(1501)
p15_N  <- 1000
p15_y  <- c(rnorm(700, 50, 10),        # responders-to-be
            rnorm(300, 60, 10))        # non-responders-to-be
p15_truth <- mean(p15_y)

# Suppose 40% respond
p15_n_resp <- 400
p15_sample_resp <- sample(1:700, p15_n_resp)   # only from responders

p15_est <- mean(p15_y[p15_sample_resp])
round(c(truth = p15_truth, estimate = p15_est,
        bias = p15_est - p15_truth), 4)
```

## 15.3 Weighting adjustment

Give responders weights inversely proportional to their response probability.

**Weight = 1 / (response rate)** for each adjustment cell.

```{r}
# --- Self-contained: weighted estimate ---
set.seed(1502)
p15_N <- 1000
p15_group <- sample(c("A", "B"), p15_N, replace = TRUE)
p15_y <- ifelse(p15_group == "A", rnorm(p15_N, 50, 10), rnorm(p15_N, 60, 10))

# Response rates differ by group
p15_resp_prob <- ifelse(p15_group == "A", 0.8, 0.4)
p15_responded <- runif(p15_N) < p15_resp_prob
p15_sample <- which(p15_responded)

# Unweighted (biased)
p15_unw <- mean(p15_y[p15_sample])

# Weighted by 1/response-prob
p15_w <- 1 / p15_resp_prob[p15_sample]
p15_wtd <- sum(p15_y[p15_sample] * p15_w) / sum(p15_w)

round(c(truth = mean(p15_y),
        unweighted = p15_unw,
        weighted = p15_wtd), 4)
```

## 15.4 Summary

| Type | Effect | Adjustment |
|---|---|---|
| MCAR | Small bias | Ignore or drop |
| MAR | Moderate bias | Weighting, imputation |
| MNAR | Large bias | Model non-response, sensitivity analysis |

**Always report:** response rate, potential bias.

---

# Part 16 — Bootstrap Variance Estimation

## 16.1 Why bootstrap?

For complex designs (cluster, two-stage, PPS), the theoretical variance may be hard to derive. Bootstrap gives an empirical estimate by resampling.

## 16.2 SRS bootstrap

**Resample with replacement** from the sample, compute the statistic, repeat $B$ times. The variance of the bootstrap replicates approximates the sampling variance.

```{r}
# --- Self-contained: bootstrap variance for SRS ---
set.seed(1601)
p16_N <- 500
p16_pop <- rnorm(p16_N, 50, 10)
p16_n <- 50
p16_srs <- sample(p16_pop, p16_n)

p16_B <- 2000
p16_boot_means <- replicate(p16_B, {
  s <- sample(p16_srs, p16_n, replace = TRUE)
  mean(s)
})

p16_boot_var <- var(p16_boot_means)
p16_theory_var <- (1 - p16_n / p16_N) * var(p16_srs) / p16_n

round(c(theory = p16_theory_var, bootstrap = p16_boot_var), 4)
```

**Note:** Standard bootstrap doesn't account for FPC. For design-based bootstrap, use a scaled version or a resampling scheme that respects the design.

## 16.3 Bootstrap CI

**Percentile CI:** the 2.5% and 97.5% quantiles of the bootstrap distribution.

```{r}
# --- Self-contained: bootstrap CI ---
set.seed(1602)
p16_pop <- rnorm(500, 50, 10)
p16_n <- 50
p16_srs <- sample(p16_pop, p16_n)

p16_B <- 2000
p16_boot_means <- replicate(p16_B, mean(sample(p16_srs, p16_n, replace = TRUE)))

p16_ci_pct <- quantile(p16_boot_means, c(0.025, 0.975))

# Basic bootstrap CI
p16_ci_basic <- 2 * mean(p16_srs) - rev(p16_ci_pct)

round(c(percentile_lo = p16_ci_pct[1],
        percentile_hi = p16_ci_pct[2],
        basic_lo = p16_ci_basic[1],
        basic_hi = p16_ci_basic[2]), 4)
```

## 16.4 Cluster bootstrap

For cluster sampling, **resample clusters** (not units):

```{r}
# --- Self-contained: cluster bootstrap ---
set.seed(1603)
p16_N_clusters <- 50
p16_cluster_size <- 20
p16_cluster_effect <- rnorm(p16_N_clusters, 0, 3)
p16_pop <- rep(p16_cluster_effect, each = p16_cluster_size) +
          rnorm(p16_N_clusters * p16_cluster_size, 50, 5)
p16_cluster_id <- rep(1:p16_N_clusters, each = p16_cluster_size)

p16_m <- 10
p16_sampled_clusters <- sample(1:p16_N_clusters, p16_m)

# Bootstrap: resample clusters
p16_B <- 2000
p16_boot_totals <- replicate(p16_B, {
  boot_clusters <- sample(p16_sampled_clusters, replace = TRUE)
  sum(sapply(boot_clusters, function(k) sum(p16_pop[p16_cluster_id == k])))
})

# Estimator: total
p16_totals <- sapply(p16_sampled_clusters, function(k) sum(p16_pop[p16_cluster_id == k]))
p16_tau_hat <- p16_N_clusters * mean(p16_totals)
p16_boot_se <- sd(p16_boot_totals) / sqrt(p16_m)

round(c(tau_hat = p16_tau_hat, se_theory = sqrt(p16_N_clusters^2 *
                                                (1 - p16_m / p16_N_clusters) *
                                                var(p16_totals) / p16_m),
        se_boot = p16_boot_se), 4)
```

## 16.5 Bootstrap summary

| Design | Resample |
|---|---|
| SRS | Units with replacement |
| Stratified | Within each stratum |
| Cluster | Whole clusters with replacement |
| Two-stage | PSUs, then SSUs within |

---

# Part 17 — Theory Q&A

**Q: What is the difference between a parameter and a statistic?**
A: A parameter is a fixed (unknown) population value. A statistic is a random variable computed from a sample that estimates the parameter.

**Q: What is the finite population correction and why does it matter?**
A: FPC = $1 - n/N$. It reduces the variance of the sample mean when the sample is a large fraction of the population. Ignoring it overestimates the variance.

**Q: What is the difference between sampling with and without replacement?**
A: With replacement, units can repeat; without replacement, each unit appears at most once. SRS usually means without replacement.

**Q: What is the bias of the sample mean under SRS?**
A: Zero. Under SRS, $E[\bar{y}] = \mu$.

**Q: What is the variance of the sample mean under SRS?**
A: $(1 - n/N) S^2 / n$.

**Q: What is stratified sampling and why use it?**
A: Divide the population into strata, sample independently within each. Use it to reduce variance by ensuring all subgroups are represented, and to control the sample size per stratum.

**Q: What is proportional allocation?**
A: Sample size per stratum is proportional to stratum size: $n_h = n N_h / N$.

**Q: What is Neyman allocation?**
A: $n_h = n N_h S_h / \sum_j N_j S_j$. Minimizes variance for fixed $n$ with equal costs.

**Q: What is optimal allocation with costs?**
A: $n_h = n \frac{N_h S_h/\sqrt{c_h}}{\sum_j N_j S_j/\sqrt{c_j}}$. Accounts for different per-unit costs across strata.

**Q: When does stratification beat SRS?**
A: Always (weakly). Strictly when between-stratum variance is large relative to within-stratum variance.

**Q: What is systematic sampling and when is it dangerous?**
A: Select every $k$-th unit after a random start. Dangerous when the frame has periodic patterns that align with $k$.

**Q: What is cluster sampling?**
A: Sample whole groups (clusters) and measure all units within each sampled cluster. Used when travel/setup cost dominates.

**Q: What is the design effect?**
A: $\text{deff} = \text{Var}(\hat{\theta}_{design}) / \text{Var}(\hat{\theta}_{SRS})$. Values > 1 mean the design is less efficient than SRS.

**Q: What is the intraclass correlation?**
A: The correlation between units within the same cluster. High $\rho$ means clusters are homogeneous, so cluster sampling is inefficient.

**Q: What is two-stage sampling?**
A: Sample PSUs (clusters) at stage 1, then SSUs within each selected PSU at stage 2. More flexible than one-stage cluster sampling.

**Q: What is the Horvitz-Thompson estimator?**
A: $\hat{\tau}_{HT} = \sum_{i \in S} y_i/\pi_i$. Unbiased for the population total for any design, using inclusion probabilities.

**Q: What is PPS sampling?**
A: Sample units with probability proportional to a size measure $x_i$. Efficient when sizes vary widely.

**Q: What is the ratio estimator?**
A: $\hat{\mu}_R = (\bar{y}/\bar{x}) \mu_x$. Uses a known auxiliary mean $\mu_x$ to improve precision.

**Q: What is the regression estimator?**
A: $\hat{\mu}_{lr} = \bar{y} + b(\mu_x - \bar{x})$. Uses the linear regression of $y$ on $x$ to adjust the estimate.

**Q: What is the difference between ratio and regression estimators?**
A: Ratio assumes the line passes through the origin; regression doesn't. Regression is more general but requires more computation.

**Q: When is the median a better estimator than the mean?**
A: When the population is skewed or has outliers. The median is robust to extreme values.

**Q: What is the coverage probability?**
A: The proportion of confidence intervals that contain the true parameter over repeated sampling. A 95% CI should have coverage ≈ 0.95.

**Q: What is the difference between design-based and model-based inference?**
A: Design-based: the population is fixed, randomness comes from sampling. Model-based: the population is random, drawn from a superpopulation.

**Q: What is the effective sample size?**
A: $n_{\text{eff}} = n/\text{deff}$. The size of an SRS sample that would give the same precision.

**Q: How do you handle non-response?**
A: Report the response rate, check if non-response is related to the outcome (MCAR/MAR/MNAR), use weighting or imputation.

**Q: Why does clustering inflate variance?**
A: Because units within a cluster tend to be similar, they provide less independent information than $N_i$ separate SRS units.

---

# Part 18 — Exam Scenarios

## 18.1 "Estimate the population mean under SRS"

1. Sample $n$ units.
2. Compute $\bar{y} = \text{mean}(s)$.
3. Compute $\hat{V} = (1 - n/N) s^2/n$.
4. Build CI: $\bar{y} \pm z\sqrt{\hat{V}}$.
5. Report estimate + SE + CI.

## 18.2 "Compare SRS and stratified variance"

1. Compute theoretical variance of SRS.
2. Compute theoretical variance of stratified (with the allocation).
3. Compare ratio: $\text{Var}_{st}/\text{Var}_{SRS}$.
4. Report which design is more efficient.

## 18.3 "Optimal allocation"

1. List $N_h$ and $S_h$ for each stratum.
2. Compute $n_h = n N_h S_h / \sum N_j S_j$.
3. Round to integers, ensure total = $n$.
4. Compute variance with the resulting allocation.

## 18.4 "Bias and MSE of an estimator"

1. Simulate $B$ samples.
2. Compute the estimator for each.
3. Bias = mean(estimates) − truth.
4. Variance = var(estimates).
5. MSE = mean((estimates − truth)^2).

## 18.5 "Coverage of a CI"

1. Simulate $B$ samples.
2. For each, compute the CI.
3. Count how many CIs contain the true parameter.
4. Report coverage.

## 18.6 "Sample size for target precision"

1. Identify target (variance, margin of error, or coverage).
2. Get $S^2$ from pilot or guess.
3. Compute $n_0 = z^2 S^2/e^2$.
4. Adjust with FPC: $n = n_0/(1 + n_0/N)$.
5. Round up.

## 18.7 "Design effect"

1. Compute variance under your design.
2. Compute variance under SRS with same $n$.
3. $\text{deff} = V_{design}/V_{SRS}$.
4. $n_{\text{eff}} = n/\text{deff}$.

## 18.8 "Cluster sampling"

1. Compute cluster totals.
2. HT or HH estimator.
3. Variance from cluster totals.
4. Design effect from intraclass correlation.

## 18.9 "Ratio/regression estimator"

1. Compute $\bar{y}$ and $\bar{x}$ from the sample.
2. Use known $\mu_x$ to compute $\hat{\mu}_R$ or $\hat{\mu}_{lr}$.
3. Compute residuals, then variance.
4. Compare to SRS.

## 18.10 "Non-response adjustment"

1. Identify response rates by group.
2. Compute weights = 1/response probability.
3. Compute weighted estimate.
4. Report difference from unweighted.

---

# Part 19 — Quick Reference

| Task | Command |
|---|---|
| SRS sample | `sample(pop, n, replace = FALSE)` |
| Sample with replacement | `sample(pop, n, replace = TRUE)` |
| Weighted sample | `sample(pop, n, prob = w, replace = TRUE)` |
| Sample mean | `mean(s)` |
| Sample variance | `var(s)` |
| FPC | `(1 - n/N)` |
| Var of sample mean | `(1 - n/N) * var(s) / n` |
| Total estimate | `N * mean(s)` |
| Var of total | `N^2 * (1 - n/N) * var(s) / n` |
| Proportion | `mean(binary_s)` |
| Prop variance | `(1 - n/N) * p * (1-p) / (n-1)` |
| Stratified mean | `sum(W_h * Xbar_h)` |
| Stratified var | `sum(W_h^2 * (1 - n_h/N_h) * s2_h / n_h)` |
| Proportional allocation | `n_h = round(n * N_h / N)` |
| Neyman allocation | `n_h = round(n * N_h * S_h / sum(N_h * S_h))` |
| Optimal with costs | `n_h = round(n * (N_h*S_h/sqrt(c_h)) / sum(...))` |
| Systematic sample | `idx = seq(start, by = k, length.out = n)` |
| Successive-difference var | `sum(diff(s)^2) / (2*n*(n-1)) * (1 - n/N)` |
| Cluster estimator | ratio of sums (y_total / size_total) |
| Cluster variance | `M^2 * (1 - m/M) * var(cluster_totals) / m` |
| Two-stage var | between + within components |
| Intraclass correlation | `(MSB - MSW) / (MSB + (n_i - 1)*MSW)` |
| Design effect | `deff = Var_design / Var_SRS` |
| Effective sample size | `n_eff = n / deff` |
| Ratio estimator | `mean(y) / mean(x) * mu_x` |
| Regression estimator | `mean(y) + b * (mu_x - mean(x))` |
| HT estimator (total) | `sum(y_i / pi_i)` |
| PPS-HH estimator | `mean(y_i / p_i)` |
| Bias | `mean(estimates) - truth` |
| Variance | `var(estimates)` |
| MSE | `mean((estimates - truth)^2)` |
| Relative efficiency | `MSE_2 / MSE_1` |
| Coverage | `mean(truth >= ci_lo & truth <= ci_hi)` |
| Sample size for margin | `z^2 * S^2 / e^2`, adjust with FPC |
| Sample size for proportion | `z^2 * p * (1-p) / e^2`, adjust |
| Sample size for target var | `S^2 / (V + S^2/N)` |
| Bootstrap CI | `quantile(boot_est, c(0.025, 0.975))` |
| Cluster bootstrap | resample whole clusters |

---

# Part 20 — Common Mistakes

1. **Forgetting `replace = FALSE`** for SRS without replacement.
2. **Using `sd(s)` instead of `var(s)`** in variance formulas.
3. **Confusing $N$ and $n$** — population vs sample size.
4. **Forgetting the FPC** when the sample is a large fraction of $N$.
5. **Not setting `set.seed()`** in simulations — results change.
6. **Using theoretical formulas** when the question asks for empirical estimates.
7. **Confusing bias and variance.**
8. **Rounding $n_h$ incorrectly** in stratified sampling — ensure $\sum n_h = n$.
9. **Reporting only the point estimate** without SE or CI.
10. **Using the wrong CI** — use $t$ when $n$ is small and $S^2$ is estimated.
11. **Assuming normality for small $n$** — use bootstrap or exact methods.
12. **Ignoring non-response bias.**
13. **Using SRS formulas for cluster sampling** — deff must be applied.
14. **Forgetting that HT is unbiased but can have high variance.**
15. **Mixing up design weights ($1/\pi_i$) and size weights ($1/p_i$).**
16. **Using PPS with replacement but applying without-replacement formulas.**
17. **Not checking that $n_h \geq 2$** for variance estimation.
18. **Using the wrong SE for the total** — remember to multiply by $N^2$.
19. **Forgetting to check that $\sum n_h$ equals the target $n$.**
20. **Interpreting CI coverage as certainty** — it's a long-run property.

---

# Part 21 — The Single Most Important Idea

Every sampling question reduces to:

1. **Design:** SRS, stratified, systematic, cluster, two-stage, PPS?
2. **Estimator:** mean, total, proportion, ratio?
3. **Variance:** apply the design-specific formula, include FPC.
4. **CI:** normal or t, with the estimated SE.
5. **Compare:** bias, MSE, relative efficiency, design effect.
6. **Simulate** when theory is hard or you want to validate.

If you can compute a design-correct estimator, its variance, and a CI, you can answer any sampling question.

---

*End of cheat sheet. Good luck in the competition.*
````

---

## What is covered

| Section | Topics |
|---|---|
| Part 0 | 30-second strategy |
| Part 1 | Foundations: parameters, estimators, FPC, bias/MSE |
| Part 2 | Building finite populations (continuous, stratified, discrete) |
| Part 3 | SRS: sample, mean, variance with FPC, CI, total, proportion, sample size |
| Part 4 | Stratified: proportional, Neyman, optimal with costs, all formulas |
| Part 5 | Systematic: sampling interval, variance approximations |
| Part 6 | Cluster sampling: estimators, variance, deff, intraclass correlation |
| Part 7 | Two-stage sampling: PSU/SSU, variance decomposition |
| Part 8 | Ratio and regression estimators |
| Part 9 | Bias, variance, MSE simulations, relative efficiency |
| Part 10 | Coverage probability simulations |
| Part 11 | Sample size determination (all designs) |
| Part 12 | Design effects and effective sample size |
| Part 13 | Horvitz-Thompson estimator |
| Part 14 | Probability proportional to size (PPS) |
| Part 15 | Non-response and weighting |
| Part 16 | Bootstrap variance and CI |
| Part 17 | Theory Q&A (25 questions) |
| Part 18 | Exam scenarios (10 step-by-step answers) |
| Part 19 | Quick reference table |
| Part 20 | Common mistakes (20 traps) |
| Part 21 | Single most important idea |

**Design consistency:** Every chunk is self-contained with section-prefixed names (`p3_pop`, `p6_m`, `p11_n_h`). Copy-paste any chunk into a fresh R session and it works.