---
title: "The Ultimate Simulation Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(knitr)
```

# How to Use This Cheat Sheet

| Topic | Section |
|---|---|
| 30-second strategy | 0 |
| Foundations: what and why | 1 |
| Random number generation | 2 |
| Distributions reference | 3 |
| Monte Carlo estimation | 4 |
| Bias, variance, MSE | 5 |
| Confidence intervals & coverage | 6 |
| Bootstrap | 7 |
| Permutation tests | 8 |
| Power simulation | 9 |
| Sample size via simulation | 10 |
| Simulating specific designs | 11 |
| Custom distributions (inverse, rejection) | 12 |
| Monte Carlo integration | 13 |
| Markov chains & random walks | 14 |
| Variance reduction | 15 |
| Jackknife & cross-validation | 16 |
| Simulating hypothesis tests under $H_0$ | 17 |
| Efficiency in R | 18 |
| Theory Q&A | 19 |
| Exam scenarios | 20 |
| Quick reference | 21 |
| Common mistakes | 22 |
| Single most important idea | 23 |

**Golden rule:** every chunk is self-contained.

---

# Part 0 — The 30-Second Strategy

For **any** simulation question, run this pipeline:

1. **What is being estimated or tested?** A mean, a variance, a bias, a p-value, a coverage, a power?
2. **What is the data-generating process (DGP)?** Write it down in code once.
3. **Choose $B$, the number of replications.** Rule of thumb: $B \geq 1000$ for estimation, $B \geq 5000$ for p-values/coverage.
4. **Loop or replicate:** generate data, compute the statistic, store the result.
5. **Summarise:** `mean`, `sd`, `quantile`.
6. **Report:** the estimated quantity, its Monte Carlo SE, and an interpretation.

```r
set.seed(1)
B <- 5000
results <- replicate(B, {
  # 1. Generate data
  # 2. Compute statistic
  # 3. Return it
})
mean(results); sd(results)
```

If you can do these six steps, you can answer any simulation question.

---

# Part 1 — Foundations

## 1.1 What is simulation?

**Simulation** (Monte Carlo) means running a **random experiment** many times to estimate a quantity that is hard to compute analytically.

**Three uses:**
1. **Estimate** quantities with no closed form (MSE of median, coverage of a complex CI).
2. **Validate** theory (does my CI achieve 95% coverage?).
3. **Explore** designs before collecting data (power, sample size).

## 1.2 The data-generating process (DGP)

Every simulation starts with a **recipe** for producing data:

- **Distribution:** $y \sim N(\mu, \sigma^2)$.
- **Sample size:** $n$.
- **Design:** one group, multiple groups, paired, time series.

Write the DGP in code **once**, then wrap it in `replicate()`.

## 1.3 Reproducibility

**Always** call `set.seed()` before generating random numbers:

```{r}
# --- Self-contained: reproducibility ---
set.seed(101)
p1_a <- rnorm(3)

set.seed(101)
p1_b <- rnorm(3)

identical(p1_a, p1_b)
```

**Rule:** Choose a seed (any integer), record it, and don't change it between runs.

## 1.4 Why randomness needs many runs

A single random sample gives one noisy estimate. The **law of large numbers** says the average over $B$ runs converges to the truth.

```{r}
# --- Self-contained: convergence ---
set.seed(102)
p1_means <- cumsum(rnorm(5000)) / 1:5000   # running mean of N(0,1)
tail(p1_means, 5)
```

**Interpretation:** The running mean settles near 0 as $B$ grows.

## 1.5 The Monte Carlo error

The standard error of a Monte Carlo estimate of a mean is:
$$\text{MC-SE} = \frac{s}{\sqrt{B}}$$
where $s$ = sample SD of the $B$ simulated values.

**Rule of thumb:** To halve the MC-SE, quadruple $B$.

## 1.6 Notation

| Symbol | Meaning |
|---|---|
| $B$ | Number of replications |
| $\hat{\theta}_b$ | Estimate from replication $b$ |
| $\hat{\theta}$ | True estimand (usually unknown) |
| $\bar{\hat{\theta}}$ | Average over replications |
| $s$ | SD of the $B$ estimates |
| MC-SE | Monte Carlo standard error $= s/\sqrt{B}$ |

## 1.7 The three quantities you simulate

1. **Bias:** $\widehat{\text{Bias}} = \bar{\hat{\theta}} - \theta$.
2. **Variance:** $\widehat{\text{Var}} = s^2$.
3. **MSE:** $\widehat{\text{MSE}} = \text{Bias}^2 + \text{Var}$.

Where $\theta$ is the true parameter you chose when setting up the DGP.

---

# Part 2 — Random Number Generation

## 2.1 Seeding

```{r}
# --- Self-contained: set.seed ---
set.seed(201)
p2_x <- rnorm(5)
p2_x
```

**Rules:**
- One seed per simulation study.
- If you change the seed, the answer changes.
- Never reseed inside the loop.

## 2.2 Basic generators

| Function | Distribution |
|---|---|
| `rnorm(n, mean, sd)` | Normal |
| `runif(n, min, max)` | Uniform |
| `rbinom(n, size, prob)` | Binomial |
| `rpois(n, lambda)` | Poisson |
| `rexp(n, rate)` | Exponential |
| `rgamma(n, shape, rate)` | Gamma |
| `rbeta(n, a, b)` | Beta |
| `rchisq(n, df)` | Chi-square |
| `rt(n, df)` | t |
| `rf(n, df1, df2)` | F |
| `rmultinom(n, size, prob)` | Multinomial |
| `sample(x, n, replace)` | From a vector |

## 2.3 Reproducibility across functions

```{r}
# --- Self-contained: same seed, different generators ---
set.seed(202)
p2_a <- rnorm(3)

set.seed(202)
p2_b <- rnorm(3)
p2_c <- runif(3)   # advances RNG state

identical(p2_a, p2_b)
```

**Note:** Different generators consume the RNG stream differently. Use one seed per experiment.

## 2.4 Sampling

```{r}
# --- Self-contained: sampling options ---
set.seed(203)
p2_pop <- 1:10

sample(p2_pop, 5)                        # without replacement
sample(p2_pop, 5, replace = TRUE)        # with replacement
sample(p2_pop, 5, prob = 1:10)           # weighted
sample(c("H", "T"), 10, replace = TRUE,
       prob = c(0.5, 0.5))               # coin flips
```

## 2.5 Sampling with strata (stratified)

```{r}
# --- Self-contained: stratified sampling ---
set.seed(204)
p2_strata <- c(rep("A", 100), rep("B", 200), rep("C", 300))
p2_values <- rnorm(600, 50, 10)

# Proportional allocation: n = 60
p2_n  <- 60
p2_N  <- table(p2_strata)
p2_nh <- round(p2_n * p2_N / sum(p2_N))

p2_sample <- unlist(lapply(names(p2_N), function(h) {
  idx <- which(p2_strata == h)
  sample(idx, p2_nh[h])
}))
length(p2_sample); table(p2_strata[p2_sample])
```

## 2.6 Shuffling and permutation

```{r}
# --- Self-contained: shuffling ---
set.seed(205)
p2_x <- 1:10
sample(p2_x)                # permutation
sample(p2_x)[1:3]           # random subset
```

## 2.7 Random splits for cross-validation

```{r}
# --- Self-contained: k-fold indices ---
set.seed(206)
p2_n <- 100
p2_K <- 5
p2_folds <- sample(rep(1:p2_K, length.out = p2_n))
table(p2_folds)
```

## 2.8 Reproducible random sampling for train/test

```{r}
# --- Self-contained: train/test split ---
set.seed(207)
p2_df <- tibble(id = 1:100, y = rnorm(100))
p2_idx <- sample(1:nrow(p2_df), size = 0.8 * nrow(p2_df))
p2_train <- p2_df[p2_idx, ]
p2_test  <- p2_df[-p2_idx, ]
nrow(p2_train); nrow(p2_test)
```

---

# Part 3 — Distributions Reference

For every distribution R provides four functions:

| Prefix | Function | Example |
|---|---|---|
| `d` | Density / PMF | `dnorm(0)` |
| `p` | CDF | `pnorm(1.96)` |
| `q` | Quantile (inverse CDF) | `qnorm(0.975)` |
| `r` | Random draw | `rnorm(5)` |

## 3.1 Normal

```{r}
# --- Self-contained: normal ---
dnorm(0)               # density at 0
pnorm(1.96)            # P(X <= 1.96)
qnorm(0.975)           # 97.5% quantile = 1.96
set.seed(301); rnorm(3)
```

## 3.2 Uniform

```{r}
# --- Self-contained: uniform ---
dunif(0.5)                # density at 0.5
punif(0.5)                # P(X <= 0.5)
qunif(0.5)                # median
set.seed(302); runif(3)
```

## 3.3 Binomial

```{r}
# --- Self-contained: binomial ---
dbinom(3, size = 10, prob = 0.5)   # P(X = 3)
pbinom(3, size = 10, prob = 0.5)   # P(X <= 3)
qbinom(0.95, size = 10, prob = 0.5)
set.seed(303); rbinom(5, size = 10, prob = 0.5)
```

## 3.4 Poisson

```{r}
# --- Self-contained: Poisson ---
dpois(2, lambda = 3)                # P(X = 2)
ppois(2, lambda = 3)                # P(X <= 2)
qpois(0.95, lambda = 3)
set.seed(304); rpois(5, lambda = 3)
```

## 3.5 Exponential

```{r}
# --- Self-contained: exponential ---
dexp(1, rate = 2)
pexp(1, rate = 2)
qexp(0.5, rate = 2)
set.seed(305); rexp(5, rate = 2)
```

## 3.6 Chi-square, t, F

```{r}
# --- Self-contained: chisq, t, F ---
pchisq(3.84, df = 1)          # ≈ 0.95
qt(0.975, df = 10)            # 97.5% t quantile
qf(0.95, df1 = 2, df2 = 10)   # 95% F quantile
set.seed(306); rchisq(3, df = 5); rt(3, df = 10); rf(3, df1 = 2, df2 = 5)
```

## 3.7 Multivariate normal

```{r}
# --- Self-contained: MVN ---
set.seed(307)
p3_mu    <- c(0, 0)
p3_Sigma <- matrix(c(1, 0.7, 0.7, 1), nrow = 2)

# Manual MVN draw
p3_z <- matrix(rnorm(10 * 2), ncol = 2)
p3_x <- p3_z %*% chol(p3_Sigma) + matrix(p3_mu, nrow = 10, ncol = 2, byrow = TRUE)
round(head(p3_x, 3), 3)
```

**Why `chol`:** The Cholesky factor of $\Sigma$ gives the "square root" of the covariance matrix.

## 3.8 Truncated distributions

```{r}
# --- Self-contained: truncated normal (rejection) ---
set.seed(308)
p3_n <- 1000
p3_x <- rnorm(p3_n)
p3_trunc <- p3_x[p3_x > 0]     # keep only values > 0
length(p3_trunc)
```

## 3.9 Discrete distributions

```{r}
# --- Self-contained: general discrete ---
set.seed(309)
p3_vals   <- c(0, 1, 2, 5)
p3_probs  <- c(0.2, 0.4, 0.3, 0.1)

p3_draws <- sample(p3_vals, 1000, replace = TRUE, prob = p3_probs)
table(p3_draws) / 1000
```

## 3.10 Verifying a generator

```{r}
# --- Self-contained: verify generator ---
set.seed(310)
p3_x <- rnorm(10000, mean = 5, sd = 2)
c(mean = mean(p3_x), sd = sd(p3_x))
```

**Interpretation:** The empirical mean and SD should be close to the theoretical values (5 and 2).

---

# Part 4 — Monte Carlo Estimation

## 4.1 Estimating a mean

**Goal:** Estimate $E[X]$ for $X \sim N(50, 100)$.

```{r}
# --- Self-contained: MC estimate of a mean ---
set.seed(401)
p4_B <- 5000
p4_x <- rnorm(p4_B, mean = 50, sd = 10)
mean(p4_x)
sd(p4_x) / sqrt(p4_B)
```

**Interpretation:** Estimate = mean of the draws. MC-SE = `sd/sqrt(B)`. The theoretical answer is 50.

## 4.2 Estimating a variance

```{r}
# --- Self-contained: MC estimate of variance ---
set.seed(402)
p4_B <- 5000
p4_x <- rnorm(p4_B, mean = 50, sd = 10)
var(p4_x)
```

**Theoretical:** 100.

## 4.3 Estimating a probability

**Goal:** $P(X > 2)$ where $X \sim N(0, 1)$.

```{r}
# --- Self-contained: MC estimate of a probability ---
set.seed(403)
p4_B <- 10000
p4_x <- rnorm(p4_B)
mean(p4_x > 2)              # estimated probability
sqrt(mean(p4_x > 2) * (1 - mean(p4_x > 2)) / p4_B)   # MC-SE
```

**Theoretical:** 1 − Φ(2) = `r 1-pnorm(2)`.

## 4.4 Estimating a quantile

```{r}
# --- Self-contained: MC estimate of a quantile ---
set.seed(404)
p4_B <- 10000
p4_x <- rnorm(p4_B)
quantile(p4_x, 0.975)
```

**Theoretical:** `qnorm(0.975)` = 1.96.

## 4.5 Estimating a nonlinear function

**Goal:** $E[X^2]$ for $X \sim U(0, 1)$.

```{r}
# --- Self-contained: MC estimate of a nonlinear function ---
set.seed(405)
p4_B <- 5000
p4_x <- runif(p4_B)
mean(p4_x^2)
```

**Theoretical:** $E[X^2] = 1/3 \approx 0.333$.

## 4.6 Estimating a tail probability

```{r}
# --- Self-contained: rare event ---
set.seed(406)
p4_B <- 100000
p4_x <- rnorm(p4_B)
mean(p4_x > 4)              # P(X > 4) ≈ 3.17e-5
```

**Warning:** For very rare events, use importance sampling (Part 15) to reduce variance.

## 4.7 Estimating a joint probability

**Goal:** $P(X > 0, Y > 0)$ where $(X, Y)$ are jointly normal with $\rho = 0.5$.

```{r}
# --- Self-contained: joint probability ---
set.seed(407)
p4_B <- 10000
p4_Sigma <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
p4_z <- matrix(rnorm(p4_B * 2), ncol = 2)
p4_x <- p4_z %*% chol(p4_Sigma)
mean(p4_x[,1] > 0 & p4_x[,2] > 0)
```

**Interpretation:** Estimate of the joint probability.

## 4.8 Monte Carlo summary table

| Quantity | Estimator | MC-SE |
|---|---|---|
| Mean | `mean(x)` | `sd(x)/sqrt(B)` |
| Variance | `var(x)` | `sd(x^2)/sqrt(B) - Var` (approx) |
| Probability | `mean(x > c)` | `sqrt(p(1-p)/B)` |
| Quantile | `quantile(x, p)` | via bootstrap |
| $\int f(x) dx$ | `mean(f(x))` | `sd(f(x))/sqrt(B)` |

## 4.9 Report template

> "Monte Carlo with $B = 10000$ replications estimates the mean at $\hat{\theta} = 49.98$ with Monte Carlo SE $0.10$. The theoretical value is 50."

---

# Part 5 — Bias, Variance, MSE

## 5.1 Setup

Choose a **true parameter** $\theta$. Estimate it $B$ times. Compute bias, variance, MSE.

$$\text{Bias} = \bar{\hat{\theta}} - \theta, \quad \text{Var} = s^2, \quad \text{MSE} = \text{Bias}^2 + \text{Var}$$

## 5.2 Bias of the sample mean (should be 0)

```{r}
# --- Self-contained: bias of the sample mean ---
set.seed(501)
p5_B  <- 5000
p5_n  <- 30
p5_mu <- 10

p5_means <- replicate(p5_B, mean(rnorm(p5_n, mean = p5_mu)))

c(bias = mean(p5_means) - p5_mu,
  var  = var(p5_means),
  mse  = mean((p5_means - p5_mu)^2))
```

**Theoretical:** Bias = 0, Var = σ²/n = 1/30 ≈ 0.033, MSE = 0.033.

## 5.3 Bias of the sample median

```{r}
# --- Self-contained: bias of the median ---
set.seed(502)
p5_B  <- 5000
p5_n  <- 30
p5_mu <- 0

p5_med <- replicate(p5_B, median(rnorm(p5_n, mean = p5_mu)))

c(bias = mean(p5_med) - p5_mu,
  var  = var(p5_med),
  mse  = mean((p5_med - p5_mu)^2))
```

**Interpretation:** Median may be slightly biased for small $n$; variance is larger than mean's variance.

## 5.4 Bias of the sample variance (using $n$)

If you use $\frac{1}{n}\sum(y_i - \bar{y})^2$, the estimator is biased.

```{r}
# --- Self-contained: biased variance estimator ---
set.seed(503)
p5_B <- 5000
p5_n <- 20
p5_sigma2 <- 4

p5_biased <- replicate(p5_B, {
  y <- rnorm(p5_n, sd = sqrt(p5_sigma2))
  mean((y - mean(y))^2)
})

c(bias = mean(p5_biased) - p5_sigma2)
```

**Theoretical:** Bias = $-\sigma^2/n$ = −0.2.

## 5.5 Bias of the ratio estimator

```{r}
# --- Self-contained: bias of ratio estimator ---
set.seed(504)
p5_B  <- 5000
p5_n  <- 30
p5_N  <- 500
p5_x  <- rnorm(p5_N, mean = 20, sd = 4)
p5_y  <- 3 * p5_x + rnorm(p5_N, sd = 3)
p5_mux <- mean(p5_x)
p5_muy <- mean(p5_y)

p5_ratio <- replicate(p5_B, {
  idx <- sample(p5_N, p5_n)
  mean(p5_y[idx]) / mean(p5_x[idx]) * p5_mux
})

c(true = p5_muy,
  bias = mean(p5_ratio) - p5_muy,
  var  = var(p5_ratio),
  mse  = mean((p5_ratio - p5_muy)^2))
```

**Interpretation:** Ratio estimator is biased but often has smaller MSE than the sample mean.

## 5.6 Comparing two estimators

```{r}
# --- Self-contained: mean vs median ---
set.seed(505)
p5_B  <- 5000
p5_n  <- 30
p5_mu <- 0

p5_mean <- replicate(p5_B, mean(rnorm(p5_n)))
p5_med  <- replicate(p5_B, median(rnorm(p5_n)))

tibble(
  Estimator = c("Mean", "Median"),
  Bias      = c(mean(p5_mean) - p5_mu, mean(p5_med) - p5_mu),
  Variance  = c(var(p5_mean), var(p5_med)),
  MSE       = c(mean((p5_mean - p5_mu)^2), mean((p5_med - p5_mu)^2))
) %>% kable(digits = 5)
```

**Interpretation:** For normal data, mean has smaller MSE. For heavy-tailed data, median wins.

## 5.7 Relative efficiency

$$\text{RE} = \frac{\text{MSE}(\hat{\theta}_2)}{\text{MSE}(\hat{\theta}_1)}$$

```{r}
# --- Self-contained: relative efficiency ---
set.seed(506)
p5_mean <- replicate(5000, mean(rnorm(30)))
p5_med  <- replicate(5000, median(rnorm(30)))

mse_mean <- mean(p5_mean^2)
mse_med  <- mean(p5_med^2)
c(RE_median_vs_mean = mse_med / mse_mean,
  RE_mean_vs_median = mse_mean / mse_med)
```

**Interpretation:** If RE > 1, the numerator is more efficient.

## 5.8 Report template

> "Over 5000 replications of sample size 30, the sample mean has bias ≈ 0.001, variance 0.033, MSE 0.033. The sample median has bias ≈ 0.002, variance 0.049, MSE 0.049. The mean is more efficient (RE ≈ 0.67)."

---

# Part 6 — Confidence Intervals & Coverage

## 6.1 Coverage of a CI

**Coverage** = fraction of CIs that contain the true parameter over many samples.

**Target:** 95% coverage for a 95% CI.

```{r}
# --- Self-contained: coverage of a t-based CI ---
set.seed(601)
p6_B  <- 5000
p6_n  <- 25
p6_mu <- 10
p6_sd <- 2

p6_cover <- replicate(p6_B, {
  y <- rnorm(p6_n, mean = p6_mu, sd = p6_sd)
  ci <- t.test(y)$conf.int
  ci[1] <= p6_mu & p6_mu <= ci[2]
})

mean(p6_cover)
```

**Interpretation:** Should be close to 0.95. If it's 0.94 or 0.96, fine.

## 6.2 Coverage with a bad CI (wrong SE)

```{r}
# --- Self-contained: coverage with wrong SE ---
set.seed(602)
p6_B  <- 5000
p6_n  <- 25
p6_mu <- 10
p6_sd <- 2

p6_cover_bad <- replicate(p6_B, {
  y <- rnorm(p6_n, mean = p6_mu, sd = p6_sd)
  z <- qnorm(0.975)
  se_wrong <- sd(y) / sqrt(p6_n - 1)   # WRONG denominator
  ci <- mean(y) + c(-1, 1) * z * se_wrong
  ci[1] <= p6_mu & p6_mu <= ci[2]
})

mean(p6_cover_bad)
```

**Interpretation:** Coverage is slightly off — demonstrates why SE formula matters.

## 6.3 Coverage with FPC (finite population)

```{r}
# --- Self-contained: coverage with FPC ---
set.seed(603)
p6_B  <- 5000
p6_N  <- 200
p6_n  <- 40
p6_pop <- rnorm(p6_N, mean = 50, sd = 10)
p6_mu  <- mean(p6_pop)
p6_z   <- qnorm(0.975)

p6_cover_fpc <- replicate(p6_B, {
  s <- sample(p6_pop, p6_n)
  se <- sqrt((1 - p6_n / p6_N) * var(s) / p6_n)
  ci <- mean(s) + c(-1, 1) * p6_z * se
  ci[1] <= p6_mu & p6_mu <= ci[2]
})

mean(p6_cover_fpc)
```

**Interpretation:** With FPC, coverage ≈ 0.95 and CI is shorter than without FPC.

## 6.4 Coverage of a bootstrap CI

```{r}
# --- Self-contained: bootstrap percentile CI coverage ---
set.seed(604)
p6_B  <- 2000
p6_n  <- 30
p6_mu <- 5

p6_cover_boot <- replicate(p6_B, {
  y <- rnorm(p6_n, mean = p6_mu)
  boot_means <- replicate(500, mean(sample(y, replace = TRUE)))
  ci <- quantile(boot_means, c(0.025, 0.975))
  ci[1] <= p6_mu & p6_mu <= ci[2]
})

mean(p6_cover_boot)
```

**Interpretation:** Percentile bootstrap CI has ~95% coverage for the mean of a normal distribution.

## 6.5 CI width as a function of $n$

```{r}
# --- Self-contained: CI width vs n ---
set.seed(605)
p6_n_grid <- c(10, 20, 50, 100, 500)

p6_widths <- sapply(p6_n_grid, function(nn) {
  mean(replicate(1000, diff(t.test(rnorm(nn))$conf.int)))
})

tibble(n = p6_n_grid, width = p6_widths) %>% kable(digits = 3)
```

**Interpretation:** CI width shrinks like $1/\sqrt{n}$.

## 6.6 Report template

> "Over 5000 replications, the 95% CI has empirical coverage of 0.948, consistent with the nominal 95% level. Average CI width is 1.02."

---

# Part 7 — Bootstrap

## 7.1 The idea

**Resample with replacement** from the observed data, compute the statistic, repeat $B$ times. The distribution of bootstrap replicates approximates the sampling distribution.

## 7.2 Bootstrap SE of a mean

```{r}
# --- Self-contained: bootstrap SE ---
set.seed(701)
p7_y <- rnorm(50, mean = 10, sd = 2)

p7_B <- 2000
p7_boot_means <- replicate(p7_B, mean(sample(p7_y, replace = TRUE)))

c(se_boot = sd(p7_boot_means),
  se_theory = sd(p7_y) / sqrt(length(p7_y)))
```

**Interpretation:** Bootstrap SE matches theoretical SE.

## 7.3 Bootstrap CI: percentile method

```{r}
# --- Self-contained: percentile bootstrap CI ---
set.seed(702)
p7_y <- rnorm(50, mean = 10, sd = 2)

p7_boot_means <- replicate(2000, mean(sample(p7_y, replace = TRUE)))
quantile(p7_boot_means, c(0.025, 0.975))
```

**Interpretation:** 95% CI is the 2.5% and 97.5% quantiles.

## 7.4 Bootstrap CI: basic (reverse percentile)

```{r}
# --- Self-contained: basic bootstrap CI ---
set.seed(703)
p7_y <- rnorm(50, mean = 10, sd = 2)

p7_boot_means <- replicate(2000, mean(sample(p7_y, replace = TRUE)))
p7_pct <- quantile(p7_boot_means, c(0.025, 0.975))
# Basic CI: 2*theta_hat - upper, 2*theta_hat - lower
p7_basic <- 2 * mean(p7_y) - rev(p7_pct)
p7_basic
```

## 7.5 Bootstrap CI: BCa (bias-corrected accelerated)

Not built into base R but easy to code approximately. The percentile CI is usually fine.

## 7.6 Bootstrap for a correlation

```{r}
# --- Self-contained: bootstrap correlation ---
set.seed(704)
p7_df <- tibble(x = rnorm(100), y = rnorm(100))
p7_boot_cor <- replicate(2000, {
  idx <- sample(nrow(p7_df), replace = TRUE)
  cor(p7_df$x[idx], p7_df$y[idx])
})
c(est = cor(p7_df$x, p7_df$y),
  lo = quantile(p7_boot_cor, 0.025),
  hi = quantile(p7_boot_cor, 0.975))
```

## 7.7 Bootstrap for a regression slope

```{r}
# --- Self-contained: bootstrap slope ---
set.seed(705)
p7_df <- tibble(x = rnorm(100), y = 2 + 3 * x + rnorm(100))
p7_boot_slope <- replicate(2000, {
  idx <- sample(nrow(p7_df), replace = TRUE)
  coef(lm(y ~ x, data = p7_df[idx, ]))[2]
})
c(est = coef(lm(y ~ x, data = p7_df))[2],
  lo = quantile(p7_boot_slope, 0.025),
  hi = quantile(p7_boot_slope, 0.975))
```

## 7.8 Bootstrap for a proportion

```{r}
# --- Self-contained: bootstrap proportion ---
set.seed(706)
p7_y <- rbinom(100, 1, 0.4)
p7_boot_p <- replicate(2000, mean(sample(p7_y, replace = TRUE)))
c(est = mean(p7_y),
  lo = quantile(p7_boot_p, 0.025),
  hi = quantile(p7_boot_p, 0.975))
```

## 7.9 Block bootstrap for time series

When observations are autocorrelated, resample **blocks** of consecutive observations:

```{r}
# --- Self-contained: block bootstrap ---
set.seed(707)
p7_ts <- arima.sim(model = list(ar = 0.7), n = 100)
p7_B  <- 2000
p7_L  <- 10      # block length
p7_n  <- length(p7_ts)

p7_boot_mean <- replicate(p7_B, {
  n_blocks <- ceiling(p7_n / p7_L)
  starts <- sample(1:(p7_n - p7_L + 1), n_blocks, replace = TRUE)
  blocks <- unlist(lapply(starts, function(s) p7_ts[s:(s + p7_L - 1)]))
  mean(blocks[1:p7_n])
})
c(est = mean(p7_ts),
  lo = quantile(p7_boot_mean, 0.025),
  hi = quantile(p7_boot_mean, 0.975))
```

## 7.10 Parametric bootstrap

When you have a model, simulate from the fitted model:

```{r}
# --- Self-contained: parametric bootstrap ---
set.seed(708)
p7_df <- tibble(x = rnorm(100), y = 2 + 3 * x + rnorm(100))
p7_fit <- lm(y ~ x, data = p7_df)
p7_sigma_hat <- summary(p7_fit)$sigma
p7_beta_hat  <- coef(p7_fit)

p7_boot <- replicate(2000, {
  p7_df$y_sim <- p7_beta_hat[1] + p7_beta_hat[2] * p7_df$x +
                 rnorm(nrow(p7_df), sd = p7_sigma_hat)
  coef(lm(y_sim ~ x, data = p7_df))[2]
})
c(est = p7_beta_hat[2],
  lo = quantile(p7_boot, 0.025),
  hi = quantile(p7_boot, 0.975))
```

## 7.11 Cluster bootstrap

Resample **whole clusters** (e.g., schools, subjects):

```{r}
# --- Self-contained: cluster bootstrap ---
set.seed(709)
p7_clusters <- 30
p7_sizes    <- sample(5:15, p7_clusters, replace = TRUE)
p7_df <- tibble(
  cluster = rep(1:p7_clusters, times = p7_sizes),
  y = rnorm(sum(p7_sizes)) + rep(rnorm(p7_clusters, 0, 2), times = p7_sizes)
)

p7_unique_clusters <- unique(p7_df$cluster)
p7_boot <- replicate(2000, {
  picked <- sample(p7_unique_clusters, replace = TRUE)
  y <- unlist(lapply(picked, function(k) p7_df$y[p7_df$cluster == k]))
  mean(y)
})
c(est = mean(p7_df$y),
  lo = quantile(p7_boot, 0.025),
  hi = quantile(p7_boot, 0.975))
```

## 7.12 Bootstrap summary

| Purpose | Method |
|---|---|
| Standard error | `sd(boot_stats)` |
| CI (symmetric) | `mean ± 1.96 * sd(boot_stats)` |
| CI (percentile) | `quantile(boot_stats, c(0.025, 0.975))` |
| CI (basic) | `2*theta_hat - rev(quantile)` |
| Time series | Block bootstrap |
| Clustered | Cluster bootstrap |
| Regression | Resample residuals (parametric) or rows (non-parametric) |

## 7.13 Report template

> "Bootstrap with 2000 replications gives SE = 0.29 and a 95% percentile CI of [9.42, 10.55], consistent with the t-based CI [9.44, 10.56]."

---

# Part 8 — Permutation Tests

## 8.1 The idea

Under $H_0$ (no effect), the group labels are arbitrary. Shuffle them, recompute the test statistic, and compare to the observed value.

## 8.2 Two-sample permutation test

```{r}
# --- Self-contained: permutation test for two means ---
set.seed(801)
p8_x <- rnorm(30, mean = 10)
p8_y <- rnorm(30, mean = 11)

p8_obs <- mean(p8_x) - mean(p8_y)
p8_pooled <- c(p8_x, p8_y)
p8_B <- 5000

p8_perms <- replicate(p8_B, {
  idx <- sample(length(p8_pooled), length(p8_x))
  mean(p8_pooled[idx]) - mean(p8_pooled[-idx])
})

p8_pval <- mean(abs(p8_perms) >= abs(p8_obs))
p8_pval
```

**Interpretation:** $p$-value = fraction of permutations with a statistic as extreme as observed.

## 8.3 Permutation test for correlation

```{r}
# --- Self-contained: permutation test for correlation ---
set.seed(802)
p8_x <- rnorm(50)
p8_y <- 0.5 * p8_x + rnorm(50, sd = sqrt(1 - 0.25))

p8_obs <- cor(p8_x, p8_y)
p8_B <- 5000

p8_perms <- replicate(p8_B, {
  cor(p8_x, sample(p8_y))
})

mean(abs(p8_perms) >= abs(p8_obs))
```

## 8.4 Permutation test for a regression slope

```{r}
# --- Self-contained: permutation test for slope ---
set.seed(803)
p8_df <- tibble(x = rnorm(60), y = 2 + 0.5 * x + rnorm(60))
p8_obs <- coef(lm(y ~ x, data = p8_df))[2]

p8_B <- 5000
p8_perms <- replicate(p8_B, {
  coef(lm(y ~ sample(x), data = p8_df))[2]
})

mean(abs(p8_perms) >= abs(p8_obs))
```

## 8.5 Paired permutation test

```{r}
# --- Self-contained: paired permutation test ---
set.seed(804)
p8_diff <- rnorm(30, mean = 0.5)
p8_obs  <- mean(p8_diff)

p8_perms <- replicate(5000, mean(sample(c(-1, 1), length(p8_diff),
                                        replace = TRUE) * p8_diff))
mean(abs(p8_perms) >= abs(p8_obs))
```

## 8.6 Permutation test for ANOVA

```{r}
# --- Self-contained: permutation F-test ---
set.seed(805)
p8_df <- tibble(
  group = factor(rep(c("A", "B", "C"), each = 20)),
  y = c(rnorm(20, 10), rnorm(20, 12), rnorm(20, 14))
)

p8_obs <- summary(aov(y ~ group, data = p8_df))[[1]][["F value"]][1]
p8_B <- 5000

p8_perms <- replicate(p8_B, {
  y_perm <- sample(p8_df$y)
  summary(aov(y_perm ~ group, data = p8_df))[[1]][["F value"]][1]
})

mean(p8_perms >= p8_obs)
```

## 8.7 Report template

> "The permutation test with 5000 permutations gives $p = 0.013$, consistent with the parametric $t$-test ($p = 0.012$)."

---

# Part 9 — Power Simulation

## 9.1 What is power?

**Power** = probability of rejecting $H_0$ when $H_0$ is false.

## 9.2 Power of a two-sample t-test

```{r}
# --- Self-contained: power of two-sample t-test ---
set.seed(901)
p9_B  <- 2000
p9_n  <- 30
p9_d  <- 0.5    # Cohen's d

p9_reject <- replicate(p9_B, {
  x <- rnorm(p9_n, mean = 0)
  y <- rnorm(p9_n, mean = p9_d)
  t.test(x, y)$p.value < 0.05
})

mean(p9_reject)
```

**Interpretation:** Power ≈ 0.42 for this configuration.

## 9.3 Power as a function of effect size

```{r}
# --- Self-contained: power curve ---
set.seed(902)
p9_d_grid <- seq(0, 1, by = 0.1)
p9_n <- 30

p9_power <- sapply(p9_d_grid, function(d) {
  mean(replicate(1000, {
    t.test(rnorm(p9_n), rnorm(p9_n, mean = d))$p.value < 0.05
  }))
})

tibble(d = p9_d_grid, power = round(p9_power, 3)) %>% kable()
```

**Interpretation:** Power increases with effect size.

## 9.4 Power as a function of $n$

```{r}
# --- Self-contained: power vs n ---
set.seed(903)
p9_n_grid <- c(10, 20, 30, 50, 100)
p9_d <- 0.5

p9_power <- sapply(p9_n_grid, function(nn) {
  mean(replicate(1000, {
    t.test(rnorm(nn), rnorm(nn, mean = p9_d))$p.value < 0.05
  }))
})

tibble(n = p9_n_grid, power = round(p9_power, 3)) %>% kable()
```

**Interpretation:** Power increases with $n$.

## 9.5 Power of a chi-square test

```{r}
# --- Self-contained: power of chi-square ---
set.seed(904)
p9_n  <- 200
p9_p1 <- 0.5
p9_p2 <- 0.3

p9_reject <- replicate(2000, {
  g1 <- rbinom(p9_n, 1, p9_p1)
  g2 <- rbinom(p9_n, 1, p9_p2)
  tab <- table(g1, g2)
  chisq.test(tab)$p.value < 0.05
})
mean(p9_reject)
```

## 9.6 Power of ANOVA

```{r}
# --- Self-contained: power of one-way ANOVA ---
set.seed(905)
p9_n <- 30
p9_means <- c(10, 11, 12)

p9_reject <- replicate(2000, {
  g <- factor(rep(c("A", "B", "C"), each = p9_n))
  y <- c(rnorm(p9_n, p9_means[1]),
         rnorm(p9_n, p9_means[2]),
         rnorm(p9_n, p9_means[3]))
  summary(aov(y ~ g))[[1]][["Pr(>F)"]][1] < 0.05
})
mean(p9_reject)
```

## 9.7 Power of a regression test

```{r}
# --- Self-contained: power of regression slope ---
set.seed(906)
p9_n <- 50
p9_beta <- 0.3

p9_reject <- replicate(2000, {
  x <- rnorm(p9_n)
  y <- p9_beta * x + rnorm(p9_n)
  summary(lm(y ~ x))$coefficients["x", "Pr(>|t|)"] < 0.05
})
mean(p9_reject)
```

## 9.8 Power curve plot

```{r}
# --- Self-contained: plot power curve ---
set.seed(907)
p9_d_grid <- seq(0, 1, by = 0.05)
p9_power <- sapply(p9_d_grid, function(d) {
  mean(replicate(500, t.test(rnorm(30), rnorm(30, d))$p.value < 0.05))
})

plot(p9_d_grid, p9_power, type = "b",
     xlab = "Effect size (Cohen's d)", ylab = "Power",
     main = "Power curve for two-sample t-test (n = 30)")
abline(h = 0.80, lty = 2, col = "red")
```

**Interpretation:** The horizontal line marks 80% power. Effect size for 80% power ≈ 0.75.

## 9.9 Report template

> "With n = 30 per group and d = 0.5, simulated power is 0.42. To reach 80% power, effect size must be at least d = 0.75 or n must be doubled."

---

# Part 10 — Sample Size via Simulation

## 10.1 Find $n$ for target power

```{r}
# --- Self-contained: sample size for 80% power ---
set.seed(1001)
p10_d <- 0.5
p10_target <- 0.80

p10_n <- 5
repeat {
  pw <- mean(replicate(500, t.test(rnorm(p10_n), rnorm(p10_n, p10_d))$p.value < 0.05))
  if (pw >= p10_target) break
  p10_n <- p10_n + 2
}
p10_n
```

**Interpretation:** Minimum $n$ per group for 80% power.

## 10.2 Binary search for $n$

```{r}
# --- Self-contained: binary search for n ---
set.seed(1002)
p10_d <- 0.5
p10_target <- 0.80

p10_power_at <- function(n) {
  mean(replicate(500, t.test(rnorm(n), rnorm(n, p10_d))$p.value < 0.05))
}

lo <- 5; hi <- 200
while (hi - lo > 1) {
  mid <- floor((lo + hi) / 2)
  if (p10_power_at(mid) >= p10_target) hi <- mid else lo <- mid
}
hi
```

**Why binary search:** Faster than linear scan.

## 10.3 Sample size for a proportion test

```r
# --- Self-contained: sample size for two proportions ---
set.seed(1003)
p10_p1 <- 0.5
p10_p2 <- 0.3
p10_target <- 0.80

p10_power_at <- function(n) {
  mean(replicate(500, {
    x <- c(sum(rbinom(n, 1, p10_p1)), sum(rbinom(n, 1, p10_p2)))
    prop.test(x, c(n, n))$p.value < 0.05
  }))
}

p10_n <- 10
while (p10_power_at(p10_n) < p10_target) p10_n <- p10_n + 5
p10_n
```

## 10.4 Sample size for a chi-square

```r
# --- Self-contained: sample size for chi-square ---
set.seed(1004)
p10_target <- 0.80

p10_power_at <- function(n) {
  mean(replicate(500, {
    g <- factor(rep(c("A", "B", "C"), each = n))
    y <- c(rnorm(n, 0), rnorm(n, 0.5), rnorm(n, 1.0))
    summary(aov(y ~ g))[[1]][["Pr(>F)"]][1] < 0.05
  }))
}

p10_n <- 5
while (p10_power_at(p10_n) < p10_target) p10_n <- p10_n + 2
p10_n
```

## 10.5 Report template

> "Simulation shows that n = 64 per group gives 80% power for detecting d = 0.5 at α = 0.05. This matches the analytic formula (pwr.t.test) which gives n = 63.8."

---

# Part 11 — Simulating Specific Designs

## 11.1 Simulating a regression dataset

```{r}
# --- Self-contained: regression data ---
set.seed(1101)
p11_n <- 100
p11_df <- tibble(
  x1 = rnorm(p11_n),
  x2 = rnorm(p11_n),
  y  = 2 + 3 * x1 - 1.5 * x2 + rnorm(p11_n)
)
head(p11_df)
```

## 11.2 Simulating a logistic regression dataset

```r
# --- Self-contained: logistic data ---
set.seed(1102)
p11_n <- 500
p11_df <- tibble(
  x = rnorm(p11_n),
  p = plogis(-1 + 1.5 * x),
  y = rbinom(p11_n, 1, p)
)
table(p11_df$y)
```

## 11.3 Simulating a Poisson regression dataset

```r
# --- Self-contained: Poisson data ---
set.seed(1103)
p11_n <- 300
p11_df <- tibble(
  x = rnorm(p11_n),
  y = rpois(p11_n, exp(1 + 0.5 * x))
)
head(p11_df)
```

## 11.4 Simulating a one-way ANOVA

```r
# --- Self-contained: one-way ANOVA data ---
set.seed(1104)
p11_k  <- 3
p11_n  <- 20
p11_mu <- c(10, 12, 15)

p11_df <- tibble(
  group = factor(rep(LETTERS[1:p11_k], each = p11_n)),
  y = rep(p11_mu, each = p11_n) + rnorm(p11_k * p11_n)
)
head(p11_df)
```

## 11.5 Simulating a two-way ANOVA

```r
# --- Self-contained: two-way ANOVA data ---
set.seed(1105)
p11_df <- expand.grid(
  A = factor(c("a1", "a2")),
  B = factor(c("b1", "b2")),
  rep = 1:20
)
p11_df$y <- 10 +
            2 * (p11_df$A == "a2") +
            1 * (p11_df$B == "b2") +
            1.5 * (p11_df$A == "a2" & p11_df$B == "b2") +
            rnorm(nrow(p11_df))
```

## 11.6 Simulating repeated measures

```r
# --- Self-contained: repeated measures data ---
set.seed(1106)
p11_subj <- 20
p11_times <- 3
p11_effect <- rep(rnorm(p11_subj, 0, 1), each = p11_times)

p11_df <- expand.grid(
  subject = factor(1:p11_subj),
  time    = factor(c("t1", "t2", "t3"))
) %>%
  mutate(y = 10 + as.numeric(time) * 1.5 + p11_effect + rnorm(nrow(.)))
head(p11_df)
```

## 11.7 Simulating an AR(1) time series

```{r}
# --- Self-contained: AR(1) series ---
set.seed(1107)
p11_ts <- arima.sim(model = list(ar = 0.7), n = 200)
plot(p11_ts, main = "Simulated AR(1)")
```

## 11.8 Simulating a random walk

```{r}
# --- Self-contained: random walk ---
set.seed(1108)
p11_rw <- cumsum(rnorm(200))
plot(p11_rw, type = "l", main = "Random walk")
```

## 11.9 Simulating an ARIMA(p,d,q)

```{r}
# --- Self-contained: ARIMA(1,1,1) ---
set.seed(1109)
p11_arima <- arima.sim(model = list(ar = 0.5, ma = 0.3), n = 200)
p11_arima_diff <- diff(p11_arima)   # if you want to see the difference
plot(p11_arima, main = "Simulated ARMA(1,1)")
```

## 11.10 Simulating a seasonal series

```{r}
# --- Self-contained: seasonal series ---
set.seed(1110)
p11_t <- 1:120
p11_seas <- 10 + 0.3 * p11_t +
            rep(c(1, 2, 3, 2, 1, 0, -1, -2, -1, 0, 1, 2), 10) +
            rnorm(120)
p11_ts <- ts(p11_seas, start = c(2000, 1), frequency = 12)
plot(p11_ts, main = "Seasonal series")
```

## 11.11 Simulating stratified data

```r
# --- Self-contained: stratified data ---
set.seed(1111)
p11_N_h <- c(100, 200, 300)
p11_mu_h <- c(10, 12, 15)
p11_sd_h <- c(2, 3, 4)

p11_pop <- unlist(mapply(rnorm, n = p11_N_h, mean = p11_mu_h, sd = p11_sd_h))
p11_strata <- rep(1:3, times = p11_N_h)
length(p11_pop)
```

## 11.12 Simulating clustered data

```r
# --- Self-contained: clustered data ---
set.seed(1112)
p11_M <- 30                # clusters
p11_m <- 10                # units per cluster
p11_cluster_effect <- rnorm(p11_M, 0, 2)

p11_df <- tibble(
  cluster = rep(1:p11_M, each = p11_m),
  y = rep(p11_cluster_effect, each = p11_m) + rnorm(p11_M * p11_m)
)
```

## 11.13 Simulating a contingency table

```r
# --- Self-contained: contingency table ---
set.seed(1113)
p11_df <- tibble(
  A = sample(c("a1", "a2"), 300, replace = TRUE),
  B = sample(c("b1", "b2"), 300, replace = TRUE)
)
table(p11_df$A, p11_df$B)
```

## 11.14 Simulating multivariate normal

```r
# --- Self-contained: MVN data ---
set.seed(1114)
p11_mu    <- c(0, 0)
p11_Sigma <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
p11_z <- matrix(rnorm(200), ncol = 2)
p11_x <- p11_z %*% chol(p11_Sigma) + matrix(p11_mu, nrow = 100, ncol = 2, byrow = TRUE)
cov(p11_x)
```

---

# Part 12 — Custom Distributions

## 12.1 Inverse-transform sampling

If $F$ is the CDF, then $F^{-1}(U) \sim F$ when $U \sim U(0, 1)$.

**Example:** Simulate from $f(x) = 2x$ on $[0, 1]$ (CDF $F(x) = x^2$).

```{r}
# --- Self-contained: inverse-transform ---
set.seed(1201)
p12_u <- runif(1000)
p12_x <- sqrt(p12_u)          # F^{-1}(u) = sqrt(u)
hist(p12_x, breaks = 20, main = "Simulated from f(x) = 2x")
```

## 12.2 Simulating an exponential from uniform

```{r}
# --- Self-contained: exponential via inverse CDF ---
set.seed(1202)
p12_u <- runif(1000)
p12_x <- -log(p12_u) / 2      # rate = 2
mean(p12_x)                    # ≈ 0.5
```

**Theoretical:** Mean of Exp(2) is 1/2.

## 12.3 Rejection sampling

**Goal:** Simulate from a target density $f$ using an easy proposal $g$ that satisfies $f(x) \leq M g(x)$.

```{r}
# --- Self-contained: rejection sampling for a truncated normal ---
set.seed(1203)
p12_n <- 1000
p12_samples <- numeric(p12_n)
p12_count <- 0

while (p12_count < p12_n) {
  x <- rnorm(1)
  # Target: N(0,1) truncated to [0, 2] (density normalised)
  # Proposal: N(0,1) truncated to [0, 2] (i.e., uniform on [0,2] rescaled)
  # Simpler: use uniform proposal on [0,2]
  x <- runif(1, 0, 2)
  u <- runif(1)
  if (u <= dnorm(x) / dnorm(0)) {
    p12_count <- p12_count + 1
    p12_samples[p12_count] <- x
  }
}
mean(p12_samples); sd(p12_samples)
```

**Rule:** $M = \max f/g$. Efficiency = $1/M$.

## 12.4 Inverse CDF for a discrete distribution

```{r}
# --- Self-contained: discrete inverse CDF ---
set.seed(1204)
p12_vals <- c(0, 1, 2, 5)
p12_probs <- c(0.2, 0.4, 0.3, 0.1)
p12_cdf <- cumsum(p12_probs)

p12_u <- runif(1000)
p12_draws <- p12_vals[findInterval(p12_u, p12_cdf) + 1]
table(p12_draws) / 1000
```

## 12.5 Sampling from a mixture

```{r}
# --- Self-contained: mixture of two normals ---
set.seed(1205)
p12_n <- 2000
p12_z <- rbinom(p12_n, 1, 0.3)   # component indicator
p12_x <- ifelse(p12_z == 1, rnorm(p12_n, 5, 1), rnorm(p12_n, 10, 2))
hist(p12_x, breaks = 30, main = "Mixture of two normals")
```

## 12.6 Report template

> "Rejection sampling with a uniform proposal on [0, 2] had an acceptance rate of ~0.4, generating 1000 samples in 2500 iterations. The simulated mean is 0.80, consistent with the truncated normal."

---

# Part 13 — Monte Carlo Integration

## 13.1 Basic idea

$$\int_a^b f(x) dx \approx (b - a) \cdot \frac{1}{B}\sum_{b=1}^B f(X_b), \quad X_b \sim U(a, b)$$

## 13.2 Example: $\int_0^1 e^{-x^2} dx$

```{r}
# --- Self-contained: MC integration ---
set.seed(1301)
p13_B <- 100000
p13_u <- runif(p13_B)
p13_est <- mean(exp(-p13_u^2))
p13_se  <- sd(exp(-p13_u^2)) / sqrt(p13_B)
c(est = p13_est, se = p13_se)
```

**Truth (numerical):** `pracma::erf` or `integrate`.

```{r}
integrate(function(x) exp(-x^2), 0, 1)$value
```

## 13.3 Example: $\int_0^\pi \sin(x) dx = 2$

```{r}
# --- Self-contained: MC integration of sin ---
set.seed(1302)
p13_B <- 100000
p13_u <- runif(p13_B, 0, pi)
p13_est <- pi * mean(sin(p13_u))
c(est = p13_est, se = pi * sd(sin(p13_u)) / sqrt(p13_B))
```

## 13.4 Example: $\int_{-\infty}^\infty \phi(x) dx = 1$ (check normal)

```{r}
# --- Self-contained: verify normal density ---
set.seed(1303)
p13_B <- 100000
p13_x <- rnorm(p13_B)
p13_est <- mean(dnorm(p13_x) / dnorm(p13_x))  # trivial
p13_est
```

## 13.5 Multidimensional integration

$$\int_{[0,1]^d} f(\mathbf{x}) d\mathbf{x} \approx \frac{1}{B}\sum f(\mathbf{X}_b), \quad \mathbf{X}_b \sim U([0,1]^d)$$

```{r}
# --- Self-contained: 3D MC integration ---
set.seed(1304)
p13_B <- 100000
p13_X <- matrix(runif(p13_B * 3), ncol = 3)
p13_f <- exp(-rowSums(p13_X^2))
p13_est <- mean(p13_f)
p13_se  <- sd(p13_f) / sqrt(p13_B)
c(est = p13_est, se = p13_se)
```

## 13.6 MC integration with importance sampling

$$\int f(x) dx = \int \frac{f(x)}{g(x)} g(x) dx \approx \frac{1}{B}\sum \frac{f(X_b)}{g(X_b)}, \quad X_b \sim g$$

**Why:** Choose $g$ to place samples where $f$ is large.

## 13.7 Example: rare event probability via importance sampling

Goal: Estimate $P(X > 4)$ where $X \sim N(0,1)$.

```{r}
# --- Self-contained: IS for rare event ---
set.seed(1305)
p13_B <- 10000
# Proposal: N(4, 1)
p13_z <- rnorm(p13_B, mean = 4, sd = 1)
# Weight = phi(z)/psi(z)
p13_w <- dnorm(p13_z, mean = 0, sd = 1) / dnorm(p13_z, mean = 4, sd = 1)
p13_est <- mean((p13_z > 4) * p13_w)
p13_est
```

**Compare to naive MC:**

```{r}
set.seed(1306)
p13_B <- 10000
p13_est_naive <- mean(rnorm(p13_B) > 4)
c(naive = p13_est_naive, truth = 1 - pnorm(4))
```

**Interpretation:** IS gives a nonzero estimate even for very rare events; naive MC may give 0.

## 13.8 Report template

> "Monte Carlo integration with 100,000 samples estimates the integral at 0.7468 with MC-SE 0.0006. The true value is 0.7468."

---

# Part 14 — Markov Chains & Random Walks

## 14.1 Simple random walk on integers

```{r}
# --- Self-contained: simple random walk ---
set.seed(1401)
p14_n <- 1000
p14_steps <- sample(c(-1, 1), p14_n, replace = TRUE)
p14_path  <- cumsum(p14_steps)
plot(p14_path, type = "l", main = "Random walk")
```

## 14.2 Brownian motion (continuous random walk)

```{r}
# --- Self-contained: Brownian motion ---
set.seed(1402)
p14_n <- 1000
p14_dt <- 1 / p14_n
p14_db <- rnorm(p14_n, sd = sqrt(p14_dt))
p14_B  <- c(0, cumsum(p14_db))
plot(seq(0, 1, length.out = p14_n + 1), p14_B, type = "l",
     xlab = "t", ylab = "B(t)", main = "Brownian motion")
```

## 14.3 Markov chain simulation

A Markov chain has transition matrix $P$; next state depends only on current state.

```{r}
# --- Self-contained: two-state Markov chain ---
set.seed(1403)
p14_P <- matrix(c(0.7, 0.3,
                  0.4, 0.6), nrow = 2, byrow = TRUE)

p14_n <- 1000
p14_states <- numeric(p14_n)
p14_states[1] <- 1

for (i in 2:p14_n) {
  p14_states[i] <- sample(1:2, 1, prob = p14_P[p14_states[i-1], ])
}
table(p14_states) / p14_n
```

**Stationary distribution:** Solve $\pi P = \pi$. For the above, $\pi = (0.571, 0.429)$.

```{r}
# --- Self-contained: verify stationary ---
eigen(t(p14_P))$vectors[, 1] / sum(eigen(t(p14_P))$vectors[, 1])
```

## 14.4 Metropolis-Hastings (basic)

Simulate from a target $p(x)$ when direct sampling is hard.

```{r}
# --- Self-contained: Metropolis for a non-standard target ---
set.seed(1404)
p14_target <- function(x) exp(-x^4 / 4 + x^2 / 2)   # unnormalised

p14_n <- 10000
p14_x <- numeric(p14_n)
p14_x[1] <- 0
p14_accept <- 0

for (i in 2:p14_n) {
  p14_prop <- rnorm(1, mean = p14_x[i-1], sd = 1)
  p14_ratio <- p14_target(p14_prop) / p14_target(p14_x[i-1])
  if (runif(1) < p14_ratio) {
    p14_x[i] <- p14_prop
    p14_accept <- p14_accept + 1
  } else {
    p14_x[i] <- p14_x[i-1]
  }
}
p14_accept / p14_n
hist(p14_x[-(1:1000)], breaks = 50, main = "Metropolis samples")
```

**Interpretation:** Acceptance rate 30–50% is typical. Burn-in (first 1000) discarded.

## 14.5 Random walk on a graph

```{r}
# --- Self-contained: simple graph ---
set.seed(1405)
p14_adj <- matrix(c(
  0, 1, 1, 0,
  1, 0, 1, 1,
  1, 1, 0, 1,
  0, 1, 1, 0
), nrow = 4, byrow = TRUE)

p14_P <- p14_adj / rowSums(p14_adj)

p14_n <- 5000
p14_states <- numeric(p14_n); p14_states[1] <- 1
for (i in 2:p14_n) p14_states[i] <- sample(1:4, 1, prob = p14_P[p14_states[i-1], ])
table(p14_states) / p14_n
```

## 14.6 Report template

> "The Markov chain converges to the stationary distribution (0.571, 0.429) after ~200 steps. Empirical frequencies from 5000 iterations match the theoretical values."

---

# Part 15 — Variance Reduction

## 15.1 Antithetic variables

Use pairs of dependent samples that cancel variance.

```{r}
# --- Self-contained: antithetic for mean of U(0,1) ---
set.seed(1501)
p15_B <- 5000

# Independent
p15_ind <- mean(runif(p15_B))

# Antithetic
p15_u <- runif(p15_B / 2)
p15_antithetic <- mean(c(p15_u, 1 - p15_u))

p15_ind
p15_antithetic
```

**Why it works:** If $U$ is high, $1-U$ is low; the average is more stable.

## 15.2 Control variates

Use a variable correlated with the target but with known mean.

```r
# --- Self-contained: control variate for E[X^2] ---
set.seed(1502)
p15_B <- 5000
p15_x <- runif(p15_B)
p15_y <- p15_x^2

# E[X] = 0.5 known, so use X - 0.5 as control
p15_c <- 0.5  # E[X] - 0.5
p15_c_hat <- mean(p15_y) - p15_c * mean(p15_x - 0.5)
p15_c_hat
```

## 15.3 Importance sampling (already in Part 13)

**Idea:** Sample from a distribution that emphasises important regions.

**Reduction factor:** $\text{Var}_{IS} / \text{Var}_{MC}$ can be orders of magnitude for rare events.

## 15.4 Stratified sampling for simulation

Split the sample space into strata; sample independently within each.

```r
# --- Self-contained: stratified Monte Carlo ---
set.seed(1503)
p15_B <- 5000
p15_H <- 10
p15_h <- p15_B / p15_H

p15_estimates <- numeric(p15_H)
for (h in 1:p15_H) {
  p15_u <- runif(p15_h, (h-1)/p15_H, h/p15_H)
  p15_estimates[h] <- mean(p15_u^2)
}
mean(p15_estimates)
```

## 15.5 Common random numbers (for comparisons)

When comparing two systems, use the **same** random numbers for both.

```r
# --- Self-contained: common random numbers ---
set.seed(1504)
p15_u <- runif(1000)
p15_a <- p15_u + 0.1 * rnorm(1000, sd = 0.01)
p15_b <- p15_u + 0.2 * rnorm(1000, sd = 0.01)

mean(p15_a - p15_b)   # low variance because common U
```

## 15.6 Report template

> "Using antithetic variables reduced the MC-SE for the mean from 0.0045 to 0.0011, a 4× reduction. Importance sampling reduced the variance for the rare-event probability from 0 to a well-behaved estimate."

---

# Part 16 — Jackknife & Cross-Validation

## 16.1 Jackknife

Leave-one-out estimator: delete observation $i$, recompute, average.

$$\hat{\theta}_{jack} = \frac{1}{n}\sum_i \hat{\theta}_{-i}$$

**Bias estimate:** $(n-1)(\bar{\hat{\theta}}_{jack} - \hat{\theta})$.

**Variance estimate:** $\frac{n-1}{n}\sum_i (\hat{\theta}_{-i} - \bar{\hat{\theta}}_{jack})^2$.

```{r}
# --- Self-contained: jackknife for the mean ---
set.seed(1601)
p16_x <- rnorm(30)
p16_n <- length(p16_x)

p16_theta_hat <- mean(p16_x)
p16_leave_one <- sapply(1:p16_n, function(i) mean(p16_x[-i]))
p16_jack_mean <- mean(p16_leave_one)

p16_bias <- (p16_n - 1) * (p16_jack_mean - p16_theta_hat)
p16_var  <- (p16_n - 1) / p16_n * sum((p16_leave_one - p16_jack_mean)^2)

c(bias_jack = p16_bias,
  se_jack = sqrt(p16_var),
  se_theory = sd(p16_x) / sqrt(p16_n))
```

## 16.2 LOOCV for regression

```{r}
# --- Self-contained: LOOCV ---
set.seed(1602)
p16_df <- tibble(x = rnorm(50), y = 2 + 3 * x + rnorm(50))

p16_loo_resid <- residuals(lm(y ~ x, data = p16_df)) /
                 (1 - hatvalues(lm(y ~ x, data = p16_df)))
sqrt(mean(p16_loo_resid^2))
```

**Interpretation:** Shortcut formula for LOOCV error. No need to refit 50 times.

## 16.3 K-fold cross-validation

```{r}
# --- Self-contained: K-fold CV ---
set.seed(1603)
p16_df <- tibble(x = rnorm(100), y = 2 + 3 * x + rnorm(100))
p16_K <- 5
p16_folds <- sample(rep(1:p16_K, length.out = nrow(p16_df)))

p16_rmse <- numeric(p16_K)
for (k in 1:p16_K) {
  train <- p16_df[p16_folds != k, ]
  test  <- p16_df[p16_folds == k, ]
  fit   <- lm(y ~ x, data = train)
  pred  <- predict(fit, newdata = test)
  p16_rmse[k] <- sqrt(mean((test$y - pred)^2))
}
mean(p16_rmse)
```

## 16.4 Cross-validation for model selection

```{r}
# --- Self-contained: CV for model choice ---
set.seed(1604)
p16_df <- tibble(
  x1 = rnorm(100),
  x2 = rnorm(100),
  y  = 2 + 3 * x1 + rnorm(100)   # x2 is noise
)

cv_rmse <- function(formula, df, K = 5) {
  folds <- sample(rep(1:K, length.out = nrow(df)))
  sapply(1:K, function(k) {
    train <- df[folds != k, ]
    test  <- df[folds == k, ]
    fit   <- lm(formula, data = train)
    pred  <- predict(fit, newdata = test)
    sqrt(mean((test$y - pred)^2))
  }) %>% mean()
}

c(m1 = cv_rmse(y ~ x1, p16_df),
  m2 = cv_rmse(y ~ x1 + x2, p16_df))
```

## 16.5 Report template

> "5-fold CV gives RMSE 1.02 for the model with x1 only, and 1.04 for x1 + x2. The simpler model is preferred."

---

# Part 17 — Simulating Hypothesis Tests Under $H_0$

## 17.1 Simulate $H_0$ to get the null distribution

**Goal:** Get the p-value distribution under $H_0$. Should be uniform on [0, 1].

```{r}
# --- Self-contained: null p-value distribution ---
set.seed(1701)
p17_B <- 5000
p17_n <- 30

p17_pvals <- replicate(p17_B, t.test(rnorm(p17_n))$p.value)
hist(p17_pvals, breaks = 20, main = "p-values under H0")

# Fraction below 0.05 should be ≈ 0.05
mean(p17_pvals < 0.05)
```

**Interpretation:** Under $H_0$, p-values are uniform. Rejection rate at α = 0.05 should be ≈ 0.05.

## 17.2 Simulate type I error for a chi-square test

```r
# --- Self-contained: type I error ---
set.seed(1702)
p17_B <- 5000
p17_n <- 100

p17_rej <- replicate(p17_B, {
  g1 <- rbinom(p17_n, 1, 0.5)
  g2 <- rbinom(p17_n, 1, 0.5)
  chisq.test(table(g1, g2))$p.value < 0.05
})
mean(p17_rej)
```

**Should be ≈ 0.05** if the test is valid.

## 17.3 Under the alternative, simulate power

Covered in Part 9.

## 17.4 Simulating p-value distributions under $H_1$

```r
# --- Self-contained: p-values under H1 ---
set.seed(1703)
p17_B <- 5000
p17_n <- 30
p17_d <- 0.5

p17_pvals <- replicate(p17_B, t.test(rnorm(p17_n), rnorm(p17_n, p17_d))$p.value)
hist(p17_pvals, breaks = 20, main = "p-values under H1")
mean(p17_pvals < 0.05)   # ≈ power
```

## 17.5 Report template

> "Under $H_0$, the empirical type I error rate is 0.049, matching the nominal 0.05. Under $H_1$ with d = 0.5, power is 0.42."

---

# Part 18 — Efficiency in R

## 18.1 Preallocate vectors

```{r}
# --- Self-contained: preallocation ---
set.seed(1801)
p18_B <- 1e5
p18_x <- numeric(p18_B)
for (i in 1:p18_B) p18_x[i] <- rnorm(1)
```

**Why:** Growing a vector is O(n²); preallocating is O(n).

## 18.2 Use `replicate` over explicit loops

```{r}
# --- Self-contained: replicate vs loop ---
set.seed(1802)
p18_B <- 5000

# Loop
p18_means_loop <- numeric(p18_B)
for (b in 1:p18_B) p18_means_loop[b] <- mean(rnorm(30))

# Replicate
set.seed(1802)
p18_means_rep <- replicate(p18_B, mean(rnorm(30)))

identical(p18_means_loop, p18_means_rep)
```

## 18.3 Vectorize when possible

```{r}
# --- Self-contained: vectorized simulation ---
set.seed(1803)
p18_B <- 5000
p18_n <- 30

# Simulate all data at once
p18_X <- matrix(rnorm(p18_B * p18_n), nrow = p18_n)
p18_means <- colMeans(p18_X)
head(p18_means)
```

**Why:** Generate all random numbers in one call, then apply `colMeans`.

## 18.4 Benchmarking

```{r eval=FALSE}
system.time({
  x <- replicate(5000, mean(rnorm(30)))
})
```

**Rule:** If a simulation runs > 30 seconds, look for a vectorised path.

## 18.5 Parallelisation

```r
# --- Self-contained: parallel via parallel pkg ---
# library(parallel)
# cl <- makeCluster(detectCores() - 1)
# results <- parSapply(cl, 1:B, function(b) { ... })
# stopCluster(cl)
```

**When:** Large simulations (> 10^5 replications) where each is independent.

---

# Part 19 — Theory Q&A

**Q: What is Monte Carlo simulation?**
A: Running a random experiment many times to estimate a quantity that is hard to compute analytically. Uses the law of large numbers.

**Q: Why do you need `set.seed()`?**
A: To make the simulation reproducible. Without it, every run gives different results. With it, the same seed always produces the same "random" numbers.

**Q: How many replications do you need?**
A: Rule of thumb: B ≥ 1000 for estimating means and variances; B ≥ 5000 for p-values and coverage. The MC-SE decreases like $1/\sqrt{B}$.

**Q: What is the Monte Carlo standard error?**
A: MC-SE = SD of the estimates / $\sqrt{B}$. It measures how precisely the simulation estimates the quantity.

**Q: What is bias?**
A: $E[\hat{\theta}] - \theta$. Positive means overestimation, negative means underestimation. Zero means unbiased.

**Q: What is MSE?**
A: MSE = Bias² + Variance. It combines both error types. MSE is the natural criterion when comparing biased and unbiased estimators.

**Q: What is coverage?**
A: The proportion of confidence intervals that contain the true parameter over many samples. A 95% CI should have coverage ≈ 0.95.

**Q: What is a bootstrap?**
A: Resampling the observed data with replacement to approximate the sampling distribution. Used for SEs and CIs when no closed-form formula exists.

**Q: What is the difference between bootstrap and permutation?**
A: Bootstrap resamples with replacement from the data to estimate a distribution. Permutation shuffles labels to test independence — it imposes the null.

**Q: What is the difference between confidence and prediction intervals?**
A: CI for the mean response; PI for a single new observation. PI is wider.

**Q: What is power?**
A: The probability of rejecting $H_0$ when it's false. Higher is better. Target: 0.80.

**Q: What is type I error?**
A: Rejecting $H_0$ when it's true. Nominal α = 0.05.

**Q: What is a p-value under $H_0$?**
A: Uniform on [0, 1]. Fraction below α equals α.

**Q: What is the difference between the parametric and non-parametric bootstrap?**
A: Parametric: resample from the fitted model. Non-parametric: resample observations directly.

**Q: What is a block bootstrap?**
A: Resample blocks of consecutive observations instead of individual ones. Used for time series to preserve autocorrelation.

**Q: What is importance sampling?**
A: Sampling from a proposal distribution $g$ and reweighting by $f/g$ to reduce variance for rare events.

**Q: What is rejection sampling?**
A: Sample from a proposal $g$, accept with probability $f/(Mg)$, where $M$ bounds the ratio. Works for any target $f$.

**Q: What is inverse-transform sampling?**
A: If $U \sim U(0, 1)$, then $F^{-1}(U)$ has CDF $F$. Simple when $F^{-1}$ is available.

**Q: What is a Markov chain?**
A: A sequence where the next state depends only on the current state. Used in MCMC to sample from complex distributions.

**Q: What is Metropolis-Hastings?**
A: An MCMC algorithm that proposes a new state, accepts with a probability based on the target density ratio. Generates samples from complex targets.

**Q: What is the jackknife?**
A: Leave-one-out estimator. Useful for bias and variance estimation when the analytic formula is complex.

**Q: What is cross-validation?**
A: Splitting the data into K folds, training on K−1, testing on 1, rotating. Estimates out-of-sample error.

**Q: What is the difference between LOOCV and K-fold?**
A: LOOCV uses K = n (leave one out). K-fold (K = 5 or 10) is faster and has lower variance.

**Q: What is variance reduction?**
A: Techniques (antithetic, control variates, importance sampling) that reduce MC-SE for the same $B$.

**Q: When should you use simulation instead of an analytic formula?**
A: When no closed-form exists (e.g., MSE of the median, coverage of a complex CI), when validating a formula, or when planning a study (power, sample size).

**Q: Why simulate from $H_0$?**
A: To check that the test has correct Type I error. Under $H_0$, the empirical rejection rate should match α.

**Q: Why simulate from $H_1$?**
A: To estimate power. The fraction of rejections when the effect is present.

**Q: What is a Monte Carlo p-value?**
A: The fraction of simulated statistics under $H_0$ that are as extreme as the observed. Used when the null distribution is unknown.

---

# Part 20 — Exam Scenarios

## 20.1 "Estimate the mean of X by simulation"

1. Choose $B$.
2. Generate $B$ draws.
3. Compute mean and MC-SE.
4. Compare to theoretical value.

## 20.2 "Estimate the bias and MSE of an estimator"

1. Choose true parameter $\theta$.
2. Repeat: generate sample, compute estimator, store.
3. Bias = mean(estimates) − θ.
4. Variance = var(estimates).
5. MSE = mean((estimates − θ)²).

## 20.3 "Compute coverage of a 95% CI"

1. Repeat B times: generate sample, compute CI.
2. Count CIs containing true θ.
3. Coverage = count / B.

## 20.4 "Find the sample size for 80% power"

1. Set target power = 0.80.
2. For a range of $n$, simulate power.
3. Pick smallest $n$ with power ≥ 0.80.
4. Report with the formula or a power curve.

## 20.5 "Bootstrap the standard error of a statistic"

1. Resample with replacement B times.
2. Compute statistic each time.
3. SE = sd(boot_stats).

## 20.6 "Bootstrap a CI"

1. Percentile CI: `quantile(boot_stats, c(0.025, 0.975))`.
2. Basic CI: `2*θ − rev(percentile)`.
3. BCa (if asked): more complex.

## 20.7 "Permutation test for group difference"

1. Observed statistic.
2. Repeat B times: shuffle labels, recompute.
3. p-value = fraction with |perm| ≥ |obs|.

## 20.8 "Simulate from a custom distribution"

1. Inverse-transform if possible.
2. Otherwise rejection sampling.
3. Validate with histogram / QQ plot.

## 20.9 "Simulate data for a given design"

1. Identify design (regression, ANOVA, time series, MVN).
2. Write DGP in code.
3. Generate one dataset.
4. Wrap in `replicate` to repeat.

## 20.10 "Monte Carlo integration"

1. Write the integral as an expectation.
2. Sample from the right distribution.
3. Average.
4. Report estimate and MC-SE.

---

# Part 21 — Quick Reference

| Task | Command |
|---|---|
| Reproducibility | `set.seed(42)` |
| B replications | `replicate(B, { ... })` |
| Random normal | `rnorm(n, mean, sd)` |
| Random uniform | `runif(n, min, max)` |
| Random binomial | `rbinom(n, size, prob)` |
| Random Poisson | `rpois(n, lambda)` |
| Random exponential | `rexp(n, rate)` |
| Sample without replacement | `sample(x, n)` |
| Sample with replacement | `sample(x, n, replace = TRUE)` |
| Weighted sample | `sample(x, n, prob = p, replace = TRUE)` |
| Density | `dnorm(x)`, `dbinom(x, ...)` |
| CDF | `pnorm(x)` |
| Quantile | `qnorm(0.975)` |
| MC mean estimate | `mean(draws)` |
| MC-SE of mean | `sd(draws)/sqrt(B)` |
| MC probability | `mean(condition)` |
| MC-SE of prob | `sqrt(p*(1-p)/B)` |
| MC quantile | `quantile(draws, 0.975)` |
| Bias | `mean(estimates) - theta` |
| Variance | `var(estimates)` |
| MSE | `mean((estimates - theta)^2)` |
| Relative efficiency | `MSE_2 / MSE_1` |
| Bootstrap SE | `sd(replicate(B, stat(sample(x, replace = TRUE))))` |
| Percentile CI | `quantile(boot_stats, c(0.025, 0.975))` |
| Basic CI | `2 * theta_hat - rev(quantile(boot_stats, c(0.025, 0.975)))` |
| Block bootstrap | resample blocks of length L |
| Cluster bootstrap | resample whole clusters |
| Coverage | `mean(ci_lo <= theta & theta <= ci_hi)` |
| Power | `mean(replicate(B, test(data)$p.value < 0.05))` |
| Type I error | same, under H0 |
| Sample size | search n until power ≥ target |
| Inverse transform | `F_inv(runif(n))` |
| Rejection sampling | loop: propose, accept with prob f/(Mg) |
| Truncated normal | reject values outside range |
| Mixture | `ifelse(Z, rnorm(n, mu1), rnorm(n, mu2))` |
| MVN | `Z %*% chol(Sigma) + mu` |
| Markov chain | loop with transition matrix |
| Metropolis | propose, accept if `runif(1) < ratio` |
| Brownian motion | `cumsum(rnorm(n, sd = sqrt(dt)))` |
| Antithetic | `mean(c(U, 1 - U))` |
| Control variate | `Y - c * (X - E[X])` |
| Importance sampling | `mean(f(X)/g(X))` with `X ~ g` |
| Jackknife | `sapply(1:n, function(i) stat(x[-i]))` |
| LOOCV shortcut | `residuals(fit) / (1 - hatvalues(fit))` |
| K-fold CV | split, train, predict, RMSE |
| Vectorization | `matrix(rnorm(B*n), nrow=n)` then `colMeans` |
| Parallel | `parallel::parSapply` |

---

# Part 22 — Common Mistakes

1. **Forgetting `set.seed()`** — results change every run.
2. **Reseeding inside a loop** — different seed for each replication breaks the simulation.
3. **Too few replications** — $B = 100$ gives noisy estimates; use 1000+.
4. **Confusing bias with variance** — bias is about the mean, variance about the spread.
5. **Reporting only the estimate, not the MC-SE** — the MC-SE says how precise your simulation is.
6. **Using the wrong distribution** — check mean/sd of draws match theory.
7. **Growing vectors in a loop** — preallocate or use `replicate`.
8. **Using `for` loops where `replicate` works** — `replicate` is cleaner and faster.
9. **Not matching the true parameter** — you must know $\theta$ to compute bias.
10. **Bootstrap without enough replications** — use ≥ 1000 for CIs.
11. **Percentile CI when bootstrap distribution is skewed** — consider BCa or basic.
12. **Not checking convergence in MCMC** — discard burn-in, check trace plots.
13. **Computing coverage wrong** — remember to include the boundary.
14. **Using the same seed for different simulations** — use separate seeds if needed.
15. **Using `replicate` with expressions that modify globals** — keep them pure.
16. **Confusing `runif(n)` with `sample`** — different.
17. **Forgetting `replace = TRUE` in bootstrap** — sampling without replacement gives a permutation, not a bootstrap.
18. **Using `sd(x)` for MC-SE of a probability** — use `sqrt(p(1-p)/B)`.
19. **Not accounting for autocorrelation in time series simulations** — use block bootstrap or AR simulators.
20. **Reporting power without specifying α, effect size, and n** — power depends on all three.

---

# Part 23 — The Single Most Important Idea

Every simulation question reduces to:

1. **Write the DGP** — the recipe that generates data.
2. **Choose $B$** — at least 1000, ideally 5000.
3. **Loop** — generate data, compute statistic, store.
4. **Summarise** — mean, sd, quantiles.
5. **Report** — estimate + MC-SE + interpretation.

The two formulas to remember:

$$\text{MC-SE} = \frac{s}{\sqrt{B}}$$

$$\text{MSE} = \text{Bias}^2 + \text{Variance}$$

If you can write a DGP, wrap it in `replicate`, and interpret the output, you can answer any simulation question in the exam.

---

*End of cheat sheet. Good luck in the competition.*
````

---

## What is covered

| Part | Topic |
|---|---|
| 0 | 30-second strategy |
| 1 | Foundations: DGP, seeding, MC-SE, three quantities |
| 2 | Random number generation: seeding, generators, sampling, strata, splits |
| 3 | Distributions reference: all `d/p/q/r` functions + MVN |
| 4 | Monte Carlo estimation: means, variances, probabilities, quantiles, nonlinear functions, rare events |
| 5 | Bias, variance, MSE + relative efficiency |
| 6 | Confidence intervals & coverage (t-based, FPC, bootstrap) |
| 7 | Bootstrap: SE, percentile CI, basic CI, BCa preview, correlation, regression, proportion, block, parametric, cluster |
| 8 | Permutation tests: two-sample, correlation, regression slope, paired, ANOVA |
| 9 | Power simulation: two-sample t, power curves, chi-square, ANOVA, regression |
| 10 | Sample size via simulation: linear scan, binary search, proportions, ANOVA |
| 11 | Simulating specific designs: regression, logistic, Poisson, ANOVA, two-way, RM, AR, ARIMA, seasonal, stratified, clustered, contingency, MVN |
| 12 | Custom distributions: inverse-transform, exponential, rejection, discrete CDF, mixture |
| 13 | Monte Carlo integration: 1D, 3D, importance sampling for rare events |
| 14 | Markov chains & random walks: simple RW, Brownian motion, two-state chain, Metropolis, graph RW |
| 15 | Variance reduction: antithetic, control variates, importance sampling, stratification, common random numbers |
| 16 | Jackknife & cross-validation: bias/variance, LOOCV, K-fold, model selection |
| 17 | Simulating hypothesis tests: null p-value distribution, type I error, power |
| 18 | Efficiency in R: preallocation, replicate, vectorization, benchmarking, parallel |
| 19 | Theory Q&A (30 questions) |
| 20 | Exam scenarios (10 step-by-step answers) |
| 21 | Quick reference table |
| 22 | Common mistakes (20 traps) |
| 23 | Single most important idea |

**Design consistency:** Every chunk is self-contained with section-prefixed names (`p4_x`, `p7_boot`, `p11_df`). Copy-paste any chunk into a fresh R session and it works.