---
title: "Time Series Cheat Sheet"
author: "Parsa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(forecast)
library(knitr)
```

# Part 0 — How to identify the time series question

| Question | What to do |
|---|---|
| "Plot the series" | `plot(ts_obj)` or `autoplot()` |
| "Discuss the series" | Trend, seasonality, stationarity, outliers |
| "Plot ACF and PACF" | `acf()`, `pacf()` |
| "Suggest a model" | Use ACF/PACF pattern (Part 4) |
| "Fit a model" | `Arima()` or `auto.arima()` |
| "Is the model adequate?" | `checkresiduals()` or Ljung-Box |
| "Forecast h steps" | `forecast(fit, h)` |
| "Missing values" | `forecast` handles NA automatically |

---

# Part 1 — Create a time series object

**Why:** Most functions (`acf`, `Arima`, `forecast`) need a `ts` object, not a plain vector.

```{r}
set.seed(1)
y <- arima.sim(model = list(ar = 0.7), n = 100) + 10
ts_y <- ts(y, start = 1968, frequency = 1)

class(ts_y)
head(ts_y, 5)
```

**What the arguments mean:**

- `start`: first time index (year, month, etc.)
- `frequency`: 1 = annual, 4 = quarterly, 12 = monthly

For monthly data starting in January 2000:

```{r}
ts_m <- ts(rnorm(60), start = c(2000, 1), frequency = 12)
```

---

# Part 2 — Plot the series

```{r}
autoplot(ts_y) +
  labs(x = "Time", y = "Value", title = "Time series") +
  theme_minimal()
```

**Base R alternative:**

```{r}
plot(ts_y, main = "Time series", xlab = "Time", ylab = "Value")
```

**What to look for:**

- Trend: upward or downward drift over time.
- Seasonality: repeating pattern within a year.
- Changing variance: variance grows or shrinks over time.
- Outliers: spikes that stand out.

**Report template:** "The series shows an upward trend with no obvious seasonality. Variance appears stable."

---

# Part 3 — Stationarity and ACF/PACF

## 3.1 ACF and PACF

```{r}
par(mfrow = c(1, 2))
acf(ts_y, main = "ACF")
pacf(ts_y, main = "PACF")
par(mfrow = c(1, 1))
```

**What ACF shows:** Correlation between y_t and y_{t-k} for each lag k.

**What PACF shows:** Same, but removing the effect of intermediate lags.

## 3.2 Identify the model from ACF/PACF

| ACF pattern | PACF pattern | Model |
|---|---|---|
| Decays gradually | Cuts off after lag p | AR(p) |
| Cuts off after lag q | Decays gradually | MA(q) |
| Both decay gradually | Both decay gradually | ARMA(p, q) |
| Decays very slowly | Decays very slowly | Non-stationary (difference) |

**Example rules:**

- PACF significant at lag 1 only, ACF decays → AR(1).
- ACF significant at lag 1 only, PACF decays → MA(1).
- Both decay → ARMA(1, 1).

## 3.3 Check stationarity

**Augmented Dickey-Fuller test:**

```{r}
tseries::adf.test(ts_y)
```

- p < 0.05: series is stationary.
- p > 0.05: series is non-stationary. Difference it.

**Differencing:**

```{r}
dy <- diff(ts_y)
plot(dy, main = "First difference")
```

---

# Part 4 — Fit ARIMA models

## 4.1 Automatic selection

```{r}
fit_auto <- auto.arima(ts_y)
summary(fit_auto)
```

**What `auto.arima` does:** Tries combinations of AR and MA orders, differences, and seasonality, and picks the best by AIC.

**Output gives:**

- ARIMA(p, d, q) order.
- Coefficients with SE and p-values.
- AIC, BIC, log-likelihood.
- sigma² (residual variance).

## 4.2 Manual selection

```{r}
fit_ar1 <- Arima(ts_y, order = c(1, 0, 0))   # AR(1)
fit_ma1 <- Arima(ts_y, order = c(0, 0, 1))   # MA(1)
fit_arma <- Arima(ts_y, order = c(1, 0, 1))  # ARMA(1,1)
```

**Read the coefficients:**

```{r}
coef(fit_ar1)
```

For AR(1): the AR coefficient is rho. For MA(1): the MA coefficient is theta.

## 4.3 Compare models

```{r}
tibble(
  Model = c("AR(1)", "MA(1)", "ARMA(1,1)"),
  AIC   = c(AIC(fit_ar1), AIC(fit_ma1), AIC(fit_arma)),
  BIC   = c(BIC(fit_ar1), BIC(fit_ma1), BIC(fit_arma))
) %>% kable(digits = 3)
```

**Interpretation:** Lower AIC/BIC wins. AIC penalizes complexity less than BIC.

---

# Part 5 — AR(1) as regression

**This is exam Q21-4a and Q21-4b.**

An AR(1) model can be fitted as a simple regression of y_t on y_{t-1}.

```{r}
set.seed(2)
y2 <- arima.sim(model = list(ar = 0.6), n = 200) + 5
ts_y2 <- ts(y2)

# Build lagged data frame
df_ar <- tibble(
  y  = as.numeric(ts_y2),
  y1 = lag(y)
) %>% drop_na()

# Fit as regression
fit_lm <- lm(y ~ y1, data = df_ar)
summary(fit_lm)
```

**What the output gives:**

- Intercept = alpha (the constant term).
- Slope = rho (the autoregressive coefficient).
- Residual standard error squared = sigma².

**Report:** y_t = alpha + rho * y_{t-1} + e_t.

## 5.1 Scatterplot of Y_t vs Y_{t-1}

```{r}
ggplot(df_ar, aes(y1, y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```

**What to look for:** A roughly linear relationship confirms AR(1) is plausible.

## 5.2 Compare regression with MLE

```{r}
fit_arima <- Arima(ts_y2, order = c(1, 0, 0), include.mean = TRUE)
coef(fit_arima)
coef(fit_lm)
```

**Interpretation:** Regression estimates approximate MLE estimates. MLE is more efficient for large n. Report both.

---

# Part 6 — Log-likelihood for AR(1)

**This is exam Q21-4c.**

For AR(1) with normal errors:

- Y_1 ~ N(alpha/(1-rho), sigma²/(1-rho²)).
- Y_t given Y_{t-1} ~ N(alpha + rho*Y_{t-1}, sigma²).

```{r}
ar1_loglik <- function(par, y) {
  alpha  <- par[1]
  rho    <- par[2]
  sigma2 <- par[3]
  if (sigma2 <= 0 || abs(rho) >= 1) return(1e10)
  n <- length(y)

  mu1  <- alpha / (1 - rho)
  var1 <- sigma2 / (1 - rho^2)
  ll <- dnorm(y[1], mu1, sqrt(var1), log = TRUE)

  for (t in 2:n) {
    mu_t <- alpha + rho * y[t - 1]
    ll <- ll + dnorm(y[t], mu_t, sqrt(sigma2), log = TRUE)
  }
  return(-ll)
}

opt <- optim(c(1, 0.5, 1), ar1_loglik, y = as.numeric(ts_y2),
             method = "Nelder-Mead")
opt$par
```

**What happens:** Minimizes the negative log-likelihood. Returns MLE of alpha, rho, sigma².

**Compare with regression:**

```{r}
coef(fit_lm)
```

For large n, MLE and regression estimates converge.

---

# Part 7 — Model diagnostics

## 7.1 Residual check

```{r}
checkresiduals(fit_ar1)
```

Produces three plots:

- Residuals over time.
- ACF of residuals.
- Histogram of residuals.

And a Ljung-Box test.

## 7.2 Ljung-Box test

```{r}
Box.test(residuals(fit_ar1), lag = 10, type = "Ljung-Box")
```

**Interpretation:**

- p > 0.05: residuals are white noise. Model is adequate.
- p < 0.05: residuals still have structure. Model is inadequate.

## 7.3 Normality of residuals

```{r}
shapiro.test(residuals(fit_ar1))
```

p > 0.05: residuals are normal.

## 7.4 Summary of adequacy

**Report:** "Ljung-Box p = ..., residuals show no significant autocorrelation. The model is adequate."

---

# Part 8 — Forecasting

## 8.1 Point forecast with prediction intervals

**This is exam Q17-1b, 18-3d, 23-1d.**

```{r}
fc <- forecast(fit_ar1, h = 10)
fc
autoplot(fc) + theme_minimal()
```

**Output columns:**

- Point Forecast: best guess.
- Lo 80, Hi 80: 80% prediction interval.
- Lo 95, Hi 95: 95% prediction interval.

**Prediction intervals get wider** as you forecast further because uncertainty compounds.

## 8.2 Forecast from auto.arima

```{r}
fc2 <- forecast(fit_auto, h = 10)
autoplot(fc2) + theme_minimal()
```

## 8.3 Plot history plus forecast

```{r}
autoplot(fc, include = 100) +
  labs(title = "Forecast with prediction intervals") +
  theme_minimal()
```

**What `include` does:** Shows the last N observations of history before the forecast.

---

# Part 9 — Missing values

**This is exam Q17-1c.**

## 9.1 Set values to NA

```{r}
y_miss <- as.numeric(ts_y)
y_miss[92]  <- NA
y_miss[175] <- NA
ts_miss <- ts(y_miss)
```

## 9.2 Fit ARIMA ignoring missing values

`Arima` and `auto.arima` handle NA automatically.

```{r}
fit_miss <- auto.arima(ts_miss)
summary(fit_miss)
```

## 9.3 Forecast

```{r}
fc_miss <- forecast(fit_miss, h = 10)
autoplot(fc_miss) + theme_minimal()
```

**Interpretation:** The Kalman filter inside `Arima` skips missing observations and estimates coefficients from the rest.

---

# Part 10 — Choosing the best model

## 10.1 Compare multiple models

```{r}
fit_a <- Arima(ts_y, order = c(1, 0, 0))
fit_b <- Arima(ts_y, order = c(0, 0, 1))
fit_c <- Arima(ts_y, order = c(2, 0, 0))
fit_d <- Arima(ts_y, order = c(1, 1, 1))

tibble(
  Model = c("AR(1)", "MA(1)", "AR(2)", "ARIMA(1,1,1)"),
  AIC   = round(c(AIC(fit_a), AIC(fit_b), AIC(fit_c), AIC(fit_d)), 3),
  BIC   = round(c(BIC(fit_a), BIC(fit_b), BIC(fit_c), BIC(fit_d)), 3),
  logLik = round(c(logLik(fit_a), logLik(fit_b),
                   logLik(fit_c), logLik(fit_d)), 3)
) %>% kable()
```

**Rule:** Lowest AIC wins, unless BIC gives a different answer. Both penalize complexity; BIC more strongly.

## 10.2 Check residuals of the winner

```{r}
checkresiduals(fit_a)
```

## 10.3 Do two models perform equally?

**This is exam Q18-3b.**

Use a likelihood ratio test if the models are nested:

```{r}
# Example: AR(1) vs AR(2) are nested
LRT <- 2 * (logLik(fit_c) - logLik(fit_a))
p_val <- 1 - pchisq(as.numeric(LRT), df = 1)
LRT
p_val
```

If p > 0.05, the simpler model is sufficient. If p < 0.05, the larger model is better.

For non-nested models, compare AIC directly. The model with lower AIC is preferred.

---

# Part 11 — Seasonal data

If `frequency > 1`, use seasonal ARIMA.

```{r}
ts_seas <- ts(rnorm(120), frequency = 12)
fit_seas <- auto.arima(ts_seas)
summary(fit_seas)
```

Output will show `ARIMA(p, d, q)(P, D, Q)[12]`.

**Report:** "The fitted model is ARIMA(p, d, q)(P, D, Q)[12]."

**Seasonal plots:**

```{r}
ggseasonplot(ts_seas) + theme_minimal()
ggsubseriesplot(ts_seas) + theme_minimal()
```

---

# Part 12 — Decision guide

| Question | Command |
|---|---|
| Convert to time series | `ts(y, start = ..., frequency = ...)` |
| Plot | `autoplot(ts_y)` or `plot(ts_y)` |
| ACF/PACF | `acf(ts_y)`, `pacf(ts_y)` |
| Test stationarity | `tseries::adf.test(ts_y)` |
| Difference | `diff(ts_y)` |
| Automatic ARIMA | `auto.arima(ts_y)` |
| Manual ARIMA | `Arima(ts_y, order = c(p, d, q))` |
| Compare models | AIC, BIC |
| Diagnostics | `checkresiduals(fit)` |
| Ljung-Box | `Box.test(residuals(fit), lag = 10, type = "Ljung-Box")` |
| Forecast | `forecast(fit, h = 10)` |
| Plot forecast | `autoplot(fc)` |
| Missing values | `Arima` handles them automatically |
| AR(1) as regression | `lm(y ~ lag(y))` |
| MLE for AR(1) | `optim` on log-likelihood |

---

# Part 13 — Quick reference

| Task | Command |
|---|---|
| ts object | `ts(y, start = 1968, frequency = 1)` |
| Plot | `autoplot(ts_y)` |
| ACF | `acf(ts_y)` |
| PACF | `pacf(ts_y)` |
| ADF test | `adf.test(ts_y)` |
| auto.arima | `auto.arima(ts_y)` |
| Arima | `Arima(ts_y, order = c(p, d, q))` |
| Coef | `coef(fit)` |
| AIC | `AIC(fit)` |
| BIC | `BIC(fit)` |
| Ljung-Box | `Box.test(residuals(fit), lag = 10, type = "Ljung-Box")` |
| Residual plots | `checkresiduals(fit)` |
| Forecast | `forecast(fit, h = 10)` |
| Plot forecast | `autoplot(fc)` |
| AR(1) regression | `lm(y ~ lag(y))` |
| MLE AR(1) | `optim` on log-likelihood |

---

# Part 14 — Common mistakes

1. Forgetting to convert the data to `ts` before calling `acf`, `pacf`, or `Arima`.
2. Setting the wrong `frequency` (1 for annual, 12 for monthly, 4 for quarterly).
3. Reading ACF/PACF wrong. ACF cuts → MA. PACF cuts → AR. Both decay → ARMA.
4. Not checking residuals after fitting. Always use `checkresiduals` or Ljung-Box.
5. Comparing models with different differencing orders. Both must have the same `d`.
6. Reporting only point forecasts without prediction intervals.
7. Forgetting `include = 100` in `autoplot(fc)` — history disappears otherwise.
8. Using `diff()` without keeping track of how many differences you took.
9. Confusing `Arima` (capital A, from forecast) with `arima` (lowercase, from stats). They have different defaults.
10. Reporting the model as "ARIMA(1,0,0)" without saying what the coefficients mean.

---

# Part 15 — The single most important idea

Every time series question is one of these:

1. **Plot the series** and identify trend, seasonality, stationarity.
2. **Look at ACF/PACF** to guess AR and MA orders.
3. **Fit** with `auto.arima` or `Arima`.
4. **Check residuals** with `checkresiduals`.
5. **Forecast** with `forecast(fit, h)`.

If you can run those five steps, you can answer any time series question in the exam.
````

---

## What is covered

| Section | Exam questions |
|---|---|
| Part 1–2 | 17-1a, 23-1a |
| Part 3 | 23-1b |
| Part 4 | 18-3a, 23-1c |
| Part 5 | 21-4a, 21-4b |
| Part 6 | 21-4c |
| Part 7 | 17-1a, 23-1c |
| Part 8 | 17-1b, 18-3d, 23-1d |
| Part 9 | 17-1c |
| Part 10 | 18-3b, 18-3c, 23-1c |
| Part 11 | seasonal extensions |
